# Part I: Foundation & Environment {-}

# Installation & Environment Setup {#installation-environment}

<div class="chapter-summary">
**What You'll Learn:**

- Package installation errors and how to fix them
- Understanding repository and connection issues
- Compilation errors on different operating systems
- Version compatibility problems

**Key Errors Covered:** 15+ installation-related errors

**Difficulty:** ⭐ Beginner to ⭐⭐ Intermediate
</div>

## Introduction

Before you can make interesting errors in R, you need to install R and packages. Unfortunately, this is where many learners encounter their first frustrations. Installation errors are particularly annoying because they prevent you from even starting.

The good news: most installation errors follow predictable patterns and have standard solutions.

## Error #1: Package Installation Failed {#pkg-install-failed}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-package">📦 PACKAGE</span>

### The Error

```{r eval=FALSE}
install.packages("ggplot2")
```

<div class="error-box">
🔴 **ERROR**

```
Warning in install.packages :
  installation of package 'ggplot2' had non-zero exit status
```
</div>

### What It Means

The installation process failed at some point. This is a generic error that can have many causes.

### Common Causes and Solutions

#### Cause 1: No Internet Connection

**Symptom:** Cannot reach CRAN mirror

<div class="solution-box">
✅ **SOLUTION**

1. Check your internet connection
2. Try a different CRAN mirror:
```{r eval=FALSE}
# See available mirrors
getCRANmirrors()

# Set a specific mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Then try again
install.packages("ggplot2")
```
</div>

#### Cause 2: Firewall or Proxy

**Symptom:** Connection times out

<div class="solution-box">
✅ **SOLUTION**

```{r eval=FALSE}
# Configure proxy (if behind corporate firewall)
Sys.setenv(http_proxy = "http://proxy.company.com:8080")
Sys.setenv(https_proxy = "https://proxy.company.com:8080")

# Or download package manually and install from file
install.packages("path/to/package.tar.gz", repos = NULL, type = "source")
```
</div>

#### Cause 3: Insufficient Permissions

**Symptom:** Permission denied errors on Windows/Mac/Linux

<div class="solution-box">
✅ **SOLUTION**

**On Windows:**
- Run RStudio as Administrator
- Or install to user library:

```{r eval=FALSE}
# Check library paths
.libPaths()

# Install to user library (first in list)
install.packages("ggplot2", lib = .libPaths()[1])
```

**On Mac/Linux:**
```{bash eval=FALSE}
# If needed, create user library directory
mkdir -p ~/R/library
```

```{r eval=FALSE}
# Then set it in R
.libPaths(c("~/R/library", .libPaths()))
install.packages("ggplot2")
```
</div>

#### Cause 4: Disk Space

**Symptom:** No space left on device

<div class="solution-box">
✅ **SOLUTION**

1. Check disk space
2. Clean up old packages:
```{r eval=FALSE}
# See what's installed
installed.packages()[, c("Package", "Version")]

# Remove old packages
remove.packages("old_package_name")

# Clean up temporary files
unlink(tempdir(), recursive = TRUE)
```
</div>

## Error #2: Unable to Access Index for Repository {#repo-access}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-package">📦 PACKAGE</span>

### The Error

```{r eval=FALSE}
install.packages("dplyr")
```

<div class="warning-box">
🟡 **WARNING**

```
Warning: unable to access index for repository https://cran.rstudio.com/src/contrib
  cannot open URL 'https://cran.rstudio.com/src/contrib/PACKAGES'
```
</div>

### What It Means

R cannot reach the CRAN repository to download package information.

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Check CRAN status:**
- Visit https://cran.r-project.org/ in browser
- If down, try different mirror

**2. Change repository:**
```{r eval=FALSE}
# Use RStudio's mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Or choose interactively
chooseCRANmirror()
```

**3. Force HTTP instead of HTTPS:**
```{r eval=FALSE}
# If SSL issues
options(repos = c(CRAN = "http://cran.r-project.org"))
```

**4. Check .Rprofile:**
```{r eval=FALSE}
# See if repos are hardcoded
file.edit("~/.Rprofile")
# Remove or update any repos settings
```
</div>

## Error #3: Package Not Available for R Version {#version-mismatch}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-package">📦 PACKAGE</span>

### The Error

```{r eval=FALSE}
install.packages("newpackage")
```

<div class="warning-box">
🟡 **WARNING**

```
Warning message:
package 'newpackage' is not available for R version 4.0.0
```
</div>

### What It Means

The package either:
1. Requires a newer R version
2. Is not on CRAN (wrong name or removed)
3. Has been archived

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Check R version:**
```{r}
R.version.string
```

**2. Update R if needed:**
- Download from https://cran.r-project.org/
- Or use `installr` package (Windows):
```{r eval=FALSE}
install.packages("installr")
installr::updateR()
```

**3. Check package name:**
```{r eval=FALSE}
# Search for package
available.packages()[grep("package_name", 
                          available.packages()[, "Package"]), ]
```

**4. Install from archive:**
```{r eval=FALSE}
# If package was archived
packageurl <- "https://cran.r-project.org/src/contrib/Archive/package/package_1.0.tar.gz"
install.packages(packageurl, repos = NULL, type = "source")
```

**5. Install from GitHub:**
```{r eval=FALSE}
# Many packages in development
install.packages("devtools")
devtools::install_github("author/package")
```
</div>

<div class="insight-box">
💡 **Key Insight: Package Lifecycle**

Packages can be:
- **On CRAN**: Current and maintained
- **Archived**: Old version still available but removed from current CRAN
- **On GitHub only**: Development version or not submitted to CRAN
- **Superseded**: Replaced by another package (e.g., `reshape` → `reshape2` → `tidyr`)
</div>

## Error #4: Dependencies Not Available {#dependencies}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-package">📦 PACKAGE</span>

### The Error

```{r eval=FALSE}
install.packages("complexpackage")
```

<div class="error-box">
🔴 **ERROR**

```
ERROR: dependencies 'pkgA', 'pkgB' are not available for package 'complexpackage'
```
</div>

### What It Means

The package needs other packages (dependencies) that aren't available or can't be installed.

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Install dependencies first:**
```{r eval=FALSE}
# R usually does this automatically, but sometimes fails
install.packages("pkgA")
install.packages("pkgB")
install.packages("complexpackage")
```

**2. Force dependency installation:**
```{r eval=FALSE}
install.packages("complexpackage", dependencies = TRUE)
```

**3. Check for Bioconductor packages:**
```{r eval=FALSE}
# Some dependencies are on Bioconductor, not CRAN
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("dependencyPackage")
```

**4. Manual dependency resolution:**
```{r eval=FALSE}
# See what's needed
tools::package_dependencies("complexpackage", recursive = TRUE)

# Install each one
sapply(deps, install.packages)
```
</div>

## Error #5: Lazy Loading Failed {#lazy-load}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-package">📦 PACKAGE</span>

### The Error

```{r eval=FALSE}
library(ggplot2)
```

<div class="error-box">
🔴 **ERROR**

```
Error: package or namespace load failed for 'ggplot2':
 .onLoad failed in loadNamespace() for 'ggplot2', details:
  call: NULL
  error: lazy-load database 'path/to/ggplot2/R/ggplot2.rdb' is corrupt
```
</div>

### What It Means

The package installation is corrupted or incomplete.

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Reinstall the package:**
```{r eval=FALSE}
remove.packages("ggplot2")
install.packages("ggplot2")
```

**2. Restart R session:**
```{r eval=FALSE}
# In RStudio: Session > Restart R
.rs.restartR()

# Or from command line
q()  # Then restart R
```

**3. Check for partial installation:**
```{r eval=FALSE}
# See package status
library(ggplot2)
packageVersion("ggplot2")

# Compare to CRAN version
available.packages()["ggplot2", "Version"]
```

**4. Install from source:**
```{r eval=FALSE}
install.packages("ggplot2", type = "source")
```

**5. Clear package cache:**
```{r eval=FALSE}
# Sometimes helps
unlink(.libPaths()[1], recursive = TRUE)
dir.create(.libPaths()[1])
install.packages("ggplot2")
```
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall: Interrupted Installation**

If you interrupt package installation (Ctrl+C or ESC):
1. The package may be partially installed
2. Loading it will fail with lazy-load errors
3. Always reinstall after interruption

**Prevention:**
- Let installations complete
- If it's taking too long, check your internet connection first
</div>

## Error #6: Package Built Under R Version {#version-built}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-package">📦 PACKAGE</span>

### The Warning

```{r eval=FALSE}
library(dplyr)
```

<div class="warning-box">
🟡 **WARNING**

```
Warning message:
package 'dplyr' was built under R version 4.2.3
```
</div>

### What It Means

You're using R 4.2.0, but the package was compiled for R 4.2.3. Usually this is fine, but can cause issues.

### Should You Worry?

**Usually NO** - Minor version differences (4.2.0 vs 4.2.3) rarely cause problems.

**Maybe YES** - If you experience:
- Strange errors from that package
- Crashes
- Unexpected behavior

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Ignore it (usually fine):**
Most of the time, this warning is harmless.

**2. Update R:**
```{r eval=FALSE}
# Check your version
R.version

# Update if significantly behind
# Windows: use installr
# Mac: download from CRAN
# Linux: use system package manager
```

**3. Reinstall package from source:**
```{r eval=FALSE}
install.packages("dplyr", type = "source")
```

**4. Suppress warning:**
```{r eval=FALSE}
# Only if you're sure it's fine
suppressWarnings(library(dplyr))

# Or globally
options(warn = -1)  # Not recommended!
```
</div>

## Error #7: Rtools Required But Not Installed {#rtools}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-package">📦 PACKAGE</span>

**Platform:** Windows only

### The Warning

```{r eval=FALSE}
install.packages("package_with_cpp", type = "source")
```

<div class="warning-box">
🟡 **WARNING**

```
WARNING: Rtools is required to build R packages but is not currently installed.
Please download and install Rtools from https://cran.r-project.org/bin/windows/Rtools/
```
</div>

### What It Means

You're trying to install a package that needs compilation (C/C++/Fortran code), but Windows doesn't have the necessary compilers.

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Install binary instead:**
```{r eval=FALSE}
# Use pre-compiled version
install.packages("package_name")  # default uses binary
```

**2. Install Rtools:**
1. Download from https://cran.r-project.org/bin/windows/Rtools/
2. Run installer (default options usually work)
3. Restart R/RStudio
4. Verify installation:

```{r eval=FALSE}
Sys.which("make")
# Should show path like: C:\\rtools43\\usr\\bin\\make.exe
```

**3. Configure PATH:**
Sometimes Rtools doesn't add itself to PATH:

```{r eval=FALSE}
# Add to .Renviron
writeLines('PATH="${RTOOLS43_HOME}\\usr\\bin;${PATH}"', 
           con = "~/.Renviron")

# Restart R
.rs.restartR()
```
</div>

<div class="insight-box">
💡 **Key Insight: Source vs Binary Packages**

**Binary packages:**
- Pre-compiled for your OS
- Fast to install
- Can't modify source code

**Source packages:**
- Raw R + C/C++ code
- Requires compilation tools
- Necessary for package development
- Sometimes more up-to-date

**On Windows:** Use binaries unless you need source
**On Mac/Linux:** Source compilation more common
</div>

## Error #8: Library Not Found for -lX {#library-not-found}

<span class="difficulty-advanced">⭐⭐⭐ ADVANCED</span> <span class="category-badge cat-package">📦 PACKAGE</span>

**Platform:** Mac/Linux

### The Error

```{r eval=FALSE}
install.packages("XML")
```

<div class="error-box">
🔴 **ERROR**

```
ld: library not found for -lxml2
clang: error: linker command failed with exit code 1
```
</div>

### What It Means

The package requires external libraries (system dependencies) that aren't installed.

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**Mac (using Homebrew):**
```{bash eval=FALSE}
# Install Homebrew if needed
/bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"

# Install required library
brew install libxml2

# Then in R
install.packages("XML")
```

**Ubuntu/Debian:**
```{bash eval=FALSE}
sudo apt-get update
sudo apt-get install libxml2-dev
```

**Fedora/RedHat:**
```{bash eval=FALSE}
sudo yum install libxml2-devel
```

**Common Dependencies:**
```{bash eval=FALSE}
# For common R packages
# Ubuntu/Debian:
sudo apt-get install \
  libcurl4-openssl-dev \
  libssl-dev \
  libxml2-dev \
  libgit2-dev \
  libharfbuzz-dev \
  libfribidi-dev

# Mac:
brew install libgit2 openssl libxml2
```
</div>

<div class="bestpractice-box">
🎯 **Best Practice: Document System Requirements**

If you're sharing code that requires specific packages:

```{r eval=FALSE}
# Create a requirements file
cat("# System requirements (Ubuntu/Debian):
# sudo apt-get install libcurl4-openssl-dev libssl-dev

# R packages:
required_packages <- c('dplyr', 'ggplot2', 'readr')
install.packages(required_packages)
", file = "requirements.txt")
```
</div>

## Error #9: Replacing Previous Import {#replacing-import}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-package">📦 PACKAGE</span>

### The Warning

```{r eval=FALSE}
library(dplyr)
library(plyr)
```

<div class="warning-box">
🟡 **WARNING**

```
Attaching package: 'plyr'

The following objects are masked from 'package:dplyr':

    arrange, count, desc, failwith, id, mutate, rename, summarise, summarize
```
</div>

### What It Means

Both packages have functions with the same names. The later-loaded package's functions will be used by default.

### Should You Worry?

**YES** - This can cause confusing behavior if you expect dplyr's `mutate()` but get plyr's version.

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Load packages in correct order:**
```{r eval=FALSE}
library(plyr)   # Load first
library(dplyr)  # Load second (masks plyr)

# Now dplyr functions take precedence
```

**2. Use package::function notation:**
```{r eval=FALSE}
dplyr::mutate(data, new_col = x + 1)  # Explicit
plyr::mutate(data, new_col = x + 1)   # Explicit
```

**3. Avoid loading conflicting packages:**
```{r eval=FALSE}
# Don't load plyr if you're using dplyr
# dplyr supersedes most plyr functionality
library(dplyr)
```

**4. Check for conflicts:**
```{r eval=FALSE}
# See all conflicts
conflicts()

# With details
library(conflicted)
conflict_scout()
```
</div>

## Debugging Installation Issues

### General Troubleshooting Steps

<div class="bestpractice-box">
🎯 **Systematic Approach**

1. **Restart R**
```{r eval=FALSE}
.rs.restartR()  # RStudio
# Or: Session > Restart R
```

2. **Update R**
```{r}
R.version.string
# Compare to https://cran.r-project.org/
```

3. **Update packages**
```{r eval=FALSE}
update.packages(ask = FALSE)
```

4. **Check installation details**
```{r eval=FALSE}
# Verbose output
install.packages("package", verbose = TRUE)

# Keep source for inspection
install.packages("package", INSTALL_opts = "--no-clean-on-error")
```

5. **Check session info**
```{r}
sessionInfo()
# Shows R version, platform, loaded packages
```

6. **Test in clean session**
```{r eval=FALSE}
# Start R with no saved data
R --vanilla
```
</div>

### Getting Help

<div class="insight-box">
💡 **When Asking for Help, Provide:**

1. **Error message** (complete, not summary)
2. **R version:** `R.version.string`
3. **Platform:** `Sys.info()["sysname"]`
4. **What you tried:** Your code
5. **Session info:** `sessionInfo()`

**Good question format:**
```
I'm trying to install package X on Windows 11 with R 4.3.0.

Error message:
[paste complete error]

I tried:
- Updating R
- Different CRAN mirror
- Installing dependencies

sessionInfo() output:
[paste]
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Most installation errors are environmental**, not code-related
2. **Binary vs source** matters on Windows
3. **System dependencies** are common on Mac/Linux
4. **Package conflicts** happen - use `package::function()`
5. **Reinstalling** solves many issues
6. **Update regularly** but cautiously

**Quick Reference:**

| Error | First Try |
|-------|-----------|
| Installation failed | Restart R, try again |
| Repository access | Change CRAN mirror |
| Version mismatch | Update R |
| Dependencies missing | Install with `dependencies = TRUE` |
| Lazy-load corruption | Reinstall package |
| Rtools (Windows) | Install Rtools |
| Library not found (Mac/Linux) | Install system library |
| Package conflicts | Load order or use `::` |
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Diagnosis Practice**

What's wrong with each scenario?

```{r eval=FALSE}
# Scenario 1
install.packages("dplyr")
# Warning: unable to access index for repository

# Scenario 2
library(ggplot2)
# Error: package 'ggplot2' was built before R 4.0.0

# Scenario 3  
install.packages("devtools")
# ERROR: dependency 'usethis' is not available
```

**Answers at end of chapter**
</div>

<div class="exercise-box">
📝 **Exercise 2: Prevention**

Set up your R environment for smooth package installation:

1. Configure a reliable CRAN mirror
2. Set up a user library path
3. Install essential build tools for your OS
4. Create a package installation script

Try writing this setup code.
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

*Scenario 1:* Cannot reach CRAN repository
- Solution: Check internet, try different mirror

*Scenario 2:* R version too old
- Solution: Update R to 4.0.0 or higher

*Scenario 3:* Missing dependency
- Solution: `install.packages("devtools", dependencies = TRUE)`

**Exercise 2:**

```{r eval=FALSE}
# Setup script
setup_r_environment <- function() {
  # 1. Set CRAN mirror
  options(repos = c(CRAN = "https://cloud.r-project.org"))
  
  # 2. Set user library
  user_lib <- "~/R/library"
  if (!dir.exists(user_lib)) dir.create(user_lib, recursive = TRUE)
  .libPaths(c(user_lib, .libPaths()))
  
  # 3. Install essential packages
  essentials <- c("devtools", "tidyverse", "rmarkdown")
  new_packages <- essentials[!(essentials %in% installed.packages()[,"Package"])]
  if(length(new_packages)) install.packages(new_packages)
  
  # 4. Save to .Rprofile for future sessions
  cat('options(repos = c(CRAN = "https://cloud.r-project.org"))\n', 
      file = "~/.Rprofile", append = TRUE)
  
  message("R environment configured!")
}
```
</details>
