# Objects and Variables {#objects-variables}

<div class="chapter-summary">
**What You'll Learn:**

- How R stores and finds objects
- The most common beginner errors
- Understanding R's case sensitivity and naming rules
- Scoping basics that prevent errors

**Key Errors Covered:** 10+ object-related errors

**Difficulty:** ⭐ Beginner
</div>

## Introduction

Every R session starts with creating objects. And every R learner's journey starts with `object 'x' not found`. This error is so universal that it deserves deep understanding.

```{r}
# Your first R command
x <- 42
x
```

Simple, right? Until it's not. Let's explore what goes wrong.

## Error #1: `object 'x' not found` {#error-object-not-found}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-scope">🔗 SCOPING</span>

### The Error

```{r error=TRUE}
# Try to use an object that doesn't exist
my_number + 5
```

<div class="error-box">
🔴 **ERROR**

```
Error: object 'my_number' not found
```
</div>

### What It Means

R looked for an object named `my_number` in:
1. The current environment (`.GlobalEnv`)
2. All loaded packages (in search path order)
3. The base environment

...and couldn't find it anywhere.

### Common Causes

#### Cause 1: Typo in Object Name

```{r error=TRUE}
my_variable <- 10
my_varaible + 5  # Notice the typo: varaible vs variable
```

💡 **Key Insight:** R is case-sensitive!

```{r error=TRUE}
Temperature <- 98.6
temperature  # Wrong case!
```

```{r}
# These are THREE different objects:
myVariable <- 1
MyVariable <- 2
myvariable <- 3

ls()  # All three exist
```

#### Cause 2: Never Created the Object

```{r error=TRUE}
# Forgot to run this line:
# result <- 100 * 2

# Trying to use it:
result / 4
```

#### Cause 3: Object Created in Different Scope

```{r error=TRUE}
my_function <- function() {
  local_var <- 42
  print(local_var)  # Works inside function
}

my_function()
local_var  # Doesn't exist outside function!
```

#### Cause 4: Cleared Environment

```{r}
x <- 10
x  # Exists

rm(x)  # Removed it

# Now:
```

```{r error=TRUE}
x  # Gone!
```

#### Cause 5: Running Lines Out of Order

```{r eval=FALSE}
# If you run line 10 before line 5:

# Line 5
my_data <- read.csv("data.csv")

# Line 10 (run this first by mistake)
summary(my_data)  # Error! my_data doesn't exist yet
```

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Check spelling carefully:**
```{r}
my_variable <- 10
my_variable + 5  # Correct spelling
```

**2. Verify object exists:**
```{r}
# List all objects
ls()

# Check if specific object exists
exists("my_variable")

# Search in environment
grep("var", ls(), value = TRUE)  # Find objects with "var"
```

**3. Use RStudio's autocomplete:**
- Type the first few letters
- Press Tab to see available objects

**4. Check your environment pane:**
In RStudio, look at the Environment pane (top-right) to see all objects.

**5. Run all necessary code:**
Make sure you've executed all lines that create the objects you need.

**6. Restart and run from top:**
```{r eval=FALSE}
# Clear everything and start fresh
rm(list = ls())
# Or: Ctrl+Shift+F10 in RStudio (restart R)
```
</div>

<div class="bestpractice-box">
🎯 **Best Practices**

1. **Use consistent naming:** Choose a style and stick to it
   - `snake_case` (recommended for R)
   - `camelCase`
   - Avoid `dot.case` (can be confusing with S3 methods)

2. **Meaningful names:** `temperature_celsius` > `temp` > `t`

3. **Avoid similar names:** 
```{r eval=FALSE}
# Confusing:
data1 <- ...
data2 <- ...
data_new <- ...
data_final <- ...
data_final2 <- ...

# Better:
raw_data <- ...
clean_data <- ...
analyzed_data <- ...
```

4. **Run scripts top-to-bottom:** Your script should work when run fresh
</div>

## Error #2: `could not find function "x"` {#error-function-not-found}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-package">📦 PACKAGE</span>

### The Error

```{r error=TRUE}
# Try to use a function from unloaded package
read_csv("data.csv")
```

<div class="error-box">
🔴 **ERROR**

```
Error: could not find function "read_csv"
```
</div>

### What It Means

R can't find a function with that name. Functions are searched in:
1. The current environment
2. All loaded packages
3. **NOT** inside unloaded packages

### Common Causes

#### Cause 1: Package Not Loaded

```{r eval=FALSE}
# tidyverse not loaded
read_csv("data.csv")  # Error!

# Solution:
library(readr)
read_csv("data.csv")  # Works!
```

#### Cause 2: Package Not Installed

```{r eval=FALSE, error=TRUE}
library(somepackage)  # If not installed, this errors
```

```
Error in library(somepackage) : 
  there is no package called 'somepackage'
```

#### Cause 3: Typo in Function Name

```{r error=TRUE}
# Base R function is read.csv (dot), not read_csv (underscore)
meen(c(1, 2, 3))  # Typo: mean
```

#### Cause 4: Object Overwrote Function

```{r}
# Created an object with same name as a function
mean <- 42
mean
```

```{r error=TRUE}
# Now the function is gone!
mean(c(1, 2, 3))  
```

<div class="error-box">
🔴 **ERROR**

```
Error in mean(c(1, 2, 3)) : could not find function "mean"
```

Wait, what? We just used `mean`! But we overwrote it with the number 42.
</div>

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Load the required package:**
```{r eval=FALSE}
library(readr)
read_csv("data.csv")
```

**2. Install then load:**
```{r eval=FALSE}
install.packages("readr")
library(readr)
```

**3. Use package::function notation:**
```{r eval=FALSE}
# Use function without loading entire package
readr::read_csv("data.csv")

# Always works, no library() needed
dplyr::mutate(data, new_col = x + 1)
```

**4. Check function spelling:**
```{r eval=FALSE}
# Base R uses dots
read.csv("data.csv")  # Note the dot

# tidyverse uses underscores  
library(readr)
read_csv("data.csv")  # Note the underscore
```

**5. Remove conflicting object:**
```{r}
# If you accidentally overwrote:
mean <- 42  # Bad!
rm(mean)    # Remove the object
mean(c(1, 2, 3))  # Now the function works
```

**6. Find where function lives:**
```{r eval=FALSE}
# Search for functions
??read_csv
help.search("read_csv")

# Find package
find("read_csv")
```
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall: Overwriting Functions**

Never name objects after common functions:

**Avoid naming objects:**
- `mean`, `sum`, `length`, `data`, `df`, `c`, `t`, `T`, `F`
- `matrix`, `list`, `vector`, `table`
- `plot`, `points`, `lines`
- Any function you use regularly!

**If you accidentally do it:**
```{r eval=FALSE}
# You created: sum <- 100
sum(1:10)  # Error!

# Fix:
rm(sum)  # Remove it
sum(1:10)  # Works now

# Or restart R session: Ctrl+Shift+F10
```
</div>

## Error #3: `unexpected symbol in "x"` {#error-unexpected-symbol}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-syntax">🔤 SYNTAX</span>

### The Error

```{r error=TRUE}
my variable <- 10
```

<div class="error-box">
🔴 **ERROR**

```
Error: unexpected symbol in "my variable"
```
</div>

### What It Means

R's parser encountered something it didn't expect. Usually a space or character where it shouldn't be.

### Common Causes

#### Cause 1: Space in Variable Name

```{r error=TRUE}
my variable <- 10  # Spaces not allowed!
```

```{r}
# Solutions:
my_variable <- 10  # underscore
myVariable <- 10   # camelCase
my.variable <- 10  # dot (less recommended)
```

#### Cause 2: Missing Operator

```{r error=TRUE}
x <- 5
y <- 10
result <- x y  # Missing operator!
```

```{r}
# Fix:
result <- x * y  # or +, -, /, etc.
```

#### Cause 3: Two Statements on One Line

```{r error=TRUE}
x <- 5 y <- 10  # Two assignments without separator
```

```{r}
# Solutions:
x <- 5; y <- 10  # Semicolon separator

# Or on separate lines (preferred):
x <- 5
y <- 10
```

#### Cause 4: Incorrect String Quotes

```{r error=TRUE}
text <- "Hello world"  # Unmatched quotes
```

```{r}
# Fix:
text <- "Hello world"  # Matching quotes
```

#### Cause 5: Copying from Formatted Text

```{r eval=FALSE}
# Copying from Word/PDF might include smart quotes or special characters
name <- "John"  # These aren't regular quotes!
```

```{r}
# Should be:
name <- "John"  # Regular ASCII quotes
```

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Use underscores for multi-word names:**
```{r}
total_sales <- 1000
average_temperature <- 72.5
```

**2. Add missing operators:**
```{r}
a <- 10
b <- 5
c <- a + b  # Not: c <- a b
```

**3. Separate statements:**
```{r}
# One per line:
x <- 1
y <- 2

# Or use semicolon:
x <- 1; y <- 2
```

**4. Use consistent quotes:**
```{r}
# Choose one style and stick to it:
text1 <- "double quotes"  # More common in R
text2 <- 'single quotes'  # Also fine

# They're equivalent:
identical(text1, "double quotes")
identical(text2, 'single quotes')
```
</div>

## Error #4: `unexpected '=' in "x"` {#error-unexpected-equal}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-syntax">🔤 SYNTAX</span>

### The Error

```{r error=TRUE}
x = 5 = y
```

<div class="error-box">
🔴 **ERROR**

```
Error: unexpected '=' in "x = 5 ="
```
</div>

### What It Means

You used `=` where R didn't expect it.

### Common Causes

#### Cause 1: Chained Assignment (doesn't work like math)

```{r error=TRUE}
# Trying to set x and y to 5 (like math: x = y = 5)
x = 5 = y  # Doesn't work!
```

```{r}
# Solutions:
x <- y <- 5  # This works
# Or:
x <- 5
y <- 5
```

#### Cause 2: Using = Instead of ==

```{r error=TRUE}
x <- 10
if (x = 10) {  # Assignment, not comparison!
  print("yes")
}
```

```{r}
# Fix: use == for comparison
if (x == 10) {
  print("yes")
}
```

#### Cause 3: Wrong Context for =

```{r eval=FALSE}
# Inside function arguments, = is fine:
mean(x = c(1, 2, 3))  # OK

# For assignment, <- is clearer:
x <- c(1, 2, 3)  # Better style
```

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Use <- for assignment:**
```{r}
# Preferred in R
x <- 10
```

**2. Use == for comparison:**
```{r}
x <- 10
x == 10  # TRUE
x == 5   # FALSE
```

**3. Use = only in function arguments:**
```{r}
# Good:
mean(x = c(1, 2, 3), na.rm = TRUE)

# Also fine, but <- preferred for assignment:
x <- c(1, 2, 3)
mean(x, na.rm = TRUE)
```
</div>

<div class="insight-box">
💡 **Key Insight: <- vs = **

```{r}
# Both work for assignment:
x <- 10
x = 10

# But <- is preferred because:
# 1. Clearer intent (unambiguous assignment)
# 2. Works everywhere
# 3. R community standard

# = can be ambiguous:
mean(x = 1:10)  # Named argument (good)
x = 1:10        # Assignment (works, but <- preferred)
```

**Best Practice:** Use `<-` for assignment, `=` for function arguments
</div>

## Error #5: `object of type 'closure' is not subsettable` {#error-closure-not-subsettable}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
mean[1]  # Trying to subset the mean function
```

<div class="error-box">
🔴 **ERROR**

```
Error in mean[1] : object of type 'closure' is not subsettable
```
</div>

### What It Means

"Closure" = function. You're trying to use `[` on a function, which doesn't make sense.

### Common Causes

#### Cause 1: Forgot to Call the Function

```{r error=TRUE}
data <- c(1, 2, 3, 4, 5)
result <- mean  # Forgot parentheses!
result[1]  # Trying to subset the function itself
```

```{r}
# Fix:
result <- mean(data)  # Call the function
result  # Now it's a number
```

#### Cause 2: Accidentally Accessed Function Instead of Object

```{r error=TRUE}
# You have an object named 'data'
data <- data.frame(x = 1:5, y = 6:10)

# But there's also a function called 'data'
data[1, ]  # If you somehow reference the function...
```

This is rare, but shows why naming is important.

#### Cause 3: Function Name Typo Leads to Another Function

```{r error=TRUE}
c <- c(1, 2, 3)  # Bad! 'c' is a function
c[1]  # Now trying to subset the c() function
```

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Call the function:**
```{r}
# Wrong:
data <- c(1, 2, 3)
result <- mean
result[1]  # Error

# Right:
result <- mean(data)  # Call it
result  # Number, can't subset but don't need to
```

**2. Check if it's a function:**
```{r}
is.function(mean)  # TRUE
is.function(c(1,2,3))  # FALSE
```

**3. Don't overwrite function names:**
```{r eval=FALSE}
# Bad:
c <- c(1, 2, 3)
t <- read.csv("data.csv")
data <- process_data()

# Good:
my_vector <- c(1, 2, 3)
my_data <- read.csv("data.csv")
processed_data <- process_data()
```
</div>

## Error #6: `cannot change value of locked binding` {#error-locked-binding}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-scope">🔗 SCOPING</span>

### The Error

```{r error=TRUE}
T <- FALSE  # Trying to change T
```

<div class="error-box">
🔴 **ERROR**

```
Error in T <- FALSE : cannot change value of locked binding for 'T'
```
</div>

### What It Means

You're trying to modify a protected object. Some objects are locked to prevent accidental changes.

### Common Protected Objects

```{r eval=FALSE}
# These are locked:
T  # TRUE
F  # FALSE
pi
letters
LETTERS
month.name
month.abb
```

### Why This Protection Exists

```{r eval=FALSE}
# Imagine if you could do this:
T <- FALSE
F <- TRUE

# Now:
if (T) {  # But T is FALSE!
  print("This won't print")
}

# Chaos!
```

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Use a different name:**
```{r}
# Don't:
# T <- FALSE

# Do:
my_test <- FALSE
temperature <- 72
```

**2. Use full names (better practice anyway):**
```{r}
# Instead of T/F:
my_var <- TRUE   # Not T
other_var <- FALSE  # Not F

# More readable and no risk of confusion
```

**3. Never use T/F in code:**
```{r}
# Bad:
if (x > 0) T else F

# Good:
if (x > 0) TRUE else FALSE
```
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall: T and F**

**Never use `T` and `F` as shortcuts for `TRUE` and `FALSE`**

Why?
- They can be overwritten (in older R)
- Less readable
- Can cause subtle bugs

```{r eval=FALSE}
# Dangerous:
result <- T  # What if someone changed T?

# Safe:
result <- TRUE  # Can never be changed
```
</div>

## Understanding Variable Scope

<div class="insight-box">
💡 **Key Insight: Where Variables Live**

```{r}
# Global environment
x <- 10

my_function <- function() {
  # Function environment
  y <- 20
  
  # Can see global x
  print(x)
  
  # Can see local y
  print(y)
}

my_function()

# Global can't see local y
```

```{r error=TRUE}
print(y)  # Error!
```

**Scoping Rules:**
1. Look in current environment
2. Look in parent environment
3. Keep going up until found (or not)
</div>

## Naming Conventions

<div class="bestpractice-box">
🎯 **R Naming Best Practices**

**Valid names:**
```{r eval=FALSE}
# Letters, numbers, dots, underscores
my_variable
myVariable
my.variable
my_variable2
var_123
.hidden_var  # Starts with dot (not shown by ls())
```

**Invalid names:**
```{r eval=FALSE}
# Can't start with number
2var <- 10  # ERROR

# Can't have spaces
my var <- 10  # ERROR

# Can't use special characters
my-var <- 10  # ERROR (minus sign)
my$var <- 10  # ERROR
my@var <- 10  # ERROR
```

**Reserved words (can't use):**
```{r eval=FALSE}
# These are taken:
if, else, repeat, while, function, for, in, next, break
TRUE, FALSE, NULL, Inf, NaN, NA, NA_integer_, NA_real_, NA_complex_, NA_character_
```

**Recommended style:**
```{r}
# Snake case (recommended for R)
total_sales <- 1000
avg_temperature <- 72
customer_age <- 35

# CamelCase (also fine)
totalSales <- 1000
avgTemperature <- 72
customerAge <- 35

# Whatever you choose, be consistent!
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **R is case-sensitive**: `X` ≠ `x`
2. **Check spelling**: Most "object not found" errors are typos
3. **Load packages**: Functions need `library()` or `::`
4. **Don't overwrite functions**: Avoid naming objects after functions
5. **Use `<-` for assignment**: Clearer than `=`
6. **No spaces in names**: Use `_` or camelCase
7. **Check environment**: Use `ls()` and RStudio's Environment pane
8. **Never use T/F**: Always write TRUE/FALSE

**Quick Fixes:**

| Error | Likely Cause | Fix |
|-------|--------------|-----|
| object not found | Typo or not created | Check spelling, use `ls()` |
| function not found | Package not loaded | `library()` or `package::function()` |
| unexpected symbol | Space in name | Use `_` instead |
| unexpected = | Used = instead of == | Use `==` for comparison |
| closure not subsettable | Forgot () on function | Add parentheses |
| locked binding | Tried to change T/F | Use different name |
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Spot the Errors**

What's wrong with each line?

```{r eval=FALSE}
# 1
my result <- 100

# 2
Mean <- mean(c(1, 2, 3))
meen(c(4, 5, 6))

# 3
x = 10
if (x = 10) print("yes")

# 4
c <- c(1, 2, 3)
c[1]

# 5
library(dplyr)
select(data, column1)  # data doesn't exist

# 6
T <- FALSE
```
</div>

<div class="exercise-box">
📝 **Exercise 2: Debug This Script**

Fix all errors:

```{r eval=FALSE}
# Load data
my data <- read.csv("results.csv")

# Calculate mean
average <- Mean(my data$value)

# Compare
if (average = 50) {
  print("Average is 50")
}

# Store result
T <- average > 50
```
</div>

<div class="exercise-box">
📝 **Exercise 3: Good Names**

Rename these variables following best practices:

```{r eval=FALSE}
# Current (bad) names:
x <- data.frame(...)
data <- read.csv("sales.csv")
t <- TRUE
result1 <- ...
result2 <- ...
finaldata <- ...
```
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r eval=FALSE}
# 1 - Space in name
my_result <- 100  # or myResult

# 2 - Typo in function name
Mean <- mean(c(1, 2, 3))  # OK (but confusing name)
mean(c(4, 5, 6))  # Fix: mean not meen

# 3 - Used = instead of ==
x <- 10
if (x == 10) print("yes")

# 4 - Overwrote c() function
my_vector <- c(1, 2, 3)
my_vector[1]

# 5 - Object 'data' doesn't exist (need to create first)
# Also need to create it before this line

# 6 - Can't change T
my_test <- FALSE  # Use different name
```

**Exercise 2:**

```{r eval=FALSE}
# Load data
my_data <- read.csv("results.csv")  # Underscore, not space

# Calculate mean
average <- mean(my_data$value)  # mean not Mean

# Compare
if (average == 50) {  # == not =
  print("Average is 50")
}

# Store result
is_above_50 <- average > 50  # Don't use T
```

**Exercise 3:**

```{r eval=FALSE}
# Better names:
sales_dataframe <- data.frame(...)  # or just sales_data
sales_data <- read.csv("sales.csv")
is_valid <- TRUE  # or use full TRUE
raw_results <- ...
clean_results <- ...
final_sales_data <- ...  # or analyzed_sales
```
</details>
