# Part II: Data Types & Coercion {-}

# Type Mismatch Errors {#type-mismatch}

<div class="chapter-summary">
**What You'll Learn:**

- Understanding R's type system
- How coercion works (and fails)
- Type checking and conversion
- Common type mismatch scenarios
- How to prevent type errors

**Key Errors Covered:** 20+ type-related errors

**Difficulty:** ⭐ Beginner to ⭐⭐ Intermediate
</div>

## Introduction

R is **dynamically typed** but **strongly typed**. This means:
- You don't declare types (dynamic)
- But types matter for operations (strong)

```{r}
x <- 5        # R figures out it's numeric
y <- "5"      # R figures out it's character
```

But try to mix them:

```{r error=TRUE}
x + y  # Error!
```

Understanding type errors is fundamental to R mastery. This chapter covers every type mismatch you'll encounter.

## R's Basic Types

<div class="insight-box">
💡 **Key Insight: The Six Atomic Types**

R has six atomic (fundamental) types:

```{r}
# 1. Logical
is_true <- TRUE
typeof(is_true)

# 2. Integer
age <- 25L  # Note the L
typeof(age)

# 3. Double (numeric)
price <- 19.99
typeof(price)

# 4. Character
name <- "Alice"
typeof(name)

# 5. Complex
z <- 3 + 2i
typeof(z)

# 6. Raw (rarely used)
raw_byte <- charToRaw("A")
typeof(raw_byte)
```

**Most common:** logical, integer, double, character
</div>

## Error #1: `non-numeric argument to binary operator` {#non-numeric-binary}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
"10" + 5
```

<div class="error-box">
🔴 **ERROR**

```
Error in "10" + 5 : non-numeric argument to binary operator
```
</div>

### What It Means

You tried to use a mathematical operator (+, -, *, /, ^, %%, %/%) with something that isn't a number.

**Binary operator** = operator that works on two things (left + right)

### Common Causes

#### Cause 1: Character That Looks Like Number

```{r error=TRUE}
# Read from CSV without proper type specification
age <- "25"  # Actually character!
age + 10     # Error
```

```{r}
# Check type
class(age)
is.numeric(age)
is.character(age)
```

#### Cause 2: Factor Instead of Numeric

```{r error=TRUE}
# Factors are secretly integers with labels
scores <- factor(c("90", "85", "95"))
scores + 10  # Error!
```

```{r}
class(scores)
typeof(scores)  # "integer" but can't do math on it!
```

#### Cause 3: Missing Data Coerced to Character

```{r error=TRUE}
# One NA can turn everything to character
values <- c(10, 20, NA, 40)
values <- as.character(values)  # Accidentally
values[1] + 5  # Error!
```

#### Cause 4: Logical in Math (This Actually Works!)

```{r}
# Wait, this works?
TRUE + 5   # TRUE becomes 1
FALSE + 5  # FALSE becomes 0

# This is by design - logical coerces to numeric
sum(c(TRUE, FALSE, TRUE))  # Counts TRUEs
```

But mixing with character doesn't:

```{r error=TRUE}
"TRUE" + 5  # Character, not logical
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Convert to Numeric**

```{r}
# Basic conversion
age <- "25"
age <- as.numeric(age)
age + 10

# Check before converting
if (is.character(age)) {
  age <- as.numeric(age)
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Handle Factors Correctly**

```{r}
# Wrong way:
scores <- factor(c("90", "85", "95"))
as.numeric(scores)  # Gives factor levels (1,2,3), not values!

# Right way:
as.numeric(as.character(scores))  # Convert to char first

# Better way:
as.numeric(levels(scores)[scores])
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Read Data with Correct Types**

```{r eval=FALSE}
# Base R - specify column types
data <- read.csv("file.csv", 
                colClasses = c("numeric", "character", "numeric"))

# tidyverse - specify on read
library(readr)
data <- read_csv("file.csv",
                col_types = cols(
                  age = col_double(),
                  name = col_character(),
                  score = col_double()
                ))
```
</div>

<div class="solution-box">
✅ **SOLUTION 4: Safe Conversion with Error Handling**

```{r}
safe_as_numeric <- function(x) {
  result <- suppressWarnings(as.numeric(x))
  
  if (all(is.na(result)) && !all(is.na(x))) {
    warning("Conversion produced all NAs - check your data")
  }
  
  return(result)
}

# Test
safe_as_numeric("25")      # Works
safe_as_numeric("abc")     # Warning + NA
safe_as_numeric(c("1", "2", "three"))  # Partial conversion
```
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall: Silent Failures**

```{r}
# This looks like it worked...
x <- c("1", "2", "3", "four")
x <- as.numeric(x)
x
```

**Problem:** "four" became NA silently!

**Solution:** Check for NAs after conversion:
```{r}
if (any(is.na(x))) {
  warning("Some values couldn't be converted")
}
```
</div>

## Error #2: `non-numeric argument to mathematical function` {#non-numeric-math}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
sqrt("16")
```

<div class="error-box">
🔴 **ERROR**

```
Error in sqrt("16") : non-numeric argument to mathematical function
```
</div>

### What It Means

Mathematical functions (sqrt, log, exp, sin, cos, etc.) need numbers, not characters or other types.

### Common Functions That Give This Error

```{r error=TRUE}
# All of these error with character input:
sqrt("16")
log("10")
exp("2")
abs("-5")
round("3.14")
floor("4.7")
ceiling("4.2")
```

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Convert before calling function:**
```{r}
sqrt(as.numeric("16"))
log(as.numeric("10"))
```

**2. Vectorized conversion and operation:**
```{r}
values <- c("16", "25", "36")
sqrt(as.numeric(values))
```

**3. Use type-safe reading:**
```{r eval=FALSE}
# When reading data
data <- read.csv("data.csv", stringsAsFactors = FALSE)
data$numeric_col <- as.numeric(data$numeric_col)
```
</div>

## Error #3: `(list) object cannot be coerced to type 'double'` {#list-coerce-double}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
my_list <- list(a = 1, b = 2, c = 3)
sum(my_list)
```

<div class="error-box">
🔴 **ERROR**

```
Error in sum(my_list) : invalid 'type' (list) of argument
```
</div>

### What It Means

You're trying to do mathematical operations on a list, which is a container that can hold anything. R can't automatically convert a list to numbers.

### Common Causes

#### Cause 1: Using List Instead of Vector

```{r error=TRUE}
# List (wrong for math)
numbers_list <- list(1, 2, 3, 4, 5)
mean(numbers_list)  # Error!
```

```{r}
# Vector (right for math)
numbers_vec <- c(1, 2, 3, 4, 5)
mean(numbers_vec)  # Works!
```

#### Cause 2: Extracting From Data Frame Incorrectly

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10)

# Single bracket returns data frame (list-based)
sum(df[1])  # Error - still a data frame

# Double bracket returns vector
sum(df[[1]])  # Works!

# Dollar sign returns vector
sum(df$x)  # Works!
```

#### Cause 3: List Column in Data Frame

```{r error=TRUE}
# Modern R can have list columns
df <- data.frame(id = 1:3)
df$values <- list(c(1,2), c(3,4), c(5,6))

# Can't do math on list column
sum(df$values)  # Error!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Convert List to Vector**

```{r}
my_list <- list(a = 1, b = 2, c = 3)

# Unlist to vector
unlist(my_list)
sum(unlist(my_list))

# Or use do.call
do.call(sum, my_list)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use Correct Extraction**

```{r}
df <- data.frame(x = 1:5, y = 6:10)

# Good ways:
sum(df$x)      # Dollar sign
sum(df[[1]])   # Double bracket
sum(df[, 1])   # Bracket with comma

# Bad way:
# sum(df[1])   # Single bracket = data frame
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Handle List Columns**

```{r}
df <- data.frame(id = 1:3)
df$values <- list(c(1,2), c(3,4), c(5,6))

# Apply operation to each list element
sapply(df$values, sum)
lapply(df$values, mean)

# Or unnest first (tidyverse)
library(tidyr)
df %>% unnest(values)
```
</div>

<div class="insight-box">
💡 **Key Insight: List vs Vector**

```{r}
# Vector: All same type
vec <- c(1, 2, 3)
typeof(vec)
class(vec)

# List: Can mix types
lst <- list(1, "two", TRUE)
typeof(lst)
class(lst)

# Data frame: Special list of vectors
df <- data.frame(x = 1:3, y = 4:6)
typeof(df)  # "list"!
class(df)   # "data.frame"

# Single bracket keeps structure
df[1]      # Data frame (list)
df[[1]]    # Vector
```
</div>

## Error #4: `invalid type (closure) for variable 'X'` {#invalid-type-closure}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
# Accidentally using a function as data
data <- data.frame(x = 1:5)
plot(mean, data$x)  # mean is the function!
```

<div class="error-box">
🔴 **ERROR**

```
Error in plot.xy(xy.coords(x, y), type = type, ...) : 
  invalid type (closure) for variable 'mean'
```
</div>

### What It Means

"Closure" = function. You passed a function where R expected data.

### Common Causes

#### Cause 1: Forgot to Call Function

```{r error=TRUE}
numbers <- 1:10
plot(mean, numbers)  # Passed function itself
```

```{r}
# Fix: Call the function
plot(mean(numbers), numbers)
```

#### Cause 2: Variable Name Same as Function

```{r error=TRUE}
# Created variable named 'c'
c <- 100
data <- c(1, 2, 3)  # Now c() function is masked!

# Later, someone tries to use the function
# But 'c' is now the number 100
```

```{r}
# Check what something is
is.function(mean)
is.function(100)
```

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Call the function (add parentheses):**
```{r eval=FALSE}
# Wrong:
plot(mean, data)

# Right:
plot(mean(data), ...)
```

**2. Don't name variables after functions:**
```{r}
# Bad:
# mean <- 42
# sum <- 100
# data <- my_data

# Good:
average_value <- 42
total_sum <- 100
my_data <- ...
```

**3. Remove conflicting variable:**
```{r}
# If you accidentally created:
# sum <- 100

# Remove it:
rm(sum)

# Now sum() function works again
sum(1:10)
```
</div>

## Error #5: `cannot coerce class "X" to a data.frame` {#cannot-coerce-df}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
# Trying to convert incompatible type
my_func <- function() { return(42) }
as.data.frame(my_func)
```

<div class="error-box">
🔴 **ERROR**

```
Error in as.data.frame.default(my_func) : 
  cannot coerce class '"function"' to a data.frame
```
</div>

### Common Causes

#### Cause 1: Wrong Object Type

```{r error=TRUE}
# Can't convert function
as.data.frame(mean)

# Can't convert environment
as.data.frame(.GlobalEnv)
```

#### Cause 2: Wrong List Structure

```{r error=TRUE}
# Uneven list lengths
bad_list <- list(a = 1:3, b = 1:5)
as.data.frame(bad_list)  # Error - different lengths!
```

```{r}
# Must be same length or length 1
good_list <- list(a = 1:3, b = 4:6)
as.data.frame(good_list)

# Or use recycling
recycled_list <- list(a = 1:3, b = 1)  # b recycled
as.data.frame(recycled_list)
```

#### Cause 3: Matrix With Wrong Type

```{r error=TRUE}
# Some object types don't convert cleanly
nested <- list(list(1, 2), list(3, 4))
as.data.frame(nested)  # Error - nested lists!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Fix List Structure**

```{r}
# Uneven lengths - fix it
bad_list <- list(a = 1:3, b = 1:5)

# Option 1: Trim to shortest
min_len <- min(lengths(bad_list))
fixed_list <- lapply(bad_list, function(x) x[1:min_len])
as.data.frame(fixed_list)

# Option 2: Pad with NA
max_len <- max(lengths(bad_list))
fixed_list <- lapply(bad_list, function(x) {
  c(x, rep(NA, max_len - length(x)))
})
as.data.frame(fixed_list)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Convert Correctly**

```{r}
# From matrix
mat <- matrix(1:6, nrow = 2)
as.data.frame(mat)

# From vector with names
vec <- c(a = 1, b = 2, c = 3)
as.data.frame(as.list(vec))

# From nested list - flatten first
nested <- list(list(1, 2), list(3, 4))
flat <- unlist(nested, recursive = FALSE)
# Or handle differently depending on structure
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Check Before Converting**

```{r}
safe_as_df <- function(x) {
  # Check if it's already a data frame
  if (is.data.frame(x)) return(x)
  
  # Check if it's a matrix
  if (is.matrix(x)) return(as.data.frame(x))
  
  # Check if it's a list with equal lengths
  if (is.list(x)) {
    lens <- lengths(x)
    if (length(unique(lens)) == 1 || all(lens == 1 | lens == max(lens))) {
      return(as.data.frame(x))
    } else {
      stop("List elements have incompatible lengths: ", 
           paste(lens, collapse = ", "))
    }
  }
  
  # Try generic conversion
  tryCatch(
    as.data.frame(x),
    error = function(e) {
      stop("Cannot convert ", class(x), " to data.frame: ", e$message)
    }
  )
}

# Test
safe_as_df(list(a = 1:3, b = 4:6))  # Works
```
</div>

## Error #6: `NAs introduced by coercion` {#nas-coercion}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Warning (Usually)

```{r}
as.numeric(c("1", "2", "three", "4"))
```

<div class="warning-box">
🟡 **WARNING**

```
Warning message:
NAs introduced by coercion
```
</div>

### What It Means

R tried to convert something to numeric, but some values couldn't be converted, so they became NA.

### Common Scenarios

#### Scenario 1: Text in Numeric Column

```{r}
# Data entry errors
scores <- c("90", "85", "N/A", "92", "absent")
as.numeric(scores)

# Check which became NA
is.na(as.numeric(scores))
```

#### Scenario 2: Special Characters

```{r}
# Currency symbols
prices <- c("$10.99", "$25.50", "$8.75")
as.numeric(prices)  # All become NA!

# Need to remove $ first
as.numeric(gsub("\\$", "", prices))
```

#### Scenario 3: Scientific Notation Issues

```{r}
# Usually these work fine
as.numeric("1.5e-10")  # Scientific notation OK

# But typos don't
as.numeric("1.5E-10a")  # Typo creates NA
```

#### Scenario 4: Factors with Text Levels

```{r}
# Factor with non-numeric levels
responses <- factor(c("Yes", "No", "Yes", "Maybe"))
as.numeric(responses)  # Gives factor codes (1,2,1,3), not what you want

# And trying to convert to the levels gives NA
as.numeric(as.character(responses))
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Clean Data First**

```{r}
# Remove non-numeric characters
dirty <- c("$10.99", "€25.50", "8.75")

# Remove currency symbols
clean <- gsub("[^0-9.]", "", dirty)
as.numeric(clean)

# More robust cleaning
clean_numeric <- function(x) {
  # Remove everything except numbers, decimal, minus
  cleaned <- gsub("[^0-9.-]", "", x)
  as.numeric(cleaned)
}

clean_numeric(c("$10.99", "-25.5%", "8 dollars"))
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Handle NAs Explicitly**

```{r}
values <- c("1", "2", "three", "4")
converted <- as.numeric(values)

# Check which failed
failed <- is.na(converted) & !is.na(values)
if (any(failed)) {
  message("Could not convert: ", paste(values[failed], collapse = ", "))
}

# Or replace NAs with default
converted[is.na(converted)] <- 0
converted
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use readr's parse_number()**

```{r}
library(readr)

# Automatically extracts numbers
parse_number("$10.99")
parse_number("Price: $25.50")
parse_number("8.75%")

# Vector
parse_number(c("$10.99", "€25.50", "8.75"))
```
</div>

<div class="bestpractice-box">
🎯 **Best Practice: Validate After Coercion**

```{r}
coerce_with_validation <- function(x, to = "numeric") {
  original <- x
  
  if (to == "numeric") {
    converted <- as.numeric(x)
  } else if (to == "integer") {
    converted <- as.integer(x)
  } else {
    stop("Unsupported conversion type")
  }
  
  # Count NAs
  original_nas <- sum(is.na(original))
  new_nas <- sum(is.na(converted))
  introduced_nas <- new_nas - original_nas
  
  if (introduced_nas > 0) {
    warning(introduced_nas, " NAs introduced by coercion")
    failed_values <- original[is.na(converted) & !is.na(original)]
    message("Failed to convert: ", 
            paste(head(failed_values, 5), collapse = ", "),
            if(length(failed_values) > 5) "..." else "")
  }
  
  return(converted)
}

# Test
coerce_with_validation(c("1", "2", "three", "4"))
```
</div>

## Error #7: `character string is not in a standard unambiguous format` {#date-parse-error}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
as.Date("2024/13/01")  # Month 13 doesn't exist
```

<div class="error-box">
🔴 **ERROR**

```
Error in charToDate(x) : 
  character string is not in a standard unambiguous format
```
</div>

### What It Means

You're trying to convert a string to a Date, but R can't figure out the format, or the date is invalid.

### Common Causes

#### Cause 1: Wrong Date Format

```{r error=TRUE}
# American format (month/day/year)
as.Date("12/25/2024")  # R expects YYYY-MM-DD
```

```{r}
# Specify format
as.Date("12/25/2024", format = "%m/%d/%Y")
```

#### Cause 2: Invalid Date

```{r error=TRUE}
as.Date("2024-02-30")  # February doesn't have 30 days
as.Date("2024-13-01")  # Month 13 doesn't exist
```

#### Cause 3: Ambiguous Format

```{r error=TRUE}
# Is this Jan 2 or Feb 1?
as.Date("01/02/2024")  # R gets confused
```

```{r}
# Be explicit
as.Date("01/02/2024", format = "%m/%d/%Y")  # Jan 2
as.Date("01/02/2024", format = "%d/%m/%Y")  # Feb 1
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Specify Format**

```{r}
# Common formats
as.Date("2024-12-25")  # ISO format (default)
as.Date("12/25/2024", format = "%m/%d/%Y")
as.Date("25/12/2024", format = "%d/%m/%Y")
as.Date("Dec 25, 2024", format = "%b %d, %Y")
as.Date("December 25, 2024", format = "%B %d, %Y")
```

**Format codes:**
- `%Y` = 4-digit year (2024)
- `%y` = 2-digit year (24)
- `%m` = numeric month (12)
- `%d` = day of month (25)
- `%b` = abbreviated month (Dec)
- `%B` = full month (December)
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use lubridate (Easier!)**

```{r}
library(lubridate)

# Auto-detect common formats
ymd("2024-12-25")
mdy("12/25/2024")
dmy("25/12/2024")
mdy("Dec 25, 2024")

# Vector of dates
dates <- c("2024-12-25", "2024/01/15", "2024.06.30")
ymd(dates)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Handle Parse Failures**

```{r}
dates <- c("2024-12-25", "invalid", "2024-02-30", "2024-01-15")

# Base R - NAs for failures
parsed <- as.Date(dates)  # Warnings
parsed

# lubridate - shows which failed
library(lubridate)
parsed <- ymd(dates, quiet = FALSE)
parsed

# Custom handling
safe_parse_date <- function(x, format = "%Y-%m-%d") {
  result <- as.Date(x, format = format)
  
  # Report failures
  failed <- is.na(result) & !is.na(x)
  if (any(failed)) {
    message("Failed to parse ", sum(failed), " dates:")
    message(paste(x[failed], collapse = ", "))
  }
  
  return(result)
}

safe_parse_date(dates)
```
</div>

## Type Checking Functions

<div class="bestpractice-box">
🎯 **Best Practice: Check Types Before Operating**

```{r}
# Checking functions
is.numeric(5)       # TRUE for integer or double
is.integer(5L)      # TRUE only for integer
is.double(5.0)      # TRUE only for double
is.character("5")   # TRUE for character
is.logical(TRUE)    # TRUE for logical
is.factor(factor(1:3))  # TRUE for factor

# Getting type info
typeof(5)           # "double"
class(5)            # "numeric"
mode(5)             # "numeric"

# More specific checks
is.na(NA)           # TRUE for NA
is.null(NULL)       # TRUE for NULL
is.nan(NaN)         # TRUE for NaN (not a number)
is.infinite(Inf)    # TRUE for Inf
is.finite(5)        # TRUE for normal numbers

# Structure checks
is.vector(c(1,2,3))      # TRUE
is.list(list(1,2))       # TRUE
is.matrix(matrix(1:4, 2, 2))  # TRUE
is.data.frame(data.frame(x=1:3))  # TRUE
is.array(array(1:8, dim=c(2,2,2)))  # TRUE
```
</div>

## Type Conversion Functions

<div class="insight-box">
💡 **Key Insight: Conversion Functions**

```{r}
# To numeric
as.numeric("5")
as.integer("5")
as.double("5.5")

# To character
as.character(5)
as.character(TRUE)

# To logical
as.logical(1)        # TRUE
as.logical(0)        # FALSE
as.logical("TRUE")   # TRUE
as.logical("T")      # TRUE

# To factor
as.factor(c("A", "B", "A"))

# Special conversions
as.Date("2024-01-15")
as.POSIXct("2024-01-15 10:30:00")
```

**Coercion Hierarchy:**
logical → integer → double → character

Everything can become character!
```{r}
c(TRUE, 1L, 1.5, "text")  # All become character
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **R has 6 atomic types**: logical, integer, double, character, complex, raw
2. **Check types before operations**: Use `typeof()`, `class()`, `is.*()` functions
3. **Explicit is better than implicit**: Use `as.numeric()` rather than hoping
4. **Watch for silent failures**: Check for NAs after coercion
5. **Factors are tricky**: Convert to character before numeric
6. **Lists aren't vectors**: Use `unlist()` or `[[]]` extraction
7. **Specify date formats**: Don't rely on auto-detection
8. **Use lubridate for dates**: Much easier than base R

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| non-numeric argument to binary operator | Character in math | `as.numeric()` |
| non-numeric argument to math function | Character in function | `as.numeric()` |
| (list) cannot be coerced | Wrong structure | `unlist()` or `[[]]` |
| invalid type (closure) | Function instead of data | Call function or rename variable |
| cannot coerce to data.frame | Incompatible type | Fix structure or use correct conversion |
| NAs introduced by coercion | Invalid values | Clean data first |
| character string not in standard format | Date parse failure | Specify format or use lubridate |

**Type Checking Checklist:**
```{r eval=FALSE}
# Before doing math:
is.numeric(x)

# Before subsetting:
is.vector(x) || is.list(x)

# Before data frame operations:
is.data.frame(df)

# After conversion:
any(is.na(result))
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Type Detective**

What's wrong and how do you fix it?

```{r eval=FALSE}
# Scenario 1
age <- "25"
next_year <- age + 1

# Scenario 2
scores <- factor(c("90", "85", "95"))
average <- mean(as.numeric(scores))

# Scenario 3
df <- data.frame(x = 1:5)
total <- sum(df[1])

# Scenario 4
dates <- c("2024-01-15", "15/01/2024", "Jan 15 2024")
parsed <- as.Date(dates)
```
</div>

<div class="exercise-box">
📝 **Exercise 2: Type Conversion**

Write a function that:
1. Takes a vector of any type
2. Tries to convert to numeric
3. Reports which values failed
4. Returns numeric vector with NAs for failures
5. Provides a summary of conversions
</div>

<div class="exercise-box">
📝 **Exercise 3: Real Data**

You receive this data:

```{r eval=FALSE}
sales <- c("$1,234.56", "$987.65", "N/A", "$2,345.67", "pending")
dates <- c("01/15/2024", "2024-02-20", "Mar 15, 2024")
```

Clean and convert both to appropriate types.
</div>

<div class="exercise-box">
📝 **Exercise 4: Data Frame Types**

Debug this code:

```{r eval=FALSE}
df <- data.frame(
  id = 1:3,
  value = c("100", "200", "300"),
  date = c("2024-01-15", "2024-02-20", "2024-03-25")
)

# Want to do:
df$value_doubled <- df$value * 2
df$days_since <- Sys.Date() - df$date
```

Fix the types so operations work.
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r eval=FALSE}
# Scenario 1 - Character in math
age <- "25"
age <- as.numeric(age)  # Fix
next_year <- age + 1

# Scenario 2 - Factor to numeric wrong way
scores <- factor(c("90", "85", "95"))
# Wrong: as.numeric(scores) gives 1,2,3
# Right:
scores_num <- as.numeric(as.character(scores))
average <- mean(scores_num)

# Scenario 3 - Single bracket returns data frame
df <- data.frame(x = 1:5)
# Wrong: df[1] is still data frame
# Right:
total <- sum(df[[1]])  # or sum(df$x)

# Scenario 4 - Mixed date formats
dates <- c("2024-01-15", "15/01/2024", "Jan 15 2024")
# Need different formats for each
library(lubridate)
parsed <- c(ymd("2024-01-15"), 
            dmy("15/01/2024"),
            mdy("Jan 15 2024"))
```

**Exercise 2:**

```{r}
smart_numeric_convert <- function(x) {
  # Store original
  original <- x
  original_class <- class(x)
  
  # Attempt conversion
  converted <- suppressWarnings(as.numeric(x))
  
  # Identify failures
  original_na <- is.na(original)
  new_na <- is.na(converted)
  failures <- new_na & !original_na
  
  # Report
  cat("Conversion Summary:\n")
  cat("  Original type:", original_class, "\n")
  cat("  Total values:", length(x), "\n")
  cat("  Successful:", sum(!new_na), "\n")
  cat("  Failed:", sum(failures), "\n")
  cat("  Already NA:", sum(original_na), "\n\n")
  
  if (any(failures)) {
    cat("Failed values:\n")
    print(head(original[failures], 10))
  }
  
  return(converted)
}

# Test
smart_numeric_convert(c("1", "2", "three", "4", "five"))
```

**Exercise 3:**

```{r}
library(readr)
library(lubridate)

# Clean sales
sales <- c("$1,234.56", "$987.65", "N/A", "$2,345.67", "pending")

# Remove currency and commas, handle text
sales_clean <- gsub("[$,]", "", sales)
sales_num <- suppressWarnings(as.numeric(sales_clean))
sales_num[is.na(sales_num)]  <- 0  # Or handle differently

# Clean dates
dates <- c("01/15/2024", "2024-02-20", "Mar 15, 2024")

# Try multiple formats
dates_parsed <- as.Date(parse_date_time(dates, 
                                        orders = c("mdy", "ymd", "bdy")))

# Result
data.frame(
  sales = sales_num,
  date = dates_parsed
)
```

**Exercise 4:**

```{r}
df <- data.frame(
  id = 1:3,
  value = c("100", "200", "300"),
  date = c("2024-01-15", "2024-02-20", "2024-03-25"),
  stringsAsFactors = FALSE
)

# Fix types
df$value <- as.numeric(df$value)
df$date <- as.Date(df$date)

# Now operations work
df$value_doubled <- df$value * 2
df$days_since <- as.numeric(Sys.Date() - df$date)

df
```
</details>
