# Vectors & Recycling {#vectors-recycling}

<div class="chapter-summary">
**What You'll Learn:**

- How R's vector recycling works
- When recycling helps and when it hurts
- Length mismatch errors
- Replacement length errors
- Vectorization best practices

**Key Errors Covered:** 15+ recycling and length errors

**Difficulty:** ⭐ Beginner to ⭐⭐ Intermediate
</div>

## Introduction

R's superpower is **vectorization** - operations work on entire vectors at once. But with this power comes a quirky feature called **recycling** that causes endless confusion.

```{r}
# Simple vectorization
c(1, 2, 3) + c(10, 20, 30)
```

But what about this?

```{r}
# Different lengths!
c(1, 2, 3, 4) + c(10, 20)
```

It works! But is this what you wanted? Let's explore when recycling helps and when it causes errors.

## Understanding Recycling

<div class="insight-box">
💡 **Key Insight: The Recycling Rule**

When vectors of different lengths are used together, **R repeats the shorter one** to match the longer one.

```{r}
# What happens:
c(1, 2, 3, 4) + c(10, 20)

# R expands to:
c(1, 2, 3, 4) + c(10, 20, 10, 20)
#                        ↑   ↑  recycled!
```

**Works smoothly when:**
- One vector is length 1 (scalar)
- Lengths are multiples (2 and 4, 3 and 6)

**Warns when:**
- Lengths aren't multiples (3 and 5)

**Errors when:**
- Replacement context and lengths don't match
</div>

## Error #1: `longer object length is not a multiple` {#recycling-warning}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-dimension">📏 LENGTH</span>

### The Warning

```{r}
c(1, 2, 3) + c(10, 20, 30, 40, 50)
```

<div class="warning-box">
🟡 **WARNING**

```
Warning message:
In c(1, 2, 3) + c(10, 20, 30, 40, 50) :
  longer object length is not a multiple of shorter object length
```
</div>

### What It Means

R is recycling, but the lengths don't match evenly. This usually indicates a mistake.

### Common Causes

#### Cause 1: Data Mismatch

```{r}
# You have 100 observations
data <- rnorm(100)

# But only 3 group labels
groups <- c("A", "B", "C")

# Recycling happens
combined <- data.frame(value = data, group = groups)
```

The warning tells you: "Hey, are you sure about this?"

#### Cause 2: Filtering Gone Wrong

```{r}
x <- 1:10
y <- 1:7  # Oops, lost some values

# Operations warn
x + y
x * y
```

#### Cause 3: Unintended Partial Matching

```{r}
treatment <- c("Drug", "Placebo")
outcomes <- rnorm(25)  # 25 subjects

# Assigning treatment to outcomes
data.frame(outcome = outcomes, treatment = treatment)
```

Warning: 25 is not a multiple of 2!

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Fix the Lengths**

```{r}
# Original problem
x <- 1:10
y <- 1:7

# Option A: Trim to match
min_len <- min(length(x), length(y))
x[1:min_len] + y[1:min_len]

# Option B: Extend with NA
y_extended <- c(y, rep(NA, length(x) - length(y)))
x + y_extended

# Option C: Explicit recycling (if intentional)
y_recycled <- rep(y, length.out = length(x))
x + y_recycled
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check Lengths Before Operating**

```{r}
safe_operation <- function(x, y, op = `+`) {
  if (length(x) != length(y)) {
    # Check if one is length 1 (scalar - OK)
    if (length(x) == 1 || length(y) == 1) {
      return(op(x, y))
    }
    
    # Check if lengths are multiples
    if (max(length(x), length(y)) %% min(length(x), length(y)) != 0) {
      warning("Lengths are not multiples: ", 
              length(x), " and ", length(y))
    }
  }
  
  return(op(x, y))
}

# Test
safe_operation(1:10, 1:7, `+`)  # Warns
safe_operation(1:10, 1:5, `+`)  # No warning (10/5 = 2)
safe_operation(1:10, 2, `+`)    # No warning (scalar)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use rep() Explicitly**

```{r}
# Make intention clear
x <- 1:12
pattern <- c(1, 2, 3)

# Explicit recycling
y <- rep(pattern, length.out = length(x))
x + y

# Or with times
y <- rep(pattern, times = length(x) / length(pattern))
x + y
```
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall: Silent Recycling with Multiples**

```{r}
# No warning when lengths are multiples!
x <- 1:6
y <- c(10, 20, 30)  # 6 is multiple of 3

result <- x + y
result

# R expanded y to: c(10, 20, 30, 10, 20, 30)
# Was this intended?
```

**Always check:** Just because it doesn't warn doesn't mean it's correct!
</div>

## Error #2: `replacement has X rows, data has Y` {#replacement-rows}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-dimension">📏 LENGTH</span>

### The Error

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10)
df$z <- 1:3  # Wrong length!
```

<div class="error-box">
🔴 **ERROR**

```
Error in `$<-.data.frame`(`*tmp*`, z, value = 1:3) : 
  replacement has 3 rows, data has 5
```
</div>

### What It Means

You're trying to add/replace a column, but the number of values doesn't match the number of rows.

### Common Causes

#### Cause 1: Wrong Length Column

```{r error=TRUE}
df <- data.frame(id = 1:10)

# Calculated something with wrong length
summary_values <- c(100, 200, 300)  # Only 3 values

# Try to add as column
df$summary <- summary_values  # Error!
```

#### Cause 2: Filtered Data Reassignment

```{r error=TRUE}
df <- data.frame(x = 1:10, y = rnorm(10))

# Filter
subset_df <- df[df$y > 0, ]  # Maybe 6 rows

# Create column for subset
new_values <- 1:6

# Try to add to original
df$new <- new_values  # Error! Original has 10 rows
```

#### Cause 3: Aggregation Length Mismatch

```{r error=TRUE}
# 20 observations
df <- data.frame(
  id = 1:20,
  group = rep(c("A", "B"), each = 10)
)

# Aggregate to 2 values (one per group)
group_means <- tapply(df$id, df$group, mean)

# Try to add back to original
df$group_mean <- group_means  # Error! 2 values, 20 rows
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Match the Length**

```{r}
df <- data.frame(id = 1:10)
summary_values <- c(100, 200, 300)

# Recycle explicitly
df$summary <- rep(summary_values, length.out = nrow(df))

# Or extend with NA
df$summary <- c(summary_values, rep(NA, nrow(df) - length(summary_values)))
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use Merge/Join for Aggregates**

```{r}
# Original data
df <- data.frame(
  id = 1:20,
  group = rep(c("A", "B"), each = 10),
  value = rnorm(20)
)

# Aggregate
group_summary <- aggregate(value ~ group, df, mean)
names(group_summary)[2] <- "group_mean"

# Merge back
df <- merge(df, group_summary, by = "group")
head(df)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: dplyr Way (Cleaner)**

```{r}
library(dplyr)

df <- data.frame(
  id = 1:20,
  group = rep(c("A", "B"), each = 10),
  value = rnorm(20)
)

# Add group mean to each row
df <- df %>%
  group_by(group) %>%
  mutate(group_mean = mean(value)) %>%
  ungroup()

head(df)
```
</div>

## Error #3: `number of items to replace is not a multiple` {#replacement-multiple}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-dimension">📏 LENGTH</span>

### The Error

```{r error=TRUE}
x <- 1:10
x[1:7] <- c(100, 200)  # 7 positions, 2 values
```

<div class="error-box">
🔴 **ERROR**

```
Error in x[1:7] <- c(100, 200) : 
  number of items to replace is not a multiple of replacement length
```
</div>

### What It Means

You're replacing a subset, but the lengths don't match evenly (not multiples).

### When This Happens

```{r error=TRUE}
# Replacing 10 items with 3 values
x <- 1:10
x[] <- c(1, 2, 3)  # 10 is not a multiple of 3

# Replacing 7 items with 2 values
x[1:7] <- c(10, 20)  # 7 is not a multiple of 2
```

But these work:

```{r}
# Length 1 always works
x <- 1:10
x[1:7] <- 99
x

# Multiples work
x <- 1:10
x[1:6] <- c(10, 20, 30)  # 6 is multiple of 3
x
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Make Lengths Match**

```{r}
x <- 1:10

# Option A: Recycle explicitly
replacement <- rep(c(100, 200), length.out = 7)
x[1:7] <- replacement
x

# Option B: Subset to match
x <- 1:10
x[1:2] <- c(100, 200)  # Only replace 2
x
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use ifelse() for Conditional Replacement**

```{r}
x <- 1:10

# Replace first 7 with pattern
x <- ifelse(seq_along(x) <= 7, 
            rep(c(100, 200), length.out = length(x))[seq_along(x)], 
            x)
x
```
</div>

## Error #4: `replacement has length zero` {#replacement-zero}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-dimension">📏 LENGTH</span>

### The Error

```{r error=TRUE}
x <- 1:5
x[3] <- c()  # Empty vector!
```

<div class="error-box">
🔴 **ERROR**

```
Error in x[3] <- c() : replacement has length zero
```
</div>

### What It Means

You're trying to replace elements with an empty vector (length 0).

### Common Causes

#### Cause 1: Empty Filter Result

```{r error=TRUE}
df <- data.frame(x = 1:10, y = letters[1:10])

# Filter returns empty
subset_values <- df$x[df$y == "z"]  # No "z", returns numeric(0)

# Try to use for replacement
df$new[1:5] <- subset_values  # Error!
```

#### Cause 2: Function Returns Empty

```{r error=TRUE}
get_values <- function(condition) {
  if (condition) {
    return(1:5)
  } else {
    return(numeric(0))  # Oops!
  }
}

x <- 1:10
x[1:5] <- get_values(FALSE)  # Error!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Before Replacing**

```{r}
x <- 1:10
replacement <- numeric(0)  # Empty

# Check first
if (length(replacement) > 0) {
  x[1:length(replacement)] <- replacement
} else {
  message("No replacement values")
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use NA as Default**

```{r}
get_values_safe <- function(condition) {
  if (condition) {
    return(1:5)
  } else {
    return(NA)  # Or a default value
  }
}

x <- 1:10
x[1:5] <- get_values_safe(FALSE)  # Works, assigns NA
x
```
</div>

## Vectorization Best Practices

<div class="bestpractice-box">
🎯 **Best Practice: Length-Safe Operations**

```{r}
# 1. Check lengths match
operate_safely <- function(x, y, fun) {
  if (length(x) != length(y)) {
    stop("Vectors must be same length. Got ", 
         length(x), " and ", length(y))
  }
  fun(x, y)
}

# 2. Use recycling intentionally (scalars only)
add_scalar <- function(vec, scalar) {
  stopifnot(length(scalar) == 1)
  vec + scalar
}

# 3. Document recycling behavior
#' Add vectors with explicit recycling
#' @param x numeric vector
#' @param y numeric vector (will be recycled to length of x)
add_with_recycling <- function(x, y) {
  if (length(y) == 1) {
    return(x + y)  # Scalar - always OK
  }
  
  y_recycled <- rep(y, length.out = length(x))
  return(x + y_recycled)
}
```
</div>

## Understanding Vector Operations

<div class="insight-box">
💡 **Key Insight: What Gets Recycled**

```{r}
# Arithmetic operators
1:4 + c(10, 20)           # Addition
1:4 - c(10, 20)           # Subtraction
1:4 * c(2, 3)             # Multiplication
1:4 / c(2, 4)             # Division

# Logical operators
c(TRUE, FALSE) & c(TRUE, TRUE, FALSE, FALSE)
c(TRUE, FALSE) | c(FALSE, FALSE, TRUE, TRUE)

# Comparison operators
1:6 > c(2, 4, 6)          # Recycles both

# Assignment
x <- 1:12
x[] <- c(1, 2, 3)         # Recycles to 12
x
```

**Key point:** Recycling happens in MANY contexts!
</div>

## Edge Cases and Gotchas

### Gotcha #1: Matrix Recycling

```{r}
# Matrices recycle by column!
matrix(1:2, nrow = 3, ncol = 4)
```

Warning appears because 12 (3×4) is not multiple of 2.

### Gotcha #2: Data Frame Column Recycling

```{r}
# This works - length 1 always recycles
df <- data.frame(
  x = 1:5,
  y = 10  # Recycled to 5
)
df

# This works - multiple lengths
df <- data.frame(
  x = 1:6,
  y = c(1, 2)  # Recycled to 6
)
df
```

```{r error=TRUE}
# This fails - not a multiple
df <- data.frame(
  x = 1:5,
  y = c(1, 2)  # 5 is not multiple of 2
)
```

### Gotcha #3: List Operations Don't Recycle

```{r error=TRUE}
# Vectors recycle
c(1, 2) + c(10, 20, 30)  # Works (with warning)

# Lists don't
list(1, 2) + list(10, 20, 30)  # Error!
```

Lists need explicit handling:

```{r}
x <- list(1, 2, 3)
y <- list(10, 20)

# Use Map or mapply
Map(`+`, x, rep(y, length.out = length(x)))
```

## Debugging Recycling Issues

<div class="insight-box">
💡 **Debugging Checklist**

```{r}
# 1. Check lengths
x <- 1:10
y <- 1:7
length(x)
length(y)

# 2. Check if they're multiples
max(length(x), length(y)) %% min(length(x), length(y))
# 0 = clean multiple, anything else = partial recycling

# 3. Visualize recycling
rep(y, length.out = length(x))

# 4. Test operation
tryCatch(
  x + y,
  warning = function(w) {
    message("Warning caught: ", w$message)
  }
)

# 5. Check for unexpected conversions
class(x); typeof(x)
class(y); typeof(y)
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Recycling is automatic**: R repeats shorter vectors to match longer ones
2. **Warnings appear**: When lengths aren't multiples (except scalars)
3. **Scalars always work**: Length 1 recycles to any length
4. **Check before operating**: Use `length()` to verify matches
5. **Explicit is better**: Use `rep()` to show intent
6. **Data frames are strict**: Column lengths must match (or be length 1)
7. **Errors vs warnings**: Replacement operations error, arithmetic operations warn

**Quick Reference:**

| Situation | Behavior |
|-----------|----------|
| Same length | No recycling needed |
| One is length 1 | Silent recycling (scalar) |
| Lengths are multiples | Silent recycling (e.g., 2 and 6) |
| Lengths not multiples | Warning + recycling (e.g., 3 and 7) |
| Replacement, wrong length | Error (not multiples) |
| Replacement, length 0 | Error |
| Data frame column | Error if not length 1 or nrow |

**Prevention:**

```{r eval=FALSE}
# Always check
stopifnot(length(x) == length(y))

# Or use scalars only
stopifnot(length(y) == 1)

# Or recycle explicitly
y <- rep(y, length.out = length(x))
```

**Remember:** No warning doesn't mean correct! Multiples recycle silently.
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Predict the Outcome**

What will happen? Will it work, warn, or error?

```{r eval=FALSE}
# A
c(1, 2, 3, 4) + c(10, 20)

# B
c(1, 2, 3, 4, 5) + c(10, 20)

# C
df <- data.frame(x = 1:10)
df$y <- c(1, 2, 3, 4, 5)

# D
x <- 1:12
x[] <- c(1, 2, 3, 4)

# E
matrix(1:5, nrow = 5, ncol = 5)
```
</div>

<div class="exercise-box">
📝 **Exercise 2: Fix the Code**

Debug these recycling problems:

```{r eval=FALSE}
# Problem 1
students <- 1:25
groups <- c("A", "B", "C")
data.frame(student = students, group = groups)

# Problem 2
values <- rnorm(100)
weights <- c(1, 2, 3)
weighted <- values * weights

# Problem 3
df <- data.frame(id = 1:20)
summary_stats <- c(mean = 50, sd = 10, n = 20)
df$mean <- summary_stats["mean"]
```
</div>

<div class="exercise-box">
📝 **Exercise 3: Safe Operations**

Write a function `safe_add(x, y)` that:
1. Checks if lengths match
2. If not, asks user what to do:
   - Error
   - Recycle shorter
   - Trim longer
   - Extend with NA
3. Performs the operation
4. Returns result with attribute showing what was done
</div>

<div class="exercise-box">
📝 **Exercise 4: Real World**

You have exam scores for 100 students across 4 quarters:

```{r eval=FALSE}
scores_q1 <- rnorm(100, mean = 75, sd = 10)
scores_q2 <- rnorm(98, mean = 78, sd = 10)   # 2 students dropped
scores_q3 <- rnorm(102, mean = 80, sd = 10)  # 2 new students
scores_q4 <- rnorm(100, mean = 82, sd = 10)
```

Create a data frame with:
- All students who completed at least one quarter
- NA for missing scores
- Calculate average score per student
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
# A - Works, silent (4 is multiple of 2)
c(1, 2, 3, 4) + c(10, 20)

# B - Works, warns (5 not multiple of 2)
c(1, 2, 3, 4, 5) + c(10, 20)

# C - Errors (10 not multiple of 5)
tryCatch(
  data.frame(x = 1:10, y = c(1, 2, 3, 4, 5)),
  error = function(e) message("Error: ", e$message)
)

# D - Works, silent (12 is multiple of 4)
x <- 1:12
x[] <- c(1, 2, 3, 4)
x

# E - Works, silent (25 is multiple of 5)
matrix(1:5, nrow = 5, ncol = 5)
```

**Exercise 2:**

```{r}
# Problem 1 - Recycle groups explicitly
students <- 1:25
groups <- c("A", "B", "C")
data.frame(
  student = students, 
  group = rep(groups, length.out = length(students))
)

# Problem 2 - Make intention clear
values <- rnorm(100)
weights <- c(1, 2, 3)
weights_full <- rep(weights, length.out = length(values))
weighted <- values * weights_full

# Problem 3 - Extract scalar properly
df <- data.frame(id = 1:20)
summary_stats <- c(mean = 50, sd = 10, n = 20)
df$mean <- summary_stats[["mean"]]  # Single value
```

**Exercise 3:**

```{r}
safe_add <- function(x, y, action = c("error", "recycle", "trim", "extend")) {
  action <- match.arg(action)
  
  if (length(x) == length(y)) {
    result <- x + y
    attr(result, "action") <- "none_needed"
    return(result)
  }
  
  if (action == "error") {
    stop("Lengths don't match: ", length(x), " vs ", length(y))
  }
  
  if (action == "recycle") {
    max_len <- max(length(x), length(y))
    x <- rep(x, length.out = max_len)
    y <- rep(y, length.out = max_len)
    result <- x + y
    attr(result, "action") <- "recycled"
  }
  
  if (action == "trim") {
    min_len <- min(length(x), length(y))
    result <- x[1:min_len] + y[1:min_len]
    attr(result, "action") <- "trimmed"
  }
  
  if (action == "extend") {
    max_len <- max(length(x), length(y))
    x <- c(x, rep(NA, max_len - length(x)))
    y <- c(y, rep(NA, max_len - length(y)))
    result <- x + y
    attr(result, "action") <- "extended"
  }
  
  return(result)
}

# Test
safe_add(1:5, 1:3, "recycle")
```

**Exercise 4:**

```{r}
# Create scores with different lengths
set.seed(123)
scores_q1 <- rnorm(100, mean = 75, sd = 10)
scores_q2 <- rnorm(98, mean = 78, sd = 10)
scores_q3 <- rnorm(102, mean = 80, sd = 10)
scores_q4 <- rnorm(100, mean = 82, sd = 10)

# Find max number of students
max_students <- max(length(scores_q1), length(scores_q2), 
                   length(scores_q3), length(scores_q4))

# Extend all to max length with NA
extend_with_na <- function(x, target_len) {
  c(x, rep(NA, target_len - length(x)))
}

# Create data frame
df <- data.frame(
  student_id = 1:max_students,
  q1 = extend_with_na(scores_q1, max_students),
  q2 = extend_with_na(scores_q2, max_students),
  q3 = extend_with_na(scores_q3, max_students),
  q4 = extend_with_na(scores_q4, max_students)
)

# Calculate average (ignoring NAs)
df$average <- rowMeans(df[, c("q1", "q2", "q3", "q4")], na.rm = TRUE)

# Keep only students with at least one score
df <- df[!is.nan(df$average), ]

head(df)
```
</details>
