# NA, NULL, NaN, Inf {#na-null-nan-inf}

<div class="chapter-summary">
**What You'll Learn:**

- The four "missing" value types in R
- Critical differences between NA, NULL, NaN, and Inf
- Common errors with missing data
- How to handle missing values properly
- Testing and detecting special values

**Key Errors Covered:** 15+ missing data errors

**Difficulty:** ⭐ Beginner to ⭐⭐ Intermediate
</div>

## Introduction

R has **four special "non-values"** that trip up everyone:

```{r}
NA    # Not Available
NULL  # Nothing
NaN   # Not a Number
Inf   # Infinity
```

They look similar but behave very differently, and confusing them causes endless errors.

<div class="insight-box">
💡 **Key Insight: The Four Special Values**

```{r}
# NA - Missing data (most common)
ages <- c(25, 30, NA, 35)  # One age is missing

# NULL - Absence of value (empty)
result <- NULL  # No result yet

# NaN - Invalid math result
0 / 0  # Undefined

# Inf - Infinite value
1 / 0  # Positive infinity
-1 / 0 # Negative infinity
```

**Think of it this way:**
- **NA**: "I don't know what this is"
- **NULL**: "There's nothing here"
- **NaN**: "This calculation doesn't make sense"
- **Inf**: "This is beyond measurement"
</div>

## NA: Not Available

### What is NA?

```{r}
# NA is a placeholder for missing data
x <- c(1, 2, NA, 4, 5)
x

# It has a type!
typeof(NA)              # logical
typeof(NA_real_)        # double
typeof(NA_integer_)     # integer
typeof(NA_character_)   # character
```

## Error #1: `missing values where TRUE/FALSE needed` {#na-logical}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
x <- NA
if (x > 5) {
  print("Large")
}
```

<div class="error-box">
🔴 **ERROR**

```
Error in if (x > 5) { : missing value where TRUE/FALSE needed
```
</div>

### What It Means

`if()` needs TRUE or FALSE, but got NA. R won't guess which branch to take.

### Common Causes

#### Cause 1: Comparison with NA

```{r}
x <- c(1, 2, NA, 4, 5)

# Comparisons with NA give NA
x > 3

# Using in if()
```

```{r error=TRUE}
if (x[3] > 3) {  # x[3] is NA
  print("yes")
}
```

#### Cause 2: Missing Data in Condition

```{r error=TRUE}
df <- data.frame(
  age = c(25, NA, 35),
  name = c("Alice", "Bob", "Charlie")
)

# This creates NA in condition
if (df$age[2] > 30) {
  print("Senior")
}
```

#### Cause 3: Function Returns NA

```{r error=TRUE}
get_value <- function() {
  return(NA)
}

if (get_value() == 10) {
  print("Ten!")
}
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Test for NA First**

```{r}
x <- NA

# Check for NA before using
if (!is.na(x) && x > 5) {
  print("Large")
} else {
  print("Not large or is NA")
}

# Or handle NA explicitly
if (is.na(x)) {
  print("Missing value")
} else if (x > 5) {
  print("Large")
} else {
  print("Small")
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use isTRUE()**

```{r}
x <- NA

# isTRUE() converts NA to FALSE
if (isTRUE(x > 5)) {
  print("Large")
}

# Equivalent to:
if (!is.na(x) && x > 5) {
  print("Large")
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use %in% for Comparisons**

```{r}
x <- NA

# == with NA gives NA
x == 5  # NA

# %in% handles NA better
x %in% c(5)  # FALSE

# But still need to check:
5 %in% c(NA, 5)  # TRUE
NA %in% c(NA, 5) # TRUE
```
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall: && vs &**

```{r}
x <- c(NA, 2, 3)

# Single & returns vector with NA
x > 1 & x < 5

# Double && errors on vector
```

```{r error=TRUE}
if (x > 1 && x < 5) {  # Error!
  print("yes")
}
```

**For if():** Use `&&` but check for NA first

**For vectorized ops:** Use `&` and handle NAs appropriately
</div>

## Error #2: `missing values and NaN's not allowed` {#na-not-allowed}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-math">🧮 MATH</span>

### The Error

```{r error=TRUE}
x <- c(1, 2, NA, 4, 5)
var(x)  # Without na.rm
```

<div class="error-box">
🔴 **ERROR**

```
Error in var(x) : missing values and NaN's not allowed if 'na.rm' is FALSE
```
</div>

### What It Means

Some functions refuse to work with NA unless you explicitly tell them how to handle it.

### Common Functions With This Error

```{r error=TRUE}
x <- c(1, 2, NA, 4, 5)

# These error without na.rm:
var(x)
sd(x)
cov(x, x)
cor(x, x)
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use na.rm = TRUE**

```{r}
x <- c(1, 2, NA, 4, 5)

# Remove NAs for calculation
var(x, na.rm = TRUE)
sd(x, na.rm = TRUE)
mean(x, na.rm = TRUE)
sum(x, na.rm = TRUE)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Remove NAs First**

```{r}
x <- c(1, 2, NA, 4, 5)

# Complete cases only
x_complete <- x[!is.na(x)]
var(x_complete)

# Or use na.omit()
x_complete <- na.omit(x)
var(x_complete)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Replace NAs**

```{r}
x <- c(1, 2, NA, 4, 5)

# Replace with mean
x_filled <- x
x_filled[is.na(x_filled)] <- mean(x, na.rm = TRUE)
var(x_filled)

# Replace with median (more robust)
x_filled <- x
x_filled[is.na(x_filled)] <- median(x, na.rm = TRUE)
var(x_filled)
```
</div>

## Error #3: `'x' contains missing values` {#x-contains-na}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-math">🧮 MATH</span>

### The Error

```{r error=TRUE}
x <- c(1, 2, NA, 4)
y <- c(2, 3, 4, 5)
t.test(x, y)
```

<div class="error-box">
🔴 **ERROR**

```
Error in t.test.default(x, y) : 'x' contains missing values
```
</div>

### What It Means

Statistical tests often require complete data and won't work with NAs.

### Common Functions

```{r error=TRUE}
x <- c(1, 2, NA, 4, 5)
y <- c(2, 3, 4, 5, 6)

# These need complete data:
t.test(x, y)
wilcox.test(x, y)
chisq.test(x, y)
```

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Remove NAs from both vectors:**
```{r}
x <- c(1, 2, NA, 4, 5)
y <- c(2, 3, 4, 5, 6)

# Find complete cases
complete <- complete.cases(x, y)
x_clean <- x[complete]
y_clean <- y[complete]

t.test(x_clean, y_clean)
```

**2. Use na.action:**
```{r}
# For functions that support it
df <- data.frame(x = c(1, 2, NA, 4), y = c(2, 3, 4, 5))
t.test(x ~ 1, data = df, na.action = na.omit)
```

**3. Impute missing values (advanced):**
```{r eval=FALSE}
library(mice)
imputed <- mice(data, m = 5, method = "pmm")
```
</div>

## NULL: The Absence of Value

### What is NULL?

```{r}
# NULL means "nothing"
x <- NULL
x

# Length zero
length(NULL)

# NULL disappears in vectors
c(1, 2, NULL, 4)

# Different from NA
c(1, 2, NA, 4)
```

## Error #4: `argument is of length zero` {#length-zero}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-dimension">📏 LENGTH</span>

### The Error

```{r error=TRUE}
x <- NULL
if (x > 5) {
  print("yes")
}
```

<div class="error-box">
🔴 **ERROR**

```
Error in if (x > 5) { : argument is of length zero
```
</div>

### What It Means

You're trying to use NULL in a context that needs at least one value.

### Common Causes

#### Cause 1: Uninitialized Variable

```{r error=TRUE}
# Forgot to assign
result <- NULL

# Try to use
if (result == 100) {
  print("success")
}
```

#### Cause 2: Empty Subset

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10)

# Filter returns nothing
subset_val <- df$x[df$x > 10]  # numeric(0)
subset_val <- NULL  # Or explicitly NULL

# Try to use
if (subset_val > 0) {
  print("positive")
}
```

#### Cause 3: Function Returns NULL

```{r error=TRUE}
get_value <- function(x) {
  if (x > 10) {
    return(x * 2)
  }
  # Implicitly returns NULL when x <= 10
}

result <- get_value(5)  # NULL

if (result > 0) {
  print("positive")
}
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check for NULL**

```{r}
x <- NULL

# Check before using
if (!is.null(x) && x > 5) {
  print("Large")
} else {
  print("NULL or not large")
}

# Or use default
if (is.null(x)) {
  x <- 0  # Default value
}

if (x > 5) {
  print("Large")
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check Length**

```{r}
x <- NULL

# Check length instead
if (length(x) > 0 && x > 5) {
  print("Large")
}

# Works for both NULL and empty vectors
y <- numeric(0)
if (length(y) > 0 && y[1] > 5) {
  print("Large")
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use Default Values**

```{r}
get_value_safe <- function(x) {
  if (x > 10) {
    return(x * 2)
  } else {
    return(0)  # Default instead of NULL
  }
}

result <- get_value_safe(5)
if (result > 0) {
  print("positive")
}
```
</div>

## NaN: Not a Number

### What is NaN?

```{r}
# Result of invalid math
0 / 0
Inf - Inf
Inf / Inf

# NaN is a special type of NA
is.na(NaN)   # TRUE
is.nan(NA)   # FALSE
```

## Error #5: `NaNs produced` {#nans-produced}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-math">🧮 MATH</span>

### The Warning

```{r}
sqrt(-1)
log(-1)
0/0
```

<div class="warning-box">
🟡 **WARNING**

```
Warning message:
In sqrt(-1) : NaNs produced
```
</div>

### What It Means

You performed a mathematically invalid operation. Result is NaN.

### Common Causes

#### Cause 1: Negative Square Root

```{r}
# Real square root of negative number doesn't exist
sqrt(-1)
sqrt(c(1, 4, -9, 16))
```

#### Cause 2: Log of Negative/Zero

```{r}
# Log of negative or zero
log(0)   # -Inf
log(-1)  # NaN

# Log of vector with negatives
values <- c(1, 10, -5, 100)
log(values)
```

#### Cause 3: Invalid Division

```{r}
# 0/0 is undefined
0 / 0

# Inf operations
Inf - Inf
Inf / Inf
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Input Before Operation**

```{r}
safe_sqrt <- function(x) {
  if (any(x < 0, na.rm = TRUE)) {
    warning("Negative values found, returning NA for those")
  }
  result <- sqrt(x)
  return(result)
}

safe_sqrt(c(1, 4, -9, 16))
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Handle NaN After Operation**

```{r}
x <- c(1, -4, 9, -16)
result <- sqrt(x)

# Replace NaN with NA (more standard)
result[is.nan(result)] <- NA
result

# Or with a specific value
result <- sqrt(x)
result[is.nan(result)] <- 0
result
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use Complex Numbers for Sqrt**

```{r}
# For genuinely needing complex results
x <- -1
sqrt(as.complex(x))  # 0+1i

# Vector
x <- c(1, -4, 9, -16)
sqrt(as.complex(x))
```
</div>

## Inf: Infinity

### What is Inf?

```{r}
# Division by zero (non-zero)
1 / 0   # Inf
-1 / 0  # -Inf

# Very large calculations
exp(1000)  # Inf

# Inf in comparisons
Inf > 1e100  # TRUE
-Inf < -1e100  # TRUE
```

## Error #6: `infinite or missing values in 'x'` {#inf-in-x}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-math">🧮 MATH</span>

### The Error

```{r error=TRUE}
x <- c(1, 2, Inf, 4)
y <- c(2, 3, 4, 5)
cor(x, y)
```

<div class="error-box">
🔴 **ERROR**

```
Error in cor(x, y) : infinite or missing values in 'x'
```
</div>

### What It Means

Some functions can't handle infinite values and need finite numbers.

### Solutions

<div class="solution-box">
✅ **SOLUTIONS**

**1. Check for and remove Inf:**
```{r}
x <- c(1, 2, Inf, 4, -Inf, 5)

# Find finite values
is.finite(x)

# Keep only finite
x_finite <- x[is.finite(x)]
y_finite <- y[is.finite(x)]
cor(x_finite, y_finite)
```

**2. Replace Inf with large number:**
```{r}
x <- c(1, 2, Inf, 4, -Inf, 5)

# Replace Inf with max/min of finite values
x_fixed <- x
x_fixed[x == Inf] <- max(x[is.finite(x)]) * 10
x_fixed[x == -Inf] <- min(x[is.finite(x)]) * 10
x_fixed
```

**3. Check before calculation:**
```{r}
safe_cor <- function(x, y) {
  finite_both <- is.finite(x) & is.finite(y)
  
  if (sum(finite_both) < 2) {
    stop("Need at least 2 finite pairs")
  }
  
  cor(x[finite_both], y[finite_both])
}
```
</div>

## Testing for Special Values

<div class="bestpractice-box">
🎯 **Best Practice: Comprehensive Testing**

```{r}
# Test functions
x <- c(1, NA, NaN, Inf, -Inf, 0)

# Individual tests
is.na(x)       # TRUE for NA and NaN
is.nan(x)      # TRUE only for NaN
is.infinite(x) # TRUE for Inf and -Inf
is.finite(x)   # TRUE for normal numbers

# NULL is different
y <- NULL
is.null(y)     # TRUE
is.na(y)       # logical(0) - no value to test

# Combined checks
is_valid <- function(x) {
  !is.na(x) & !is.nan(x) & is.finite(x)
}

is_valid(x)
```

**Decision Tree:**
```{r eval=FALSE}
if (is.null(x)) {
  # Handle NULL - no value at all
} else if (is.nan(x)) {
  # Handle NaN - invalid math result
} else if (is.infinite(x)) {
  # Handle Inf/-Inf - beyond limits
} else if (is.na(x)) {
  # Handle NA - missing data
} else {
  # Normal value
}
```
</div>

## Handling Missing Data Strategies

<div class="insight-box">
💡 **Key Insight: Missing Data Strategies**

**1. Complete Case Analysis (Listwise Deletion)**
```{r}
df <- data.frame(
  x = c(1, 2, NA, 4, 5),
  y = c(10, NA, 30, 40, 50)
)

# Keep only complete rows
complete.cases(df)
df_complete <- df[complete.cases(df), ]
df_complete
```

**2. Available Case Analysis (Pairwise Deletion)**
```{r}
# Use all available data for each calculation
cor(df, use = "pairwise.complete.obs")
```

**3. Imputation (Replacing with Estimates)**
```{r}
# Replace with mean (simple)
df$x[is.na(df$x)] <- mean(df$x, na.rm = TRUE)

# Replace with median (robust to outliers)
df$y[is.na(df$y)] <- median(df$y, na.rm = TRUE)
```

**4. Keep as NA (Most Honest)**
```{r}
# Just handle NAs in analysis
mean(df$x, na.rm = TRUE)
```

**When to use each:**
- **Complete case**: When data missing completely at random (MCAR)
- **Pairwise**: When you want to use all available information
- **Imputation**: When you have good reason to estimate missing values
- **Keep NA**: When missingness is informative
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Four special values**: NA (missing), NULL (nothing), NaN (invalid math), Inf (infinite)
2. **NA vs NULL**: NA is a placeholder in a vector, NULL is absence of vector
3. **Test before using**: Always check `is.na()`, `is.null()`, `is.finite()`
4. **Use isTRUE()**: For conditions that might be NA
5. **na.rm = TRUE**: Most statistical functions need this with NAs
6. **NaN from invalid math**: sqrt(-1), 0/0, etc.
7. **Inf from overflow**: 1/0, exp(1000), etc.

**Quick Reference:**

| Value | Test | Meaning | Example |
|-------|------|---------|---------|
| NA | `is.na()` | Missing data | Survey non-response |
| NULL | `is.null()` | No value | Uninitialized variable |
| NaN | `is.nan()` | Invalid math | 0/0 |
| Inf | `is.infinite()` | Infinite | 1/0 |

**In if() statements:**

```{r eval=FALSE}
# ❌ Dangerous
if (x > 5) { }  # Errors if x is NA or NULL

# ✅ Safe
if (!is.na(x) && x > 5) { }
if (isTRUE(x > 5)) { }
if (length(x) > 0 && !is.na(x) && x > 5) { }
```

**With functions:**

```{r eval=FALSE}
# ❌ May error
mean(x)
var(x)
cor(x, y)

# ✅ Handle missing
mean(x, na.rm = TRUE)
var(x[is.finite(x)])
cor(x, y, use = "complete.obs")
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Identify the Type**

What are these and why?

```{r eval=FALSE}
# A
x <- c()
typeof(x)

# B  
y <- sqrt(-1)
class(y)

# C
z <- 1/0
is.finite(z)

# D
w <- c(1, 2, NA, 4)
w == NA
```
</div>

<div class="exercise-box">
📝 **Exercise 2: Fix the Code**

Debug these:

```{r eval=FALSE}
# Problem 1
data <- c(1, 2, NA, 4, 5)
if (mean(data) > 3) {
  print("High average")
}

# Problem 2
get_score <- function(x) {
  if (x > 10) {
    return(x * 2)
  }
}
score <- get_score(5)
if (score > 10) {
  print("High score")
}

# Problem 3
values <- c(10, 20, 0, 30)
log_values <- log(values)
mean(log_values)
```
</div>

<div class="exercise-box">
📝 **Exercise 3: Robust Function**

Write `robust_mean(x)` that:
1. Handles NA, NULL, NaN, Inf
2. Reports how many of each were found
3. Calculates mean of valid values
4. Returns list with mean and diagnostics
</div>

<div class="exercise-box">
📝 **Exercise 4: Data Cleaning**

You have survey data:

```{r eval=FALSE}
survey <- data.frame(
  age = c(25, NA, 35, -999, 40),  # -999 = missing
  income = c(50000, 75000, 0, 80000, NA),  # 0 = refused
  satisfaction = c(5, 3, NA, 4, 99)  # 99 = invalid
)
```

Clean it:
1. Convert -999 to NA
2. Convert 0 in income to NA
3. Convert 99 in satisfaction to NA
4. Calculate complete case statistics
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
# A - Empty vector
x <- c()
typeof(x)  # "logical" (default empty type)

# B - NaN from invalid math
y <- sqrt(-1)
class(y)  # "numeric"
is.nan(y)  # TRUE

# C - Inf from division by zero
z <- 1/0
is.finite(z)  # FALSE (Inf is not finite)

# D - Comparison with NA gives NA
w <- c(1, 2, NA, 4)
w == NA  # All NA! Use is.na() instead
is.na(w)  # Correct way
```

**Exercise 2:**

```{r}
# Problem 1 - Need na.rm
data <- c(1, 2, NA, 4, 5)
mean_val <- mean(data, na.rm = TRUE)
if (!is.na(mean_val) && mean_val > 3) {
  print("High average")
}

# Problem 2 - Check for NULL
get_score <- function(x) {
  if (x > 10) {
    return(x * 2)
  } else {
    return(0)  # Return 0 instead of NULL
  }
}
score <- get_score(5)
if (!is.null(score) && score > 10) {
  print("High score")
}

# Problem 3 - Handle -Inf from log(0)
values <- c(10, 20, 0, 30)
log_values <- log(values)
log_values[is.infinite(log_values)] <- NA
mean(log_values, na.rm = TRUE)
```

**Exercise 3:**

```{r}
robust_mean <- function(x) {
  # Initialize diagnostics
  diagnostics <- list(
    total = length(x),
    null = is.null(x),
    na = 0,
    nan = 0,
    inf = 0,
    valid = 0
  )
  
  # Check for NULL
  if (is.null(x)) {
    return(list(mean = NULL, diagnostics = diagnostics))
  }
  
  # Count special values
  diagnostics$na <- sum(is.na(x) & !is.nan(x))
  diagnostics$nan <- sum(is.nan(x))
  diagnostics$inf <- sum(is.infinite(x))
  
  # Find valid values
  valid <- x[!is.na(x) & !is.nan(x) & is.finite(x)]
  diagnostics$valid <- length(valid)
  
  # Calculate mean
  if (length(valid) == 0) {
    mean_val <- NA
  } else {
    mean_val <- mean(valid)
  }
  
  # Return
  list(
    mean = mean_val,
    diagnostics = diagnostics
  )
}

# Test
robust_mean(c(1, 2, NA, NaN, Inf, 5, -Inf))
```

**Exercise 4:**

```{r}
survey <- data.frame(
  age = c(25, NA, 35, -999, 40),
  income = c(50000, 75000, 0, 80000, NA),
  satisfaction = c(5, 3, NA, 4, 99)
)

# Clean data
clean_survey <- survey

# Convert -999 to NA in age
clean_survey$age[clean_survey$age == -999] <- NA

# Convert 0 to NA in income
clean_survey$income[clean_survey$income == 0] <- NA

# Convert 99 to NA in satisfaction
clean_survey$satisfaction[clean_survey$satisfaction == 99] <- NA

# Complete case analysis
clean_survey_complete <- clean_survey[complete.cases(clean_survey), ]

# Statistics
list(
  n_complete = nrow(clean_survey_complete),
  mean_age = mean(clean_survey_complete$age),
  mean_income = mean(clean_survey_complete$income),
  mean_satisfaction = mean(clean_survey_complete$satisfaction)
)
```
</details>
