# Part III: Indexing & Subsetting {-}

# Subscript & Dimension Errors {#subscript-dimension}

<div class="chapter-summary">
**What You'll Learn:**

- How R's indexing system works
- Understanding dimensions and subscripts
- Common indexing mistakes
- Negative vs positive indexing
- Logical indexing pitfalls
- Matrix and array subsetting

**Key Errors Covered:** 18+ indexing errors

**Difficulty:** ⭐ Beginner to ⭐⭐ Intermediate
</div>

## Introduction

R's indexing system is powerful but confusing. You'll see these errors constantly:

```{r error=TRUE}
x <- 1:5
x[10]  # Wait, this works?
```

```{r}
x[10]  # Returns NA, not an error!
```

But this doesn't:

```{r error=TRUE}
x[[10]]  # Now it's an error
```

Understanding R's indexing is **critical** for avoiding errors. Let's master it.

## R's Indexing Methods

<div class="insight-box">
💡 **Key Insight: Five Ways to Index**

```{r}
x <- c(10, 20, 30, 40, 50)

# 1. Positive integers (positions)
x[c(1, 3, 5)]

# 2. Negative integers (exclusion)
x[-c(2, 4)]

# 3. Logical vectors
x[c(TRUE, FALSE, TRUE, FALSE, TRUE)]

# 4. Names (if vector has names)
names(x) <- c("a", "b", "c", "d", "e")
x[c("a", "c", "e")]

# 5. Empty (returns all)
x[]
```

**Each method has different error patterns!**
</div>

## Error #1: `subscript out of bounds` {#subscript-out-bounds}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-dimension">📏 DIMENSION</span>

### The Error

```{r error=TRUE}
x <- 1:5
x[[10]]  # Double bracket
```

<div class="error-box">
🔴 **ERROR**

```
Error in x[[10]] : subscript out of bounds
```
</div>

### What It Means

You're trying to access an element beyond the vector's length using `[[]]`.

### Single vs Double Bracket

```{r}
x <- 1:5

# Single bracket: Returns NA, no error
x[10]

# Double bracket: Errors
```

```{r error=TRUE}
x[[10]]
```

**Why the difference?**
- `[` can return multiple elements or NA
- `[[` must return exactly one element

### Common Causes

#### Cause 1: Off-by-One Error

```{r error=TRUE}
scores <- c(85, 90, 95)

# Loop goes too far
for (i in 1:4) {  # Only 3 elements!
  print(scores[[i]])
}
```

#### Cause 2: Wrong Length Assumption

```{r error=TRUE}
data <- c(10, 20, 30)

# Assumed it had 5 elements
first_five <- data[[1:5]]  # Error on 4th
```

#### Cause 3: After Filtering

```{r error=TRUE}
values <- 1:10
large_values <- values[values > 100]  # Empty!

# Try to access first element
large_values[[1]]  # Out of bounds (length 0)
```

#### Cause 4: List Indexing

```{r error=TRUE}
my_list <- list(a = 1, b = 2, c = 3)

# Trying to access 4th element
my_list[[4]]  # Only 3 elements
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Length First**

```{r}
x <- 1:5
index <- 10

# Safe access
if (index <= length(x)) {
  x[[index]]
} else {
  message("Index ", index, " is out of bounds")
  NA
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use Single Bracket**

```{r}
x <- 1:5

# Single bracket returns NA instead of error
x[10]  # NA

# Good for loops where you want to continue
for (i in 1:10) {
  val <- x[i]
  if (!is.na(val)) {
    print(val)
  }
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Safe Indexing Function**

```{r}
safe_extract <- function(x, i, default = NA) {
  if (i < 1 || i > length(x)) {
    return(default)
  }
  return(x[[i]])
}

# Test
x <- 1:5
safe_extract(x, 3)   # 3
safe_extract(x, 10)  # NA
safe_extract(x, 10, default = 0)  # 0
```
</div>

<div class="solution-box">
✅ **SOLUTION 4: Use seq_along() in Loops**

```{r}
values <- c(10, 20, 30)

# Wrong: assumes length
for (i in 1:5) {
  # Error on 4th iteration
}

# Right: uses actual length
for (i in seq_along(values)) {
  print(values[[i]])  # Safe
}

# Even safer: iterate over values directly
for (val in values) {
  print(val)
}
```
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall: Empty Vectors**

```{r error=TRUE}
# Filter returns empty
x <- 1:10
big_numbers <- x[x > 100]  # numeric(0)

length(big_numbers)  # 0

# This errors!
big_numbers[[1]]

# Always check
if (length(big_numbers) > 0) {
  big_numbers[[1]]
} else {
  NA
}
```
</div>

## Error #2: `undefined columns selected` {#undefined-columns}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-dimension">📏 DIMENSION</span>

### The Error

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10)
df[, "z"]  # Column doesn't exist
```

<div class="error-box">
🔴 **ERROR**

```
Error in `[.data.frame`(df, , "z") : undefined columns selected
```
</div>

### What It Means

You're trying to select columns that don't exist in the data frame.

### Common Causes

#### Cause 1: Typo in Column Name

```{r error=TRUE}
df <- data.frame(age = c(25, 30, 35), name = c("A", "B", "C"))

# Typo: "agee" instead of "age"
df[, "agee"]
```

#### Cause 2: Case Sensitivity

```{r error=TRUE}
df <- data.frame(Age = c(25, 30, 35))

# Wrong case
df[, "age"]  # Error! It's "Age" not "age"
```

#### Cause 3: Column Doesn't Exist Yet

```{r error=TRUE}
df <- data.frame(x = 1:5)

# Trying to select column before creating it
df[, c("x", "y")]  # "y" doesn't exist
```

#### Cause 4: After Subsetting

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10, z = 11:15)

# Select some columns
df_subset <- df[, c("x", "y")]

# Try to access z (no longer exists)
df_subset[, "z"]
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Column Names**

```{r}
df <- data.frame(age = c(25, 30, 35), name = c("A", "B", "C"))

# List all columns
names(df)
colnames(df)

# Check if column exists
"age" %in% names(df)  # TRUE
"agee" %in% names(df) # FALSE

# Safe selection
col_name <- "age"
if (col_name %in% names(df)) {
  df[, col_name]
} else {
  message("Column ", col_name, " not found")
  NULL
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use exists in dplyr**

```{r}
library(dplyr)

df <- data.frame(x = 1:5, y = 6:10)

# Select only existing columns
cols_to_select <- c("x", "z", "y")
existing_cols <- cols_to_select[cols_to_select %in% names(df)]

df %>% select(all_of(existing_cols))
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Safe Column Selection Function**

```{r}
safe_select_cols <- function(df, cols) {
  # Check which columns exist
  existing <- cols[cols %in% names(df)]
  missing <- cols[!cols %in% names(df)]
  
  if (length(missing) > 0) {
    warning("Columns not found: ", paste(missing, collapse = ", "))
  }
  
  if (length(existing) == 0) {
    return(data.frame())  # Empty data frame
  }
  
  return(df[, existing, drop = FALSE])
}

# Test
df <- data.frame(x = 1:5, y = 6:10)
safe_select_cols(df, c("x", "z", "y"))
```
</div>

## Error #3: `incorrect number of dimensions` {#incorrect-dimensions}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-dimension">📏 DIMENSION</span>

### The Error

```{r error=TRUE}
x <- 1:5  # Vector (1D)
x[1, 2]   # Using 2D indexing on 1D object
```

<div class="error-box">
🔴 **ERROR**

```
Error in x[1, 2] : incorrect number of dimensions
```
</div>

### What It Means

You're using the wrong number of indices for the object's dimensions.

### Understanding Dimensions

```{r}
# Vector: 1 dimension
vec <- 1:5
length(vec)
dim(vec)  # NULL

# Matrix: 2 dimensions
mat <- matrix(1:6, nrow = 2, ncol = 3)
dim(mat)  # 2 3

# Array: 3+ dimensions
arr <- array(1:24, dim = c(2, 3, 4))
dim(arr)  # 2 3 4

# Data frame: 2 dimensions (special)
df <- data.frame(x = 1:3, y = 4:6)
dim(df)  # 3 2
```

### Common Causes

#### Cause 1: Treating Vector as Matrix

```{r error=TRUE}
x <- c(10, 20, 30, 40, 50)

# Vector needs 1 index
x[3]  # Correct

# Not 2 indices
x[1, 3]  # Error!
```

#### Cause 2: Treating Matrix as Vector

```{r error=TRUE}
mat <- matrix(1:6, nrow = 2, ncol = 3)

# Matrix needs 2 indices
mat[1, 2]  # Correct

# Or can use 1 index (treats as vector)
mat[5]  # Also works! (column-major order)

# But this is confusing:
mat[1]  # First element, not first row
```

#### Cause 3: After Subsetting

```{r error=TRUE}
mat <- matrix(1:12, nrow = 3, ncol = 4)

# Extract one column (becomes vector!)
col1 <- mat[, 1]
class(col1)  # "numeric" (not matrix)

# Now 1D, can't use 2D indexing
col1[1, 1]  # Error!
```

#### Cause 4: List vs Data Frame Confusion

```{r error=TRUE}
# List: 1D (use single bracket or [[]])
my_list <- list(a = 1:3, b = 4:6)
my_list[[1]]     # Correct
my_list[1, 2]    # Error!

# Data frame: 2D (use row, col)
df <- data.frame(a = 1:3, b = 4:6)
df[1, 2]         # Correct
df[[1]]          # Also works (returns column)
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Dimensions Before Indexing**

```{r}
x <- 1:5

# Check what you have
ndims <- length(dim(x))  # 0 for vector

if (is.null(dim(x))) {
  # Vector: use 1 index
  x[3]
} else if (length(dim(x)) == 2) {
  # Matrix/data frame: use 2 indices
  x[1, 3]
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Preserve Dimensions with drop = FALSE**

```{r}
mat <- matrix(1:12, nrow = 3, ncol = 4)

# Default: drops to vector
col1 <- mat[, 1]
class(col1)  # "numeric"

# Preserve matrix structure
col1 <- mat[, 1, drop = FALSE]
class(col1)  # "matrix"
dim(col1)    # 3 1

# Now can still use 2D indexing
col1[1, 1]
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use Appropriate Functions**

```{r}
mat <- matrix(1:12, nrow = 3, ncol = 4)

# For vectors: use vector operations
vec <- 1:5
vec[3]
vec[c(1, 3, 5)]

# For matrices: use matrix operations
mat[1, ]      # First row
mat[, 2]      # Second column
mat[1:2, 3:4] # Submatrix

# For data frames: mix of both
df <- data.frame(x = 1:5, y = 6:10)
df[, "x"]     # Column (becomes vector)
df[, "x", drop = FALSE]  # Column (stays data frame)
df$x          # Column (vector)
```
</div>

## Error #4: `incorrect number of subscripts on matrix` {#incorrect-subscripts-matrix}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-dimension">📏 DIMENSION</span>

### The Error

```{r error=TRUE}
mat <- matrix(1:6, nrow = 2, ncol = 3)
mat[1, 2, 3]  # Too many indices!
```

<div class="error-box">
🔴 **ERROR**

```
Error in mat[1, 2, 3] : incorrect number of subscripts on matrix
```
</div>

### What It Means

Matrix needs exactly 2 indices (or 1), but you provided a different number.

### Correct Matrix Indexing

```{r}
mat <- matrix(1:6, nrow = 2, ncol = 3)
mat

# Correct ways:
mat[1, 2]      # Single element
mat[1, ]       # Entire row
mat[, 2]       # Entire column
mat[1:2, 2:3]  # Submatrix

# Also works (treats as vector):
mat[5]         # 5th element (column-major)

# Wrong:
# mat[1, 2, 3]  # Too many indices
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use Correct Number of Indices**

```{r}
mat <- matrix(1:6, nrow = 2, ncol = 3)

# For matrix: [row, col]
mat[1, 2]

# For array: [dim1, dim2, dim3, ...]
arr <- array(1:24, dim = c(2, 3, 4))
arr[1, 2, 3]
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check Object Type First**

```{r}
check_and_subset <- function(x, ...) {
  indices <- list(...)
  
  if (is.matrix(x)) {
    if (length(indices) > 2) {
      stop("Matrix needs 1 or 2 indices, got ", length(indices))
    }
    return(do.call(`[`, c(list(x), indices)))
  } else if (is.array(x)) {
    expected <- length(dim(x))
    if (length(indices) != expected) {
      stop("Array needs ", expected, " indices, got ", length(indices))
    }
    return(do.call(`[`, c(list(x), indices)))
  } else {
    return(x[[indices[[1]]]])
  }
}
```
</div>

## Error #5: `only 0's may be mixed with negative subscripts` {#zero-negative-mix}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-syntax">🔤 SYNTAX</span>

### The Error

```{r error=TRUE}
x <- 1:10
x[c(-1, 5)]  # Can't mix negative and positive!
```

<div class="error-box">
🔴 **ERROR**

```
Error in x[c(-1, 5)] : only 0's may be mixed with negative subscripts
```
</div>

### What It Means

R won't let you mix negative indices (exclusion) with positive indices (selection).

### Negative Indexing Rules

```{r}
x <- 1:10

# Positive: select elements
x[c(1, 3, 5)]

# Negative: exclude elements
x[-c(1, 3, 5)]

# Zero: ignored
x[c(0, 1, 3)]  # Same as x[c(1, 3)]

# Can mix zero with negative
x[c(0, -1, -3)]  # Same as x[-c(1, 3)]
```

```{r error=TRUE}
# CANNOT mix positive and negative
x[c(-1, 5)]  # Error!
x[c(1, -5)]  # Error!
```

### Why This Rule?

```{r eval=FALSE}
# Ambiguous meaning:
x[c(-1, 5)]

# Does this mean:
# "Select 5th, excluding 1st"?
# "Exclude 1st, but also select 5th"?

# R refuses to guess!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use Only Positive or Only Negative**

```{r}
x <- 1:10

# Want elements 2-10 (exclude 1st)?
# Use negative:
x[-1]

# Want elements except 1 and 3?
# Use negative:
x[-c(1, 3)]

# Want only 1, 3, 5?
# Use positive:
x[c(1, 3, 5)]
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Convert to Logical**

```{r}
x <- 1:10

# Want: "not 1, not 3, but yes 5"
# Create logical vector
indices <- rep(TRUE, length(x))
indices[c(1, 3)] <- FALSE  # Exclude these
indices[5] <- TRUE          # Include this (already TRUE)

x[indices]
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use setdiff()**

```{r}
x <- 1:10

# Want all except positions 1 and 3
exclude <- c(1, 3)
keep <- setdiff(seq_along(x), exclude)
x[keep]

# More complex: all except 1 and 3, but must include 5
exclude <- c(1, 3)
include <- 5
keep <- union(setdiff(seq_along(x), exclude), include)
keep <- sort(unique(keep))
x[keep]
```
</div>

## Error #6: `negative length vectors are not allowed` {#negative-length}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-dimension">📏 LENGTH</span>

### The Error

```{r error=TRUE}
n <- -5
x <- numeric(n)  # Can't have negative length!
```

<div class="error-box">
🔴 **ERROR**

```
Error in numeric(n) : negative length vectors are not allowed
```
</div>

### What It Means

You're trying to create a vector with negative length, which is impossible.

### Common Causes

#### Cause 1: Calculation Error

```{r error=TRUE}
n_start <- 5
n_end <- 3

# Calculation gives negative
n <- n_end - n_start  # -2
result <- numeric(n)  # Error!
```

#### Cause 2: User Input

```{r error=TRUE}
create_vector <- function(n) {
  numeric(n)
}

create_vector(-5)  # Error!
```

#### Cause 3: Filtering Gone Wrong

```{r error=TRUE}
data <- c(10, 20, 30)
threshold <- 50

# No values meet criteria
n_above <- sum(data > threshold)  # 0

# Then you subtract
n_below <- length(data) - n_above - 5  # Negative!

result <- numeric(n_below)  # Error!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Validate Before Creating**

```{r}
create_vector_safe <- function(n, default_value = 0) {
  if (n < 0) {
    warning("Negative length requested: ", n, ". Using 0.")
    n <- 0
  }
  
  if (n == 0) {
    return(numeric(0))
  }
  
  return(rep(default_value, n))
}

# Test
create_vector_safe(5)
create_vector_safe(-5)
create_vector_safe(0)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use max() for Safety**

```{r}
n_start <- 5
n_end <- 3

# Ensure non-negative
n <- max(0, n_end - n_start)
result <- numeric(n)  # Safe
result
```
</div>

## Logical Indexing

<div class="insight-box">
💡 **Key Insight: Logical Indexing**

```{r}
x <- c(10, 20, 30, 40, 50)

# Logical vector
x > 25

# Use for indexing
x[x > 25]

# Multiple conditions
x[x > 25 & x < 45]

# With which()
which(x > 25)
x[which(x > 25)]
```

**Important:** Logical indexing with NA creates NA in result!

```{r}
x <- c(10, NA, 30, 40)
x > 25  # Has NA

# Result includes NA
x[x > 25]

# Remove NA from condition
x[which(x > 25)]  # which() drops NA
```
</div>

## Error #7: `[ ] with missing values only allowed for atomic vectors` {#bracket-missing-values}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
my_list <- list(a = 1, b = 2, c = 3)
indices <- c(1, NA, 3)
my_list[indices]
```

<div class="error-box">
🔴 **ERROR**

```
Error in my_list[indices] : 
  [ ] with missing values only allowed for atomic vectors
```
</div>

### What It Means

You can use NA in indices for vectors, but not for lists or data frames without special handling.

### Atomic vs Non-Atomic

```{r}
# Atomic vector: NA indexing works
vec <- c(10, 20, 30)
vec[c(1, NA, 3)]  # Returns with NA

# List: NA indexing fails
my_list <- list(a = 1, b = 2, c = 3)
```

```{r error=TRUE}
my_list[c(1, NA, 3)]  # Error!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Remove NAs from Indices**

```{r}
my_list <- list(a = 1, b = 2, c = 3)
indices <- c(1, NA, 3)

# Remove NAs
clean_indices <- indices[!is.na(indices)]
my_list[clean_indices]
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use Logical Indexing**

```{r}
my_list <- list(a = 1, b = 2, c = 3)
indices <- c(1, NA, 3)

# Convert to logical
logical_indices <- seq_along(my_list) %in% indices
my_list[logical_indices]
```
</div>

## Matrix Indexing Special Cases

<div class="bestpractice-box">
🎯 **Best Practice: Matrix Indexing Mastery**

```{r}
mat <- matrix(1:12, nrow = 3, ncol = 4)
mat

# Single index: column-major order
mat[5]  # Row 2, Column 2

# Row, column
mat[2, 2]

# Entire row
mat[2, ]

# Entire column
mat[, 2]

# Multiple rows/columns
mat[1:2, 2:3]

# Logical indexing (rows)
mat[mat[, 1] > 1, ]

# Negative indexing
mat[-1, ]     # All but first row
mat[, -c(1, 4)]  # All but first and last column

# Preserve matrix structure
mat[, 2]              # Becomes vector
mat[, 2, drop = FALSE]  # Stays matrix
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **[ ] vs [[]]**: Single bracket is lenient (returns NA), double bracket is strict (errors)
2. **Check bounds**: Always validate indices before using [[]]
3. **Dimensions matter**: Vectors (1D), matrices (2D), arrays (3D+)
4. **Negative indexing**: Can't mix with positive (except 0)
5. **Logical indexing**: Watch for NA in conditions
6. **drop = FALSE**: Preserves matrix structure
7. **seq_along()**: Safer than 1:length() for loops

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| subscript out of bounds | Index > length with [[]] | Check length first |
| undefined columns | Column name doesn't exist | Check with `%in% names()` |
| incorrect number of dimensions | Wrong # of indices | Match object dimensions |
| incorrect number of subscripts | Too many/few indices | Matrix needs 2, array needs n |
| only 0's may be mixed with negative | Positive + negative indices | Use one or the other |
| negative length vectors | Tried length < 0 | Validate with `max(0, n)` |
| [ ] with missing values | NA index on list | Remove NAs or use logical |

**Safe Indexing Checklist:**

```{r eval=FALSE}
# Before indexing:
length(x)              # Check size
dim(x)                 # Check dimensions
names(x)               # Check names (if using)
seq_along(x)           # Safe iteration
index %in% seq_along(x)  # Validate index

# For conditional indexing:
which(condition)       # Drops NA automatically
x[!is.na(x) & x > 0]  # Handle NA explicitly
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Predict the Outcome**

What happens? Error, NA, or value?

```{r eval=FALSE}
# A
x <- 1:5
x[10]

# B
x[[10]]

# C
mat <- matrix(1:6, nrow = 2)
mat[3, 1]

# D
x[c(-1, 5)]

# E
my_list <- list(a = 1, b = 2)
my_list[c(1, NA)]
```
</div>

<div class="exercise-box">
📝 **Exercise 2: Fix the Code**

Debug these indexing problems:

```{r eval=FALSE}
# Problem 1
scores <- c(85, 90, 95)
top_score <- scores[[4]]

# Problem 2
df <- data.frame(x = 1:5, y = 6:10)
result <- df[, "z"]

# Problem 3
vec <- 1:10
subset <- vec[c(-1, -2, 5, 6)]

# Problem 4
mat <- matrix(1:6, nrow = 2)
col2 <- mat[, 2]
element <- col2[1, 1]
```
</div>

<div class="exercise-box">
📝 **Exercise 3: Safe Indexing Function**

Write `safe_index(x, i)` that:
1. Works with vectors, lists, matrices
2. Never errors on out-of-bounds
3. Returns NA for invalid indices
4. Handles both [ ] and [[ ]] style
5. Reports what went wrong
</div>

<div class="exercise-box">
📝 **Exercise 4: Matrix Subsetting**

Given a matrix, write functions to:
1. Get elements on the diagonal
2. Get upper triangle (above diagonal)
3. Get lower triangle (below diagonal)
4. Get border elements (edges only)
5. Handle any matrix size
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
# A - Returns NA (single bracket is lenient)
x <- 1:5
x[10]

# B - Errors (double bracket is strict)
tryCatch(x[[10]], error = function(e) "ERROR")

# C - Errors (only 2 rows)
mat <- matrix(1:6, nrow = 2)
tryCatch(mat[3, 1], error = function(e) "ERROR")

# D - Errors (can't mix positive and negative)
tryCatch(x[c(-1, 5)], error = function(e) "ERROR")

# E - Errors (lists don't allow NA indices)
my_list <- list(a = 1, b = 2)
tryCatch(my_list[c(1, NA)], error = function(e) "ERROR")
```

**Exercise 2:**

```{r}
# Problem 1 - Out of bounds
scores <- c(85, 90, 95)
# Fix: Check length
if (4 <= length(scores)) {
  top_score <- scores[[4]]
} else {
  top_score <- NA
}

# Problem 2 - Column doesn't exist
df <- data.frame(x = 1:5, y = 6:10)
# Fix: Check column exists
if ("z" %in% names(df)) {
  result <- df[, "z"]
} else {
  result <- NULL
}

# Problem 3 - Mixing positive and negative
vec <- 1:10
# Fix: Use only negative
subset <- vec[-c(1, 2)]
# Or only positive
subset <- vec[c(5, 6)]

# Problem 4 - Vector can't use 2D indexing
mat <- matrix(1:6, nrow = 2)
col2 <- mat[, 2, drop = FALSE]  # Keep as matrix
element <- col2[1, 1]
# Or:
col2 <- mat[, 2]  # Vector
element <- col2[1]  # 1D indexing
```

**Exercise 3:**

```{r}
safe_index <- function(x, i, double_bracket = FALSE) {
  # Handle different object types
  if (is.null(x)) {
    message("Object is NULL")
    return(NULL)
  }
  
  # Get valid range
  max_index <- if (is.list(x)) {
    length(x)
  } else if (!is.null(dim(x))) {
    length(x)  # For matrices, treat as vector
  } else {
    length(x)
  }
  
  # Check index validity
  if (any(is.na(i))) {
    message("Index contains NA")
    i <- i[!is.na(i)]
  }
  
  if (length(i) == 0) {
    message("No valid indices")
    return(if (double_bracket) NA else x[integer(0)])
  }
  
  if (any(i < 1 | i > max_index)) {
    invalid <- i[i < 1 | i > max_index]
    message("Invalid indices: ", paste(invalid, collapse = ", "))
    i <- i[i >= 1 & i <= max_index]
  }
  
  if (length(i) == 0) {
    return(NA)
  }
  
  # Extract
  if (double_bracket) {
    if (length(i) > 1) {
      message("Double bracket with multiple indices, using first")
      i <- i[1]
    }
    return(x[[i]])
  } else {
    return(x[i])
  }
}

# Test
x <- 1:5
safe_index(x, 3)
safe_index(x, 10)
safe_index(x, c(1, 10, 3))
```

**Exercise 4:**

```{r}
# Get diagonal elements
get_diagonal <- function(mat) {
  if (!is.matrix(mat)) stop("Input must be a matrix")
  n <- min(nrow(mat), ncol(mat))
  mat[cbind(1:n, 1:n)]
}

# Get upper triangle
get_upper_tri <- function(mat, include_diag = FALSE) {
  if (!is.matrix(mat)) stop("Input must be a matrix")
  mat[upper.tri(mat, diag = include_diag)]
}

# Get lower triangle
get_lower_tri <- function(mat, include_diag = FALSE) {
  if (!is.matrix(mat)) stop("Input must be a matrix")
  mat[lower.tri(mat, diag = include_diag)]
}

# Get border elements
get_border <- function(mat) {
  if (!is.matrix(mat)) stop("Input must be a matrix")
  nr <- nrow(mat)
  nc <- ncol(mat)
  
  if (nr == 1 || nc == 1) {
    return(as.vector(mat))
  }
  
  c(
    mat[1, ],                    # Top row
    mat[nr, ],                   # Bottom row
    mat[2:(nr-1), 1],           # Left column (excluding corners)
    mat[2:(nr-1), nc]           # Right column (excluding corners)
  )
}

# Test
mat <- matrix(1:20, nrow = 4, ncol = 5)
mat
get_diagonal(mat)
get_upper_tri(mat)
get_lower_tri(mat)
get_border(mat)
```
</details>
