# Part IV: Data Frames & Matrices {-}

# Data Frame Construction {#dataframe-construction}

<div class="chapter-summary">
**What You'll Learn:**

- How data frames are constructed
- Row and column length requirements
- Type coercion during construction
- Row names and their pitfalls
- Converting between structures

**Key Errors Covered:** 15+ construction errors

**Difficulty:** ⭐ Beginner to ⭐⭐ Intermediate
</div>

## Introduction

Data frames are R's workhorse for tabular data. But creating them can be tricky:

```{r error=TRUE}
# This looks innocent...
data.frame(x = 1:3, y = 1:5)
```

<div class="error-box">
🔴 **ERROR**

```
Error in data.frame(x = 1:3, y = 1:5) : 
  arguments imply differing number of rows: 3, 5
```
</div>

Let's master data frame construction and avoid all the common pitfalls.

## Data Frame Basics

<div class="insight-box">
💡 **Key Insight: What is a Data Frame?**

A data frame is a **list of vectors** with special properties:

```{r}
# Create a data frame
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  score = c(85, 90, 95)
)

# It's actually a list!
typeof(df)
is.list(df)

# But special
class(df)
is.data.frame(df)

# Each column is a vector
df$name
df$age

# All columns must have same length
length(df$name)
length(df$age)
```

**Key requirements:**
1. All columns must be same length (or length 1)
2. Each column must be a vector (atomic or list)
3. Row names must be unique
4. Column names should be unique (R allows duplicates but it's confusing)
</div>

## Error #1: `arguments imply differing number of rows` {#differing-rows}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-dimension">📏 DIMENSION</span>

### The Error

```{r error=TRUE}
data.frame(
  x = 1:5,
  y = 1:3
)
```

<div class="error-box">
🔴 **ERROR**

```
Error in data.frame(x = 1:5, y = 1:3) : 
  arguments imply differing number of rows: 5, 3
```
</div>

### What It Means

You're trying to create a data frame with columns of different lengths that aren't compatible.

### The Recycling Rule

Data frames allow recycling, but only in specific cases:

```{r}
# Length 1 always recycles
data.frame(
  x = 1:5,
  y = 10      # Recycled to 5
)

# Multiples work
data.frame(
  x = 1:6,
  y = c(1, 2)  # Recycled to 6 (multiple of 2)
)
```

```{r error=TRUE}
# Non-multiples fail
data.frame(
  x = 1:5,
  y = 1:3      # 5 is not a multiple of 3
)
```

### Common Causes

#### Cause 1: Filtered Data

```{r error=TRUE}
# Start with same length
ids <- 1:10
values <- rnorm(10)

# Filter one but not the other
filtered_values <- values[values > 0]  # Might be 6 elements

# Try to combine
data.frame(id = ids, value = filtered_values)  # Error!
```

#### Cause 2: Calculation Gone Wrong

```{r error=TRUE}
# Generate data
x <- 1:5
y <- x * 2

# Accidentally modify one
y <- y[-3]  # Remove one element

# Try to create data frame
data.frame(x = x, y = y)
```

#### Cause 3: Reading Different Sources

```{r error=TRUE}
# Simulating reading from different sources
col1 <- readLines(textConnection("a\nb\nc\nd\ne"))
col2 <- readLines(textConnection("1\n2\n3"))  # Shorter!

# Try to combine
data.frame(col1 = col1, col2 = col2)
```

#### Cause 4: List Column with Wrong Length

```{r error=TRUE}
# Regular columns
df <- data.frame(id = 1:5)

# Try to add list column of wrong length
df$nested <- list(a = 1:3, b = 4:6)  # Length 2, not 5!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Match Lengths Before Creating**

```{r}
ids <- 1:10
values <- rnorm(10)
filtered_values <- values[values > 0]

# Option A: Filter both the same way
keep <- values > 0
data.frame(
  id = ids[keep],
  value = filtered_values
)

# Option B: Extend shorter with NA
max_len <- max(length(ids), length(filtered_values))
ids_ext <- c(ids, rep(NA, max_len - length(ids)))
val_ext <- c(filtered_values, rep(NA, max_len - length(filtered_values)))
data.frame(id = ids_ext, value = val_ext)

# Option C: Trim longer to match
min_len <- min(length(ids), length(filtered_values))
data.frame(
  id = ids[1:min_len],
  value = filtered_values[1:min_len]
)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check Lengths First**

```{r}
safe_data_frame <- function(...) {
  # Get all arguments
  args <- list(...)
  
  # Get lengths
  lens <- sapply(args, length)
  
  # Check compatibility
  max_len <- max(lens)
  valid <- lens == 1 | lens == max_len | max_len %% lens == 0
  
  if (!all(valid)) {
    invalid_lens <- unique(lens[!valid])
    stop("Incompatible lengths: ", paste(lens, collapse = ", "))
  }
  
  # Create data frame
  data.frame(..., stringsAsFactors = FALSE)
}

# Test
safe_data_frame(x = 1:5, y = 10)      # Works
safe_data_frame(x = 1:6, y = c(1,2))  # Works
```

```{r error=TRUE}
safe_data_frame(x = 1:5, y = 1:3)     # Clear error
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use tidyverse for Better Errors**

```{r}
library(tibble)

# tibble is stricter - no recycling except length 1
tibble(
  x = 1:5,
  y = 10      # Length 1 OK
)
```

```{r error=TRUE}
# Doesn't allow multiple recycling
tibble(
  x = 1:6,
  y = c(1, 2)  # Error! Clearer message
)
```
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall: Silent Recycling**

```{r}
# This works but may not be intended!
df <- data.frame(
  group = 1:12,
  label = c("A", "B", "C")  # Recycled 4 times
)
df

# Was this intended? Hard to tell!
# Explicit is better:
df <- data.frame(
  group = 1:12,
  label = rep(c("A", "B", "C"), times = 4)
)
```
</div>

## Error #2: `row names supplied are of wrong length` {#rownames-wrong-length}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-dimension">📏 DIMENSION</span>

### The Error

```{r error=TRUE}
data.frame(
  x = 1:5,
  y = 6:10,
  row.names = c("a", "b", "c")  # Only 3 names for 5 rows!
)
```

<div class="error-box">
🔴 **ERROR**

```
Error in data.frame(x = 1:5, y = 6:10, row.names = c("a", "b", "c")) : 
  row names supplied are of wrong length
```
</div>

### What It Means

Row names must be exactly one per row. No more, no less.

### Row Names Basics

```{r}
# Row names are optional
df <- data.frame(x = 1:3, y = 4:6)
rownames(df)  # Default: "1", "2", "3"

# Can set custom row names
df <- data.frame(
  x = 1:3,
  y = 4:6,
  row.names = c("first", "second", "third")
)
rownames(df)

# Or set after creation
df <- data.frame(x = 1:3, y = 4:6)
rownames(df) <- c("a", "b", "c")
rownames(df)
```

### Common Causes

#### Cause 1: Wrong Count

```{r error=TRUE}
df <- data.frame(x = 1:10, y = 11:20)

# Too few row names
rownames(df) <- c("a", "b", "c")
```

#### Cause 2: After Subsetting

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10)
original_names <- c("a", "b", "c", "d", "e")
rownames(df) <- original_names

# Subset data frame
df_subset <- df[1:3, ]
rownames(df_subset)  # Kept original names (good)

# But if you try to reuse original names on subset:
rownames(df_subset) <- original_names  # Error! 5 names for 3 rows
```

#### Cause 3: From External Source

```{r error=TRUE}
data_values <- data.frame(x = 1:5, y = 6:10)
labels <- c("a", "b", "c")  # From somewhere else, wrong length

rownames(data_values) <- labels
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Match Length**

```{r}
df <- data.frame(x = 1:5, y = 6:10)
names_available <- c("a", "b", "c")

# Option A: Extend with numbers
all_names <- c(names_available, 
               paste0("row", (length(names_available)+1):nrow(df)))
rownames(df) <- all_names

# Option B: Use only available, add column instead
df$label <- c(names_available, rep(NA, nrow(df) - length(names_available)))
df
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check Before Assigning**

```{r}
safe_set_rownames <- function(df, names) {
  if (length(names) != nrow(df)) {
    warning("Row names length (", length(names), 
            ") doesn't match rows (", nrow(df), ")")
    return(df)
  }
  
  rownames(df) <- names
  return(df)
}

# Test
df <- data.frame(x = 1:5, y = 6:10)
df <- safe_set_rownames(df, c("a", "b", "c"))  # Warning, unchanged
df <- safe_set_rownames(df, letters[1:5])      # Works
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use Column Instead**

```{r}
# Modern best practice: avoid row names
df <- data.frame(
  id = c("a", "b", "c", "d", "e"),
  x = 1:5,
  y = 6:10
)

# Use first column as identifier
df

# Can still use for indexing
df[df$id == "c", ]
```
</div>

<div class="bestpractice-box">
🎯 **Best Practice: Avoid Row Names**

Row names are a legacy feature. Modern R style:

```{r}
# Old style (avoid)
df <- data.frame(x = 1:3, y = 4:6, row.names = c("a", "b", "c"))

# New style (prefer)
df <- data.frame(
  id = c("a", "b", "c"),
  x = 1:3,
  y = 4:6
)

# Tibbles don't even support row names!
library(tibble)
tibble(id = c("a", "b", "c"), x = 1:3, y = 4:6)
```

**Why?**
- Row names are easily lost in operations
- Harder to work with programmatically
- Not supported by modern tidyverse
- Column is more explicit and flexible
</div>

## Error #3: `duplicate row.names are not allowed` {#duplicate-rownames}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-syntax">🔤 SYNTAX</span>

### The Error

```{r error=TRUE}
data.frame(
  x = 1:3,
  y = 4:6,
  row.names = c("a", "b", "a")  # Duplicate!
)
```

<div class="error-box">
🔴 **ERROR**

```
Error in data.frame(x = 1:3, y = 4:6, row.names = c("a", "b", "a")) : 
  duplicate row.names: a
```
</div>

### What It Means

Row names must be unique identifiers. Duplicates aren't allowed.

### Common Causes

#### Cause 1: Non-Unique IDs

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10)
ids <- c("sample1", "sample2", "sample2", "sample3", "sample4")

rownames(df) <- ids  # Error!
```

#### Cause 2: After Combining Data Frames

```{r error=TRUE}
df1 <- data.frame(x = 1:3, row.names = c("a", "b", "c"))
df2 <- data.frame(x = 4:6, row.names = c("c", "d", "e"))  # "c" repeats!

# rbind checks for this
rbind(df1, df2)  # Error!
```

#### Cause 3: Default Row Names from Subsetting

```{r error=TRUE}
df <- data.frame(x = 1:10, row.names = letters[1:10])

# Take rows
df1 <- df[1:3, ]
df2 <- df[2:4, ]  # Overlaps with df1

# Try to combine
rbind(df1, df2)  # Error! "b" and "c" appear twice
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Make Unique**

```{r}
df <- data.frame(x = 1:5, y = 6:10)
ids <- c("sample1", "sample2", "sample2", "sample3", "sample4")

# Make unique automatically
unique_ids <- make.unique(ids, sep = "_")
unique_ids

rownames(df) <- unique_ids
df
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Reset Row Names When Combining**

```{r}
df1 <- data.frame(x = 1:3, row.names = c("a", "b", "c"))
df2 <- data.frame(x = 4:6, row.names = c("c", "d", "e"))

# Remove row names before combining
rownames(df1) <- NULL
rownames(df2) <- NULL
rbind(df1, df2)

# Or use row.names = FALSE
df_combined <- rbind(df1, df2)  # Still errors

# Better: let R assign new row names
df1_clean <- df1
df2_clean <- df2
rownames(df1_clean) <- NULL
rownames(df2_clean) <- NULL
rbind(df1_clean, df2_clean)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Check for Duplicates First**

```{r}
safe_set_rownames_unique <- function(df, names) {
  if (anyDuplicated(names)) {
    dupes <- names[duplicated(names)]
    warning("Duplicate row names found: ", 
            paste(unique(dupes), collapse = ", "))
    names <- make.unique(names, sep = "_")
  }
  
  rownames(df) <- names
  return(df)
}

# Test
df <- data.frame(x = 1:5)
df <- safe_set_rownames_unique(df, c("a", "b", "a", "c", "d"))
rownames(df)
```
</div>

## Error #4: `invalid type (list) for variable` {#invalid-type-list}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
# Trying to create column from nested list
data.frame(
  id = 1:3,
  values = list(1:3, 4:6, 7:9)  # List column - old data.frame rejects
)
```

<div class="error-box">
🔴 **ERROR** (in older R versions or strict mode)

```
Error in data.frame(...) : invalid type (list) for variable 'values'
```
</div>

### What It Means

Traditional `data.frame()` doesn't easily support list columns. Each column should be an atomic vector.

### Modern Solution: List Columns

```{r}
# Modern R allows this with I()
df <- data.frame(
  id = 1:3,
  values = I(list(1:3, 4:6, 7:9))
)
df

# Access list column
df$values[[1]]

# Tibbles make it easier
library(tibble)
tibble(
  id = 1:3,
  values = list(1:3, 4:6, 7:9)
)
```

### Common Causes

#### Cause 1: Nested Data

```{r error=TRUE}
# Have nested data
nested <- list(
  list(x = 1, y = 2),
  list(x = 3, y = 4),
  list(x = 5, y = 6)
)

# Try to put in data frame
data.frame(
  id = 1:3,
  data = nested  # Old error
)
```

#### Cause 2: Split/Group Result

```{r error=TRUE}
# Split creates list
values <- 1:12
groups <- rep(1:3, each = 4)
split_data <- split(values, groups)

# Try to put in data frame directly
data.frame(
  group = 1:3,
  values = split_data  # May error
)
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use I() to Protect List**

```{r}
# Wrap in I() to inhibit conversion
df <- data.frame(
  id = 1:3,
  values = I(list(1:3, 4:6, 7:9))
)

# Or use list column explicitly
df <- data.frame(id = 1:3)
df$values <- list(1:3, 4:6, 7:9)
df
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use Tibble (Easier)**

```{r}
library(tibble)

# Tibbles naturally support list columns
tib <- tibble(
  id = 1:3,
  values = list(1:3, 4:6, 7:9),
  nested = list(
    list(a = 1, b = 2),
    list(a = 3, b = 4),
    list(a = 5, b = 6)
  )
)
tib

# Clean syntax
tib$values
tib$nested[[1]]
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Unnest if Possible**

```{r}
# If you don't need list column, unnest
library(tidyr)

# Start with list column
df <- tibble(
  id = 1:3,
  values = list(1:3, 4:6, 7:9)
)

# Unnest to regular columns
unnest(df, values)
```
</div>

## Error #5: `cannot coerce class X to a data.frame` {#cannot-coerce-df-dataframe}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
# Try to convert function to data frame
my_func <- function(x) x + 1
as.data.frame(my_func)
```

<div class="error-box">
🔴 **ERROR**

```
Error in as.data.frame.default(my_func) : 
  cannot coerce class '"function"' to a data.frame
```
</div>

### What It Means

You're trying to convert something to a data frame that can't be converted.

### Common Causes

#### Cause 1: Wrong Object Type

```{r error=TRUE}
# Can't convert these:
as.data.frame(mean)           # Function
as.data.frame(environment())  # Environment
```

#### Cause 2: Incompatible Structure

```{r error=TRUE}
# Uneven list
irregular <- list(
  a = 1:3,
  b = 1:5,  # Different length!
  c = 1:2
)

as.data.frame(irregular)
```

#### Cause 3: After Failed Operation

```{r error=TRUE}
# Operation returns something unexpected
result <- try(stop("Error"), silent = TRUE)
class(result)  # "try-error"

as.data.frame(result)  # Can't convert error object
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Type First**

```{r}
safe_as_dataframe <- function(x) {
  # Check if already data frame
  if (is.data.frame(x)) {
    return(x)
  }
  
  # Check if matrix
  if (is.matrix(x)) {
    return(as.data.frame(x))
  }
  
  # Check if list with equal lengths
  if (is.list(x)) {
    lens <- lengths(x)
    if (length(unique(lens)) == 1 || all(lens == 1 | lens == max(lens))) {
      return(as.data.frame(x, stringsAsFactors = FALSE))
    } else {
      stop("List has incompatible lengths: ", paste(lens, collapse = ", "))
    }
  }
  
  # Check if vector
  if (is.atomic(x)) {
    return(data.frame(value = x, stringsAsFactors = FALSE))
  }
  
  stop("Cannot convert ", class(x)[1], " to data frame")
}

# Test
safe_as_dataframe(1:5)
safe_as_dataframe(matrix(1:6, 2, 3))
safe_as_dataframe(list(a = 1:3, b = 4:6))
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Fix Structure First**

```{r}
# Uneven list
irregular <- list(
  a = 1:3,
  b = 1:5,
  c = 1:2
)

# Option A: Pad with NA
max_len <- max(lengths(irregular))
regular <- lapply(irregular, function(x) {
  c(x, rep(NA, max_len - length(x)))
})
as.data.frame(regular)

# Option B: Trim to shortest
min_len <- min(lengths(irregular))
regular <- lapply(irregular, function(x) x[1:min_len])
as.data.frame(regular)
```
</div>

## Stringsasfactors Historical Issue

<div class="pitfall-box">
⚠️ **Historical Pitfall: stringsAsFactors**

In R < 4.0.0, `stringsAsFactors = TRUE` was default:

```{r}
# R < 4.0.0 behavior
df_old <- data.frame(
  name = c("Alice", "Bob"),
  age = c(25, 30),
  stringsAsFactors = TRUE  # Was default
)
class(df_old$name)  # "factor" (was default)

# R >= 4.0.0 behavior
df_new <- data.frame(
  name = c("Alice", "Bob"),
  age = c(25, 30)
  # stringsAsFactors = FALSE is now default
)
class(df_new$name)  # "character"
```

**Best Practice:** Always specify explicitly:

```{r eval=FALSE}
# For compatibility across R versions
data.frame(..., stringsAsFactors = FALSE)

# Or use tibble (never converts to factor)
tibble(...)
```
</div>

## Creating Data Frames: All Methods

<div class="bestpractice-box">
🎯 **Best Practice: Data Frame Creation Methods**

```{r}
# Method 1: data.frame() - base R
df1 <- data.frame(
  x = 1:3,
  y = 4:6,
  stringsAsFactors = FALSE
)

# Method 2: tibble() - modern tidyverse
library(tibble)
df2 <- tibble(
  x = 1:3,
  y = 4:6
)

# Method 3: From matrix
mat <- matrix(1:6, nrow = 2, ncol = 3)
df3 <- as.data.frame(mat)

# Method 4: From list
lst <- list(x = 1:3, y = 4:6)
df4 <- as.data.frame(lst)

# Method 5: From vectors
x <- 1:3
y <- 4:6
df5 <- data.frame(x, y)

# Method 6: Reading data
df6 <- read.csv("file.csv")  # Base R
# df7 <- read_csv("file.csv")  # readr (tibble)

# Method 7: tribble() - row-wise
library(tibble)
df8 <- tribble(
  ~x, ~y,
   1,  4,
   2,  5,
   3,  6
)
```

**When to use each:**
- `data.frame()`: Base R compatibility needed
- `tibble()`: Modern code, better defaults
- `as.data.frame()`: Converting from other structures
- `tribble()`: Small data, readable layout
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **All columns must be same length** (or length 1 for recycling)
2. **Row names must be unique** (if used at all)
3. **Avoid row names** in modern code - use column instead
4. **List columns need I()** in data.frame(), or use tibble
5. **stringsAsFactors = FALSE** for compatibility
6. **Check lengths before construction**
7. **tibble is stricter** and usually better

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| differing number of rows | Unequal lengths | Match lengths or use length 1 |
| row names wrong length | Row names ≠ nrows | Provide correct number |
| duplicate row.names | Non-unique names | Use make.unique() or remove |
| invalid type (list) | List column in old data.frame | Use I() or tibble |
| cannot coerce to data.frame | Wrong structure | Check type and fix structure |

**Construction Checklist:**

```{r eval=FALSE}
# Before creating data frame:
lengths(list_of_vectors)     # Check all lengths
anyDuplicated(row_names)     # Check for duplicates
class(each_column)           # Verify types

# Safe construction:
tibble(...)                  # Stricter checking
data.frame(..., stringsAsFactors = FALSE)  # Explicit

# After construction:
str(df)                      # Verify structure
anyDuplicated(rownames(df))  # Check row names
```

**Modern Best Practices:**

```{r eval=FALSE}
# ✅ Good
tibble(x = 1:5, y = 10)                    # Stricter
data.frame(x = 1:5, stringsAsFactors = FALSE)  # Explicit

# ❌ Avoid
data.frame(x = 1:5)                        # stringsAsFactors unclear
df <- data.frame(x = 1:5, row.names = names)  # Row names fragile
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Diagnose the Error**

What's wrong and how do you fix it?

```{r eval=FALSE}
# Scenario 1
df <- data.frame(
  id = 1:10,
  group = c("A", "B", "C")
)

# Scenario 2
df <- data.frame(
  x = 1:5,
  y = 6:10,
  row.names = c("a", "b", "c", "a", "e")
)

# Scenario 3
df1 <- data.frame(x = 1:3, row.names = c("a", "b", "c"))
df2 <- data.frame(x = 4:6, row.names = c("b", "c", "d"))
combined <- rbind(df1, df2)

# Scenario 4
data.frame(
  id = 1:3,
  data = list(
    c(1, 2, 3),
    c(4, 5),
    c(6, 7, 8, 9)
  )
)
```
</div>

<div class="exercise-box">
📝 **Exercise 2: Safe Constructor**

Write `safe_df(...)` that:
1. Checks all vectors are same length or length 1
2. Warns about recycling
3. Checks for duplicate names
4. Returns tibble or data.frame
5. Handles list columns properly
</div>

<div class="exercise-box">
📝 **Exercise 3: Fix Irregular Data**

You have:
```{r eval=FALSE}
data <- list(
  id = 1:5,
  name = c("Alice", "Bob", "Charlie"),
  score = c(85, 90, 95, 88)
)
```

Create a data frame handling the unequal lengths gracefully.
</div>

<div class="exercise-box">
📝 **Exercise 4: Combine with Row Names**

You have multiple data frames with overlapping row names:
```{r eval=FALSE}
df1 <- data.frame(x = 1:3, row.names = c("a", "b", "c"))
df2 <- data.frame(y = 4:6, row.names = c("b", "c", "d"))
df3 <- data.frame(z = 7:9, row.names = c("c", "d", "e"))
```

Combine them into one data frame keeping all data.
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
# Scenario 1 - Length mismatch (10 vs 3)
# Fix: Recycle explicitly or trim
df <- data.frame(
  id = 1:10,
  group = rep(c("A", "B", "C"), length.out = 10)
)

# Scenario 2 - Duplicate row name "a"
df <- data.frame(
  x = 1:5,
  y = 6:10,
  row.names = make.unique(c("a", "b", "c", "a", "e"))
)

# Scenario 3 - Overlapping row names
df1 <- data.frame(x = 1:3, row.names = c("a", "b", "c"))
df2 <- data.frame(x = 4:6, row.names = c("b", "c", "d"))
# Option A: Remove row names
combined <- rbind(
  data.frame(x = df1$x),
  data.frame(x = df2$x)
)
# Option B: Keep as column
combined <- rbind(
  data.frame(id = rownames(df1), x = df1$x),
  data.frame(id = rownames(df2), x = df2$x)
)

# Scenario 4 - List column (different lengths within)
# Use tibble or I()
library(tibble)
df <- tibble(
  id = 1:3,
  data = list(
    c(1, 2, 3),
    c(4, 5),
    c(6, 7, 8, 9)
  )
)
# Or with data.frame:
df <- data.frame(id = 1:3)
df$data <- list(c(1,2,3), c(4,5), c(6,7,8,9))
```

**Exercise 2:**

```{r}
safe_df <- function(..., use_tibble = TRUE) {
  args <- list(...)
  
  # Get lengths
  lens <- sapply(args, length)
  
  # Check for issues
  max_len <- max(lens)
  
  # Check compatibility
  recyclable <- lens == 1 | lens == max_len
  
  if (!all(recyclable)) {
    # Check if multiples
    multiples <- max_len %% lens == 0
    if (!all(recyclable | multiples)) {
      stop("Incompatible lengths: ", paste(lens, collapse = ", "),
           "\nMax length: ", max_len)
    }
    warning("Recycling vectors of length ", 
            paste(unique(lens[!recyclable]), collapse = ", "),
            " to length ", max_len)
  }
  
  # Check names
  arg_names <- names(args)
  if (!is.null(arg_names) && anyDuplicated(arg_names)) {
    warning("Duplicate column names: ",
            paste(arg_names[duplicated(arg_names)], collapse = ", "))
  }
  
  # Create data frame
  if (use_tibble) {
    library(tibble)
    result <- tibble(...)
  } else {
    result <- data.frame(..., stringsAsFactors = FALSE)
  }
  
  return(result)
}

# Test
safe_df(x = 1:5, y = 10)
safe_df(x = 1:6, y = c(1, 2))  # Warning about recycling
```

**Exercise 3:**

```{r}
data <- list(
  id = 1:5,
  name = c("Alice", "Bob", "Charlie"),
  score = c(85, 90, 95, 88)
)

# Option A: Extend shorter with NA
max_len <- max(lengths(data))
data_fixed <- lapply(data, function(x) {
  c(x, rep(NA, max_len - length(x)))
})
df <- as.data.frame(data_fixed, stringsAsFactors = FALSE)
df

# Option B: Trim all to shortest
min_len <- min(lengths(data))
data_fixed <- lapply(data, function(x) x[1:min_len])
df <- as.data.frame(data_fixed, stringsAsFactors = FALSE)
df

# Option C: Use only complete cases
# (more complex - requires pairing)
```

**Exercise 4:**

```{r}
library(tibble)

df1 <- data.frame(x = 1:3, row.names = c("a", "b", "c"))
df2 <- data.frame(y = 4:6, row.names = c("b", "c", "d"))
df3 <- data.frame(z = 7:9, row.names = c("c", "d", "e"))

# Convert row names to column
df1_with_id <- tibble(id = rownames(df1), x = df1$x)
df2_with_id <- tibble(id = rownames(df2), y = df2$y)
df3_with_id <- tibble(id = rownames(df3), z = df3$z)

# Full join to keep all
library(dplyr)
result <- df1_with_id %>%
  full_join(df2_with_id, by = "id") %>%
  full_join(df3_with_id, by = "id")

result

# Alternative: using merge
result <- merge(df1_with_id, df2_with_id, by = "id", all = TRUE)
result <- merge(result, df3_with_id, by = "id", all = TRUE)
result
```
</details>
