# Column Manipulation {#column-manipulation}

<div class="chapter-summary">
**What You'll Learn:**

- Adding and removing columns safely
- Column name handling
- Type preservation during operations
- Renaming strategies
- Common manipulation pitfalls

**Key Errors Covered:** 12+ column manipulation errors

**Difficulty:** ⭐ Beginner to ⭐⭐ Intermediate
</div>

## Introduction

Working with data frame columns is a daily task, but it's full of traps:

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10)
df[, "z"]  # Typo in column name
```

```{r error=TRUE}
# Or this:
df$new_column <- 1:3  # Wrong length!
```

Let's master column manipulation to avoid these common errors.

## Column Basics

<div class="insight-box">
💡 **Key Insight: Data Frame is a List**

Understanding this is key to column operations:

```{r}
df <- data.frame(x = 1:3, y = 4:6, z = 7:9)

# Data frame is a special list
is.list(df)
length(df)  # Number of columns!

# Each column is a list element
df[[1]]     # First column (vector)
df[1]       # First column (data frame)

# Three ways to access columns:
df$x        # Dollar sign
df[["x"]]   # Double bracket
df["x"]     # Single bracket (returns data frame)

# Column names
names(df)
colnames(df)
```
</div>

## Error #1: `undefined columns selected` {#undefined-columns-select}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-dimension">📏 DIMENSION</span>

### The Error

```{r error=TRUE}
df <- data.frame(age = c(25, 30, 35), name = c("Alice", "Bob", "Charlie"))
df[, "salary"]  # Column doesn't exist
```

<div class="error-box">
🔴 **ERROR**

```
Error in `[.data.frame`(df, , "salary") : undefined columns selected
```
</div>

### What It Means

You're trying to select a column that doesn't exist in the data frame.

### Common Causes

#### Cause 1: Typo

```{r error=TRUE}
df <- data.frame(temperature = c(20, 25, 30))

# Typo
df[, "tempurature"]
```

#### Cause 2: Case Sensitivity

```{r error=TRUE}
df <- data.frame(Name = c("Alice", "Bob"))

# Wrong case
df[, "name"]  # It's "Name" not "name"
```

#### Cause 3: Column Doesn't Exist Yet

```{r error=TRUE}
df <- data.frame(x = 1:5)

# Trying to select before creating
df[, c("x", "y")]  # "y" doesn't exist
```

#### Cause 4: After Transformation

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10, z = 11:15)

# Select some columns
df_subset <- df[, c("x", "y")]

# Try to access original column
df_subset[, "z"]  # No longer exists
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Column Exists**

```{r}
df <- data.frame(age = c(25, 30, 35), name = c("Alice", "Bob", "Charlie"))

# Check before accessing
if ("salary" %in% names(df)) {
  df[, "salary"]
} else {
  message("Column 'salary' not found")
  NULL
}

# Or for multiple columns
cols_wanted <- c("age", "salary", "name")
cols_available <- cols_wanted[cols_wanted %in% names(df)]
df[, cols_available]
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use dplyr's select() with Helpers**

```{r}
library(dplyr)

df <- data.frame(age = c(25, 30, 35), name = c("Alice", "Bob", "Charlie"))

# Select only existing columns
df %>% select(any_of(c("age", "salary", "name")))

# Or with error on missing
```

```{r error=TRUE}
df %>% select(all_of(c("age", "salary")))  # Errors on missing
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Safe Column Selection Function**

```{r}
safe_select <- function(df, cols, warn = TRUE) {
  existing <- cols[cols %in% names(df)]
  missing <- cols[!cols %in% names(df)]
  
  if (length(missing) > 0 && warn) {
    warning("Columns not found: ", paste(missing, collapse = ", "))
  }
  
  if (length(existing) == 0) {
    return(data.frame())  # Empty data frame
  }
  
  return(df[, existing, drop = FALSE])
}

# Test
df <- data.frame(x = 1:5, y = 6:10)
safe_select(df, c("x", "z", "y"))
```
</div>

## Error #2: `replacement has X rows, data has Y` {#replacement-rows-column}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-dimension">📏 DIMENSION</span>

### The Error

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10)
df$z <- 1:3  # Wrong length!
```

<div class="error-box">
🔴 **ERROR**

```
Error in `$<-.data.frame`(`*tmp*`, z, value = 1:3) : 
  replacement has 3 rows, data has 5
```
</div>

### What It Means

When adding/replacing a column, the new values must match the number of rows (or be length 1).

### The Recycling Rule for Columns

```{r}
df <- data.frame(x = 1:5)

# Length 1: recycles
df$y <- 10
df

# Same length: works
df$z <- 11:15
df
```

```{r error=TRUE}
# Wrong length: errors
df$w <- 1:3
```

### Common Causes

#### Cause 1: Calculation Resulted in Wrong Length

```{r error=TRUE}
df <- data.frame(id = 1:10, value = rnorm(10))

# Filter creates shorter vector
high_values <- df$value[df$value > 0]  # Maybe 6 elements

# Try to add back
df$high <- high_values  # Error! 6 vs 10
```

#### Cause 2: Using Summary on Column

```{r error=TRUE}
df <- data.frame(
  group = rep(c("A", "B"), each = 5),
  value = 1:10
)

# Calculate group means (2 values)
group_means <- tapply(df$value, df$group, mean)

# Try to add as column
df$mean <- group_means  # Error! 2 vs 10
```

#### Cause 3: After Subsetting

```{r error=TRUE}
df <- data.frame(x = 1:10, y = 11:20)

# Subset rows
df_sub <- df[1:5, ]

# Create column for full df
new_col <- 1:5

# Try to add to original
df$new <- new_col  # Error! 5 vs 10
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Match Lengths**

```{r}
df <- data.frame(id = 1:10, value = rnorm(10))
high_values <- df$value[df$value > 0]

# Option A: Use NA for missing
df$high <- NA
df$high[df$value > 0] <- high_values
df

# Option B: Use ifelse
df$high <- ifelse(df$value > 0, df$value, NA)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use Merge for Aggregates**

```{r}
df <- data.frame(
  group = rep(c("A", "B"), each = 5),
  value = 1:10
)

# Calculate group means
group_summary <- aggregate(value ~ group, df, mean)
names(group_summary)[2] <- "group_mean"

# Merge back
df <- merge(df, group_summary, by = "group")
df
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use dplyr (Cleaner)**

```{r}
library(dplyr)

df <- data.frame(
  group = rep(c("A", "B"), each = 5),
  value = 1:10
)

# Add group mean to each row
df <- df %>%
  group_by(group) %>%
  mutate(group_mean = mean(value)) %>%
  ungroup()

df
```
</div>

## Error #3: `duplicate column names` {#duplicate-columns}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-syntax">🔤 SYNTAX</span>

### The Warning/Problem

```{r}
df <- data.frame(x = 1:3, x = 4:6, check.names = FALSE)
names(df)
```

R allows duplicate column names (with warning), but it causes problems:

```{r}
# Which x?
df$x  # Gets first one

# Confusion!
df[, "x"]  # Gets first one
```

### Why It's Dangerous

```{r}
# Create with duplicates
df <- data.frame(value = 1:3, value = 4:6, check.names = FALSE)

# Operations become unpredictable
df$value <- df$value * 2  # Which one gets modified?
df

# Selection is confusing
df[, c("value", "value")]  # Gets same column twice
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Let R Fix Names**

```{r}
# Default: R makes names unique
df <- data.frame(x = 1:3, x = 4:6)  # check.names = TRUE by default
names(df)  # "x" and "x.1"

# Or manually
names_original <- c("value", "value", "score")
names_fixed <- make.names(names_original, unique = TRUE)
names_fixed
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check and Fix Names**

```{r}
fix_duplicate_names <- function(df) {
  col_names <- names(df)
  
  if (anyDuplicated(col_names)) {
    dupes <- col_names[duplicated(col_names)]
    warning("Duplicate column names found: ", 
            paste(unique(dupes), collapse = ", "))
    names(df) <- make.names(col_names, unique = TRUE)
  }
  
  return(df)
}

# Test
df <- data.frame(x = 1:3, x = 4:6, check.names = FALSE)
df <- fix_duplicate_names(df)
names(df)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Prevent Duplicates**

```{r}
safe_add_column <- function(df, name, values) {
  if (name %in% names(df)) {
    stop("Column '", name, "' already exists. ",
         "Use a different name or remove the existing column first.")
  }
  
  df[[name]] <- values
  return(df)
}

# Test
df <- data.frame(x = 1:3)
df <- safe_add_column(df, "y", 4:6)  # Works
```

```{r error=TRUE}
df <- safe_add_column(df, "x", 7:9)  # Errors
```
</div>

## Error #4: `names attribute must be same length as vector` {#names-length-mismatch}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-dimension">📏 DIMENSION</span>

### The Error

```{r error=TRUE}
df <- data.frame(x = 1:5, y = 6:10, z = 11:15)
names(df) <- c("a", "b")  # Only 2 names for 3 columns!
```

<div class="error-box">
🔴 **ERROR**

```
Error in names(df) <- c("a", "b") : 
  'names' attribute must be the same length as the vector (3)
```
</div>

### What It Means

When setting column names, you must provide exactly one name per column.

### Common Causes

#### Cause 1: Wrong Count

```{r error=TRUE}
df <- data.frame(x = 1:3, y = 4:6, z = 7:9)

# Too few
names(df) <- c("first", "second")

# Too many
names(df) <- c("first", "second", "third", "fourth")
```

#### Cause 2: After Adding Columns

```{r error=TRUE}
df <- data.frame(x = 1:3, y = 4:6)
new_names <- c("a", "b")

# Add a column
df$z <- 7:9

# Try to use old names
names(df) <- new_names  # Error! Now 3 columns
```

#### Cause 3: From External Source

```{r error=TRUE}
df <- data.frame(matrix(1:12, nrow = 3, ncol = 4))
column_labels <- c("ID", "Value")  # Wrong number

names(df) <- column_labels
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Match Number of Names**

```{r}
df <- data.frame(x = 1:3, y = 4:6, z = 7:9)

# Provide all names
names(df) <- c("first", "second", "third")

# Or rename specific columns
names(df)[1] <- "id"
names(df)[3] <- "score"
names(df)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use Named Vector for Partial Rename**

```{r}
library(dplyr)

df <- data.frame(x = 1:3, y = 4:6, z = 7:9)

# Rename specific columns
df <- df %>% rename(id = x, score = z)
names(df)

# Or base R
names(df)[names(df) == "y"] <- "value"
names(df)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Safe Rename Function**

```{r}
safe_rename <- function(df, ...) {
  name_mapping <- list(...)
  
  for (old_name in names(name_mapping)) {
    new_name <- name_mapping[[old_name]]
    
    if (!old_name %in% names(df)) {
      warning("Column '", old_name, "' not found, skipping")
      next
    }
    
    if (new_name %in% names(df) && new_name != old_name) {
      warning("Column '", new_name, "' already exists, skipping")
      next
    }
    
    names(df)[names(df) == old_name] <- new_name
  }
  
  return(df)
}

# Test
df <- data.frame(x = 1:3, y = 4:6, z = 7:9)
df <- safe_rename(df, x = "id", z = "score", w = "missing")
names(df)
```
</div>

## Removing Columns

<div class="bestpractice-box">
🎯 **Best Practice: Removing Columns**

```{r}
df <- data.frame(x = 1:3, y = 4:6, z = 7:9)

# Method 1: Set to NULL
df$y <- NULL
df

# Method 2: Subset (keep what you want)
df <- data.frame(x = 1:3, y = 4:6, z = 7:9)
df <- df[, c("x", "z")]
df

# Method 3: Subset (exclude what you don't want)
df <- data.frame(x = 1:3, y = 4:6, z = 7:9)
df <- df[, !names(df) %in% c("y")]
df

# Method 4: dplyr select with minus
library(dplyr)
df <- data.frame(x = 1:3, y = 4:6, z = 7:9)
df <- df %>% select(-y)
df

# Method 5: Remove multiple
df <- data.frame(x = 1:3, y = 4:6, z = 7:9, w = 10:12)
df <- df %>% select(-c(y, w))
df
```

**Never do this:**
```{r eval=FALSE}
# ❌ Bad: modifies in place
df[, "y"] <- NULL  # Doesn't work as expected!

# ✅ Good: explicit assignment
df$y <- NULL
```
</div>

## Column Reordering

<div class="insight-box">
💡 **Key Insight: Reordering Columns**

```{r}
df <- data.frame(z = 7:9, x = 1:3, y = 4:6)
names(df)

# Method 1: Specify order explicitly
df <- df[, c("x", "y", "z")]
names(df)

# Method 2: Sort alphabetically
df <- df[, sort(names(df))]
names(df)

# Method 3: Move specific columns first
df <- data.frame(z = 7:9, x = 1:3, y = 4:6)
df <- df[, c("x", setdiff(names(df), "x"))]
names(df)

# Method 4: dplyr relocate
library(dplyr)
df <- data.frame(z = 7:9, x = 1:3, y = 4:6)
df <- df %>% relocate(x, y, z)
names(df)

# Or move to front/end
df <- data.frame(z = 7:9, x = 1:3, y = 4:6)
df <- df %>% relocate(x, .before = everything())
df %>% relocate(z, .after = everything())
```
</div>

## Type Preservation

<div class="pitfall-box">
⚠️ **Common Pitfall: Type Changes**

```{r}
# Start with factors
df <- data.frame(
  id = 1:3,
  category = c("A", "B", "C"),
  stringsAsFactors = TRUE
)
class(df$category)  # "factor"

# Select columns - type changes!
df_subset <- df[, "category"]
class(df_subset)  # "factor" (still)

# But extract as vector
vec <- df$category
class(vec)  # "factor"

# With drop = TRUE (default)
df_subset <- df[, "category", drop = TRUE]
class(df_subset)  # "factor" - becomes vector

# With drop = FALSE
df_subset <- df[, "category", drop = FALSE]
class(df_subset)  # "data.frame" - stays data frame
```

**Best practice:**
```{r}
# Use drop = FALSE when you want to keep data frame structure
df[, "category", drop = FALSE]

# Or use $ when you explicitly want a vector
df$category
```
</div>

## Adding Multiple Columns

<div class="bestpractice-box">
🎯 **Best Practice: Adding Multiple Columns**

```{r}
df <- data.frame(x = 1:5)

# Method 1: One at a time
df$y <- 6:10
df$z <- 11:15

# Method 2: cbind
df <- data.frame(x = 1:5)
df <- cbind(df, data.frame(y = 6:10, z = 11:15))

# Method 3: dplyr mutate
library(dplyr)
df <- data.frame(x = 1:5)
df <- df %>%
  mutate(
    y = x + 5,
    z = y + 5
  )

# Method 4: Transform base R
df <- data.frame(x = 1:5)
df <- transform(df,
  y = x + 5,
  z = y + 5  # Can reference previous
)

# Method 5: within
df <- data.frame(x = 1:5)
df <- within(df, {
  y <- x + 5
  z <- y + 5
})
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Check column exists** before accessing with `%in% names()`
2. **Match row count** when adding columns (or use length 1)
3. **Avoid duplicate names** - check with `anyDuplicated()`
4. **Provide all names** when renaming - one per column
5. **Use drop = FALSE** to preserve data frame structure
6. **$ sets to NULL** removes columns cleanly
7. **dplyr is clearer** for complex column operations

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| undefined columns selected | Column doesn't exist | Check with `%in% names()` |
| replacement has X rows | Wrong length column | Match nrows or use length 1 |
| duplicate column names | Non-unique names | Use `make.names(unique=TRUE)` |
| names attribute wrong length | Wrong # of names | Provide one per column |

**Column Operations Checklist:**

```{r eval=FALSE}
# Before accessing:
"colname" %in% names(df)      # Check exists
anyDuplicated(names(df))      # Check no duplicates

# When adding column:
length(new_values) == nrow(df) || length(new_values) == 1

# When renaming:
length(new_names) == ncol(df)
!anyDuplicated(new_names)

# Safe patterns:
df$col <- NULL                # Remove column
df[, cols, drop = FALSE]      # Keep as data frame
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
df %>% select(any_of(c("a", "b", "c")))  # Safe selection
df %>% mutate(new = old * 2)             # Add column
names(df)[names(df) == "old"] <- "new"   # Rename one

# ❌ Avoid
df[, "missing_col"]           # No check
df$new <- wrong_length_vector # No validation
names(df) <- c("a", "b")      # Partial names
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Safe Column Access**

Write a function that safely gets a column:
- Returns the column if it exists
- Returns default value if it doesn't
- Warns user about missing columns
- Handles both $ and [[ ]] style access
</div>

<div class="exercise-box">
📝 **Exercise 2: Batch Rename**

You have:
```{r eval=FALSE}
df <- data.frame(
  old_name_1 = 1:5,
  old_name_2 = 6:10,
  old_name_3 = 11:15
)
```

Write a function to rename all columns matching a pattern.
</div>

<div class="exercise-box">
📝 **Exercise 3: Safe Column Addition**

Write `add_column(df, name, values)` that:
1. Checks if name already exists
2. Validates values length
3. Handles recycling appropriately
4. Returns modified data frame
5. Gives informative errors
</div>

<div class="exercise-box">
📝 **Exercise 4: Column Audit**

Write a function that audits a data frame and reports:
- Missing column names
- Duplicate column names
- Invalid column names (non-syntactic)
- Columns with NA names
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
safe_get_column <- function(df, col, default = NULL, warn = TRUE) {
  if (!col %in% names(df)) {
    if (warn) {
      warning("Column '", col, "' not found in data frame")
    }
    return(default)
  }
  
  return(df[[col]])
}

# Test
df <- data.frame(x = 1:5, y = 6:10)
safe_get_column(df, "x")         # Returns column
safe_get_column(df, "z")         # Returns NULL with warning
safe_get_column(df, "z", default = NA, warn = FALSE)
```

**Exercise 2:**

```{r}
rename_pattern <- function(df, pattern, replacement) {
  old_names <- names(df)
  new_names <- gsub(pattern, replacement, old_names)
  
  if (identical(old_names, new_names)) {
    message("No columns matched pattern '", pattern, "'")
    return(df)
  }
  
  # Check for duplicates after rename
  if (anyDuplicated(new_names)) {
    warning("Renaming would create duplicate names, using make.unique()")
    new_names <- make.unique(new_names)
  }
  
  names(df) <- new_names
  
  # Report changes
  changed <- old_names != new_names
  if (any(changed)) {
    message("Renamed ", sum(changed), " columns:")
    for (i in which(changed)) {
      message("  ", old_names[i], " -> ", new_names[i])
    }
  }
  
  return(df)
}

# Test
df <- data.frame(
  old_name_1 = 1:5,
  old_name_2 = 6:10,
  old_name_3 = 11:15
)
df <- rename_pattern(df, "old_name_", "new_col_")
names(df)
```

**Exercise 3:**

```{r}
add_column <- function(df, name, values, overwrite = FALSE) {
  # Check if name exists
  if (name %in% names(df) && !overwrite) {
    stop("Column '", name, "' already exists. ",
         "Use overwrite = TRUE to replace.")
  }
  
  # Check length
  n_rows <- nrow(df)
  n_values <- length(values)
  
  if (n_values == n_rows) {
    # Perfect match
    df[[name]] <- values
  } else if (n_values == 1) {
    # Recycle single value
    message("Recycling single value to ", n_rows, " rows")
    df[[name]] <- values
  } else if (n_rows %% n_values == 0) {
    # Multiple recycling
    message("Recycling ", n_values, " values to ", n_rows, " rows")
    df[[name]] <- rep(values, length.out = n_rows)
  } else {
    stop("Length mismatch: values has ", n_values, 
         " elements but data frame has ", n_rows, " rows")
  }
  
  return(df)
}

# Test
df <- data.frame(x = 1:5)
df <- add_column(df, "y", 10)           # Recycles
df <- add_column(df, "z", 11:15)        # Matches
```

```{r error=TRUE}
df <- add_column(df, "w", 1:3)          # Errors
```

**Exercise 4:**

```{r}
audit_columns <- function(df) {
  col_names <- names(df)
  issues <- list()
  
  # Check for missing names
  if (any(is.na(col_names) | col_names == "")) {
    issues$missing <- which(is.na(col_names) | col_names == "")
  }
  
  # Check for duplicates
  if (anyDuplicated(col_names)) {
    dupes <- col_names[duplicated(col_names)]
    issues$duplicates <- unique(dupes)
  }
  
  # Check for invalid names (non-syntactic)
  valid <- make.names(col_names) == col_names
  if (!all(valid)) {
    issues$invalid <- col_names[!valid]
  }
  
  # Report
  if (length(issues) == 0) {
    message("✓ All column names are valid")
    return(invisible(TRUE))
  }
  
  message("Column name issues found:")
  
  if (!is.null(issues$missing)) {
    message("  Missing names at positions: ", 
            paste(issues$missing, collapse = ", "))
  }
  
  if (!is.null(issues$duplicates)) {
    message("  Duplicate names: ", 
            paste(issues$duplicates, collapse = ", "))
  }
  
  if (!is.null(issues$invalid)) {
    message("  Invalid names: ", 
            paste(issues$invalid, collapse = ", "))
    message("  Suggested: ", 
            paste(make.names(issues$invalid), collapse = ", "))
  }
  
  return(invisible(issues))
}

# Test
df_good <- data.frame(x = 1:3, y = 4:6)
audit_columns(df_good)

df_bad <- data.frame(x = 1:3, x = 4:6, `2bad` = 7:9, 
                     check.names = FALSE)
audit_columns(df_bad)
```
</details>
