# Part V: Factors {-}

# Factor Creation & Levels {#factor-creation-levels}

<div class="chapter-summary">
**What You'll Learn:**

- What factors are and why they exist
- Creating factors correctly
- Understanding levels and labels
- Ordered vs unordered factors
- Common factor creation pitfalls

**Key Errors Covered:** 15+ factor errors

**Difficulty:** ⭐⭐ Intermediate
</div>

## Introduction

Factors are R's way of representing categorical data, but they're confusing:

```{r error=TRUE}
# This looks like it should work...
grades <- factor(c("A", "B", "C"))
grades[1] <- "D"
```

<div class="warning-box">
🟡 **WARNING**

```
Warning message:
In `[<-.factor`(`*tmp*`, 1, value = "D") :
  invalid factor level, NA generated
```
</div>

```{r}
grades  # First element became NA!
```

Let's understand factors to avoid these surprises.

## What Are Factors?

<div class="insight-box">
💡 **Key Insight: Factors Are Integers in Disguise**

```{r}
# Create a factor
colors <- factor(c("red", "blue", "red", "green"))
colors

# But underneath, it's integers!
typeof(colors)    # "integer"
as.integer(colors)  # 3 1 3 2

# The labels are stored separately
levels(colors)

# Structure revealed
str(colors)
```

**Key points:**
- Factors store data as integers (1, 2, 3, ...)
- Each integer maps to a level (label)
- Levels are stored once, data stores references
- More memory-efficient for repeated values
- Used extensively in statistical modeling

**Why factors exist:**
1. Memory efficiency (repeated strings)
2. Statistical modeling (R knows it's categorical)
3. Ordering (can be ordered or unordered)
4. Validation (only valid levels allowed)
</div>

## Factor vs Character

<div class="insight-box">
💡 **Factor vs Character Comparison**

```{r}
# Character vector
char_vec <- c("red", "blue", "red", "green")
typeof(char_vec)
class(char_vec)

# Factor
fac_vec <- factor(char_vec)
typeof(fac_vec)
class(fac_vec)

# Memory difference (with many repetitions)
x_char <- rep(c("Category A", "Category B"), 10000)
x_fac <- factor(x_char)

object.size(x_char)
object.size(x_fac)  # Much smaller!

# Statistical modeling difference
df <- data.frame(
  group = factor(c("A", "B", "A", "B")),
  value = c(10, 20, 15, 25)
)

# R knows 'group' is categorical
lm(value ~ group, data = df)
```

**When to use each:**
- **Character:** Text data, unique values, will manipulate as strings
- **Factor:** Categories, repeated values, for modeling/plotting
</div>

## Error #1: `invalid factor level, NA generated` {#invalid-factor-level}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r}
sizes <- factor(c("small", "medium", "large"))
sizes[1] <- "extra-large"  # Not in levels!
```

<div class="warning-box">
🟡 **WARNING**

```
Warning message:
In `[<-.factor`(`*tmp*`, 1, value = "extra-large") :
  invalid factor level, NA generated
```
</div>

```{r}
sizes  # Became NA!
```

### What It Means

You're trying to assign a value that's not in the factor's levels. R converts it to NA instead.

### Why This Happens

```{r}
sizes <- factor(c("small", "medium", "large"))

# Only these levels exist
levels(sizes)

# Can only assign existing levels
sizes[1] <- "medium"  # OK
sizes

# New levels not allowed
sizes[2] <- "tiny"    # Warning, becomes NA
sizes
```

### Common Causes

#### Cause 1: Typo in Assignment

```{r}
colors <- factor(c("red", "blue", "green"))
levels(colors)

# Typo: "blu" instead of "blue"
colors[2] <- "blu"
colors  # NA!
```

#### Cause 2: Case Mismatch

```{r}
sizes <- factor(c("small", "medium", "large"))

# Wrong case
sizes[1] <- "Small"  # "Small" != "small"
sizes
```

#### Cause 3: Adding New Categories

```{r}
responses <- factor(c("yes", "no", "yes"))

# Try to add new response
responses[4] <- "maybe"  # "maybe" not in original levels
responses
```

#### Cause 4: Data Merge Issues

```{r}
# Original data
df1 <- data.frame(
  group = factor(c("A", "B", "C"))
)

# New data with different level
df2 <- data.frame(
  group = factor(c("D", "E"))
)

# Combine
combined <- rbind(df1, df2)
combined$group  # D and E became NA!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Add New Level First**

```{r}
sizes <- factor(c("small", "medium", "large"))

# Add new level
levels(sizes) <- c(levels(sizes), "extra-large")
levels(sizes)

# Now assignment works
sizes[1] <- "extra-large"
sizes
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Convert to Character, Modify, Convert Back**

```{r}
sizes <- factor(c("small", "medium", "large"))

# Convert to character
sizes_char <- as.character(sizes)

# Modify freely
sizes_char[1] <- "extra-large"
sizes_char[4] <- "tiny"

# Convert back to factor
sizes_new <- factor(sizes_char)
sizes_new
levels(sizes_new)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Specify All Levels Upfront**

```{r}
# Specify all possible levels when creating
sizes <- factor(
  c("small", "medium", "large"),
  levels = c("tiny", "small", "medium", "large", "extra-large")
)

levels(sizes)

# Now any level can be assigned
sizes[1] <- "extra-large"
sizes[4] <- "tiny"
sizes
```
</div>

<div class="solution-box">
✅ **SOLUTION 4: Use forcats Package (Tidyverse)**

```{r}
library(forcats)

sizes <- factor(c("small", "medium", "large"))

# Add level dynamically
sizes <- fct_expand(sizes, "extra-large", "tiny")
levels(sizes)

sizes[1] <- "extra-large"
sizes
```
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall: Silent NA Creation**

```{r}
# Create factor
status <- factor(c("active", "inactive", "active"))

# Update many values
new_values <- c("active", "paused", "inactive")
status <- new_values  # Coerces to character!

class(status)  # Not a factor anymore!

# Or if forcing to stay factor:
status <- factor(c("active", "inactive", "active"))
status[] <- new_values  # "paused" becomes NA silently!
status
```

**Always check for NAs after factor assignment:**
```{r}
if (any(is.na(status))) {
  warning("Some values became NA - check factor levels")
}
```
</div>

## Error #2: `number of levels differs` {#number-levels-differs}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
f1 <- factor(c("a", "b", "c"))
f2 <- factor(c("a", "b"))

c(f1, f2)  # Try to combine
```

<div class="warning-box">
🟡 **WARNING**

```
Warning message:
In c.factor(f1, f2) : number of levels differs
```
</div>

```{r}
# Result loses factor structure
result <- c(f1, f2)
class(result)  # "integer" not "factor"!
```

### What It Means

Combining factors with different levels produces unexpected results.

### The Problem

```{r}
f1 <- factor(c("red", "blue"))
f2 <- factor(c("green", "yellow"))

levels(f1)
levels(f2)

# Combine - loses factor structure!
combined <- c(f1, f2)
combined  # Just integers!
class(combined)
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Convert to Character First**

```{r}
f1 <- factor(c("red", "blue"))
f2 <- factor(c("green", "yellow"))

# Convert both to character
combined <- c(as.character(f1), as.character(f2))
combined

# Convert back to factor
combined <- factor(combined)
combined
levels(combined)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use Same Levels for Both**

```{r}
# Define all levels upfront
all_levels <- c("red", "blue", "green", "yellow")

f1 <- factor(c("red", "blue"), levels = all_levels)
f2 <- factor(c("green", "yellow"), levels = all_levels)

# Now same levels
identical(levels(f1), levels(f2))

# Combine works better
combined <- c(f1, f2)
combined <- factor(combined, levels = all_levels)
combined
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use forcats::fct_c()**

```{r}
library(forcats)

f1 <- factor(c("red", "blue"))
f2 <- factor(c("green", "yellow"))

# Smart concatenation
combined <- fct_c(f1, f2)
combined
levels(combined)
```
</div>

## Error #3: `contrasts can be applied only to factors with 2 or more levels` {#contrasts-levels}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-math">🧮 MATH</span>

### The Error

```{r error=TRUE}
# Factor with only one level
single_level <- factor(c("A", "A", "A", "A"))
levels(single_level)

# Try to use in model
df <- data.frame(
  group = single_level,
  value = c(10, 20, 15, 25)
)

lm(value ~ group, data = df)
```

<div class="error-box">
🔴 **ERROR**

```
Error in `contrasts<-`(`*tmp*`, value = contr.treatment(2)) : 
  contrasts can be applied only to factors with 2 or more levels
```
</div>

### What It Means

Statistical models need at least 2 levels to compare. A single-level factor can't be used as a predictor.

### Common Causes

#### Cause 1: Accidental Filtering

```{r error=TRUE}
df <- data.frame(
  treatment = factor(c("A", "B", "A", "B", "C")),
  outcome = rnorm(5)
)

# Filter to subset
df_filtered <- df[df$treatment == "A", ]
df_filtered$treatment  # Still a factor, but only one level used

# Try to model
lm(outcome ~ treatment, data = df_filtered)
```

#### Cause 2: Data Preparation Gone Wrong

```{r error=TRUE}
# Read data
responses <- factor(c("yes", "no", "maybe", "yes"))

# Remove certain responses
clean_responses <- responses[responses != "no" & responses != "maybe"]
clean_responses  # Only "yes" left

df <- data.frame(
  response = clean_responses,
  score = c(80, 90)
)

lm(score ~ response, data = df)
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Drop Unused Levels**

```{r}
df <- data.frame(
  treatment = factor(c("A", "B", "A", "B", "C")),
  outcome = rnorm(5)
)

# Filter
df_filtered <- df[df$treatment == "A", ]

# Drop unused levels
df_filtered$treatment <- droplevels(df_filtered$treatment)
levels(df_filtered$treatment)  # Only "A" now

# Model will error (only 1 level)
# But at least levels match reality
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check Before Modeling**

```{r}
check_factor_for_modeling <- function(f) {
  # Check if factor
  if (!is.factor(f)) {
    stop("Input is not a factor")
  }
  
  # Count levels with data
  level_counts <- table(f)
  levels_with_data <- sum(level_counts > 0)
  
  if (levels_with_data < 2) {
    stop("Factor has only ", levels_with_data, 
         " level(s) with data. Need at least 2 for modeling.")
  }
  
  # Check for unused levels
  if (nlevels(f) > levels_with_data) {
    message("Factor has ", nlevels(f) - levels_with_data, 
            " unused level(s). Consider droplevels().")
  }
  
  return(TRUE)
}

# Test
single <- factor(c("A", "A"))
```

```{r error=TRUE}
check_factor_for_modeling(single)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Convert to Character If Needed**

```{r}
df <- data.frame(
  treatment = factor(c("A", "B", "A", "B", "C")),
  outcome = rnorm(5)
)

df_filtered <- df[df$treatment == "A", ]

# If you don't need it as a factor, convert
df_filtered$treatment <- as.character(df_filtered$treatment)

# Or remove from model
lm(outcome ~ 1, data = df_filtered)  # Intercept-only model
```
</div>

## Creating Factors Correctly

<div class="bestpractice-box">
🎯 **Best Practice: Factor Creation**

```{r}
# Method 1: Basic factor
sizes <- factor(c("S", "M", "L", "M", "S"))
sizes

# Method 2: Specify levels explicitly
sizes <- factor(
  c("S", "M", "L"),
  levels = c("XS", "S", "M", "L", "XL")
)
sizes
levels(sizes)  # All levels present

# Method 3: With labels (different from levels)
sizes <- factor(
  c(1, 2, 3, 2, 1),
  levels = 1:5,
  labels = c("XS", "S", "M", "L", "XL")
)
sizes

# Method 4: Ordered factor
sizes <- factor(
  c("S", "M", "L", "M", "S"),
  levels = c("XS", "S", "M", "L", "XL"),
  ordered = TRUE
)
sizes
class(sizes)  # "ordered" "factor"

# Can now compare
sizes[1] < sizes[3]  # TRUE (S < L)

# Method 5: From numeric
ages_binned <- cut(
  c(15, 25, 35, 45, 55),
  breaks = c(0, 18, 30, 50, 100),
  labels = c("Youth", "Young Adult", "Middle Age", "Senior")
)
ages_binned
```
</div>

## Levels vs Labels

<div class="insight-box">
💡 **Key Insight: Levels vs Labels**

```{r}
# Levels: What you have in the data
# Labels: What you want to display

# Example: Survey responses coded as numbers
responses <- c(1, 2, 3, 2, 1, 3)

# Wrong: Just convert to factor
bad <- factor(responses)
bad  # Shows 1, 2, 3

# Right: Provide labels
good <- factor(
  responses,
  levels = 1:3,
  labels = c("Disagree", "Neutral", "Agree")
)
good  # Shows actual meanings

# The underlying data is still integers
as.integer(good)

# But displays with labels
print(good)
levels(good)
```

**Key difference:**
- **levels**: Values in your data (what it IS)
- **labels**: Display names (what you WANT TO SHOW)
</div>

## Ordered Factors

<div class="insight-box">
💡 **Ordered vs Unordered Factors**

```{r}
# Unordered (nominal)
colors <- factor(c("red", "blue", "green"))
colors
class(colors)

# Can't compare
colors[1] < colors[2]  # Not meaningful

# Ordered (ordinal)
sizes <- ordered(c("S", "M", "L", "M", "S"),
                levels = c("S", "M", "L"))
sizes
class(sizes)

# Can compare
sizes[1] < sizes[3]  # TRUE

# Or use factor with ordered = TRUE
grades <- factor(
  c("B", "A", "C", "A"),
  levels = c("F", "D", "C", "B", "A"),
  ordered = TRUE
)
grades
grades[1] < grades[2]  # TRUE (B < A)
```

**When to use ordered:**
- Size (S < M < L)
- Grade (F < D < C < B < A)
- Likert scales (Strongly Disagree < ... < Strongly Agree)
- Any natural ordering

**When NOT to use ordered:**
- Colors (no natural order)
- Categories (no natural order)
- Nominal data
</div>

## Checking and Modifying Levels

<div class="bestpractice-box">
🎯 **Best Practice: Working with Levels**

```{r}
sizes <- factor(c("S", "M", "L", "M", "S"))

# Check levels
levels(sizes)
nlevels(sizes)

# Check for specific level
"XL" %in% levels(sizes)

# Add levels
levels(sizes) <- c(levels(sizes), "XS", "XL")
levels(sizes)

# Rename levels
sizes <- factor(c("S", "M", "L"))
levels(sizes) <- c("Small", "Medium", "Large")
sizes

# Reorder levels
sizes <- factor(c("L", "S", "M"))
sizes <- factor(sizes, levels = c("S", "M", "L"))
sizes

# Drop unused levels
sizes <- factor(c("S", "M", "L"), levels = c("XS", "S", "M", "L", "XL"))
levels(sizes)  # All 5 levels

sizes <- sizes[sizes != "L"]  # Remove L observations
levels(sizes)  # Still shows L!

sizes <- droplevels(sizes)
levels(sizes)  # Now only S and M

# Collapse levels
sizes <- factor(c("XS", "S", "M", "L", "XL"))
sizes_collapsed <- fct_collapse(sizes,
  Small = c("XS", "S"),
  Medium = "M",
  Large = c("L", "XL")
)
sizes_collapsed
```
</div>

## Common Factor Mistakes

<div class="pitfall-box">
⚠️ **Pitfall 1: Converting Factor to Numeric**

```{r}
# Factor with numeric-looking levels
scores <- factor(c("90", "85", "95", "88"))
scores

# WRONG: Direct conversion
as.numeric(scores)  # Gives 4 2 5 3 (factor codes!)

# RIGHT: Convert through character
as.numeric(as.character(scores))  # 90 85 95 88

# Or use levels
as.numeric(levels(scores))[scores]  # 90 85 95 88
```
</div>

<div class="pitfall-box">
⚠️ **Pitfall 2: Unexpected Coercion**

```{r}
# Combining factor and character
fac <- factor(c("a", "b"))
char <- c("c", "d")

combined <- c(fac, char)
class(combined)  # "character" (lost factor)
combined

# Arithmetic with factors
fac <- factor(c("1", "2", "3"))
```

```{r error=TRUE}
fac + 10  # Error!
```

```{r}
# Convert first
as.numeric(as.character(fac)) + 10
```
</div>

<div class="pitfall-box">
⚠️ **Pitfall 3: Factor Subsetting Keeps All Levels**

```{r}
sizes <- factor(c("S", "M", "L", "XL"))
levels(sizes)

# Subset to only S and M
sizes_small <- sizes[sizes %in% c("S", "M")]
sizes_small

# But levels still show L and XL!
levels(sizes_small)

# Drop unused levels
sizes_small <- droplevels(sizes_small)
levels(sizes_small)
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Factors are integers with labels** - Understanding this prevents confusion
2. **Can only assign existing levels** - Add level first or convert to character
3. **Combining factors is tricky** - Use forcats or convert to character
4. **Drop unused levels** after subsetting with `droplevels()`
5. **Specify levels explicitly** when creating factors
6. **Ordered factors** for data with natural ordering
7. **Convert through character** when converting factor to numeric

**Quick Reference:**

| Error/Warning | Cause | Fix |
|---------------|-------|-----|
| invalid factor level, NA | Assigning non-existent level | Add level first or use character |
| number of levels differs | Combining different factors | Use fct_c() or same levels |
| contrasts need 2+ levels | Single-level factor in model | Check levels before modeling |
| Wrong numeric conversion | as.numeric(factor) | as.numeric(as.character(factor)) |

**Factor Operations:**

```{r eval=FALSE}
# Creation
factor(x)
factor(x, levels = ...)
factor(x, levels = ..., labels = ...)
ordered(x, levels = ...)

# Inspection
levels(f)
nlevels(f)
is.factor(f)
is.ordered(f)

# Modification
levels(f) <- new_levels
f <- droplevels(f)
f <- factor(f, levels = new_order)

# Conversion
as.character(f)
as.numeric(as.character(f))  # If numeric-like
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
factor(x, levels = all_possible_levels)  # Explicit levels
as.character(f) %>% modify() %>% factor() # Modify as character
droplevels(f)                             # After subsetting
fct_c(f1, f2)                            # Combine factors

# ❌ Avoid
as.numeric(factor_with_numbers)          # Wrong conversion
c(factor1, factor2)                      # Loses factor structure
factor(x)                                # Without explicit levels
f[f %in% subset] without droplevels()    # Unused levels remain
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Factor Conversion**

You have:
```{r eval=FALSE}
scores <- factor(c("85", "90", "95", "88", "92"))
```

1. Convert to proper numeric values
2. Bin into letter grades (A: 90-100, B: 80-89, etc.)
3. Create ordered factor of letter grades
</div>

<div class="exercise-box">
📝 **Exercise 2: Combining Factors**

You have survey data from two sources:
```{r eval=FALSE}
survey1 <- factor(c("Agree", "Disagree", "Neutral"))
survey2 <- factor(c("Strongly Agree", "Agree", "Disagree"))
```

Combine them into one factor with all response levels.
</div>

<div class="exercise-box">
📝 **Exercise 3: Factor Validation**

Write `validate_factor(f)` that checks:
1. If input is a factor
2. If it has at least 2 levels
3. If it has unused levels
4. Returns report of issues found
</div>

<div class="exercise-box">
📝 **Exercise 4: Safe Factor Assignment**

Write `safe_assign_level(f, index, value)` that:
1. Checks if value is in levels
2. Adds level if not present
3. Assigns the value
4. Returns modified factor
5. Warns about any changes made
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
scores <- factor(c("85", "90", "95", "88", "92"))

# 1. Convert to numeric
scores_num <- as.numeric(as.character(scores))
scores_num

# 2. Bin into letter grades
letter_grades <- cut(
  scores_num,
  breaks = c(0, 60, 70, 80, 90, 100),
  labels = c("F", "D", "C", "B", "A"),
  include.lowest = TRUE
)
letter_grades

# 3. Create ordered factor
letter_grades_ordered <- ordered(
  letter_grades,
  levels = c("F", "D", "C", "B", "A")
)
letter_grades_ordered

# Can now compare
letter_grades_ordered[1] < letter_grades_ordered[3]
```

**Exercise 2:**

```{r}
library(forcats)

survey1 <- factor(c("Agree", "Disagree", "Neutral"))
survey2 <- factor(c("Strongly Agree", "Agree", "Disagree"))

# Define all possible levels
all_levels <- c("Strongly Disagree", "Disagree", "Neutral", 
                "Agree", "Strongly Agree")

# Recreate with same levels
survey1 <- factor(survey1, levels = all_levels)
survey2 <- factor(survey2, levels = all_levels)

# Combine
combined <- fct_c(survey1, survey2)
combined
levels(combined)

# Alternative: convert to character first
survey1 <- factor(c("Agree", "Disagree", "Neutral"))
survey2 <- factor(c("Strongly Agree", "Agree", "Disagree"))

combined <- c(as.character(survey1), as.character(survey2))
combined <- factor(combined, levels = all_levels)
combined
```

**Exercise 3:**

```{r}
validate_factor <- function(f) {
  issues <- list()
  
  # Check if factor
  if (!is.factor(f)) {
    issues$not_factor <- paste("Input is", class(f)[1], "not factor")
    return(issues)
  }
  
  # Check number of levels
  n_levels <- nlevels(f)
  if (n_levels < 2) {
    issues$too_few_levels <- paste("Only", n_levels, "level(s). Need at least 2 for most analyses.")
  }
  
  # Check for unused levels
  used_levels <- unique(as.character(f))
  all_levels <- levels(f)
  unused <- setdiff(all_levels, used_levels)
  
  if (length(unused) > 0) {
    issues$unused_levels <- paste("Unused levels:", 
                                  paste(unused, collapse = ", "))
  }
  
  # Report
  if (length(issues) == 0) {
    message("✓ Factor validation passed")
    return(invisible(NULL))
  } else {
    message("Factor validation issues found:")
    for (name in names(issues)) {
      message("  - ", issues[[name]])
    }
    return(invisible(issues))
  }
}

# Test
good <- factor(c("A", "B", "A", "B"))
validate_factor(good)

bad <- factor(c("A", "A", "A"), levels = c("A", "B", "C"))
validate_factor(bad)
```

**Exercise 4:**

```{r}
safe_assign_level <- function(f, index, value) {
  # Validate input
  if (!is.factor(f)) {
    stop("Input must be a factor")
  }
  
  if (index < 1 || index > length(f)) {
    stop("Index out of bounds")
  }
  
  # Check if value is in levels
  if (!value %in% levels(f)) {
    message("Adding new level: '", value, "'")
    levels(f) <- c(levels(f), value)
  }
  
  # Assign
  old_value <- as.character(f[index])
  f[index] <- value
  
  if (old_value != value) {
    message("Changed position ", index, " from '", old_value, 
            "' to '", value, "'")
  }
  
  return(f)
}

# Test
sizes <- factor(c("S", "M", "L"))

# Existing level
sizes <- safe_assign_level(sizes, 1, "M")
sizes

# New level
sizes <- safe_assign_level(sizes, 2, "XL")
sizes
levels(sizes)
```
</details>
