# Factor Operations {#factor-operations}

<div class="chapter-summary">
**What You'll Learn:**

- Recoding and relabeling factors
- Grouping and collapsing levels
- Reordering factors
- Factor arithmetic issues
- Common factor operation errors

**Key Errors Covered:** 10+ factor operation errors

**Difficulty:** ⭐⭐ Intermediate
</div>

## Introduction

Working with existing factors requires care:

```{r error=TRUE}
grades <- factor(c("A", "B", "C", "B", "A"))
grades > "B"  # Try to compare
```

<div class="error-box">
🔴 **ERROR**

```
Error in Ops.factor(grades, "B") : 
  level sets of factors are different
```
</div>

Let's master factor operations to avoid these errors.

## Recoding Factors

<div class="insight-box">
💡 **Key Insight: Recoding Strategies**

```{r}
# Original factor
responses <- factor(c("Y", "N", "Y", "N", "M"))
responses

# Method 1: Relabel levels directly
responses_v1 <- responses
levels(responses_v1) <- c("Maybe", "No", "Yes")  # Alphabetical order!
responses_v1

# Method 2: Create mapping
library(forcats)
responses_v2 <- fct_recode(responses,
  "Yes" = "Y",
  "No" = "N",
  "Maybe" = "M"
)
responses_v2

# Method 3: Convert to character, recode, factor
responses_v3 <- as.character(responses)
responses_v3[responses_v3 == "Y"] <- "Yes"
responses_v3[responses_v3 == "N"] <- "No"
responses_v3[responses_v3 == "M"] <- "Maybe"
responses_v3 <- factor(responses_v3)
responses_v3

# Method 4: Use named vector mapping
recode_map <- c(Y = "Yes", N = "No", M = "Maybe")
responses_v4 <- factor(recode_map[as.character(responses)])
responses_v4
```
</div>

## Error #1: `level sets of factors are different` {#level-sets-different}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
sizes1 <- factor(c("S", "M", "L"))
sizes2 <- factor(c("M", "L", "XL"))

sizes1 == sizes2
```

<div class="error-box">
🔴 **ERROR**

```
Error in Ops.factor(sizes1, sizes2) : 
  level sets of factors are different
```
</div>

### What It Means

You're comparing factors with different level sets. R doesn't know how to match them.

### Common Causes

#### Cause 1: Comparing Different Factors

```{r error=TRUE}
color1 <- factor(c("red", "blue"))
color2 <- factor(c("green", "yellow"))

color1 == color2  # Different levels!
```

#### Cause 2: After Subsetting

```{r error=TRUE}
all_sizes <- factor(c("S", "M", "L", "XL"))

# Different subsets
small_sizes <- all_sizes[1:2]
large_sizes <- all_sizes[3:4]

# Both still have all levels
levels(small_sizes)
levels(large_sizes)

# But after droplevels...
small_sizes <- droplevels(small_sizes)
large_sizes <- droplevels(large_sizes)

levels(small_sizes)  # S, M
levels(large_sizes)  # L, XL

# Now can't compare
small_sizes[1] == large_sizes[1]
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Convert to Character**

```{r}
sizes1 <- factor(c("S", "M", "L"))
sizes2 <- factor(c("M", "L", "XL"))

# Compare as characters
as.character(sizes1) == as.character(sizes2)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Unify Levels**

```{r}
sizes1 <- factor(c("S", "M", "L"))
sizes2 <- factor(c("M", "L", "XL"))

# Find all unique levels
all_levels <- union(levels(sizes1), levels(sizes2))
all_levels

# Recreate with same levels
sizes1 <- factor(sizes1, levels = all_levels)
sizes2 <- factor(sizes2, levels = all_levels)

# Now can compare
sizes1 == sizes2
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use %in% Instead**

```{r}
sizes1 <- factor(c("S", "M", "L"))
sizes2 <- factor(c("M", "L", "XL"))

# Check membership
as.character(sizes1) %in% as.character(sizes2)
```
</div>

## Error #2: `NAs produced` in factor operations {#factor-nas-produced}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Warning

```{r}
sizes <- factor(c("S", "M", "L"))
levels(sizes) <- c("Small", "Large")  # Only 2 new names for 3 levels!
```

<div class="warning-box">
🟡 **WARNING**

```
Warning message:
In `levels<-.factor`(`*tmp*`, value = c("Small", "Large")) :
  longer object length is not a multiple of replacement length
```
</div>

```{r}
sizes  # What happened?
```

### What It Means

When recoding, if you don't provide the right number of new level names, unexpected things happen.

### The Problem

```{r}
# Original
sizes <- factor(c("S", "M", "L", "XL"))
levels(sizes)

# Try to simplify to 2 categories
levels(sizes) <- c("Small", "Large")
sizes  # Recycled! Not what we wanted
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Provide All New Names**

```{r}
sizes <- factor(c("S", "M", "L", "XL"))

# All four mappings
levels(sizes) <- c("Small", "Medium", "Large", "Extra Large")
sizes
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use fct_collapse() for Grouping**

```{r}
library(forcats)

sizes <- factor(c("S", "M", "L", "XL"))

# Collapse into groups
sizes_collapsed <- fct_collapse(sizes,
  Small = c("S", "M"),
  Large = c("L", "XL")
)
sizes_collapsed
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use fct_recode() for Precise Mapping**

```{r}
sizes <- factor(c("S", "M", "L", "XL"))

# Explicit recoding
sizes_recoded <- fct_recode(sizes,
  "Small" = "S",
  "Small" = "M",    # Map M to Small too
  "Large" = "L",
  "Large" = "XL"
)
sizes_recoded
```
</div>

## Collapsing Factor Levels

<div class="bestpractice-box">
🎯 **Best Practice: Grouping Levels**

```{r}
library(forcats)

# Original detailed categories
age_groups <- factor(c("0-10", "11-20", "21-30", "31-40", 
                      "41-50", "51-60", "61-70", "71+"))

# Method 1: fct_collapse (explicit groups)
age_collapsed <- fct_collapse(age_groups,
  Youth = c("0-10", "11-20"),
  Adult = c("21-30", "31-40", "41-50"),
  Senior = c("51-60", "61-70", "71+")
)
age_collapsed

# Method 2: fct_other (keep some, lump rest)
top_ages <- fct_other(age_groups,
  keep = c("21-30", "31-40", "41-50"),
  other_level = "Other"
)
top_ages

# Method 3: fct_lump (keep n most frequent)
responses <- factor(c("A", "A", "A", "B", "B", "C", "D", "E"))
lumped <- fct_lump(responses, n = 2)  # Keep top 2
lumped

# Method 4: fct_lump_min (minimum count)
lumped_min <- fct_lump_min(responses, min = 2)  # Keep if appears 2+ times
lumped_min
```
</div>

## Reordering Factors

<div class="insight-box">
💡 **Key Insight: Factor Ordering**

```{r}
library(forcats)

# Original (alphabetical by default)
grades <- factor(c("B", "A", "C", "A", "B"))
grades
levels(grades)

# Method 1: Specify order explicitly
grades <- factor(grades, levels = c("C", "B", "A"))
levels(grades)

# Method 2: By frequency
grades <- fct_infreq(grades)
levels(grades)  # Most common first

# Method 3: By another variable
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  score = c(95, 85, 90)
)
df$name <- factor(df$name)

# Order by score
df$name <- fct_reorder(df$name, df$score)
levels(df$name)  # Ordered by score

# Method 4: Reverse order
grades <- fct_rev(grades)
levels(grades)

# Method 5: Manual reordering
grades <- fct_relevel(grades, "A", "B", "C")
levels(grades)
```
</div>

## Error #3: `'ordered' must be a factor` {#ordered-must-be-factor}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
sizes <- c("S", "M", "L")  # Character vector
ordered(sizes)  # Try to make ordered
```

<div class="error-box">
🔴 **ERROR**

```
Error in ordered(sizes) : argument is not a factor
```
</div>

### What It Means

The `ordered()` function requires a factor, not a character vector.

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Convert to Factor First**

```{r}
sizes <- c("S", "M", "L")

# Option A: Two steps
sizes_fac <- factor(sizes)
sizes_ord <- ordered(sizes_fac, levels = c("S", "M", "L"))
sizes_ord

# Option B: Direct with levels
sizes_ord <- ordered(sizes, levels = c("S", "M", "L"))
sizes_ord
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use factor() with ordered = TRUE**

```{r}
sizes <- c("S", "M", "L")

# Directly to ordered factor
sizes_ord <- factor(sizes, levels = c("S", "M", "L"), ordered = TRUE)
sizes_ord
class(sizes_ord)
```
</div>

## Factor Arithmetic Errors

<div class="pitfall-box">
⚠️ **Pitfall: You Can't Do Math on Factors**

```{r error=TRUE}
# Numeric-looking factor
numbers <- factor(c("1", "2", "3", "4", "5"))

# Try arithmetic
numbers + 10
numbers * 2
mean(numbers)
```

**Why:** Factors are categorical, not numeric, even if they look like numbers.

**Solution:**
```{r}
# Convert first
numbers_numeric <- as.numeric(as.character(numbers))
numbers_numeric + 10
mean(numbers_numeric)
```
</div>

<div class="pitfall-box">
⚠️ **Pitfall: Factor Comparison Limitations**

```{r}
# Unordered factors can only check equality
colors <- factor(c("red", "blue", "green"))
colors[1] == colors[2]  # OK
```

```{r error=TRUE}
colors[1] < colors[2]   # Error!
```

```{r}
# Ordered factors can be compared
sizes <- ordered(c("S", "M", "L"), levels = c("S", "M", "L"))
sizes[1] < sizes[3]     # OK
```
</div>

## Useful forcats Functions

<div class="bestpractice-box">
🎯 **Best Practice: forcats Toolkit**

```{r}
library(forcats)

# Sample data
responses <- factor(c("Good", "Bad", "Good", "Excellent", "Bad", 
                     "Good", "Fair", "Excellent"))

# 1. Count levels
fct_count(responses)

# 2. Reorder by frequency
responses_freq <- fct_infreq(responses)
levels(responses_freq)

# 3. Reverse order
responses_rev <- fct_rev(responses_freq)
levels(responses_rev)

# 4. Recode specific levels
responses_clean <- fct_recode(responses,
  "Very Good" = "Excellent",
  "Not Good" = "Bad"
)
levels(responses_clean)

# 5. Lump rare levels
responses_lumped <- fct_lump(responses, n = 2, other_level = "Other")
table(responses_lumped)

# 6. Add new levels
responses_expanded <- fct_expand(responses, "Outstanding")
levels(responses_expanded)

# 7. Drop unused levels
sub <- responses[responses %in% c("Good", "Bad")]
levels(sub)  # Still has all levels
sub <- fct_drop(sub)
levels(sub)  # Only Good and Bad

# 8. Explicit ordering
responses_ordered <- fct_relevel(responses, 
  "Bad", "Fair", "Good", "Excellent")
levels(responses_ordered)
```
</div>

## Factor in Data Frames

<div class="insight-box">
💡 **Key Insight: Factors in Data Analysis**

```{r}
library(dplyr)

# Sample data
df <- data.frame(
  category = factor(c("A", "B", "A", "C", "B", "A")),
  value = c(10, 20, 15, 25, 22, 18)
)

# Factors work well with grouping
df %>%
  group_by(category) %>%
  summarise(mean_value = mean(value))

# But watch out for unused levels after filtering
df_filtered <- df %>%
  filter(category != "C")

levels(df_filtered$category)  # C still there!

# Drop unused
df_filtered <- df_filtered %>%
  mutate(category = droplevels(category))

levels(df_filtered$category)  # Now just A and B

# Or use forcats
df_filtered <- df %>%
  filter(category != "C") %>%
  mutate(category = fct_drop(category))
```
</div>

## Converting Between Types

<div class="bestpractice-box">
🎯 **Best Practice: Type Conversions**

```{r}
# Factor to character
f <- factor(c("a", "b", "c"))
as.character(f)

# Factor to numeric (for numeric-looking factors)
f_num <- factor(c("10", "20", "30"))
as.numeric(as.character(f_num))  # Correct
# NOT: as.numeric(f_num)  # Wrong! Gives 1, 2, 3

# Character to factor
ch <- c("x", "y", "z")
factor(ch)

# Numeric to factor (with labels)
nums <- c(1, 2, 3, 2, 1)
factor(nums, levels = 1:3, labels = c("Low", "Medium", "High"))

# Ordered to unordered
ord <- ordered(c("S", "M", "L"), levels = c("S", "M", "L"))
factor(ord, ordered = FALSE)

# Unordered to ordered
unord <- factor(c("S", "M", "L"))
ordered(unord, levels = c("S", "M", "L"))
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Can't compare factors with different levels** - Unify first or convert to character
2. **Recoding requires all levels** - Use fct_recode() or fct_collapse()
3. **Can't do arithmetic on factors** - Convert to numeric first
4. **Unordered factors can't use < >** - Use ordered() for ordinal data
5. **droplevels() after subsetting** - Remove unused levels
6. **forcats makes factor work easier** - Use it!
7. **Convert through character** when going to numeric

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| level sets are different | Comparing different factors | Unify levels or convert |
| NAs produced | Wrong number of level names | Use fct_recode() or provide all names |
| 'ordered' must be a factor | Using ordered() on character | Convert to factor first |
| Can't do arithmetic | Math on factor | Convert to numeric |
| Can't use < > | Unordered factor comparison | Make ordered or use character |

**forcats Essential Functions:**

```{r eval=FALSE}
# Reordering
fct_infreq()      # By frequency
fct_reorder()     # By another variable
fct_relevel()     # Manually
fct_rev()         # Reverse

# Recoding
fct_recode()      # Explicit mapping
fct_collapse()    # Group levels
fct_lump()        # Combine rare levels

# Utilities
fct_count()       # Count levels
fct_drop()        # Drop unused
fct_expand()      # Add levels
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
as.numeric(as.character(f))       # Factor to numeric
fct_recode(f, "new" = "old")     # Explicit recoding
fct_drop(f)                       # After subsetting
ordered(f, levels = c(...))       # For ordinal data

# ❌ Avoid
as.numeric(f)                     # Wrong conversion
levels(f) <- too_few_names        # Will recycle
f1 == f2                          # Without checking levels
factor(x) without levels          # Implicit ordering
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Factor Recoding**

You have:
```{r eval=FALSE}
grades <- factor(c("A+", "A", "A-", "B+", "B", "B-", "C+", "C"))
```

1. Collapse to simple letter grades (A, B, C)
2. Convert to ordered factor
3. Create numeric scale (A=4, B=3, C=2)
</div>

<div class="exercise-box">
📝 **Exercise 2: Survey Data Cleaning**

You have messy survey responses:
```{r eval=FALSE}
responses <- factor(c("yes", "Yes", "YES", "no", "No", "NO", 
                     "maybe", "Maybe"))
```

Standardize all to lowercase and create ordered factor:
Disagree < Maybe < Agree
</div>

<div class="exercise-box">
📝 **Exercise 3: Factor Comparison**

Write `safe_compare(f1, f2, op)` that:
1. Checks if factors can be compared
2. Unifies levels if needed
3. Performs comparison
4. Returns result with warnings
</div>

<div class="exercise-box">
📝 **Exercise 4: Factor Summary**

Write `factor_summary(f)` that reports:
1. Number of levels
2. Most/least common levels
3. Any unused levels
4. Whether it's ordered
5. Suggested recoding (if many levels)
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
library(forcats)

grades <- factor(c("A+", "A", "A-", "B+", "B", "B-", "C+", "C"))

# 1. Collapse to letter grades
grades_simple <- fct_collapse(grades,
  A = c("A+", "A", "A-"),
  B = c("B+", "B", "B-"),
  C = c("C+", "C")
)
grades_simple

# 2. Convert to ordered
grades_ordered <- ordered(grades_simple, levels = c("C", "B", "A"))
grades_ordered

# 3. Numeric scale
grade_to_numeric <- c(A = 4, B = 3, C = 2)
grades_numeric <- grade_to_numeric[as.character(grades_simple)]
grades_numeric
```

**Exercise 2:**

```{r}
responses <- factor(c("yes", "Yes", "YES", "no", "No", "NO", 
                     "maybe", "Maybe"))

# Standardize to lowercase
responses_clean <- tolower(as.character(responses))

# Map to agreement scale
responses_clean[responses_clean == "yes"] <- "Agree"
responses_clean[responses_clean == "no"] <- "Disagree"
responses_clean[responses_clean == "maybe"] <- "Maybe"

# Create ordered factor
responses_ordered <- ordered(
  responses_clean,
  levels = c("Disagree", "Maybe", "Agree")
)
responses_ordered
```

**Exercise 3:**

```{r}
safe_compare <- function(f1, f2, op = c("==", "!=", "<", ">", "<=", ">=")) {
  op <- match.arg(op)
  
  # Check if both are factors
  if (!is.factor(f1) || !is.factor(f2)) {
    stop("Both inputs must be factors")
  }
  
  # Check length
  if (length(f1) != length(f2)) {
    warning("Factors have different lengths: ", 
            length(f1), " vs ", length(f2))
  }
  
  # Check if levels match
  if (!identical(levels(f1), levels(f2))) {
    message("Factors have different levels. Unifying...")
    
    # Unify levels
    all_levels <- union(levels(f1), levels(f2))
    f1 <- factor(f1, levels = all_levels)
    f2 <- factor(f2, levels = all_levels)
  }
  
  # Check if ordered (for < > <= >=)
  if (op %in% c("<", ">", "<=", ">=")) {
    if (!is.ordered(f1) || !is.ordered(f2)) {
      warning("Using ordering operators on unordered factors. ",
              "Converting to ordered.")
      f1 <- ordered(f1, levels = levels(f1))
      f2 <- ordered(f2, levels = levels(f2))
    }
  }
  
  # Perform comparison
  result <- switch(op,
    "==" = f1 == f2,
    "!=" = f1 != f2,
    "<"  = f1 < f2,
    ">"  = f1 > f2,
    "<=" = f1 <= f2,
    ">=" = f1 >= f2
  )
  
  return(result)
}

# Test
f1 <- factor(c("A", "B", "C"))
f2 <- factor(c("B", "C", "D"))
safe_compare(f1, f2, "==")
```

**Exercise 4:**

```{r}
factor_summary <- function(f) {
  if (!is.factor(f)) {
    stop("Input must be a factor")
  }
  
  # Basic info
  cat("Factor Summary\n")
  cat("==============\n\n")
  
  cat("Type:", if(is.ordered(f)) "Ordered" else "Unordered", "\n")
  cat("Number of levels:", nlevels(f), "\n")
  cat("Number of observations:", length(f), "\n\n")
  
  # Level counts
  level_counts <- table(f)
  cat("Level frequencies:\n")
  print(sort(level_counts, decreasing = TRUE))
  cat("\n")
  
  # Most/least common
  cat("Most common:", names(which.max(level_counts)), 
      "(", max(level_counts), "times)\n")
  cat("Least common:", names(which.min(level_counts)), 
      "(", min(level_counts), "times)\n\n")
  
  # Unused levels
  used_levels <- levels(f)[levels(f) %in% as.character(f)]
  unused_levels <- setdiff(levels(f), used_levels)
  
  if (length(unused_levels) > 0) {
    cat("⚠ Unused levels:", paste(unused_levels, collapse = ", "), "\n")
    cat("  Consider using droplevels()\n\n")
  }
  
  # Suggestions
  if (nlevels(f) > 10) {
    cat("💡 Suggestion: Factor has", nlevels(f), "levels.\n")
    cat("   Consider grouping rare levels with fct_lump()\n")
    
    # Find rare levels (< 5% of data)
    rare <- level_counts < (0.05 * length(f))
    if (any(rare)) {
      cat("   Rare levels:", paste(names(level_counts)[rare], collapse = ", "), "\n")
    }
  }
  
  invisible(list(
    n_levels = nlevels(f),
    counts = level_counts,
    unused = unused_levels
  ))
}

# Test
grades <- factor(c("A", "B", "A", "C", "B", "A"), 
                levels = c("A", "B", "C", "D", "F"))
factor_summary(grades)
```
</details>
