# Part VI: Strings & Characters {-}

# String Basics & Common Errors {#string-basics}

<div class="chapter-summary">
**What You'll Learn:**

- Character vectors vs strings
- String creation and encoding
- Common string errors
- Quotes and escaping
- String conversion issues

**Key Errors Covered:** 15+ string errors

**Difficulty:** ⭐⭐ Intermediate
</div>

## Introduction

Strings in R seem simple but have surprising complexity:

```{r error=TRUE}
text <- "He said "Hello""  # Try to include quotes
```

<div class="error-box">
🔴 **ERROR**

```
Error: unexpected symbol in "text <- "He said "Hello"
```
</div>

Let's master string handling to avoid these pitfalls.

## String Basics

<div class="insight-box">
💡 **Key Insight: No String Type in R**

R doesn't have a separate "string" type:

```{r}
# What you think of as a "string"
text <- "hello"
typeof(text)      # "character"
class(text)       # "character"

# It's a character vector of length 1
length(text)      # 1 (one element)
nchar(text)       # 5 (five characters)

# Multiple strings
texts <- c("hello", "world")
typeof(texts)     # Still "character"
length(texts)     # 2 (two elements)
nchar(texts)      # 5 5 (characters in each)
```

**Key points:**
- R has "character vectors", not "strings"
- A "string" is a character vector of length 1
- `length()` = number of elements
- `nchar()` = number of characters in each element
</div>

## Error #1: `unexpected symbol` (quote issues) {#unexpected-symbol-quotes}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-syntax">🔤 SYNTAX</span>

### The Error

```{r error=TRUE}
text <- "She said "yes""
```

<div class="error-box">
🔴 **ERROR**

```
Error: unexpected symbol in "text <- "She said "yes"
```
</div>

### What It Means

You're trying to include quotes inside a quoted string without escaping them.

### The Problem

```{r error=TRUE}
# Double quotes inside double quotes
message <- "He said "Hello""

# Single quotes inside single quotes
message <- 'It's nice'
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use Different Quotes**

```{r}
# Use single quotes outside, double inside
message <- 'He said "Hello"'
message

# Or vice versa
message <- "It's nice"
message
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Escape Quotes**

```{r}
# Escape with backslash
message <- "He said \"Hello\""
message

# When printed, quotes show
cat(message)

# Single quote escaping
message <- 'It\'s nice'
cat(message)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use Raw Strings (R 4.0+)**

```{r}
# Raw strings (if available in your R version)
if (getRversion() >= "4.0.0") {
  message <- r"(He said "Hello" and she said 'Hi')"
  cat(message)
}
```
</div>

## Escape Sequences

<div class="insight-box">
💡 **Key Insight: Common Escape Sequences**

```{r}
# Newline
cat("Line 1\nLine 2")

# Tab
cat("Col1\tCol2\tCol3")

# Backslash itself
cat("Path: C:\\Users\\Documents")

# Quotes
cat("He said \"Hello\"")
cat('It\'s working')

# Carriage return
cat("Part 1\rPart 2")  # Overwrites

# Unicode
cat("\u03B1 \u03B2 \u03B3")  # α β γ

# Hex
cat("\x48\x65\x6C\x6C\x6F")  # Hello

# All escape sequences
cat("Newline:\nTab:\tQuote:\"Backslash:\\")
```

**Common escapes:**
- `\n` - newline
- `\t` - tab
- `\\` - backslash
- `\"` - double quote
- `\'` - single quote
- `\r` - carriage return
- `\uXXXX` - Unicode character
- `\xXX` - Hex character
</div>

## Error #2: `argument is not of mode character` {#not-mode-character}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
numbers <- c(1, 2, 3, 4, 5)
nchar(numbers)
```

<div class="error-box">
🔴 **ERROR** (in some contexts)

```
Error in nchar(numbers) : 'nchar()' requires a character vector
```
</div>

Actually, `nchar()` coerces, but other functions don't:

```{r error=TRUE}
substr(123, 1, 2)
```

<div class="error-box">
🔴 **ERROR**

```
Error in substr(123, 1, 2) : argument is not of mode character
```
</div>

### What It Means

You're passing a non-character vector to a function that requires characters.

### Common Causes

#### Cause 1: Using Numbers Directly

```{r error=TRUE}
# Try to substring a number
substr(12345, 1, 3)

# Try to split a number
strsplit(123, "")
```

#### Cause 2: After Calculation

```{r error=TRUE}
x <- 10
y <- 20
result <- x + y

# Try to manipulate as string
substr(result, 1, 1)
```

#### Cause 3: Column Type Issues

```{r error=TRUE}
df <- data.frame(id = 1:5)

# Try string operation on numeric column
substr(df$id, 1, 1)
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Convert to Character First**

```{r}
numbers <- c(1, 2, 3, 4, 5)

# Convert
numbers_char <- as.character(numbers)
nchar(numbers_char)

# Or inline
substr(as.character(12345), 1, 3)

# Extract first digit
as.numeric(substr(as.character(12345), 1, 1))
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check Type Before Operation**

```{r}
safe_substr <- function(x, start, stop) {
  if (!is.character(x)) {
    message("Converting to character")
    x <- as.character(x)
  }
  substr(x, start, stop)
}

# Test
safe_substr(12345, 1, 3)
safe_substr("hello", 1, 3)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use stringr (Auto-converts)**

```{r}
library(stringr)

# stringr functions auto-convert
str_sub(12345, 1, 3)
str_length(12345)
```
</div>

## Error #3: `invalid multibyte string` {#invalid-multibyte-string}

<span class="difficulty-advanced">⭐⭐⭐ ADVANCED</span> <span class="category-badge cat-encoding">🌐 ENCODING</span>

### The Error

```{r error=TRUE}
# Try to read file with wrong encoding
text <- readLines("file_with_utf8.txt", encoding = "latin1")
nchar(text)  # May error
```

<div class="error-box">
🔴 **ERROR**

```
Error in nchar(text) : invalid multibyte string
```
</div>

### What It Means

The string contains bytes that aren't valid in the expected encoding.

### Common Causes

#### Cause 1: Reading with Wrong Encoding

```{r eval=FALSE}
# File is UTF-8 but reading as different encoding
text <- readLines("utf8_file.txt", encoding = "latin1")

# Or vice versa
text <- readLines("latin1_file.txt", encoding = "UTF-8")
```

#### Cause 2: Locale Issues

```{r}
# Check current locale
Sys.getlocale("LC_CTYPE")

# May differ across systems
```

#### Cause 3: Pasting Invalid Bytes

```{r error=TRUE}
# Create invalid UTF-8
invalid <- rawToChar(as.raw(c(0xFF, 0xFE)))
nchar(invalid)
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Specify Correct Encoding**

```{r eval=FALSE}
# Read with correct encoding
text <- readLines("file.txt", encoding = "UTF-8")

# Or detect encoding
library(readr)
guess_encoding("file.txt")

# Then read with detected encoding
text <- readLines("file.txt", encoding = "detected_encoding")
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Convert Encoding**

```{r}
# Sample text (may need real multibyte text)
text <- "café"

# Check encoding
Encoding(text)

# Convert if needed
text_utf8 <- iconv(text, from = "latin1", to = "UTF-8")

# Or ensure UTF-8
enc2utf8(text)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Clean Invalid Characters**

```{r}
clean_string <- function(x) {
  # Try to convert to UTF-8, replacing invalid
  iconv(x, to = "UTF-8", sub = "?")
}

# Or remove invalid
clean_string_remove <- function(x) {
  iconv(x, to = "UTF-8", sub = "")
}
```
</div>

<div class="pitfall-box">
⚠️ **Platform Differences**

Encoding issues often vary by platform:

```{r eval=FALSE}
# Windows default: may be Windows-1252 or local encoding
Sys.getlocale()

# Mac/Linux default: usually UTF-8
Sys.getlocale()

# Always specify encoding when reading files:
readr::read_csv("file.csv", locale = locale(encoding = "UTF-8"))
readLines("file.txt", encoding = "UTF-8")
```
</div>

## String Creation and Manipulation

<div class="bestpractice-box">
🎯 **Best Practice: Creating and Combining Strings**

```{r}
# Creating strings
single <- "hello"
multiple <- c("hello", "world", "!")

# Combining strings
paste("hello", "world")                    # Space by default
paste("hello", "world", sep = "")          # No space
paste0("hello", "world")                   # paste with sep=""

# Collapse vector into one string
paste(c("a", "b", "c"), collapse = ", ")

# Vector operations (recycling)
paste("File", 1:5, ".txt", sep = "")

# stringr alternatives
library(stringr)
str_c("hello", "world", sep = " ")
str_c(c("a", "b", "c"), collapse = ", ")

# glue for interpolation
library(glue)
name <- "Alice"
age <- 30
glue("My name is {name} and I am {age} years old")
```
</div>

## Length vs Number of Characters

<div class="pitfall-box">
⚠️ **Common Confusion: length() vs nchar()**

```{r}
# Single string
text <- "hello"
length(text)      # 1 (one element in vector)
nchar(text)       # 5 (five characters)

# Multiple strings
texts <- c("hi", "hello", "hey")
length(texts)     # 3 (three elements)
nchar(texts)      # 2 5 3 (characters in each)

# Empty string vs NULL
empty <- ""
length(empty)     # 1 (one element)
nchar(empty)      # 0 (no characters)

nothing <- character(0)
length(nothing)   # 0 (no elements)
nchar(nothing)    # integer(0)

# Common mistake
text <- "hello world"
length(text)      # 1 (NOT 11!)
nchar(text)       # 11 (including space)

# To split into characters
strsplit(text, "")[[1]]
length(strsplit(text, "")[[1]])  # 11
```
</div>

## Empty Strings and NA

<div class="insight-box">
💡 **Key Insight: Empty vs NA vs NULL**

```{r}
# Empty string (exists but empty)
empty <- ""
length(empty)         # 1
nchar(empty)          # 0
is.na(empty)          # FALSE
empty == ""           # TRUE

# NA (missing value)
missing <- NA_character_
length(missing)       # 1
nchar(missing)        # NA (can't count characters of NA)
is.na(missing)        # TRUE

# NULL (doesn't exist)
nothing <- NULL
length(nothing)       # 0
is.null(nothing)      # TRUE

# In a vector
vec <- c("hello", "", NA, "world")
length(vec)           # 4
nchar(vec)            # 5 0 NA 5
is.na(vec)            # FALSE FALSE TRUE FALSE
vec == ""             # FALSE TRUE NA FALSE (NA propagates!)

# Testing for empty strings safely
is_empty <- function(x) {
  !is.na(x) & x == ""
}

is_empty(vec)         # FALSE TRUE FALSE FALSE
```
</div>

## Case Conversion

<div class="bestpractice-box">
🎯 **Best Practice: Case Operations**

```{r}
text <- "Hello World"

# Base R
toupper(text)
tolower(text)

# First letter (no built-in function)
capitalize <- function(x) {
  paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x)))
}
capitalize("hello")

# stringr alternatives
library(stringr)
str_to_upper(text)
str_to_lower(text)
str_to_title(text)    # Title Case

# Handle NAs better
text_with_na <- c("hello", NA, "world")
toupper(text_with_na)        # Preserves NA
str_to_upper(text_with_na)   # Also preserves NA
```
</div>

## Whitespace Issues

<div class="pitfall-box">
⚠️ **Common Pitfall: Invisible Whitespace**

```{r}
# Strings that look the same
text1 <- "hello"
text2 <- " hello"
text3 <- "hello "
text4 <- "hello\n"

# But aren't equal
text1 == text2  # FALSE (leading space)
text1 == text3  # FALSE (trailing space)
text1 == text4  # FALSE (newline)

# Hard to see!
print(text2)
print(text3)

# Better visualization
cat("[", text1, "]\n", sep = "")
cat("[", text2, "]\n", sep = "")
cat("[", text3, "]\n", sep = "")

# Trim whitespace
trimws(text2)  # Remove leading/trailing
trimws(text3)

# stringr
library(stringr)
str_trim(text2)
str_squish("hello    world")  # Remove extra internal spaces too
```
</div>

## String Comparison

<div class="insight-box">
💡 **Key Insight: String Comparison**

```{r}
# Equality
"hello" == "hello"   # TRUE
"hello" == "Hello"   # FALSE (case-sensitive)

# Lexicographic ordering
"a" < "b"           # TRUE
"apple" < "banana"  # TRUE
"10" < "2"          # TRUE (lexicographic, not numeric!)

# Vector comparison
c("a", "b") == c("a", "c")  # TRUE FALSE

# %in% for membership
"apple" %in% c("apple", "banana", "cherry")  # TRUE

# Case-insensitive comparison
tolower("Hello") == tolower("hello")  # TRUE

# Partial matching (base R)
grep("app", c("apple", "banana", "application"))  # 1 3
grepl("app", c("apple", "banana", "application")) # TRUE FALSE TRUE

# stringr
library(stringr)
str_detect(c("apple", "banana"), "app")  # TRUE FALSE
str_which(c("apple", "banana", "app"), "app")  # 1 3
```
</div>

## Type Coercion with Strings

<div class="pitfall-box">
⚠️ **Pitfall: Implicit String Coercion**

```{r}
# Combining strings and numbers
c("a", 1, "b", 2)  # All become character

# In data frames (old R)
df_old <- data.frame(
  id = 1:3,
  name = c("Alice", "Bob", "Charlie"),
  stringsAsFactors = TRUE  # Old default
)
class(df_old$name)  # "factor" (was default in R < 4.0)

# Modern R
df_new <- data.frame(
  id = 1:3,
  name = c("Alice", "Bob", "Charlie")
)
class(df_new$name)  # "character"

# Operations can coerce
x <- c(1, 2, 3)
y <- paste(x, "items")
y                   # "1 items" "2 items" "3 items"
class(y)            # "character"
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **No string type** - R has character vectors
2. **length() vs nchar()** - Elements vs characters
3. **Escape quotes** with `\` or use different quotes
4. **Encoding matters** - Specify UTF-8 when reading files
5. **Empty vs NA vs NULL** - Three different concepts
6. **Convert to character** before string operations
7. **Whitespace is invisible** - Use trimws() or str_trim()

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| unexpected symbol | Quotes not escaped | Use `\"` or different quotes |
| not of mode character | Non-character input | as.character() first |
| invalid multibyte string | Encoding mismatch | Specify correct encoding |
| Wrong comparison | Case or whitespace | tolower() and trimws() |

**Essential Functions:**

```{r eval=FALSE}
# Creation
c(), paste(), paste0(), sprintf()

# Inspection
length(), nchar(), Encoding()

# Manipulation
substr(), substring(), strsplit()
toupper(), tolower(), trimws()

# Comparison
==, %in%, grep(), grepl()

# Conversion
as.character(), toString()

# stringr equivalents (better)
str_c(), str_length(), str_sub()
str_to_upper(), str_to_lower()
str_trim(), str_squish()
str_detect(), str_which()
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
text <- 'He said "Hello"'           # Different quotes
readLines("file.txt", encoding = "UTF-8")  # Explicit encoding
trimws(text)                        # Clean whitespace
as.character(x) before string ops   # Convert first

# ❌ Avoid
text <- "He said "Hello""           # Unescaped quotes
readLines("file.txt")               # Platform-dependent encoding
Assuming no whitespace              # Invisible characters
String operations on numbers        # Type mismatch
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Quote Handling**

Create strings containing:
1. Double quotes inside single quotes
2. Single quotes inside double quotes
3. Both quote types in one string
4. A file path with backslashes
</div>

<div class="exercise-box">
📝 **Exercise 2: Length vs Characters**

You have: `texts <- c("hi", "hello", "hey")`

1. Find number of elements
2. Find characters in each
3. Find total characters
4. Find longest string
</div>

<div class="exercise-box">
📝 **Exercise 3: Clean Text**

Write `clean_text(x)` that:
1. Trims whitespace
2. Converts to consistent case
3. Removes or replaces NAs
4. Reports what was changed
</div>

<div class="exercise-box">
📝 **Exercise 4: Safe String Operations**

Write `safe_substr(x, start, stop)` that:
1. Converts to character if needed
2. Handles NAs appropriately
3. Handles out-of-bounds indices
4. Returns character vector
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
# 1. Double quotes inside single quotes
text1 <- 'She said "Hello"'
cat(text1)

# 2. Single quotes inside double quotes
text2 <- "It's a nice day"
cat(text2)

# 3. Both quote types
text3 <- "She said \"It's nice\""  # Escape double quotes
cat(text3)

# Alternative with single outside
text3_alt <- 'She said "It\'s nice"'  # Escape single quote
cat(text3_alt)

# 4. File path with backslashes
path <- "C:\\Users\\Documents\\file.txt"
cat(path)

# Or use forward slashes (works on all platforms)
path_alt <- "C:/Users/Documents/file.txt"
```

**Exercise 2:**

```{r}
texts <- c("hi", "hello", "hey")

# 1. Number of elements
num_elements <- length(texts)
num_elements

# 2. Characters in each
chars_each <- nchar(texts)
chars_each

# 3. Total characters
total_chars <- sum(nchar(texts))
total_chars

# 4. Longest string
longest <- texts[which.max(nchar(texts))]
longest

# Or get length of longest
max_length <- max(nchar(texts))
max_length

# Complete analysis
analyze_strings <- function(x) {
  list(
    n_elements = length(x),
    chars_each = nchar(x),
    total_chars = sum(nchar(x)),
    avg_chars = mean(nchar(x)),
    longest = x[which.max(nchar(x))],
    shortest = x[which.min(nchar(x))]
  )
}

analyze_strings(texts)
```

**Exercise 3:**

```{r}
clean_text <- function(x, 
                       trim = TRUE,
                       case = c("lower", "upper", "none"),
                       na_action = c("keep", "remove", "replace"),
                       na_replacement = "",
                       report = TRUE) {
  
  case <- match.arg(case)
  na_action <- match.arg(na_action)
  
  original <- x
  changes <- list()
  
  # Handle NAs
  n_na <- sum(is.na(x))
  if (n_na > 0) {
    if (na_action == "remove") {
      x <- x[!is.na(x)]
      changes$na <- paste("Removed", n_na, "NAs")
    } else if (na_action == "replace") {
      x[is.na(x)] <- na_replacement
      changes$na <- paste("Replaced", n_na, "NAs with", 
                         shQuote(na_replacement))
    } else {
      changes$na <- paste("Kept", n_na, "NAs")
    }
  }
  
  # Trim whitespace
  if (trim) {
    had_whitespace <- x != trimws(x) & !is.na(x)
    if (any(had_whitespace)) {
      x <- trimws(x)
      changes$whitespace <- paste("Trimmed whitespace from",
                                 sum(had_whitespace), "strings")
    }
  }
  
  # Case conversion
  if (case == "lower") {
    x <- tolower(x)
    changes$case <- "Converted to lowercase"
  } else if (case == "upper") {
    x <- toupper(x)
    changes$case <- "Converted to uppercase"
  }
  
  # Report
  if (report && length(changes) > 0) {
    message("Text cleaning applied:")
    for (change in changes) {
      message("  - ", change)
    }
  }
  
  return(x)
}

# Test
messy <- c("  Hello ", "WORLD", NA, " Test  ")
clean_text(messy, case = "lower", na_action = "replace", 
          na_replacement = "[missing]")
```

**Exercise 4:**

```{r}
safe_substr <- function(x, start, stop) {
  # Convert to character if needed
  if (!is.character(x)) {
    message("Converting input to character")
    x <- as.character(x)
  }
  
  # Validate indices
  if (start < 1) {
    warning("start < 1, setting to 1")
    start <- 1
  }
  
  if (stop < start) {
    warning("stop < start, returning empty strings")
    return(rep("", length(x)))
  }
  
  # Handle each element
  result <- character(length(x))
  
  for (i in seq_along(x)) {
    if (is.na(x[i])) {
      result[i] <- NA_character_
      next
    }
    
    # Get length
    len <- nchar(x[i])
    
    # Adjust stop if beyond length
    actual_stop <- min(stop, len)
    
    if (start > len) {
      result[i] <- ""
    } else {
      result[i] <- substr(x[i], start, actual_stop)
    }
  }
  
  return(result)
}

# Test
safe_substr(c("hello", "world", NA, "R"), 1, 3)
safe_substr(12345, 1, 3)  # Auto-converts
safe_substr("short", 1, 100)  # Beyond length
safe_substr("test", 10, 20)  # Start beyond length
```
</details>
