# Part VII: Functions & Programming {-}

# Function Creation & Errors {#function-creation}

<div class="chapter-summary">
**What You'll Learn:**

- Creating functions in R
- Function arguments and defaults
- Return values
- Common function creation errors
- Function documentation

**Key Errors Covered:** 18+ function errors

**Difficulty:** ⭐⭐ Intermediate to ⭐⭐⭐ Advanced
</div>

## Introduction

Functions are fundamental to R programming, but creating them has pitfalls:

```{r error=TRUE}
# Try to create a function
my_function <- function(x) {
  result <- x * 2
  # Forget to return!
}

my_function(5)  # Returns NULL!
```

Let's master function creation to avoid these issues.

## Function Basics

<div class="insight-box">
💡 **Key Insight: Functions Are Objects**

```{r}
# Functions are first-class objects
my_func <- function(x) {
  x * 2
}

# It's an object
class(my_func)
typeof(my_func)

# Can be passed around
apply_func <- function(f, value) {
  f(value)
}

apply_func(my_func, 5)

# Can be stored in lists
func_list <- list(
  double = function(x) x * 2,
  triple = function(x) x * 3
)

func_list$double(5)

# Can be returned from functions
make_multiplier <- function(n) {
  function(x) x * n
}

times_4 <- make_multiplier(4)
times_4(5)
```

**Key points:**
- Functions are objects like any other
- Can be assigned to variables
- Can be passed as arguments
- Can be returned from functions
- Can be anonymous (unnamed)
</div>

## Error #1: `could not find function` {#could-not-find-function}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-scope">🔍 SCOPE</span>

### The Error

```{r error=TRUE}
# Typo in function name
maen(c(1, 2, 3))
```

<div class="error-box">
🔴 **ERROR**

```
Error in maen(c(1, 2, 3)) : could not find function "maen"
```
</div>

### What It Means

R can't find a function with that name in the current environment or loaded packages.

### Common Causes

#### Cause 1: Typo

```{r error=TRUE}
# Spelling mistakes
summry(mtcars)
lenght(1:10)
subsett(mtcars, mpg > 20)
```

#### Cause 2: Package Not Loaded

```{r error=TRUE}
# Function from package not loaded
ggplot(mtcars, aes(x = mpg, y = hp))  # ggplot2 not loaded
```

#### Cause 3: Function Not Created Yet

```{r error=TRUE}
# Calling before definition
result <- calculate_total(10, 20)

calculate_total <- function(a, b) {
  a + b
}
```

#### Cause 4: Wrong Scope

```{r error=TRUE}
# Function created in different environment
{
  local_func <- function(x) x * 2
}

local_func(5)  # No longer accessible
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Spelling**

```{r}
# Use tab completion in RStudio
# Type first few letters and press Tab

# Check available functions
apropos("mean")  # Find functions with "mean" in name

# Search help
??mean
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Load Required Package**

```{r}
# Check if function is in a package
help.search("ggplot")

# Load package
library(ggplot2)

# Or use package::function syntax
ggplot2::ggplot(mtcars, aes(x = mpg, y = hp))
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Check Function Exists**

```{r}
safe_call <- function(func_name, ...) {
  if (!exists(func_name, mode = "function")) {
    stop("Function '", func_name, "' not found. ",
         "Did you mean: ", paste(apropos(func_name), collapse = ", "))
  }
  
  do.call(func_name, list(...))
}

# Test
safe_call("mean", c(1, 2, 3))
```

```{r error=TRUE}
safe_call("maen", c(1, 2, 3))  # Helpful error
```
</div>

## Error #2: `argument "x" is missing, with no default` {#argument-missing}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-args">📋 ARGS</span>

### The Error

```{r error=TRUE}
my_func <- function(x, y) {
  x + y
}

my_func(5)  # Missing y!
```

<div class="error-box">
🔴 **ERROR**

```
Error in my_func(5) : argument "y" is missing, with no default
```
</div>

### What It Means

You're calling a function without providing all required arguments.

### Understanding Arguments

```{r}
# Required arguments (no default)
func1 <- function(x, y) {
  x + y
}

func1(5, 10)  # Must provide both
```

```{r error=TRUE}
func1(5)      # Error!
```

```{r}
# Optional arguments (with default)
func2 <- function(x, y = 10) {
  x + y
}

func2(5, 20)  # Can override default
func2(5)      # Uses default y = 10

# All optional
func3 <- function(x = 5, y = 10) {
  x + y
}

func3()       # Uses all defaults
func3(8)      # Override x, use default y
func3(8, 12)  # Override both
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Provide All Required Arguments**

```{r}
my_func <- function(x, y) {
  x + y
}

# Call with all arguments
my_func(5, 10)

# Named arguments (order doesn't matter)
my_func(y = 10, x = 5)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Add Default Values**

```{r}
# Make some/all arguments optional
my_func <- function(x, y = 0) {
  x + y
}

my_func(5)     # Works, y defaults to 0
my_func(5, 10) # Can still override

# Can use NULL as default
my_func2 <- function(x, y = NULL) {
  if (is.null(y)) {
    y <- x  # Default to same as x
  }
  x + y
}

my_func2(5)
my_func2(5, 3)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Check Arguments**

```{r}
my_func <- function(x, y) {
  # Check if arguments provided
  if (missing(x)) {
    stop("Argument 'x' is required")
  }
  if (missing(y)) {
    message("Argument 'y' not provided, using default of 0")
    y <- 0
  }
  
  x + y
}

my_func(5)    # Warning but works
my_func(5, 3)
```

```{r error=TRUE}
my_func()     # Clear error about x
```
</div>

## Error #3: `unused argument` {#unused-argument}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-args">📋 ARGS</span>

### The Error

```{r error=TRUE}
my_func <- function(x, y) {
  x + y
}

my_func(5, 10, 15)  # Too many arguments!
```

<div class="error-box">
🔴 **ERROR**

```
Error in my_func(5, 10, 15) : unused argument (15)
```
</div>

### What It Means

You're passing more arguments than the function accepts.

### Common Causes

#### Cause 1: Extra Arguments

```{r error=TRUE}
mean(c(1, 2, NA), na.rm = TRUE, extra = "oops")
```

#### Cause 2: Wrong Argument Name

```{r error=TRUE}
mean(c(1, 2, NA), remove_na = TRUE)  # It's na.rm not remove_na
```

#### Cause 3: Positional Confusion

```{r error=TRUE}
# substr expects (x, start, stop)
substr("hello", 1, 3, "extra")
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Remove Extra Arguments**

```{r}
# Check function signature
args(mean)

# Provide correct arguments only
mean(c(1, 2, NA), na.rm = TRUE)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use ... to Accept Extra Arguments**

```{r}
# Allow any number of additional arguments
my_func <- function(x, y, ...) {
  result <- x + y
  
  # Can pass ... to other functions
  extra_args <- list(...)
  if (length(extra_args) > 0) {
    message("Ignoring extra arguments: ", 
            paste(names(extra_args), collapse = ", "))
  }
  
  result
}

my_func(5, 10)              # Works
my_func(5, 10, z = 15)      # Works, ignores z
my_func(5, 10, 15, 20)      # Works, ignores unnamed extras
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Validate Arguments**

```{r}
my_func <- function(x, y) {
  # Capture call
  call <- match.call()
  
  # Check for unexpected arguments
  valid_args <- c("x", "y")
  provided_args <- names(call)[-1]  # Remove function name
  
  invalid <- setdiff(provided_args, valid_args)
  if (length(invalid) > 0) {
    stop("Unexpected arguments: ", paste(invalid, collapse = ", "),
         "\nValid arguments are: ", paste(valid_args, collapse = ", "))
  }
  
  x + y
}

my_func(5, 10)
```

```{r error=TRUE}
my_func(5, 10, z = 15)  # Helpful error
```
</div>

## Return Values

<div class="insight-box">
💡 **Key Insight: Return Values**

```{r}
# Implicit return (last expression)
func1 <- function(x) {
  x * 2
}
func1(5)  # Returns 10

# Explicit return
func2 <- function(x) {
  return(x * 2)
}
func2(5)  # Returns 10

# Early return
func3 <- function(x) {
  if (x < 0) {
    return(0)  # Exit early
  }
  x * 2
}
func3(-5)  # Returns 0
func3(5)   # Returns 10

# No return (returns NULL)
func4 <- function(x) {
  result <- x * 2
  # Forgot to return or print result
}
func4(5)  # Returns NULL invisibly

# Multiple values (use list)
func5 <- function(x) {
  list(
    original = x,
    doubled = x * 2,
    squared = x^2
  )
}
func5(5)

# Return NULL explicitly
func6 <- function(x) {
  if (x < 0) {
    return(NULL)
  }
  x * 2
}
func6(-5)  # NULL
func6(5)   # 10
```

**Best practices:**
- Last expression is returned automatically
- Use `return()` for early exits
- Use `invisible()` for functions with side effects
- Return lists for multiple values
</div>

## Error #4: `object of type 'closure' is not subsettable` {#closure-not-subsettable}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
my_func <- function(x) x * 2

# Try to subset a function
my_func[1]
```

<div class="error-box">
🔴 **ERROR**

```
Error in my_func[1] : object of type 'closure' is not subsettable
```
</div>

### What It Means

You're trying to subset a function as if it were a vector or list. "Closure" is R's internal name for functions.

### Common Causes

#### Cause 1: Name Collision

```{r error=TRUE}
# Accidentally named data same as function
mean <- function(x) sum(x) / length(x)

# Later, try to use mean as data
mean[1]  # Error! mean is now a function
```

#### Cause 2: Forgot to Call Function

```{r error=TRUE}
data <- c(1, 2, 3, 4, 5)

# Forgot parentheses
result <- mean  # Assigns the function, not result
result[1]       # Error!
```

#### Cause 3: Wrong Object

```{r error=TRUE}
my_list <- list(a = 1, b = 2)
my_func <- function(x) x * 2

# Accidentally use function instead of list
my_func$a
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Object Type**

```{r}
my_func <- function(x) x * 2

# Check what it is
is.function(my_func)
class(my_func)

# If you need to subset, make sure you're using the right object
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Call Function Properly**

```{r}
# Call the function
result <- mean(c(1, 2, 3, 4, 5))
result

# Now can use result
result
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Avoid Name Collisions**

```{r}
# Don't overwrite common function names
# Bad
# mean <- my_data

# Good
my_mean_value <- mean(my_data)

# If you accidentally overwrote
mean <- function(x) sum(x) / length(x)

# Restore
rm(mean)  # Remove your version
mean(c(1, 2, 3))  # Uses base::mean again
```
</div>

## Function Arguments: ... (Dots)

<div class="insight-box">
💡 **Key Insight: The ... Argument**

```{r}
# ... captures any additional arguments
my_plot <- function(x, y, ...) {
  plot(x, y, ...)  # Pass ... to another function
}

# Can pass any plot arguments
my_plot(1:10, 1:10, col = "red", pch = 16, main = "Test")

# Access ... contents
my_func <- function(...) {
  args <- list(...)
  cat("Received", length(args), "arguments\n")
  args
}

my_func(a = 1, b = 2, c = 3)

# Extract specific arguments from ...
my_func2 <- function(x, ...) {
  dots <- list(...)
  
  # Get specific argument
  if ("multiplier" %in% names(dots)) {
    mult <- dots$multiplier
  } else {
    mult <- 1
  }
  
  x * mult
}

my_func2(5)
my_func2(5, multiplier = 3)

# Common use: wrapper functions
my_mean <- function(..., na.rm = FALSE) {
  # Add custom behavior
  message("Calculating mean...")
  
  # Pass to base function
  mean(..., na.rm = na.rm)
}

my_mean(c(1, 2, NA, 4), na.rm = TRUE)
```

**When to use ...:**
- Wrapper functions (pass args to another function)
- Flexible functions (accept varying arguments)
- Methods (S3/S4 generics often use ...)

**Caution:**
- Arguments after ... must be named explicitly
- Easy to make typos that go unnoticed
</div>

## Argument Matching

<div class="insight-box">
💡 **Key Insight: How R Matches Arguments**

```{r}
my_func <- function(first, second, third) {
  cat("first:", first, "\n")
  cat("second:", second, "\n")
  cat("third:", third, "\n")
}

# 1. Exact name match
my_func(first = 1, second = 2, third = 3)

# 2. Partial name match (not recommended!)
my_func(f = 1, s = 2, t = 3)

# 3. Positional match
my_func(1, 2, 3)

# 4. Mixed (named don't need to be in order)
my_func(third = 3, first = 1, second = 2)
my_func(1, third = 3, second = 2)

# Order of matching:
# 1. Exact name matches
# 2. Partial name matches  
# 3. Positional matches
```

**Best practices:**
- Use exact names for clarity
- Avoid partial matching (can cause confusion)
- Use names for all arguments after the first few
- Named arguments can be in any order
</div>

## Common Function Patterns

<div class="bestpractice-box">
🎯 **Best Practice: Function Patterns**

```{r}
# 1. Validate inputs
safe_divide <- function(x, y) {
  # Check types
  if (!is.numeric(x) || !is.numeric(y)) {
    stop("Both x and y must be numeric")
  }
  
  # Check values
  if (any(y == 0)) {
    stop("Cannot divide by zero")
  }
  
  x / y
}

safe_divide(10, 2)
```

```{r error=TRUE}
safe_divide(10, 0)
```

```{r}
# 2. Provide informative messages
verbose_mean <- function(x, na.rm = FALSE, verbose = TRUE) {
  if (verbose) {
    message("Calculating mean of ", length(x), " values")
    if (na.rm) {
      message("Removing NA values")
    }
  }
  
  mean(x, na.rm = na.rm)
}

verbose_mean(c(1, 2, NA, 4), na.rm = TRUE)

# 3. Handle edge cases
robust_max <- function(x) {
  if (length(x) == 0) {
    return(NULL)
  }
  
  if (all(is.na(x))) {
    return(NA)
  }
  
  max(x, na.rm = TRUE)
}

robust_max(numeric(0))
robust_max(c(NA, NA))
robust_max(c(1, 2, NA, 3))

# 4. Return useful objects
detailed_summary <- function(x) {
  result <- list(
    mean = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE),
    n = length(x),
    n_missing = sum(is.na(x))
  )
  
  class(result) <- "detailed_summary"
  result
}

summary_obj <- detailed_summary(c(1, 2, NA, 4, 5))
summary_obj

# 5. Use ... appropriately
flexible_plot <- function(x, y, type = "p", ...) {
  # Set defaults
  defaults <- list(
    pch = 16,
    col = "blue"
  )
  
  # Override with ...
  args <- modifyList(defaults, list(...))
  
  # Call plot
  do.call(plot, c(list(x = x, y = y, type = type), args))
}

flexible_plot(1:10, 1:10, col = "red", main = "Test")
```
</div>

## Documentation

<div class="bestpractice-box">
🎯 **Best Practice: Document Your Functions**

```{r}
#' Calculate the area of a circle
#'
#' @param radius Numeric. The radius of the circle.
#' @param units Character. The units of measurement (default: "cm").
#' @return Numeric. The area of the circle.
#' @examples
#' circle_area(5)
#' circle_area(10, units = "inches")
#' @export
circle_area <- function(radius, units = "cm") {
  if (!is.numeric(radius) || radius < 0) {
    stop("radius must be a non-negative number")
  }
  
  area <- pi * radius^2
  
  structure(
    area,
    units = units,
    class = c("circle_area", "numeric")
  )
}

# In-function comments
calculate_price <- function(base_price, tax_rate = 0.1, discount = 0) {
  # Validate inputs
  if (base_price < 0) stop("base_price cannot be negative")
  if (tax_rate < 0 || tax_rate > 1) stop("tax_rate must be between 0 and 1")
  if (discount < 0 || discount > 1) stop("discount must be between 0 and 1")
  
  # Apply discount
  discounted_price <- base_price * (1 - discount)
  
  # Add tax
  final_price <- discounted_price * (1 + tax_rate)
  
  # Return itemized result
  list(
    base_price = base_price,
    discount = discount,
    discounted_price = discounted_price,
    tax = discounted_price * tax_rate,
    final_price = final_price
  )
}
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Functions are objects** - Can be assigned, passed, returned
2. **could not find function** - Check spelling, loading, scope
3. **Provide required arguments** - Or add defaults
4. **Don't pass extra arguments** - Unless function uses ...
5. **Last expression is returned** - Or use return() explicitly
6. **closure not subsettable** - You're trying to subset a function
7. **Use ... for flexibility** - Pass extra args to other functions
8. **Validate inputs** - Check types and values

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| could not find function | Typo, not loaded, not defined | Check spelling, load package |
| argument missing | Required arg not provided | Provide arg or add default |
| unused argument | Too many args | Remove extra or use ... |
| closure not subsettable | Subsetting a function | Call function or use right object |

**Function Creation:**

```{r eval=FALSE}
# Basic function
my_func <- function(x, y) {
  x + y
}

# With defaults
my_func <- function(x, y = 0) {
  x + y
}

# With ...
my_func <- function(x, ...) {
  # Do something
  other_func(x, ...)
}

# With validation
my_func <- function(x) {
  if (!is.numeric(x)) stop("x must be numeric")
  x * 2
}

# Return values
my_func <- function(x) {
  list(result = x * 2, original = x)
}
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
function(x, y = 0)              # Sensible defaults
if (!is.numeric(x)) stop()      # Validate inputs
return(list(a = 1, b = 2))      # Multiple values in list
Use descriptive names            # clear_cache() not cc()

# ❌ Avoid
function(x)                     # No defaults when optional
No input validation             # Causes cryptic errors later
mean <- my_data                 # Overwriting function names
function(x, y, z, a, b, c)      # Too many arguments
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Safe Division Function**

Write `safe_divide(x, y)` that:
1. Checks both are numeric
2. Handles division by zero
3. Works with vectors
4. Returns informative errors
</div>

<div class="exercise-box">
📝 **Exercise 2: Flexible Summary**

Write `my_summary(x, ...)` that:
1. Calculates mean, median, sd
2. Accepts ... for additional stats
3. Handles NA values
4. Returns named list
</div>

<div class="exercise-box">
📝 **Exercise 3: Argument Validator**

Write `validate_args(func, ...)` that:
1. Checks if function exists
2. Validates argument types
3. Checks required args provided
4. Returns TRUE/FALSE with messages
</div>

<div class="exercise-box">
📝 **Exercise 4: Function Factory**

Write `make_adder(n)` that returns a function that adds n to its argument.

Example:
```{r eval=FALSE}
add_5 <- make_adder(5)
add_5(10)  # Should return 15
```
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
safe_divide <- function(x, y) {
  # Check types
  if (!is.numeric(x)) {
    stop("x must be numeric, got ", class(x)[1])
  }
  if (!is.numeric(y)) {
    stop("y must be numeric, got ", class(y)[1])
  }
  
  # Check for zero
  if (any(y == 0)) {
    warning("Division by zero detected, returning Inf/-Inf")
    # R handles this naturally, but we warn
  }
  
  # Check lengths match or can recycle
  if (length(x) != length(y) && length(x) != 1 && length(y) != 1) {
    if (max(length(x), length(y)) %% min(length(x), length(y)) != 0) {
      warning("Lengths not compatible for recycling: ",
              length(x), " and ", length(y))
    }
  }
  
  result <- x / y
  result
}

# Test
safe_divide(10, 2)
safe_divide(c(10, 20, 30), c(2, 4, 5))
safe_divide(10, 0)  # Warning
```

```{r error=TRUE}
safe_divide("10", 2)  # Error
```

**Exercise 2:**

```{r}
my_summary <- function(x, ..., na.rm = TRUE) {
  # Base statistics
  result <- list(
    n = length(x),
    n_missing = sum(is.na(x)),
    mean = mean(x, na.rm = na.rm),
    median = median(x, na.rm = na.rm),
    sd = sd(x, na.rm = na.rm),
    min = min(x, na.rm = na.rm),
    max = max(x, na.rm = na.rm)
  )
  
  # Additional statistics from ...
  extra_stats <- list(...)
  
  for (stat_name in names(extra_stats)) {
    stat_func <- extra_stats[[stat_name]]
    if (is.function(stat_func)) {
      result[[stat_name]] <- stat_func(x, na.rm = na.rm)
    }
  }
  
  class(result) <- "my_summary"
  result
}

# Print method
print.my_summary <- function(x, ...) {
  cat("Summary Statistics\n")
  cat("==================\n")
  for (name in names(x)) {
    cat(sprintf("%-12s: %s\n", name, 
                format(x[[name]], digits = 3)))
  }
  invisible(x)
}

# Test
data <- c(1, 2, NA, 4, 5, 6, 7, 8, 9, 10)
my_summary(data)

# With extra stats
my_summary(data, 
          IQR = IQR,
          mad = mad)
```

**Exercise 3:**

```{r}
validate_args <- function(func_name, ...) {
  # Check function exists
  if (!exists(func_name, mode = "function")) {
    message("✗ Function '", func_name, "' not found")
    return(FALSE)
  }
  
  func <- get(func_name, mode = "function")
  func_args <- formals(func)
  
  # Get provided arguments
  provided <- list(...)
  
  # Check required arguments (those without defaults)
  required_args <- names(func_args)[sapply(func_args, 
                                           function(x) class(x) == "name")]
  
  missing_required <- setdiff(required_args, names(provided))
  
  if (length(missing_required) > 0) {
    message("✗ Missing required arguments: ", 
            paste(missing_required, collapse = ", "))
    return(FALSE)
  }
  
  # Check for unexpected arguments
  if (!"..." %in% names(func_args)) {
    unexpected <- setdiff(names(provided), names(func_args))
    if (length(unexpected) > 0) {
      message("✗ Unexpected arguments: ", 
              paste(unexpected, collapse = ", "))
      return(FALSE)
    }
  }
  
  # Check argument types (basic)
  for (arg_name in names(provided)) {
    arg_value <- provided[[arg_name]]
    # Could add more sophisticated type checking here
  }
  
  message("✓ All validations passed for '", func_name, "'")
  return(TRUE)
}

# Test
validate_args("mean", x = c(1, 2, 3), na.rm = TRUE)
validate_args("mean", na.rm = TRUE)  # Missing x
validate_args("mean", x = c(1, 2, 3), invalid = TRUE)  # Extra arg
```

**Exercise 4:**

```{r}
make_adder <- function(n) {
  # Return a function that adds n
  function(x) {
    x + n
  }
}

# Test
add_5 <- make_adder(5)
add_5(10)

add_100 <- make_adder(100)
add_100(50)

# Multiple uses
add_5(1:10)

# Each function retains its own n
add_5(5)
add_100(5)

# More advanced: with validation
make_adder_safe <- function(n) {
  if (!is.numeric(n) || length(n) != 1) {
    stop("n must be a single numeric value")
  }
  
  function(x) {
    if (!is.numeric(x)) {
      stop("x must be numeric")
    }
    x + n
  }
}

add_3 <- make_adder_safe(3)
add_3(10)
```
</details>
