# Scoping & Environments {#scoping-environments}

<div class="chapter-summary">
**What You'll Learn:**

- How R finds objects
- Lexical scoping rules
- Environment hierarchy
- Common scoping errors
- Global vs local variables

**Key Errors Covered:** 12+ scoping errors

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Introduction

Scoping determines how R finds objects, and it's often surprising:

```{r}
x <- 10

my_func <- function() {
  x  # Where does this x come from?
}

my_func()  # Returns 10!
```

Let's understand scoping to avoid confusion.

## Environment Basics

<div class="insight-box">
💡 **Key Insight: Environments Are Like Named Lists**

```{r}
# Create an environment
my_env <- new.env()

# Add objects
my_env$x <- 10
my_env$y <- 20

# Access objects
my_env$x

# List contents
ls(my_env)

# Every function has an environment
f <- function() {
  x <- 5
  environment()
}

f_env <- f()
ls(f_env)

# Current environment
environment()  # Usually global environment

# Parent environment
parent.env(my_env)
```

**Key points:**
- Environments contain named objects
- Every function creates a new environment
- Environments have parents
- R searches up the chain
</div>

## Error #1: `object not found` (scoping) {#object-not-found-scope}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-scope">🔍 SCOPE</span>

### The Error

```{r error=TRUE}
my_func <- function() {
  y <- 20
}

my_func()

y  # Where is y?
```

<div class="error-box">
🔴 **ERROR**

```
Error in eval(expr, envir, enclos) : object 'y' not found
```
</div>

### What It Means

The object was created in a function's environment and isn't accessible outside.

### Understanding Scope

```{r}
# Global scope
x <- 10

func1 <- function() {
  # Function scope
  y <- 20
  
  # Can see global x
  cat("x:", x, "\n")
  
  # Can see local y
  cat("y:", y, "\n")
}

func1()

# x exists globally
x

# y doesn't (was in function scope)
exists("y")  # FALSE

# Example 2: Nested functions
outer_func <- function() {
  z <- 30
  
  inner_func <- function() {
    # Can see z from parent environment
    z
  }
  
  inner_func()
}

outer_func()

# z doesn't exist here
exists("z")  # FALSE
```

### Common Causes

#### Cause 1: Expecting Function to Modify Global

```{r}
x <- 10

# This doesn't modify global x!
add_ten <- function() {
  x <- x + 10
  x  # Returns 20
}

add_ten()  # Returns 20
x          # Still 10!
```

#### Cause 2: Variables Only Exist in Loop Scope

```{r error=TRUE}
for (i in 1:5) {
  temp <- i * 2
}

# temp still exists (for loop doesn't create new scope in R)
temp

# But function scope is different
func <- function() {
  for (i in 1:5) {
    temp2 <- i * 2
  }
}
func()

temp2  # Doesn't exist
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Return Values**

```{r}
x <- 10

# Return the new value
add_ten <- function(val) {
  val + 10
}

# Assign result
x <- add_ten(x)
x  # Now 20
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use <<- for Global Assignment (Caution!)**

```{r}
x <- 10

# Modifies global x
add_ten_global <- function() {
  x <<- x + 10
}

add_ten_global()
x  # Now 20

# ⚠️ Use sparingly! Can cause confusion
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Pass and Return Explicitly**

```{r}
# Best practice: explicit in and out
process_data <- function(data, operation) {
  # Do something
  result <- data * operation
  
  # Return explicitly
  return(result)
}

my_data <- 100
my_data <- process_data(my_data, 2)
my_data
```
</div>

## Lexical Scoping

<div class="insight-box">
💡 **Key Insight: Lexical Scoping**

R uses lexical (static) scoping: where a function was **defined** matters, not where it's **called**.

```{r}
x <- 10

func1 <- function() {
  x  # Uses x from where function was defined
}

func2 <- function() {
  x <- 20  # Different x
  func1()  # What does this return?
}

func2()  # Returns 10 (not 20!)

# Why? func1 looks for x where it was defined (global env)
# Not where it was called (inside func2)

# Example 2: Function factories
make_multiplier <- function(n) {
  function(x) {
    x * n  # n from make_multiplier's environment
  }
}

times_3 <- make_multiplier(3)
times_5 <- make_multiplier(5)

times_3(10)  # 30
times_5(10)  # 50

# Each function remembers its own n!
```

**Key rule:** R looks for variables in:
1. Current environment
2. Parent environment (where function was defined)
3. Parent's parent, etc.
4. Eventually global environment
5. Loaded packages
</div>

## Error #2: Unexpected Value from Outer Scope

<span class="difficulty-advanced">⭐⭐⭐ ADVANCED</span> <span class="category-badge cat-scope">🔍 SCOPE</span>

### The Problem

```{r}
x <- 999  # Global x

calculate <- function(a, b) {
  # Forgot to define x, uses global!
  a + b + x
}

calculate(1, 2)  # Returns 1002 (not 3!)
```

### What Happened

The function used a global variable unintentionally.

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Explicit Parameters**

```{r}
# Make all inputs explicit
calculate <- function(a, b, x = 0) {
  a + b + x
}

calculate(1, 2)      # Uses default x = 0
calculate(1, 2, 10)  # Uses provided x
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check for Undefined Variables**

```{r}
# Development tool
check_function_variables <- function(func) {
  # Get function body
  body_expr <- body(func)
  
  # Get all variable names (simplified)
  all_vars <- all.names(body_expr)
  
  # Get formal arguments
  args <- names(formals(func))
  
  # Find variables not in arguments
  external <- setdiff(all_vars, c(args, "function", "{", "+", "-", "*", "/"))
  
  if (length(external) > 0) {
    message("Potentially external variables: ", 
            paste(external, collapse = ", "))
  }
}

x <- 999
calculate <- function(a, b) {
  a + b + x
}

check_function_variables(calculate)  # Warns about x
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use Local Functions**

```{r}
# Keep related functions together
calculator <- local({
  # Private variable
  default_offset <- 0
  
  # Public functions
  list(
    add = function(a, b) a + b + default_offset,
    set_offset = function(value) default_offset <<- value
  )
})

calculator$add(5, 10)
calculator$set_offset(100)
calculator$add(5, 10)
```
</div>

## Global Assignment: <<-

<div class="pitfall-box">
⚠️ **Pitfall: Global Assignment**

```{r}
counter <- 0

# Bad: modifies global
increment_bad <- function() {
  counter <<- counter + 1
}

increment_bad()
counter  # 1

increment_bad()
counter  # 2

# Problem: Hard to track where changes happen
# Can cause bugs in large programs

# Better: Use closures
make_counter <- function() {
  count <- 0
  
  list(
    increment = function() {
      count <<- count + 1  # OK here (modifying enclosing env)
      count
    },
    get = function() count,
    reset = function() count <<- 0
  )
}

counter_obj <- make_counter()
counter_obj$increment()
counter_obj$increment()
counter_obj$get()
counter_obj$reset()
counter_obj$get()
```

**When to use <<-:**
- Inside closures/function factories
- For memoization
- When you truly need state

**When NOT to use <<-:**
- In regular functions (use return values instead)
- When you can pass arguments
- In package functions (very rarely appropriate)
</div>

## Environments in Functions

<div class="insight-box">
💡 **Key Insight: Function Environments**

```{r}
# Each function call creates new environment
create_accumulator <- function() {
  sum <- 0
  
  function(x) {
    sum <<- sum + x
    sum
  }
}

# Create two independent accumulators
acc1 <- create_accumulator()
acc2 <- create_accumulator()

# Each has its own sum!
acc1(5)   # 5
acc1(10)  # 15
acc2(3)   # 3
acc2(7)   # 10

acc1(0)   # 15 (independent)
acc2(0)   # 10 (independent)

# Inspect environments
environment(acc1)
environment(acc2)
# Different!

# Get value from environment
get("sum", environment(acc1))
get("sum", environment(acc2))
```
</div>

## Common Scoping Patterns

<div class="bestpractice-box">
🎯 **Best Practice: Scoping Patterns**

```{r}
# 1. Function factories
make_power <- function(n) {
  function(x) {
    x ^ n
  }
}

square <- make_power(2)
cube <- make_power(3)

square(5)
cube(5)

# 2. Memoization (caching results)
fib_memo <- local({
  cache <- list()
  
  function(n) {
    if (n <= 1) return(n)
    
    # Check cache
    key <- as.character(n)
    if (!is.null(cache[[key]])) {
      return(cache[[key]])
    }
    
    # Calculate and cache
    result <- fib_memo(n - 1) + fib_memo(n - 2)
    cache[[key]] <<- result
    result
  }
})

system.time(fib_memo(30))
system.time(fib_memo(30))  # Much faster (cached)

# 3. Private variables
create_account <- function(initial_balance = 0) {
  balance <- initial_balance  # Private
  
  list(
    deposit = function(amount) {
      if (amount <= 0) stop("Amount must be positive")
      balance <<- balance + amount
      invisible(balance)
    },
    
    withdraw = function(amount) {
      if (amount > balance) stop("Insufficient funds")
      balance <<- balance - amount
      invisible(balance)
    },
    
    get_balance = function() {
      balance
    }
  )
}

account <- create_account(100)
account$deposit(50)
account$withdraw(30)
account$get_balance()

# Can't access balance directly
# account$balance  # NULL (not accessible)

# 4. Package-like namespacing
my_package <- local({
  # Private function
  helper <- function(x) {
    x * 2
  }
  
  # Public functions
  list(
    public_func1 = function(x) {
      helper(x) + 1
    },
    
    public_func2 = function(x) {
      helper(x) - 1
    }
  )
})

my_package$public_func1(5)  # Works
# my_package$helper(5)       # NULL (private)
```
</div>

## Search Path

<div class="insight-box">
💡 **Key Insight: Search Path**

```{r}
# Where R looks for objects
search()

# Order matters!
# 1. Global environment
# 2. Loaded packages (in order)
# 3. Base packages

# Example: name conflicts
library(dplyr)

# Both have filter()
# Which one gets used?
filter  # Shows dplyr::filter

# Use package::function to be explicit
stats::filter  # Base R version
dplyr::filter  # dplyr version

# Check where function comes from
find("filter")
```
</div>

## Debugging Scope Issues

<div class="bestpractice-box">
🎯 **Best Practice: Debug Scoping**

```{r}
# 1. Check where you are
debug_env <- function() {
  cat("Current environment:\n")
  print(environment())
  
  cat("\nParent environment:\n")
  print(parent.env(environment()))
  
  cat("\nObjects in current env:\n")
  print(ls())
}

my_func <- function(x) {
  y <- 10
  debug_env()
}

my_func(5)

# 2. Trace variable lookups
where_is <- function(name) {
  env <- parent.frame()
  
  while (!identical(env, emptyenv())) {
    if (exists(name, envir = env, inherits = FALSE)) {
      return(environmentName(env))
    }
    env <- parent.env(env)
  }
  
  "Not found"
}

x <- 10

test_func <- function() {
  where_is("x")
}

test_func()

# 3. List all variables in scope
ls.all <- function() {
  # Get all environments in search path
  envs <- search()
  
  for (env_name in envs) {
    env <- as.environment(env_name)
    objs <- ls(env)
    if (length(objs) > 0) {
      cat("\n", env_name, ":\n", sep = "")
      cat("  ", paste(head(objs, 10), collapse = ", "), "\n")
      if (length(objs) > 10) {
        cat("  ... and", length(objs) - 10, "more\n")
      }
    }
  }
}

# ls.all()  # Lists everything
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Lexical scoping** - Functions use variables from where they're defined
2. **Function environments** - Each call creates new environment
3. **Search path** - R looks up through parent environments
4. **Local before global** - Local variables shadow global ones
5. **<<- for parent environment** - Use cautiously
6. **Return values preferred** - Better than global modification
7. **Closures retain environment** - Function factories work because of this

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| object not found | Variable in wrong scope | Return from function or use <<- |
| Unexpected value | Using unintended global | Make parameters explicit |
| Function modifies global | Using <<- unintentionally | Use return values |
| Name conflicts | Same name in multiple packages | Use package::function |

**Scoping Rules:**

```{r eval=FALSE}
# R looks for variables in order:
1. Current environment
2. Parent environment (where defined, not called)
3. Parent's parent
4. ... up to global environment
5. Loaded packages
6. Base package

# Assignment
x <- value    # Creates in current environment
x <<- value   # Creates in first parent with x, or global

# Accessing
x             # Searches up environments
get("x")      # Same as above
exists("x")   # Check if exists
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
function(x, y) { x + y }         # Explicit parameters
result <- my_func(data)          # Return and assign
make_counter <- function() { }   # Closures for state

# ❌ Avoid
function() { global_var + 5 }    # Implicit global use
my_func <- function() { x <<- 5 } # Modifying global
Assuming variable exists          # Check with exists()
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Scope Exploration**

Predict the output:
```{r eval=FALSE}
x <- 10

func1 <- function() {
  x <- 20
  func2()
}

func2 <- function() {
  x
}

func1()  # What does this return?
```
</div>

<div class="exercise-box">
📝 **Exercise 2: Counter Implementation**

Create a counter using closures:
1. `increment()` - adds 1
2. `decrement()` - subtracts 1
3. `get()` - returns current value
4. `reset()` - sets to 0
</div>

<div class="exercise-box">
📝 **Exercise 3: Function Factory**

Write `make_scaler(center, scale)` that returns a function that:
1. Subtracts center from input
2. Divides by scale
3. Use with built-in datasets
</div>

<div class="exercise-box">
📝 **Exercise 4: Environment Inspector**

Write `inspect_scope()` that:
1. Shows current environment
2. Lists parent environments
3. Shows variables at each level
4. Identifies potential conflicts
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
x <- 10

func1 <- function() {
  x <- 20  # Local x in func1
  func2()
}

func2 <- function() {
  x  # Looks for x where func2 was defined (global)
}

func1()  # Returns 10 (not 20!)

# Why? Lexical scoping:
# func2 was defined in global environment
# So it looks for x in global environment (x = 10)
# NOT where it was called (inside func1 with x = 20)
```

**Exercise 2:**

```{r}
make_counter <- function(initial = 0) {
  count <- initial
  
  list(
    increment = function() {
      count <<- count + 1
      invisible(count)
    },
    
    decrement = function() {
      count <<- count - 1
      invisible(count)
    },
    
    get = function() {
      count
    },
    
    reset = function() {
      count <<- initial
      invisible(count)
    }
  )
}

# Test
counter <- make_counter(10)
counter$increment()
counter$increment()
counter$get()          # 12

counter$decrement()
counter$get()          # 11

counter$reset()
counter$get()          # 10

# Create multiple independent counters
counter1 <- make_counter(0)
counter2 <- make_counter(100)

counter1$increment()
counter2$decrement()

counter1$get()  # 1
counter2$get()  # 99
```

**Exercise 3:**

```{r}
make_scaler <- function(center = 0, scale = 1) {
  # Validate inputs
  if (!is.numeric(center) || length(center) != 1) {
    stop("center must be a single numeric value")
  }
  if (!is.numeric(scale) || length(scale) != 1 || scale == 0) {
    stop("scale must be a single non-zero numeric value")
  }
  
  # Return scaling function
  function(x) {
    if (!is.numeric(x)) {
      stop("x must be numeric")
    }
    (x - center) / scale
  }
}

# Test with mtcars
mpg_mean <- mean(mtcars$mpg)
mpg_sd <- sd(mtcars$mpg)

standardize_mpg <- make_scaler(mpg_mean, mpg_sd)

# Standardize mpg
mpg_scaled <- standardize_mpg(mtcars$mpg)

# Check: should have mean ≈ 0, sd ≈ 1
mean(mpg_scaled)
sd(mpg_scaled)

# Create different scalers
scale_0_1 <- make_scaler(
  center = min(mtcars$hp),
  scale = max(mtcars$hp) - min(mtcars$hp)
)

hp_scaled <- scale_0_1(mtcars$hp)
range(hp_scaled)  # Should be 0 to 1
```

**Exercise 4:**

```{r}
inspect_scope <- function() {
  # Get calling environment
  env <- parent.frame()
  
  cat("=== Environment Inspection ===\n\n")
  
  level <- 0
  while (!identical(env, emptyenv())) {
    # Environment name
    env_name <- environmentName(env)
    if (env_name == "") {
      env_name <- paste0("<anonymous ", level, ">")
    }
    
    cat("Level", level, ":", env_name, "\n")
    
    # List objects
    objs <- ls(env, all.names = FALSE)
    if (length(objs) > 0) {
      cat("  Objects:", paste(head(objs, 5), collapse = ", "))
      if (length(objs) > 5) {
        cat(" ... +", length(objs) - 5, "more")
      }
      cat("\n")
      
      # Check for conflicts with parent
      parent_env <- parent.env(env)
      if (!identical(parent_env, emptyenv())) {
        parent_objs <- ls(parent_env, all.names = FALSE)
        conflicts <- intersect(objs, parent_objs)
        if (length(conflicts) > 0) {
          cat("  ⚠ Shadows parent objects:", 
              paste(conflicts, collapse = ", "), "\n")
        }
      }
    } else {
      cat("  (empty)\n")
    }
    
    cat("\n")
    
    # Move to parent
    env <- parent.env(env)
    level <- level + 1
    
    # Stop at global or after reasonable depth
    if (level > 10) {
      cat("... (stopping after 10 levels)\n")
      break
    }
  }
}

# Test
test_function <- function() {
  local_var <- 123
  x <- "local x"  # Shadows global x if it exists
  
  inspect_scope()
}

x <- "global x"
test_function()
```
</details>
