# Control Flow {#control-flow}

<div class="chapter-summary">
**What You'll Learn:**

- if/else statements
- for, while, repeat loops
- break and next
- switch statements
- Common control flow errors

**Key Errors Covered:** 15+ control flow errors

**Difficulty:** ⭐⭐ Intermediate
</div>

## Introduction

Control flow directs code execution, but has pitfalls:

```{r error=TRUE}
x <- 5

# Looks okay...
if (x > 3 & x < 10) {
  print("In range")
}
```

```{r}
# Works, but what about this?
x <- c(5, 15)

if (x > 3 & x < 10) {  # Warning!
  print("In range")
}
```

Let's master control flow to avoid these issues.

## if/else Basics

<div class="insight-box">
💡 **Key Insight: if Requires Single Logical**

```{r}
# Correct: single TRUE/FALSE
if (TRUE) {
  print("Yes")
}

if (5 > 3) {
  print("Five is greater")
}

# With else
x <- 10
if (x > 5) {
  print("Large")
} else {
  print("Small")
}

# else if
x <- 5
if (x > 10) {
  print("Large")
} else if (x > 5) {
  print("Medium")
} else {
  print("Small")
}

# ifelse for vectors (different!)
x <- c(3, 7, 12)
ifelse(x > 5, "Large", "Small")
```

**Key points:**
- `if` needs single logical value
- `ifelse()` is vectorized (for vectors)
- `if` can have `else` and `else if`
- Braces `{}` recommended even for single lines
</div>

## Error #1: `the condition has length > 1` {#condition-length-gt-1}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-logic">🧠 LOGIC</span>

### The Error

```{r error=TRUE}
x <- c(5, 15)

if (x > 10) {
  print("Greater than 10")
}
```

<div class="error-box">
🔴 **ERROR** (R >= 4.2)

```
Error in if (x > 10) { : the condition has length > 1
```
</div>

In older R versions, this gives a warning and uses only the first element.

### What It Means

You're using a vector in `if` when it expects a single TRUE/FALSE.

### Common Causes

#### Cause 1: Testing Vector

```{r error=TRUE}
ages <- c(15, 25, 35)

if (ages >= 18) {
  print("All adults")
}
```

#### Cause 2: Multiple Conditions

```{r error=TRUE}
x <- 5
y <- 10

# Wrong: creates vector of length 2
if (c(x > 3, y > 8)) {
  print("Both true")
}
```

#### Cause 3: Using | instead of ||

```{r}
x <- c(5, 15)

# & and | are vectorized
x > 3 & x < 10  # Vector of 2 elements

# && and || use only first element
x > 3 && x < 10  # Single logical (from first element)
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use && or || for Scalar Conditions**

```{r}
x <- 5
y <- 10

# Logical AND
if (x > 3 && y > 8) {
  print("Both conditions true")
}

# Logical OR  
if (x < 3 || y > 8) {
  print("At least one true")
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use all() or any() for Vectors**

```{r}
ages <- c(15, 25, 35)

# Check if all are adults
if (all(ages >= 18)) {
  print("All adults")
} else {
  print("Some minors")
}

# Check if any are adults
if (any(ages >= 18)) {
  print("At least one adult")
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use ifelse() for Vectorized Operations**

```{r}
ages <- c(15, 25, 35)

# Vectorized: returns vector
status <- ifelse(ages >= 18, "Adult", "Minor")
status

# Or with dplyr::case_when (cleaner for multiple conditions)
library(dplyr)
case_when(
  ages < 13 ~ "Child",
  ages < 18 ~ "Teen",
  ages < 65 ~ "Adult",
  TRUE ~ "Senior"
)
```
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall: & vs && and | vs ||**

```{r}
# For if statements (scalar)
x <- 5
y <- 10

# Use && and || (short-circuit evaluation)
if (x > 3 && y > 8) {
  print("Both true")
}

# For vectors
v1 <- c(TRUE, FALSE, TRUE)
v2 <- c(TRUE, TRUE, FALSE)

# Use & and | (element-wise)
v1 & v2
v1 | v2

# Never use && or || on vectors in if!
```

**Key difference:**
- `&` and `|`: Vectorized, return vector
- `&&` and `||`: Scalar, return single value, short-circuit
</div>

## Error #2: `argument is of length zero` {#argument-length-zero}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-logic">🧠 LOGIC</span>

### The Error

```{r error=TRUE}
x <- numeric(0)  # Empty vector

if (x > 5) {
  print("Greater")
}
```

<div class="error-box">
🔴 **ERROR**

```
Error in if (x > 5) { : argument is of length zero
```
</div>

### What It Means

You're testing an empty vector in `if`.

### Common Causes

#### Cause 1: Empty Result from Operation

```{r error=TRUE}
data <- c(1, 2, 3, 4)
filtered <- data[data > 10]  # Empty!

if (filtered > 0) {
  print("Has values")
}
```

#### Cause 2: Missing Data

```{r error=TRUE}
value <- NA
result <- value[!is.na(value)]  # Empty if all NA

if (length(result) > 0) {  # Good
  # ...
}

if (result > 0) {  # Bad! Error if empty
  # ...
}
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Length First**

```{r}
x <- numeric(0)

if (length(x) > 0 && x[1] > 5) {
  print("First element greater than 5")
}

# Or more robustly
safe_check <- function(x, threshold) {
  if (length(x) == 0) {
    return(FALSE)
  }
  x[1] > threshold
}

if (safe_check(x, 5)) {
  print("Greater")
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Provide Default**

```{r}
x <- numeric(0)

# Use default if empty
value <- if (length(x) > 0) x[1] else 0

if (value > 5) {
  print("Greater")
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use any() or all()**

```{r}
x <- numeric(0)

# any() and all() handle empty vectors
any(x > 5)  # FALSE (no elements satisfy condition)
all(x > 5)  # TRUE (all zero elements satisfy condition!)

if (any(x > 5)) {
  print("At least one greater")
}
```
</div>

## for Loops

<div class="insight-box">
💡 **Key Insight: for Loop Patterns**

```{r}
# 1. Loop over vector
for (i in 1:5) {
  print(i)
}

# 2. Loop over elements
fruits <- c("apple", "banana", "cherry")
for (fruit in fruits) {
  print(fruit)
}

# 3. Loop with indices
for (i in seq_along(fruits)) {
  cat(i, ":", fruits[i], "\n")
}

# 4. Nested loops
for (i in 1:3) {
  for (j in 1:2) {
    cat("(", i, ",", j, ") ")
  }
  cat("\n")
}

# 5. Pre-allocate results (important for performance!)
n <- 1000
result <- numeric(n)  # Pre-allocate

for (i in 1:n) {
  result[i] <- i^2
}
```

**Best practices:**
- Use `seq_along()` instead of `1:length()`
- Pre-allocate result vectors
- Consider vectorization instead
- Use `for` for side effects (plots, files)
</div>

## Error #3: `object not found` in loops {#loop-object-not-found}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-scope">🔍 SCOPE</span>

### The Error

```{r error=TRUE}
# Empty vector
values <- numeric(0)

for (i in 1:length(values)) {
  print(values[i])
}
```

<div class="error-box">
🔴 **ERROR**

```
Error in 1:length(values) : argument of length 0
```
</div>

Wait, that's different. Let me show the real issue:

```{r error=TRUE}
# This actually works (badly)
values <- numeric(0)
for (i in 1:length(values)) {  # 1:0 creates c(1, 0)
  print(i)
}
```

### The Real Problem

```{r}
# 1:length() is dangerous with empty vectors
values <- numeric(0)
1:length(values)  # c(1, 0) - not what you want!

# Safer: seq_along()
seq_along(values)  # integer(0) - correct!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use seq_along()**

```{r}
values <- numeric(0)

# Safe with empty vectors
for (i in seq_along(values)) {
  print(values[i])
}
# Doesn't iterate (correct behavior)

# Works with non-empty too
values <- c(10, 20, 30)
for (i in seq_along(values)) {
  print(values[i])
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check Length First**

```{r}
values <- numeric(0)

if (length(values) > 0) {
  for (i in 1:length(values)) {
    print(values[i])
  }
} else {
  message("No values to process")
}
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use seq_len()**

```{r}
n <- 0

# seq_len() handles zero correctly
for (i in seq_len(n)) {
  print(i)
}
# Doesn't iterate

n <- 5
for (i in seq_len(n)) {
  print(i)
}
# Iterates 1 to 5
```
</div>

## while Loops

<div class="insight-box">
💡 **Key Insight: while vs for**

```{r}
# for: known number of iterations
for (i in 1:5) {
  print(i)
}

# while: iterate until condition false
count <- 1
while (count <= 5) {
  print(count)
  count <- count + 1
}

# while with break
count <- 1
while (TRUE) {
  print(count)
  count <- count + 1
  if (count > 5) break
}

# Infinite loop danger!
# while (TRUE) {
#   # Never breaks!
# }
```

**When to use while:**
- Unknown number of iterations
- Convergence checking
- Reading until end
- Waiting for condition
</div>

## break and next

<div class="insight-box">
💡 **Key Insight: Loop Control**

```{r}
# break: exit loop immediately
for (i in 1:10) {
  if (i > 5) break
  print(i)
}
# Prints 1-5, then stops

# next: skip to next iteration
for (i in 1:10) {
  if (i %% 2 == 0) next  # Skip even numbers
  print(i)
}
# Prints only odd numbers

# Combined
for (i in 1:20) {
  if (i > 15) break       # Stop at 15
  if (i %% 2 == 0) next   # Skip evens
  print(i)
}
# Prints odd numbers up to 15

# In nested loops
for (i in 1:3) {
  for (j in 1:3) {
    if (i == j) next  # Skip diagonal
    cat("(", i, ",", j, ") ")
  }
  cat("\n")
}
```
</div>

## repeat Loops

<div class="insight-box">
💡 **Key Insight: repeat Loop**

```{r}
# repeat: infinite loop with break
count <- 0
repeat {
  count <- count + 1
  print(count)
  if (count >= 5) break
}

# Common pattern: read until done
# repeat {
#   line <- readLines(connection, n = 1)
#   if (length(line) == 0) break
#   process(line)
# }

# Convergence checking
tolerance <- 0.001
value <- 10
repeat {
  old_value <- value
  value <- value / 2 + 1
  
  if (abs(value - old_value) < tolerance) {
    break
  }
}
cat("Converged to:", value, "\n")
```
</div>

## switch Statement

<div class="bestpractice-box">
🎯 **Best Practice: switch()**

```{r}
# Cleaner than multiple if/else
operation <- "add"

result <- switch(operation,
  add = 10 + 5,
  subtract = 10 - 5,
  multiply = 10 * 5,
  divide = 10 / 5,
  "Unknown operation"  # Default
)
result

# With functions
calculate <- function(op, x, y) {
  switch(op,
    "+" = x + y,
    "-" = x - y,
    "*" = x * y,
    "/" = x / y,
    stop("Unknown operation: ", op)
  )
}

calculate("+", 10, 5)
calculate("*", 10, 5)

# Numeric switch (uses position)
type <- 2
switch(type,
  "First",    # 1
  "Second",   # 2
  "Third"     # 3
)

# Multiple cases to same result
grade <- "B"
message <- switch(grade,
  "A" = ,
  "B" = "Good job!",
  "C" = ,
  "D" = "Need improvement",
  "F" = "Failed",
  "Invalid grade"
)
message
```
</div>

## Common Loop Patterns

<div class="bestpractice-box">
🎯 **Best Practice: Loop Patterns**

```{r}
# 1. Accumulation
total <- 0
for (i in 1:10) {
  total <- total + i
}
total

# Better: use sum()
sum(1:10)

# 2. Building a result vector
n <- 5
squares <- numeric(n)  # Pre-allocate!
for (i in 1:n) {
  squares[i] <- i^2
}
squares

# Better: vectorize
(1:5)^2

# 3. Processing with indices
data <- c(10, 20, 30, 40)
for (i in seq_along(data)) {
  cat("Element", i, "is", data[i], "\n")
}

# 4. Conditional accumulation
values <- c(1, 5, 3, 8, 2, 9, 4)
count <- 0
for (val in values) {
  if (val > 5) {
    count <- count + 1
  }
}
count

# Better: use sum()
sum(values > 5)

# 5. Early exit
find_first <- function(x, threshold) {
  for (i in seq_along(x)) {
    if (x[i] > threshold) {
      return(i)
    }
  }
  return(NA)
}

find_first(c(1, 3, 7, 2, 9), 5)

# 6. Nested iteration
matrix_data <- matrix(1:9, nrow = 3)
for (i in 1:nrow(matrix_data)) {
  for (j in 1:ncol(matrix_data)) {
    cat(matrix_data[i, j], " ")
  }
  cat("\n")
}

# Better: often use apply family or vectorization
```
</div>

## Vectorization vs Loops

<div class="pitfall-box">
⚠️ **Performance: Vectorization Usually Better**

```{r}
n <- 10000

# Loop (slow)
system.time({
  result <- numeric(n)
  for (i in 1:n) {
    result[i] <- sqrt(i)
  }
})

# Vectorized (fast)
system.time({
  result <- sqrt(1:n)
})

# When to use loops:
# 1. Sequential dependencies
fibonacci <- function(n) {
  fib <- numeric(n)
  fib[1] <- 1
  fib[2] <- 1
  for (i in 3:n) {
    fib[i] <- fib[i-1] + fib[i-2]  # Depends on previous
  }
  fib
}

# 2. Side effects (printing, plotting, file I/O)
for (i in 1:3) {
  plot(1:10, main = paste("Plot", i))
  Sys.sleep(0.1)
}

# 3. Complex logic that can't be vectorized
# 4. Early termination conditions
```
</div>

## ifelse() Details

<div class="insight-box">
💡 **Key Insight: ifelse() Behavior**

```{r}
# Basic ifelse
x <- c(1, 5, 3, 8, 2)
ifelse(x > 4, "High", "Low")

# Nested ifelse
ifelse(x < 3, "Low",
       ifelse(x < 7, "Medium", "High"))

# With NAs
x_na <- c(1, 5, NA, 8, 2)
ifelse(x_na > 4, "High", "Low")  # NA stays NA

# Type coercion in ifelse
ifelse(c(TRUE, FALSE, TRUE), 1, "No")  # Coerces to character!

# More control with dplyr::case_when
library(dplyr)
case_when(
  x < 3 ~ "Low",
  x < 7 ~ "Medium",
  TRUE ~ "High"  # Default
)

# Maintains types better
case_when(
  c(TRUE, FALSE, TRUE) ~ 1L,
  TRUE ~ NA_integer_
)
```

**Prefer case_when() for:**
- Multiple conditions
- Type preservation
- Clearer code
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **if needs single logical** - Use `&&`/`||` not `&`/`|`
2. **Check length first** - Avoid length-zero errors
3. **Use seq_along()** - Not `1:length()` in loops
4. **Pre-allocate vectors** - Important for performance
5. **break exits loop** - `next` skips iteration
6. **Vectorize when possible** - Usually faster than loops
7. **ifelse() is vectorized** - Different from `if`
8. **Use case_when()** - Cleaner than nested ifelse

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| condition has length > 1 | Vector in if | Use `all()`, `any()`, or `&&`/`||` |
| argument is of length zero | Empty vector in if | Check `length()` first |
| Infinite loop | No break condition | Add break or fix condition |
| Wrong 1:length() | Empty vector | Use `seq_along()` |

**Control Flow:**

```{r eval=FALSE}
# if/else
if (condition) {
  # code
} else if (other_condition) {
  # code  
} else {
  # code
}

# ifelse (vectorized)
ifelse(test, yes, no)

# for loop
for (i in seq_along(x)) {
  # code
}

# while loop
while (condition) {
  # code
}

# repeat loop
repeat {
  # code
  if (condition) break
}

# switch
switch(value,
  case1 = result1,
  case2 = result2,
  default
)
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
if (length(x) > 0 && x[1] > 5)     # Check length
for (i in seq_along(x))             # Safe indexing
result <- numeric(n); for...       # Pre-allocate
result <- sqrt(x)                   # Vectorize when possible

# ❌ Avoid
if (x > 5)                          # Vector in if
for (i in 1:length(x))              # Fails on empty
for... result <- c(result, new)     # Growing vector (slow)
for (i in 1:n) result[i] <- x[i]^2  # Loop when vectorization works
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Safe Condition Checker**

Write `safe_if(condition, true_val, false_val)` that:
1. Checks if condition is single logical
2. Handles NA in condition
3. Returns appropriate value
4. Gives helpful errors
</div>

<div class="exercise-box">
📝 **Exercise 2: Loop Converter**

Convert this loop to vectorized code:
```{r eval=FALSE}
x <- 1:1000
result <- numeric(length(x))
for (i in seq_along(x)) {
  if (x[i] %% 2 == 0) {
    result[i] <- x[i]^2
  } else {
    result[i] <- x[i]^3
  }
}
```
</div>

<div class="exercise-box">
📝 **Exercise 3: Find First**

Write `find_first(x, condition)` that:
1. Finds first element satisfying condition
2. Returns index and value
3. Handles case where none match
4. Uses early exit for efficiency
</div>

<div class="exercise-box">
📝 **Exercise 4: Grade Classifier**

Write `classify_grade(score)` using switch() that:
1. Converts numeric score to letter grade
2. Handles vectorized input
3. Validates input range (0-100)
4. Returns appropriate grade
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
safe_if <- function(condition, true_val, false_val) {
  # Check if condition is logical
  if (!is.logical(condition)) {
    stop("Condition must be logical, got ", class(condition)[1])
  }
  
  # Check length
  if (length(condition) == 0) {
    stop("Condition has length zero")
  }
  
  if (length(condition) > 1) {
    warning("Condition has length ", length(condition), 
            ", using first element only")
    condition <- condition[1]
  }
  
  # Handle NA
  if (is.na(condition)) {
    warning("Condition is NA, returning NA")
    return(NA)
  }
  
  # Return appropriate value
  if (condition) {
    true_val
  } else {
    false_val
  }
}

# Test
safe_if(TRUE, "yes", "no")
safe_if(5 > 3, "greater", "less")
safe_if(NA, "yes", "no")  # Warning
safe_if(c(TRUE, FALSE), "yes", "no")  # Warning
```

```{r error=TRUE}
safe_if("not logical", "yes", "no")  # Error
```

**Exercise 2:**

```{r}
# Original loop
x <- 1:1000
result_loop <- numeric(length(x))
for (i in seq_along(x)) {
  if (x[i] %% 2 == 0) {
    result_loop[i] <- x[i]^2
  } else {
    result_loop[i] <- x[i]^3
  }
}

# Vectorized version 1: ifelse
result_vec1 <- ifelse(x %% 2 == 0, x^2, x^3)

# Vectorized version 2: case_when
library(dplyr)
result_vec2 <- case_when(
  x %% 2 == 0 ~ x^2,
  TRUE ~ x^3
)

# Vectorized version 3: logical indexing
result_vec3 <- numeric(length(x))
even <- x %% 2 == 0
result_vec3[even] <- x[even]^2
result_vec3[!even] <- x[!even]^3

# Verify all give same result
all.equal(result_loop, result_vec1)
all.equal(result_loop, result_vec2)
all.equal(result_loop, result_vec3)

# Compare performance
library(microbenchmark)
microbenchmark(
  loop = {
    result <- numeric(length(x))
    for (i in seq_along(x)) {
      if (x[i] %% 2 == 0) result[i] <- x[i]^2
      else result[i] <- x[i]^3
    }
  },
  ifelse = ifelse(x %% 2 == 0, x^2, x^3),
  case_when = case_when(x %% 2 == 0 ~ x^2, TRUE ~ x^3),
  logical_index = {
    result <- numeric(length(x))
    even <- x %% 2 == 0
    result[even] <- x[even]^2
    result[!even] <- x[!even]^3
  },
  times = 100
)
```

**Exercise 3:**

```{r}
find_first <- function(x, condition) {
  # Validate inputs
  if (!is.function(condition)) {
    stop("condition must be a function")
  }
  
  if (length(x) == 0) {
    return(list(index = NA, value = NA, found = FALSE))
  }
  
  # Search with early exit
  for (i in seq_along(x)) {
    if (condition(x[i])) {
      return(list(
        index = i,
        value = x[i],
        found = TRUE
      ))
    }
  }
  
  # None found
  list(index = NA, value = NA, found = FALSE)
}

# Test
data <- c(1, 3, 5, 8, 2, 9, 4)

# Find first > 5
find_first(data, function(x) x > 5)

# Find first even
find_first(data, function(x) x %% 2 == 0)

# Find first > 100 (none)
find_first(data, function(x) x > 100)

# Empty vector
find_first(numeric(0), function(x) x > 5)

# More complex condition
find_first(c("apple", "banana", "cherry"), 
          function(x) nchar(x) > 5)
```

**Exercise 4:**

```{r}
classify_grade <- function(score) {
  # Validate input
  if (!is.numeric(score)) {
    stop("Score must be numeric")
  }
  
  # Vectorized function
  sapply(score, function(s) {
    # Check range
    if (is.na(s)) {
      return(NA_character_)
    }
    
    if (s < 0 || s > 100) {
      warning("Score ", s, " is out of range (0-100)")
      return("Invalid")
    }
    
    # Classify using switch on integer ranges
    grade_num <- cut(s, 
                    breaks = c(-Inf, 60, 70, 80, 90, Inf),
                    labels = FALSE)
    
    switch(grade_num,
      "F",  # 1: 0-59
      "D",  # 2: 60-69
      "C",  # 3: 70-79
      "B",  # 4: 80-89
      "A"   # 5: 90-100
    )
  })
}

# Test
classify_grade(85)
classify_grade(c(45, 65, 75, 85, 95))
classify_grade(c(95, NA, 105, 65))

# More detailed version with +/-
classify_grade_detailed <- function(score) {
  sapply(score, function(s) {
    if (is.na(s)) return(NA_character_)
    if (s < 0 || s > 100) return("Invalid")
    
    if (s >= 97) return("A+")
    if (s >= 93) return("A")
    if (s >= 90) return("A-")
    if (s >= 87) return("B+")
    if (s >= 83) return("B")
    if (s >= 80) return("B-")
    if (s >= 77) return("C+")
    if (s >= 73) return("C")
    if (s >= 70) return("C-")
    if (s >= 67) return("D+")
    if (s >= 63) return("D")
    if (s >= 60) return("D-")
    return("F")
  })
}

classify_grade_detailed(c(98, 88, 78, 68, 58))
```
</details>
