# Part VIII: Data Manipulation (dplyr & tidyr) {-}

# Introduction to dplyr {#dplyr-intro}

<div class="chapter-summary">
**What You'll Learn:**

- dplyr basics and philosophy
- The pipe operator (%>% and |>)
- Core dplyr verbs
- Common dplyr errors
- Tidy evaluation basics

**Key Errors Covered:** 20+ dplyr errors

**Difficulty:** ⭐⭐ Intermediate
</div>

## Introduction

dplyr revolutionized R data manipulation:

```{r error=TRUE}
library(dplyr)

# Try to filter
mtcars %>% filter(mpg > 20)
```

```{r}
# Works! But errors are common...
mtcars %>% filter(mpg > 20 & cyl = 4)  # Oops!
```

Let's master dplyr to avoid these pitfalls.

## dplyr Basics

<div class="insight-box">
💡 **Key Insight: The dplyr Philosophy**

```{r message=FALSE}
library(dplyr)

# Core verbs (functions)
# 1. filter() - keep rows
# 2. select() - keep columns
# 3. mutate() - create/modify columns
# 4. arrange() - sort rows
# 5. summarize() - aggregate data
# 6. group_by() - group for operations

# Each returns a data frame
# Easy to chain with pipe

# Example data
head(mtcars)
```

**Key principles:**
- Each function does one thing well
- First argument is always data
- Returns a data frame
- Works with pipes
- NSE (non-standard evaluation) for column names
</div>

## The Pipe Operator

<div class="insight-box">
💡 **Key Insight: Pipe Operators**

```{r}
# Traditional nesting (hard to read)
result <- arrange(
  filter(
    select(mtcars, mpg, cyl, hp),
    mpg > 20
  ),
  desc(hp)
)

# With magrittr pipe %>%
result <- mtcars %>%
  select(mpg, cyl, hp) %>%
  filter(mpg > 20) %>%
  arrange(desc(hp))

# Native pipe |> (R >= 4.1)
result <- mtcars |>
  select(mpg, cyl, hp) |>
  filter(mpg > 20) |>
  arrange(desc(hp))

result

# How pipe works
# x %>% f(y) becomes f(x, y)
# x %>% f(y) %>% g(z) becomes g(f(x, y), z)

# Pipe sends left side as first argument to right side
mtcars %>% head(3)
# Same as:
head(mtcars, 3)
```

**Differences between %>% and |>:**
- `%>%` from magrittr (more flexible)
- `|>` native R (R >= 4.1, faster)
- `%>%` has more features (., shortcuts)
- `|>` simpler, no package needed
</div>

## Error #1: `object not found` with pipes {#pipe-object-not-found}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-scope">🔍 SCOPE</span>

### The Error

```{r error=TRUE}
library(dplyr)

mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl, hp) %>%
  arrange(horsepower)  # Wrong column name!
```

<div class="error-box">
🔴 **ERROR**

```
Error in arrange():
! Can't subset columns that don't exist.
✖ Column `horsepower` doesn't exist.
```
</div>

### What It Means

You're referencing a column that doesn't exist in the data.

### Common Causes

#### Cause 1: Typo in Column Name

```{r error=TRUE}
mtcars %>%
  select(mpgg, cyl)  # Typo: mpgg instead of mpg
```

#### Cause 2: Column Already Removed

```{r error=TRUE}
mtcars %>%
  select(mpg, cyl) %>%
  filter(hp > 100)  # hp not in selected columns!
```

#### Cause 3: Wrong Dataset

```{r error=TRUE}
# Expecting different columns
iris %>%
  select(mpg, cyl)  # iris doesn't have these columns
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Column Names**

```{r}
# Before piping, check what you have
names(mtcars)

# Or with glimpse
glimpse(mtcars)

# In pipe chain
mtcars %>%
  names()
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Order Operations Correctly**

```{r}
# Filter before selecting
mtcars %>%
  filter(hp > 100) %>%  # Use hp while it exists
  select(mpg, cyl)      # Then select subset

# Or keep needed columns
mtcars %>%
  select(mpg, cyl, hp) %>%  # Keep hp
  filter(hp > 100)          # Now can use it
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use Column Existence Check**

```{r}
safe_filter <- function(data, ...) {
  # Check if columns exist
  expr <- rlang::enexpr(...)
  cols_used <- all.vars(expr)
  
  missing_cols <- setdiff(cols_used, names(data))
  
  if (length(missing_cols) > 0) {
    stop("Columns not found: ", paste(missing_cols, collapse = ", "))
  }
  
  filter(data, ...)
}

# Will give helpful error
# mtcars %>% safe_filter(horsepower > 100)
```
</div>

## filter() Basics

<div class="insight-box">
💡 **Key Insight: filter() Patterns**

```{r}
# Keep rows where condition is TRUE
mtcars %>%
  filter(mpg > 20)

# Multiple conditions (AND)
mtcars %>%
  filter(mpg > 20, cyl == 4)

# Same as
mtcars %>%
  filter(mpg > 20 & cyl == 4)

# OR condition
mtcars %>%
  filter(mpg > 25 | hp > 200)

# Complex conditions
mtcars %>%
  filter(mpg > 20 & (cyl == 4 | cyl == 6))

# Using %in%
mtcars %>%
  filter(cyl %in% c(4, 6))

# Negation
mtcars %>%
  filter(!(cyl %in% c(4, 6)))

# Or
mtcars %>%
  filter(!cyl %in% c(4, 6))

# Filter with string operations
mtcars %>%
  rownames_to_column("car") %>%
  filter(grepl("Merc", car))

# Remove NAs
data_with_na <- tibble(x = c(1, 2, NA, 4))
data_with_na %>%
  filter(!is.na(x))
```
</div>

## Error #2: Using `=` instead of `==` {#filter-equals}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-syntax">🔤 SYNTAX</span>

### The Error

```{r error=TRUE}
mtcars %>%
  filter(cyl = 4)  # Wrong: = is assignment
```

<div class="error-box">
🔴 **ERROR**

```
Error in filter():
! We detected a named input.
ℹ This usually means that you've used `=` instead of `==`.
ℹ Did you mean `cyl == 4`?
```
</div>

### What It Means

You used assignment (`=`) instead of comparison (`==`).

### Solutions

<div class="solution-box">
✅ **SOLUTION: Use == for Comparison**

```{r}
# Correct: use ==
mtcars %>%
  filter(cyl == 4)

# Other comparison operators
mtcars %>%
  filter(mpg > 20)      # Greater than

mtcars %>%
  filter(mpg >= 20)     # Greater or equal

mtcars %>%
  filter(mpg < 20)      # Less than

mtcars %>%
  filter(mpg <= 20)     # Less or equal

mtcars %>%
  filter(mpg != 4)      # Not equal

# For NA, use is.na()
data_with_na <- tibble(x = c(1, NA, 3))
data_with_na %>%
  filter(is.na(x))      # Keep NAs

data_with_na %>%
  filter(!is.na(x))     # Remove NAs
```
</div>

## select() Basics

<div class="insight-box">
💡 **Key Insight: select() Patterns**

```{r}
# Select specific columns
mtcars %>%
  select(mpg, cyl, hp)

# Select range
mtcars %>%
  select(mpg:hp)

# Select by position
mtcars %>%
  select(1:3)

# Exclude columns
mtcars %>%
  select(-mpg, -cyl)

# Exclude range
mtcars %>%
  select(-(mpg:disp))

# Helper functions
# starts_with()
iris %>%
  select(starts_with("Sepal"))

# ends_with()
iris %>%
  select(ends_with("Width"))

# contains()
iris %>%
  select(contains("etal"))

# matches() - regex
mtcars %>%
  select(matches("^d"))  # Starts with 'd'

# everything() - all remaining
mtcars %>%
  select(mpg, cyl, everything())

# where() - by condition
mtcars %>%
  select(where(is.numeric))

# Rename while selecting
mtcars %>%
  select(miles_per_gallon = mpg, cylinders = cyl)

# Just rename (keep all)
mtcars %>%
  rename(miles_per_gallon = mpg)
```
</div>

## Error #3: `unused argument` in select {#select-unused-argument}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-args">📋 ARGS</span>

### The Error

```{r error=TRUE}
mtcars %>%
  select(mpg, cyl, cyl == 4)  # Can't filter in select!
```

<div class="error-box">
🔴 **ERROR**

```
Error in select():
! Can't subset columns with `cyl == 4`.
✖ `cyl == 4` must be numeric or character, not `logical`.
```
</div>

### What It Means

You're trying to use filter logic in select, or using wrong syntax.

### Common Causes

```{r error=TRUE}
# Trying to filter in select
mtcars %>%
  select(mpg, cyl > 4)

# Wrong helper syntax
mtcars %>%
  select(starts_with(Sepal))  # Missing quotes
```

### Solutions

<div class="solution-box">
✅ **SOLUTION: Separate select and filter**

```{r}
# Correct: separate operations
mtcars %>%
  select(mpg, cyl, hp) %>%
  filter(cyl == 4)

# Or filter first
mtcars %>%
  filter(cyl == 4) %>%
  select(mpg, cyl, hp)

# Use quotes with helpers
iris %>%
  select(starts_with("Sepal"))
```
</div>

## mutate() Basics

<div class="insight-box">
💡 **Key Insight: mutate() Patterns**

```{r}
# Create new column
mtcars %>%
  mutate(mpg_per_cyl = mpg / cyl) %>%
  select(mpg, cyl, mpg_per_cyl)

# Multiple new columns
mtcars %>%
  mutate(
    kpl = mpg * 0.425,
    wt_kg = wt * 453.592
  ) %>%
  select(mpg, kpl, wt, wt_kg)

# Modify existing column
mtcars %>%
  mutate(hp = hp * 0.746) %>%  # Convert to kW
  select(hp)

# Use newly created columns
mtcars %>%
  mutate(
    hp_per_cyl = hp / cyl,
    hp_ratio = hp_per_cyl / mean(hp_per_cyl)
  ) %>%
  select(hp, cyl, hp_per_cyl, hp_ratio)

# Conditional creation
mtcars %>%
  mutate(
    efficiency = case_when(
      mpg > 25 ~ "High",
      mpg > 20 ~ "Medium",
      TRUE ~ "Low"
    )
  ) %>%
  select(mpg, efficiency)

# With ifelse
mtcars %>%
  mutate(
    heavy = ifelse(wt > 3.5, "Heavy", "Light")
  ) %>%
  select(wt, heavy)

# Keep only new columns with transmute()
mtcars %>%
  transmute(
    car = rownames(mtcars),
    kpl = mpg * 0.425
  )
```
</div>

## Error #4: `object not found` in mutate {#mutate-object-not-found}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-scope">🔍 SCOPE</span>

### The Error

```{r error=TRUE}
mpg_threshold <- 20

mtcars %>%
  mutate(efficient = mpg > mpg_treshold)  # Typo!
```

<div class="error-box">
🔴 **ERROR**

```
Error in mutate():
! object 'mpg_treshold' not found
```
</div>

### What It Means

Referenced an object or column that doesn't exist.

### Common Causes

```{r error=TRUE}
# Typo in external variable
threshold <- 20
mtcars %>%
  mutate(high = mpg > threshhold)

# Typo in column name
mtcars %>%
  mutate(power_ratio = horsepower / wt)

# Using column before creating it
mtcars %>%
  mutate(
    mpg_ratio = mpg_scaled / mean(mpg_scaled),
    mpg_scaled = mpg / mean(mpg)
  )
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Names Carefully**

```{r}
# Check column names
names(mtcars)

# Check variable exists
threshold <- 20
exists("threshold")

# Correct usage
mtcars %>%
  mutate(efficient = mpg > threshold) %>%
  select(mpg, efficient)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Order Operations**

```{r}
# Create columns in right order
mtcars %>%
  mutate(
    mpg_scaled = mpg / mean(mpg),
    mpg_ratio = mpg_scaled / mean(mpg_scaled)
  ) %>%
  select(mpg, mpg_scaled, mpg_ratio)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use .data pronoun**

```{r}
# Explicit about column reference
mtcars %>%
  mutate(power_ratio = .data$hp / .data$wt) %>%
  select(hp, wt, power_ratio)

# Useful in functions
calculate_ratio <- function(data, num_col, denom_col) {
  data %>%
    mutate(ratio = .data[[num_col]] / .data[[denom_col]])
}

calculate_ratio(mtcars, "hp", "wt") %>%
  select(hp, wt, ratio)
```
</div>

## arrange() Basics

<div class="insight-box">
💡 **Key Insight: arrange() Patterns**

```{r}
# Sort ascending (default)
mtcars %>%
  arrange(mpg) %>%
  select(mpg, cyl, hp)

# Sort descending
mtcars %>%
  arrange(desc(mpg)) %>%
  select(mpg, cyl, hp)

# Multiple columns
mtcars %>%
  arrange(cyl, desc(mpg)) %>%
  select(cyl, mpg, hp)

# With NA handling
data_with_na <- tibble(
  x = c(3, 1, NA, 2),
  y = c("a", "b", "c", "d")
)

data_with_na %>%
  arrange(x)  # NAs last by default

# NAs first
data_with_na %>%
  arrange(desc(is.na(x)), x)
```
</div>

## summarize() Basics

<div class="insight-box">
💡 **Key Insight: summarize() Patterns**

```{r}
# Single summary
mtcars %>%
  summarize(avg_mpg = mean(mpg))

# Multiple summaries
mtcars %>%
  summarize(
    avg_mpg = mean(mpg),
    sd_mpg = sd(mpg),
    min_mpg = min(mpg),
    max_mpg = max(mpg),
    n = n()
  )

# With grouping (next chapter covers in detail)
mtcars %>%
  group_by(cyl) %>%
  summarize(
    avg_mpg = mean(mpg),
    count = n()
  )

# Useful summary functions
mtcars %>%
  summarize(
    mean = mean(mpg),
    median = median(mpg),
    sd = sd(mpg),
    min = min(mpg),
    max = max(mpg),
    q25 = quantile(mpg, 0.25),
    q75 = quantile(mpg, 0.75),
    n = n(),
    n_distinct = n_distinct(cyl)
  )
```
</div>

## Common Patterns

<div class="bestpractice-box">
🎯 **Best Practice: Common dplyr Workflows**

```{r}
# 1. Filter → Select → Arrange
mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl, hp) %>%
  arrange(desc(mpg))

# 2. Create variables → Summarize
mtcars %>%
  mutate(hp_per_cyl = hp / cyl) %>%
  summarize(
    avg_hp_per_cyl = mean(hp_per_cyl),
    max_hp_per_cyl = max(hp_per_cyl)
  )

# 3. Group → Summarize → Arrange
mtcars %>%
  group_by(cyl) %>%
  summarize(
    avg_mpg = mean(mpg),
    count = n()
  ) %>%
  arrange(desc(avg_mpg))

# 4. Top N
mtcars %>%
  arrange(desc(mpg)) %>%
  slice_head(n = 5)

# 5. Distinct values
mtcars %>%
  distinct(cyl, gear)

# 6. Count
mtcars %>%
  count(cyl, sort = TRUE)

# 7. Add row numbers
mtcars %>%
  mutate(row_id = row_number()) %>%
  select(row_id, everything())
```
</div>

## Error #5: Forgetting to assign or print {#forget-assign}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-usage">💡 USAGE</span>

### The Problem

```{r}
# Doesn't save result!
mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl)

# Original unchanged
head(mtcars, 2)
```

### What It Means

dplyr doesn't modify in place - must assign result.

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Assign Result**

```{r}
# Save to new variable
high_mpg <- mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl)

head(high_mpg, 3)

# Or overwrite (careful!)
mtcars_filtered <- mtcars %>%
  filter(mpg > 20)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Just Print (for exploration)**

```{r}
# Print automatically (no assignment)
mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl)

# Explicit print
mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl) %>%
  print()
```
</div>

## Debugging dplyr Chains

<div class="bestpractice-box">
🎯 **Best Practice: Debug Pipe Chains**

```{r}
# 1. Run step by step
# Start here
mtcars %>%
  filter(mpg > 20)

# Add next step
mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl, hp)

# Continue...
mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl, hp) %>%
  arrange(desc(hp))

# 2. Check intermediate results
mtcars %>%
  filter(mpg > 20) %>%
  {cat("After filter:", nrow(.), "rows\n"); .} %>%
  select(mpg, cyl, hp) %>%
  {cat("After select:", ncol(.), "cols\n"); .} %>%
  arrange(desc(hp))

# 3. Use View() in RStudio
# mtcars %>%
#   filter(mpg > 20) %>%
#   View()

# 4. Print at each step
debug_pipe <- function(data, label = NULL) {
  if (!is.null(label)) {
    cat("\n===", label, "===\n")
  }
  print(head(data, 3))
  data
}

mtcars %>%
  filter(mpg > 20) %>%
  debug_pipe("After filter") %>%
  select(mpg, cyl, hp) %>%
  debug_pipe("After select")
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **dplyr philosophy** - Simple verbs, chainable with pipes
2. **Pipe operator** - `%>%` or `|>` makes code readable
3. **filter() uses ==** - Not `=` for comparison
4. **Check column names** - Use `names()` or `glimpse()`
5. **Order matters** - Can't use columns after removing them
6. **Assign results** - dplyr doesn't modify in place
7. **Debug step-by-step** - Run pipes incrementally

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| object not found | Column doesn't exist | Check names, order operations |
| Using = not == | Assignment in filter | Use == for comparison |
| unused argument | Wrong function usage | Check function signature |
| Result not saved | No assignment | Assign with <- |

**Core Verbs:**

```{r eval=FALSE}
# filter() - keep rows
data %>% filter(condition)

# select() - keep columns
data %>% select(col1, col2)

# mutate() - create columns
data %>% mutate(new_col = expression)

# arrange() - sort rows
data %>% arrange(col)
data %>% arrange(desc(col))

# summarize() - aggregate
data %>% summarize(stat = function(col))

# Common helpers
starts_with("prefix")
ends_with("suffix")
contains("text")
where(is.numeric)
everything()
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
data %>% filter(col == value)      # Use ==
data %>% select(col1, col2)        # Select then filter if needed
names(data)                         # Check before selecting
result <- data %>% ...              # Assign result

# ❌ Avoid
data %>% filter(col = value)       # Using =
data %>% select(col) %>%           # Then try to use removed col
  filter(other_col > 5)
data %>% filter(...) # no assign    # Result lost
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Basic Pipeline**

Use `mtcars` to:
1. Filter cars with mpg > 20
2. Select mpg, cyl, hp, wt
3. Create hp_per_ton = hp / wt
4. Arrange by hp_per_ton descending
5. Show top 5
</div>

<div class="exercise-box">
📝 **Exercise 2: Error Fixing**

Fix these errors:
```{r eval=FALSE}
# Error 1
mtcars %>%
  filter(cyl = 6)

# Error 2
mtcars %>%
  select(mpg, cyl) %>%
  filter(hp > 100)

# Error 3
mtcars %>%
  mutate(ratio = mpg_per_cyl / wt) %>%
  mutate(mpg_per_cyl = mpg / cyl)
```
</div>

<div class="exercise-box">
📝 **Exercise 3: Safe Select**

Write `safe_select(data, ...)` that:
1. Checks columns exist before selecting
2. Gives helpful error if not found
3. Suggests similar names
4. Works with select helpers
</div>

<div class="exercise-box">
📝 **Exercise 4: Summary Report**

Create function `summarize_by_group(data, group_col, value_col)` that:
1. Groups by specified column
2. Summarizes mean, median, sd, min, max
3. Counts observations
4. Returns sorted by mean descending
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
result <- mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl, hp, wt) %>%
  mutate(hp_per_ton = hp / wt) %>%
  arrange(desc(hp_per_ton)) %>%
  slice_head(n = 5)

result

# Alternative: use top_n (deprecated but still works)
mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl, hp, wt) %>%
  mutate(hp_per_ton = hp / wt) %>%
  slice_max(hp_per_ton, n = 5)
```

**Exercise 2:**

```{r}
# Error 1: Use == not =
mtcars %>%
  filter(cyl == 6)

# Error 2: Keep hp column or filter before select
# Option A: Select hp too
mtcars %>%
  select(mpg, cyl, hp) %>%
  filter(hp > 100)

# Option B: Filter first
mtcars %>%
  filter(hp > 100) %>%
  select(mpg, cyl)

# Error 3: Create mpg_per_cyl before using it
mtcars %>%
  mutate(mpg_per_cyl = mpg / cyl) %>%
  mutate(ratio = mpg_per_cyl / wt)

# Or in one mutate
mtcars %>%
  mutate(
    mpg_per_cyl = mpg / cyl,
    ratio = mpg_per_cyl / wt
  )
```

**Exercise 3:**

```{r}
safe_select <- function(data, ...) {
  # Capture expressions
  cols <- rlang::enquos(...)
  
  # Get all column names in data
  data_cols <- names(data)
  
  # Evaluate each selection
  for (col_expr in cols) {
    col_name <- rlang::as_name(col_expr)
    
    # Skip if it's a helper function call
    if (grepl("^(starts_with|ends_with|contains|matches|everything|where)", 
              col_name)) {
      next
    }
    
    # Check if column exists
    if (!col_name %in% data_cols) {
      # Find similar names
      similar <- agrep(col_name, data_cols, max.distance = 0.3, value = TRUE)
      
      error_msg <- paste0("Column '", col_name, "' not found in data.")
      
      if (length(similar) > 0) {
        error_msg <- paste0(error_msg, "\nDid you mean: ",
                           paste(similar, collapse = ", "), "?")
      }
      
      error_msg <- paste0(error_msg, "\nAvailable columns: ",
                         paste(data_cols, collapse = ", "))
      
      stop(error_msg, call. = FALSE)
    }
  }
  
  # If all checks pass, do the select
  select(data, !!!cols)
}

# Test
safe_select(mtcars, mpg, cyl, hp)
```

```{r error=TRUE}
safe_select(mtcars, mpg, horsepower)  # Helpful error
```

**Exercise 4:**

```{r}
summarize_by_group <- function(data, group_col, value_col) {
  # Use across for flexible column selection
  data %>%
    group_by(across({{ group_col }})) %>%
    summarize(
      n = n(),
      mean = mean({{ value_col }}, na.rm = TRUE),
      median = median({{ value_col }}, na.rm = TRUE),
      sd = sd({{ value_col }}, na.rm = TRUE),
      min = min({{ value_col }}, na.rm = TRUE),
      max = max({{ value_col }}, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(mean))
}

# Test
summarize_by_group(mtcars, cyl, mpg)
summarize_by_group(mtcars, gear, hp)

# With iris
summarize_by_group(iris, Species, Sepal.Length)
```
</details>
