# Grouping & Summarizing {#dplyr-grouping}

<div class="chapter-summary">
**What You'll Learn:**

- group_by() mechanics
- Grouped operations
- summarize() with groups
- Common grouping errors
- Multiple grouping variables
- Ungrouping data

**Key Errors Covered:** 18+ grouping errors

**Difficulty:** ⭐⭐ Intermediate to ⭐⭐⭐ Advanced
</div>

## Introduction

Grouping is one of dplyr's most powerful features:

```{r message=FALSE}
library(dplyr)

# Without grouping
mtcars %>%
  summarize(avg_mpg = mean(mpg))

# With grouping
mtcars %>%
  group_by(cyl) %>%
  summarize(avg_mpg = mean(mpg))
```

But grouping has surprising behaviors. Let's master it!

## group_by() Basics

<div class="insight-box">
💡 **Key Insight: group_by() Creates Invisible Groups**

```{r}
# Create grouped data
grouped_data <- mtcars %>%
  group_by(cyl)

# Looks the same when printed
head(grouped_data, 3)

# But it's grouped!
class(grouped_data)
groups(grouped_data)

# Check if grouped
is_grouped_df(grouped_data)

# See grouping variables
group_vars(grouped_data)

# Count groups
n_groups(grouped_data)

# Subsequent operations work within groups
grouped_data %>%
  summarize(avg_mpg = mean(mpg))
```

**Key insight:** Grouping doesn't change data, just how operations apply!
</div>

## Error #1: Forgetting Data is Grouped {#forgot-grouped}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-logic">🧠 LOGIC</span>

### The Problem

```{r}
# Group the data
cars_grouped <- mtcars %>%
  group_by(cyl)

# Later in code... expecting ungrouped behavior
cars_grouped %>%
  mutate(rank = row_number())  # Ranks within groups!
```

### What Happened

`row_number()` operated within each group, not across all data.

### Common Causes

```{r}
# Expecting overall statistics
mtcars %>%
  group_by(cyl) %>%
  mutate(
    overall_mean = mean(mpg)  # Actually group means!
  ) %>%
  select(cyl, mpg, overall_mean)

# Expecting all rows
mtcars %>%
  group_by(cyl) %>%
  slice(1)  # Gets first row of EACH group, not overall first

# Filtering
mtcars %>%
  group_by(cyl) %>%
  filter(mpg == max(mpg))  # Max within each group
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Ungroup When Done**

```{r}
# Ungroup explicitly
result <- mtcars %>%
  group_by(cyl) %>%
  summarize(avg_mpg = mean(mpg)) %>%
  ungroup()  # Remove grouping

# Now operations are ungrouped
result %>%
  mutate(rank = row_number())  # Ranks across all rows
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check Grouping Status**

```{r}
# Always check if data is grouped
check_grouping <- function(data) {
  if (is_grouped_df(data)) {
    message("Data is grouped by: ", paste(group_vars(data), collapse = ", "))
  } else {
    message("Data is not grouped")
  }
  invisible(data)
}

mtcars %>%
  group_by(cyl) %>%
  check_grouping() %>%
  summarize(avg_mpg = mean(mpg)) %>%
  check_grouping()
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use .groups Argument**

```{r}
# Control grouping in summarize
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(
    avg_mpg = mean(mpg),
    .groups = "drop"  # Ungroup completely
  )

# Keep some grouping
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(
    avg_mpg = mean(mpg),
    .groups = "drop_last"  # Keep cyl grouping
  ) %>%
  group_vars()

# Keep all groups
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(
    avg_mpg = mean(mpg),
    .groups = "keep"
  ) %>%
  group_vars()
```
</div>

## summarize() with Groups

<div class="insight-box">
💡 **Key Insight: summarize() Reduces Groups**

```{r}
# Single grouping variable
mtcars %>%
  group_by(cyl) %>%
  summarize(
    n = n(),
    avg_mpg = mean(mpg),
    sd_mpg = sd(mpg)
  )

# Multiple grouping variables
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(
    n = n(),
    avg_mpg = mean(mpg),
    .groups = "drop"
  )

# Useful summary functions
mtcars %>%
  group_by(cyl) %>%
  summarize(
    count = n(),                    # Number of observations
    n_distinct = n_distinct(gear),  # Unique values
    mean = mean(mpg),
    median = median(mpg),
    sd = sd(mpg),
    min = min(mpg),
    max = max(mpg),
    first = first(mpg),             # First value
    last = last(mpg),               # Last value
    nth = nth(mpg, 2)               # Nth value
  )

# Multiple values from same column
mtcars %>%
  group_by(cyl) %>%
  summarize(
    mpg_stats = list(summary(mpg))  # List column
  )
```
</div>

## Error #2: `must be length 1, not ...` {#summarize-length}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
mtcars %>%
  group_by(cyl) %>%
  summarize(
    top_mpg = head(mpg, 3)  # Returns multiple values!
  )
```

<div class="error-box">
🔴 **ERROR**

```
Error in summarize():
! `top_mpg` must be size 1, not 3.
```
</div>

### What It Means

`summarize()` expects functions that return a single value per group, not vectors.

### Common Causes

```{r error=TRUE}
# Functions returning multiple values
mtcars %>%
  group_by(cyl) %>%
  summarize(all_mpg = mpg)  # All values, not one

mtcars %>%
  group_by(cyl) %>%
  summarize(quantiles = quantile(mpg))  # Returns 5 values

mtcars %>%
  group_by(cyl) %>%
  summarize(range = range(mpg))  # Returns 2 values
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use Single-Value Functions**

```{r}
# Functions that return one value
mtcars %>%
  group_by(cyl) %>%
  summarize(
    mean_mpg = mean(mpg),
    max_mpg = max(mpg),
    min_mpg = min(mpg),
    n_obs = n()
  )

# For ranges, get separately
mtcars %>%
  group_by(cyl) %>%
  summarize(
    min_mpg = min(mpg),
    max_mpg = max(mpg)
  )
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use reframe() for Multiple Rows**

```{r}
# reframe() allows multiple rows per group (dplyr >= 1.1.0)
if (packageVersion("dplyr") >= "1.1.0") {
  mtcars %>%
    group_by(cyl) %>%
    reframe(
      top_3_mpg = head(sort(mpg, decreasing = TRUE), 3)
    )
}

# For older dplyr, use summarize with list
mtcars %>%
  group_by(cyl) %>%
  summarize(
    top_3_mpg = list(head(sort(mpg, decreasing = TRUE), 3))
  ) %>%
  tidyr::unnest(top_3_mpg)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use slice() Instead**

```{r}
# Get top N rows per group with slice
mtcars %>%
  group_by(cyl) %>%
  slice_max(mpg, n = 3) %>%
  select(cyl, mpg)

# Or with arrange and slice
mtcars %>%
  group_by(cyl) %>%
  arrange(desc(mpg)) %>%
  slice(1:3) %>%
  select(cyl, mpg)
```
</div>

## mutate() with Groups

<div class="insight-box">
💡 **Key Insight: mutate() Keeps All Rows**

```{r}
# mutate() operates within groups but keeps all rows
mtcars %>%
  group_by(cyl) %>%
  mutate(
    group_mean = mean(mpg),
    deviation = mpg - group_mean
  ) %>%
  select(cyl, mpg, group_mean, deviation) %>%
  arrange(cyl, mpg)

# Ranking within groups
mtcars %>%
  group_by(cyl) %>%
  mutate(
    mpg_rank = rank(desc(mpg)),
    mpg_dense_rank = dense_rank(desc(mpg)),
    mpg_row_number = row_number(desc(mpg))
  ) %>%
  select(cyl, mpg, mpg_rank, mpg_dense_rank, mpg_row_number) %>%
  arrange(cyl, desc(mpg))

# Cumulative operations within groups
mtcars %>%
  group_by(cyl) %>%
  arrange(mpg) %>%
  mutate(
    cumsum_mpg = cumsum(mpg),
    cummax_mpg = cummax(mpg)
  ) %>%
  select(cyl, mpg, cumsum_mpg, cummax_mpg)

# Lead and lag within groups
mtcars %>%
  group_by(cyl) %>%
  arrange(mpg) %>%
  mutate(
    prev_mpg = lag(mpg),
    next_mpg = lead(mpg),
    diff_from_prev = mpg - lag(mpg)
  ) %>%
  select(cyl, mpg, prev_mpg, next_mpg, diff_from_prev)
```
</div>

## filter() with Groups

<div class="insight-box">
💡 **Key Insight: filter() with Groups**

```{r}
# Filter based on group statistics
mtcars %>%
  group_by(cyl) %>%
  filter(mpg > mean(mpg)) %>%
  select(cyl, mpg) %>%
  arrange(cyl, mpg)

# Keep only groups meeting criteria
mtcars %>%
  group_by(cyl) %>%
  filter(n() >= 10) %>%  # Only groups with 10+ observations
  select(cyl) %>%
  distinct()

# Top N per group
mtcars %>%
  group_by(cyl) %>%
  filter(rank(desc(mpg)) <= 2) %>%
  select(cyl, mpg) %>%
  arrange(cyl, desc(mpg))

# Filter by group condition
mtcars %>%
  group_by(cyl) %>%
  filter(mean(mpg) > 20) %>%  # Only groups with avg mpg > 20
  select(cyl, mpg)
```
</div>

## Multiple Grouping Variables

<div class="insight-box">
💡 **Key Insight: Multiple Groups**

```{r}
# Group by multiple variables
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(
    n = n(),
    avg_mpg = mean(mpg),
    .groups = "drop"
  )

# Order matters for operations
mtcars %>%
  group_by(cyl, gear) %>%
  mutate(
    rank_in_cyl_gear = row_number(desc(mpg))
  ) %>%
  select(cyl, gear, mpg, rank_in_cyl_gear) %>%
  arrange(cyl, gear, desc(mpg))

# Count combinations
mtcars %>%
  count(cyl, gear)

# Or with group_by + summarize
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(n = n(), .groups = "drop")
```
</div>

## Error #3: Unexpected Grouping After summarize() {#unexpected-grouping}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-logic">🧠 LOGIC</span>

### The Problem

```{r}
# Summarize with multiple groups
result <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(avg_mpg = mean(mpg))  # Message about grouping

# Still grouped!
is_grouped_df(result)
group_vars(result)

# Next operation uses this grouping
result %>%
  mutate(rank = row_number())  # Within cyl groups!
```

### What Happened

By default, `summarize()` drops the last grouping variable but keeps others.

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use .groups = "drop"**

```{r}
# Ungroup completely
result <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(
    avg_mpg = mean(mpg),
    .groups = "drop"
  )

is_grouped_df(result)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Explicit ungroup()**

```{r}
result <- mtcars %>%
  group_by(cyl, gear) %>%
  summarize(avg_mpg = mean(mpg)) %>%
  ungroup()

is_grouped_df(result)
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Understand .groups Options**

```{r}
# .groups = "drop_last" (default)
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(avg_mpg = mean(mpg), .groups = "drop_last") %>%
  group_vars()  # Still has cyl

# .groups = "keep"
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(avg_mpg = mean(mpg), .groups = "keep") %>%
  group_vars()  # Has both cyl and gear

# .groups = "drop"
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(avg_mpg = mean(mpg), .groups = "drop") %>%
  group_vars()  # No groups

# .groups = "rowwise"
mtcars %>%
  group_by(cyl, gear) %>%
  summarize(avg_mpg = mean(mpg), .groups = "rowwise") %>%
  class()
```
</div>

## Window Functions

<div class="bestpractice-box">
🎯 **Best Practice: Window Functions with Groups**

```{r}
# Ranking functions
mtcars %>%
  group_by(cyl) %>%
  mutate(
    rank = rank(desc(mpg)),           # Ties get average rank
    dense_rank = dense_rank(desc(mpg)), # Ties get same rank, no gaps
    row_number = row_number(desc(mpg)), # Ties broken arbitrarily
    percent_rank = percent_rank(desc(mpg)), # Percentile
    cume_dist = cume_dist(desc(mpg))   # Cumulative distribution
  ) %>%
  select(cyl, mpg, rank, dense_rank, row_number, percent_rank, cume_dist) %>%
  arrange(cyl, desc(mpg))

# Lead and lag
mtcars %>%
  group_by(cyl) %>%
  arrange(mpg) %>%
  mutate(
    prev_value = lag(mpg, 1),      # Previous value
    next_value = lead(mpg, 1),     # Next value
    prev_2 = lag(mpg, 2),          # 2 rows back
    change = mpg - lag(mpg, 1)     # Change from previous
  ) %>%
  select(cyl, mpg, prev_value, next_value, change)

# Cumulative functions
mtcars %>%
  group_by(cyl) %>%
  arrange(mpg) %>%
  mutate(
    cumsum = cumsum(mpg),
    cummean = cummean(mpg),
    cummin = cummin(mpg),
    cummax = cummax(mpg)
  ) %>%
  select(cyl, mpg, cumsum, cummean, cummin, cummax)

# Offset functions
mtcars %>%
  group_by(cyl) %>%
  arrange(mpg) %>%
  mutate(
    first_val = first(mpg),
    last_val = last(mpg),
    nth_val = nth(mpg, 3)
  ) %>%
  select(cyl, mpg, first_val, last_val, nth_val)
```
</div>

## count() and add_count()

<div class="insight-box">
💡 **Key Insight: Counting Shortcuts**

```{r}
# count() is shortcut for group_by + summarize + ungroup
mtcars %>%
  count(cyl)

# Equivalent to:
mtcars %>%
  group_by(cyl) %>%
  summarize(n = n()) %>%
  ungroup()

# Multiple variables
mtcars %>%
  count(cyl, gear)

# With sorting
mtcars %>%
  count(cyl, sort = TRUE)

# With weights
mtcars %>%
  count(cyl, wt = mpg)  # Sum of mpg instead of count

# add_count() keeps all rows
mtcars %>%
  add_count(cyl) %>%
  select(cyl, n, everything())

# Name the count column
mtcars %>%
  add_count(cyl, name = "n_per_cyl") %>%
  select(cyl, n_per_cyl, mpg)
```
</div>

## across() for Multiple Columns

<div class="bestpractice-box">
🎯 **Best Practice: across() for Multiple Columns**

```{r}
# Summarize multiple columns
mtcars %>%
  group_by(cyl) %>%
  summarize(
    across(c(mpg, hp, wt), mean),
    .groups = "drop"
  )

# With named functions
mtcars %>%
  group_by(cyl) %>%
  summarize(
    across(c(mpg, hp, wt), 
           list(mean = mean, sd = sd),
           .names = "{.col}_{.fn}"),
    .groups = "drop"
  )

# Using selection helpers
mtcars %>%
  group_by(cyl) %>%
  summarize(
    across(where(is.numeric), mean),
    .groups = "drop"
  )

# In mutate
mtcars %>%
  group_by(cyl) %>%
  mutate(
    across(c(mpg, hp), 
           ~ . - mean(.),
           .names = "{.col}_centered")
  ) %>%
  select(cyl, mpg, mpg_centered, hp, hp_centered)

# Multiple functions
mtcars %>%
  group_by(cyl) %>%
  summarize(
    across(c(mpg, hp), 
           list(min = min, max = max, mean = mean)),
    .groups = "drop"
  )
```
</div>

## Common Pitfalls

<div class="pitfall-box">
⚠️ **Common Grouping Pitfalls**

```{r}
# Pitfall 1: Forgetting to ungroup
bad <- mtcars %>%
  group_by(cyl) %>%
  mutate(mpg_scaled = mpg / mean(mpg))
# Still grouped!

good <- mtcars %>%
  group_by(cyl) %>%
  mutate(mpg_scaled = mpg / mean(mpg)) %>%
  ungroup()

# Pitfall 2: Summarizing already summarized data
mtcars %>%
  group_by(cyl) %>%
  summarize(avg_mpg = mean(mpg)) %>%
  summarize(overall_avg = mean(avg_mpg))  # Wrong! Treats groups equally

# Correct: weight by group size
mtcars %>%
  group_by(cyl) %>%
  summarize(
    avg_mpg = mean(mpg),
    n = n(),
    .groups = "drop"
  ) %>%
  summarize(overall_avg = sum(avg_mpg * n) / sum(n))

# Pitfall 3: Group order affecting row_number
set.seed(123)
df <- tibble(
  group = rep(c("A", "B"), each = 3),
  value = sample(1:6)
)

# Different results!
df %>% group_by(group) %>% mutate(rn = row_number())
df %>% arrange(value) %>% group_by(group) %>% mutate(rn = row_number())

# Pitfall 4: Comparing grouped vs ungrouped
mtcars %>%
  group_by(cyl) %>%
  filter(mpg == max(mpg)) %>%  # Max within each cyl
  nrow()

mtcars %>%
  filter(mpg == max(mpg)) %>%  # Overall max
  nrow()
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **group_by() is invisible** - Data looks same but behaves differently
2. **Always ungroup** - Use `ungroup()` or `.groups = "drop"`
3. **summarize() reduces groups** - Drops last group by default
4. **mutate() keeps all rows** - Operations within groups
5. **Window functions** - Work great with groups
6. **Check grouping status** - Use `group_vars()` or `is_grouped_df()`
7. **across() for multiple columns** - Apply same operation to many columns

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| Unexpected grouped behavior | Forgot data is grouped | Check with `group_vars()`, use `ungroup()` |
| must be size 1 | summarize() with vector | Use single-value function or `reframe()` |
| Wrong statistics | Operating on groups | Ungroup or use different operation |

**Grouping Operations:**

```{r eval=FALSE}
# Basic grouping
data %>% group_by(col)

# Multiple groups
data %>% group_by(col1, col2)

# Ungroup
data %>% ungroup()

# Check grouping
group_vars(data)
is_grouped_df(data)
n_groups(data)

# Summarize with groups
data %>%
  group_by(col) %>%
  summarize(stat = mean(value), .groups = "drop")

# Mutate with groups (keeps all rows)
data %>%
  group_by(col) %>%
  mutate(group_mean = mean(value))

# Filter with groups
data %>%
  group_by(col) %>%
  filter(value > mean(value))

# Count
data %>% count(col)
data %>% add_count(col)

# across for multiple columns
data %>%
  group_by(col) %>%
  summarize(across(where(is.numeric), mean))
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
data %>% group_by(col) %>% 
  summarize(stat = mean(x), .groups = "drop")  # Explicit ungrouping
data %>% group_by(col) %>% ... %>% ungroup()   # Ungroup at end
Always check: group_vars(data)                  # Verify grouping

# ❌ Avoid
data %>% group_by(col) %>% summarize(...)      # No .groups argument
Leave data grouped                               # Causes confusion later
Assume data is/isn't grouped                     # Always check
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Group Statistics**

Using `mtcars`:
1. Group by cyl
2. Calculate mean, median, sd of mpg per group
3. Add count and min/max mpg
4. Sort by mean mpg descending
5. Ensure result is ungrouped
</div>

<div class="exercise-box">
📝 **Exercise 2: Window Functions**

Using `mtcars`:
1. Group by cyl
2. Add rank of mpg within each cyl
3. Add difference from group mean
4. Add percentile within group
5. Keep only top 2 per group
</div>

<div class="exercise-box">
📝 **Exercise 3: Group Detector**

Write `check_grouped_operations(data)` that:
1. Checks if data is grouped
2. Shows what grouping variables exist
3. Warns about potential issues
4. Suggests when to ungroup
</div>

<div class="exercise-box">
📝 **Exercise 4: Grouped Summary Report**

Write `grouped_summary(data, group_vars, summary_vars)` that:
1. Groups by specified columns
2. Summarizes multiple statistics for each summary variable
3. Includes count and percentage of total
4. Returns ungrouped, sorted result
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
result <- mtcars %>%
  group_by(cyl) %>%
  summarize(
    count = n(),
    mean_mpg = mean(mpg),
    median_mpg = median(mpg),
    sd_mpg = sd(mpg),
    min_mpg = min(mpg),
    max_mpg = max(mpg),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_mpg))

result

# Verify it's ungrouped
is_grouped_df(result)
```

**Exercise 2:**

```{r}
result <- mtcars %>%
  group_by(cyl) %>%
  mutate(
    mpg_rank = rank(desc(mpg)),
    deviation_from_mean = mpg - mean(mpg),
    percentile = percent_rank(mpg)
  ) %>%
  filter(mpg_rank <= 2) %>%
  select(cyl, mpg, mpg_rank, deviation_from_mean, percentile) %>%
  arrange(cyl, mpg_rank) %>%
  ungroup()

result
```

**Exercise 3:**

```{r}
check_grouped_operations <- function(data, warn_threshold = 3) {
  cat("=== Grouping Analysis ===\n\n")
  
  # Check if grouped
  if (!is_grouped_df(data)) {
    cat("✓ Data is NOT grouped\n")
    return(invisible(data))
  }
  
  # Get grouping info
  grp_vars <- group_vars(data)
  n_grps <- n_groups(data)
  
  cat("⚠ Data IS grouped\n")
  cat("Grouping variables:", paste(grp_vars, collapse = ", "), "\n")
  cat("Number of groups:", n_grps, "\n\n")
  
  # Group sizes
  grp_sizes <- data %>%
    summarize(n = n(), .groups = "drop") %>%
    pull(n)
  
  cat("Group size summary:\n")
  cat("  Min:", min(grp_sizes), "\n")
  cat("  Max:", max(grp_sizes), "\n")
  cat("  Mean:", round(mean(grp_sizes), 1), "\n\n")
  
  # Warnings
  if (length(grp_vars) > warn_threshold) {
    warning("Many grouping variables (", length(grp_vars), 
            ") may lead to many small groups")
  }
  
  if (any(grp_sizes == 1)) {
    warning("Some groups have only 1 observation")
  }
  
  # Suggestions
  cat("Suggestions:\n")
  cat("  - Use ungroup() when done with grouped operations\n")
  cat("  - Or use .groups = 'drop' in summarize()\n")
  cat("  - Check grouping with group_vars() before operations\n")
  
  invisible(data)
}

# Test
mtcars %>%
  group_by(cyl) %>%
  check_grouped_operations()

mtcars %>%
  group_by(cyl, gear, carb) %>%
  check_grouped_operations()
```

**Exercise 4:**

```{r}
grouped_summary <- function(data, group_vars, summary_vars) {
  # Get total count for percentages
  total_n <- nrow(data)
  
  # Create summary
  result <- data %>%
    group_by(across(all_of(group_vars))) %>%
    summarize(
      count = n(),
      pct_of_total = n() / total_n * 100,
      across(
        all_of(summary_vars),
        list(
          mean = ~mean(., na.rm = TRUE),
          median = ~median(., na.rm = TRUE),
          sd = ~sd(., na.rm = TRUE),
          min = ~min(., na.rm = TRUE),
          max = ~max(., na.rm = TRUE)
        ),
        .names = "{.col}_{.fn}"
      ),
      .groups = "drop"
    ) %>%
    arrange(desc(count))
  
  result
}

# Test
grouped_summary(
  mtcars,
  group_vars = c("cyl"),
  summary_vars = c("mpg", "hp")
)

grouped_summary(
  mtcars,
  group_vars = c("cyl", "gear"),
  summary_vars = c("mpg", "hp", "wt")
)

# With iris
grouped_summary(
  iris,
  group_vars = "Species",
  summary_vars = c("Sepal.Length", "Sepal.Width", "Petal.Length")
)
```
</details>
