# Joins & Merging {#dplyr-joins}

<div class="chapter-summary">
**What You'll Learn:**

- Types of joins (inner, left, right, full)
- Join keys and matching
- Common join errors
- Multiple key joins
- Handling duplicates
- Anti and semi joins

**Key Errors Covered:** 20+ join errors

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Introduction

Joining combines data from multiple sources:

```{r message=FALSE}
library(dplyr)

# Two related datasets
customers <- tibble(
  id = 1:3,
  name = c("Alice", "Bob", "Charlie")
)

orders <- tibble(
  order_id = 1:4,
  customer_id = c(1, 1, 2, 4),
  amount = c(100, 150, 200, 75)
)

# Join them
customers %>%
  left_join(orders, by = c("id" = "customer_id"))
```

But joins have many pitfalls. Let's master them!

## Join Types

<div class="insight-box">
💡 **Key Insight: Four Main Join Types**

```{r}
# Sample data
band_members
band_instruments

# inner_join: Only matching rows
band_members %>%
  inner_join(band_instruments, by = "name")

# left_join: All from left, matching from right
band_members %>%
  left_join(band_instruments, by = "name")

# right_join: All from right, matching from left
band_members %>%
  right_join(band_instruments, by = "name")

# full_join: All rows from both
band_members %>%
  full_join(band_instruments, by = "name")
```

**Visual representation:**
- `inner_join`: ∩ (intersection)
- `left_join`: ← (left + intersection)
- `right_join`: → (right + intersection)
- `full_join`: ∪ (union, all rows)
</div>

## Error #1: `by must be supplied` {#join-by-missing}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-args">📋 ARGS</span>

### The Error

```{r error=TRUE}
df1 <- tibble(id = 1:3, x = c("a", "b", "c"))
df2 <- tibble(identifier = 1:3, y = c("d", "e", "f"))

df1 %>%
  left_join(df2)
```

<div class="error-box">
🔴 **ERROR** (dplyr >= 1.1.0)

```
Error in left_join():
! `by` must be supplied when `x` and `y` have no common variables.
```
</div>

### What It Means

The two datasets have no columns with the same name, so dplyr doesn't know how to join them.

### Common Causes

```{r error=TRUE}
# Different column names for same concept
customers <- tibble(customer_id = 1:3, name = c("A", "B", "C"))
orders <- tibble(cust_id = c(1, 1, 2), amount = c(100, 150, 200))

customers %>%
  left_join(orders)  # No common columns!

# Typo in column name
df1 <- tibble(id = 1:3, x = "a")
df2 <- tibble(ID = 1:3, y = "b")  # Different case!

df1 %>%
  left_join(df2)
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Specify Join Keys**

```{r}
# Use by argument with named vector
customers <- tibble(customer_id = 1:3, name = c("A", "B", "C"))
orders <- tibble(cust_id = c(1, 1, 2), amount = c(100, 150, 200))

customers %>%
  left_join(orders, by = c("customer_id" = "cust_id"))

# Multiple keys
df1 <- tibble(id = 1:3, type = "A", x = 1:3)
df2 <- tibble(id = 1:3, type = "A", y = 4:6)

df1 %>%
  left_join(df2, by = c("id", "type"))

# Or with different names
df1 <- tibble(id1 = 1:3, type1 = "A", x = 1:3)
df2 <- tibble(id2 = 1:3, type2 = "A", y = 4:6)

df1 %>%
  left_join(df2, by = c("id1" = "id2", "type1" = "type2"))
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Rename Columns First**

```{r}
customers <- tibble(customer_id = 1:3, name = c("A", "B", "C"))
orders <- tibble(cust_id = c(1, 1, 2), amount = c(100, 150, 200))

# Rename to match
orders_renamed <- orders %>%
  rename(customer_id = cust_id)

customers %>%
  left_join(orders_renamed, by = "customer_id")
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use join_by() (dplyr >= 1.1.0)**

```{r}
# Modern syntax (if available)
if (packageVersion("dplyr") >= "1.1.0") {
  customers %>%
    left_join(orders, join_by(customer_id == cust_id))
}
```
</div>

## Error #2: Duplicate keys creating many rows {#join-duplicates}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-logic">🧠 LOGIC</span>

### The Problem

```{r}
# Data with duplicates
customers <- tibble(
  id = c(1, 1, 2),  # Duplicate id 1!
  name = c("Alice", "Alice", "Bob")
)

orders <- tibble(
  customer_id = c(1, 1, 2),  # Duplicate id 1!
  amount = c(100, 150, 200)
)

# Join creates cartesian product
result <- customers %>%
  left_join(orders, by = c("id" = "customer_id"))

result
nrow(result)  # 5 rows from 3 + 3!
```

### What Happened

When both tables have duplicate keys, you get all combinations (cartesian product).

### Common Causes

```{r}
# Many-to-many relationship
students <- tibble(
  student_id = c(1, 1, 2, 2),
  course = c("Math", "English", "Math", "Science")
)

grades <- tibble(
  student_id = c(1, 1, 2, 2),
  grade = c("A", "B", "A", "B")
)

# Which grade goes with which course?
students %>%
  left_join(grades, by = "student_id")  # Ambiguous!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Remove Duplicates Before Join**

```{r}
customers <- tibble(
  id = c(1, 1, 2),
  name = c("Alice", "Alice", "Bob")
)

# Keep unique customers
customers_unique <- customers %>%
  distinct(id, .keep_all = TRUE)

customers_unique %>%
  left_join(orders, by = c("id" = "customer_id"))
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Check for Duplicates**

```{r}
check_join_keys <- function(data, key_cols) {
  dups <- data %>%
    group_by(across(all_of(key_cols))) %>%
    filter(n() > 1) %>%
    ungroup()
  
  if (nrow(dups) > 0) {
    warning("Duplicate keys found:\n")
    print(dups)
    return(FALSE)
  }
  
  message("No duplicate keys")
  return(TRUE)
}

check_join_keys(customers, "id")
check_join_keys(customers_unique, "id")
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Aggregate Before Join**

```{r}
# If duplicates make sense (e.g., multiple orders)
orders_summary <- orders %>%
  group_by(customer_id) %>%
  summarize(
    total_orders = n(),
    total_amount = sum(amount),
    .groups = "drop"
  )

customers_unique %>%
  left_join(orders_summary, by = c("id" = "customer_id"))
```
</div>

<div class="solution-box">
✅ **SOLUTION 4: Use relationship argument (dplyr >= 1.1.0)**

```{r}
# Specify expected relationship
if (packageVersion("dplyr") >= "1.1.0") {
  # This will error if relationship is wrong
  # customers %>%
  #   left_join(orders, 
  #            by = c("id" = "customer_id"),
  #            relationship = "one-to-one")
  
  # Specify many-to-one
  customers %>%
    left_join(orders, 
             by = c("id" = "customer_id"),
             relationship = "many-to-one")
}
```
</div>

## Multiple Key Joins

<div class="insight-box">
💡 **Key Insight: Joining on Multiple Columns**

```{r}
# Data with composite keys
sales <- tibble(
  store = c("A", "A", "B", "B"),
  product = c("X", "Y", "X", "Y"),
  quantity = c(10, 20, 15, 25)
)

prices <- tibble(
  store = c("A", "A", "B", "B"),
  product = c("X", "Y", "X", "Y"),
  price = c(5.00, 7.50, 4.50, 8.00)
)

# Join on both store and product
sales %>%
  left_join(prices, by = c("store", "product")) %>%
  mutate(revenue = quantity * price)

# With different names
prices2 <- tibble(
  location = c("A", "A", "B", "B"),
  item = c("X", "Y", "X", "Y"),
  price = c(5.00, 7.50, 4.50, 8.00)
)

sales %>%
  left_join(prices2, by = c("store" = "location", "product" = "item"))
```
</div>

## Error #3: Unexpected NA values after join {#join-unexpected-na}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-logic">🧠 LOGIC</span>

### The Problem

```{r}
customers <- tibble(
  id = 1:4,
  name = c("Alice", "Bob", "Charlie", "David")
)

orders <- tibble(
  customer_id = c(1, 1, 2, 5),  # Customer 5 doesn't exist!
  amount = c(100, 150, 200, 75)
)

# Left join
result <- customers %>%
  left_join(orders, by = c("id" = "customer_id"))

result
# Charlie and David have no orders (NA in amount)
```

### What Happened

Left join keeps all rows from left table. Unmatched rows get NA in right table columns.

### Understanding NA Sources

```{r}
# Right join - unmatched from right get NA in left columns
result_right <- customers %>%
  right_join(orders, by = c("id" = "customer_id"))

result_right
# Customer 5 order has NA in name

# Inner join - no NAs (only matching rows)
result_inner <- customers %>%
  inner_join(orders, by = c("id" = "customer_id"))

result_inner
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Choose Right Join Type**

```{r}
# If you want only matching rows
customers %>%
  inner_join(orders, by = c("id" = "customer_id"))

# If you want all customers (even without orders)
customers %>%
  left_join(orders, by = c("id" = "customer_id"))

# If you want all orders (even with unknown customers)
customers %>%
  right_join(orders, by = c("id" = "customer_id"))
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Fill NAs After Join**

```{r}
customers %>%
  left_join(orders, by = c("id" = "customer_id")) %>%
  mutate(
    amount = replace_na(amount, 0),
    order_count = if_else(is.na(amount), 0L, 1L)
  )

# Or with tidyr
library(tidyr)
customers %>%
  left_join(orders, by = c("id" = "customer_id")) %>%
  replace_na(list(amount = 0))
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Check for Unmatched Rows**

```{r}
# Find unmatched customers
customers %>%
  anti_join(orders, by = c("id" = "customer_id"))

# Find unmatched orders
orders %>%
  anti_join(customers, by = c("customer_id" = "id"))
```
</div>

## Filtering Joins

<div class="insight-box">
💡 **Key Insight: semi_join() and anti_join()**

```{r}
# semi_join: Keep rows in x that have match in y
# Like inner_join but doesn't add columns from y
band_members %>%
  semi_join(band_instruments, by = "name")

# anti_join: Keep rows in x that DON'T have match in y
band_members %>%
  anti_join(band_instruments, by = "name")

# Practical use: find missing data
customers <- tibble(id = 1:5, name = letters[1:5])
orders <- tibble(customer_id = c(1, 2, 2, 3), amount = c(100, 150, 200, 75))

# Customers with no orders
customers %>%
  anti_join(orders, by = c("id" = "customer_id"))

# Customers with orders
customers %>%
  semi_join(orders, by = c("id" = "customer_id"))
```
</div>

## Error #4: Column name conflicts {#join-name-conflicts}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-naming">🏷️ NAMING</span>

### The Problem

```{r}
df1 <- tibble(
  id = 1:3,
  value = c(10, 20, 30)
)

df2 <- tibble(
  id = 1:3,
  value = c(100, 200, 300)
)

# Both have 'value' column
result <- df1 %>%
  left_join(df2, by = "id")

result
# Creates value.x and value.y
```

### What Happened

Columns with same name (not in join key) get suffixes.

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use suffix argument**

```{r}
df1 %>%
  left_join(df2, by = "id", suffix = c("_left", "_right"))

# Or more descriptive
df1 %>%
  left_join(df2, by = "id", suffix = c("_original", "_new"))
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Rename Before Join**

```{r}
df1 %>%
  rename(value1 = value) %>%
  left_join(
    df2 %>% rename(value2 = value),
    by = "id"
  )
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Select Columns First**

```{r}
# Only keep needed columns from each
df1 %>%
  left_join(
    df2 %>% select(id, value2 = value),
    by = "id"
  )
```
</div>

## Common Join Patterns

<div class="bestpractice-box">
🎯 **Best Practice: Common Join Patterns**

```{r}
# 1. Lookup table (add descriptions)
products <- tibble(
  product_id = 1:3,
  quantity = c(10, 20, 15)
)

product_info <- tibble(
  product_id = 1:3,
  name = c("Widget", "Gadget", "Doohickey"),
  price = c(9.99, 19.99, 14.99)
)

products %>%
  left_join(product_info, by = "product_id")

# 2. Aggregate then join
orders <- tibble(
  customer_id = c(1, 1, 2, 2, 3),
  amount = c(100, 150, 200, 75, 300)
)

customers <- tibble(
  id = 1:3,
  name = c("Alice", "Bob", "Charlie")
)

order_summary <- orders %>%
  group_by(customer_id) %>%
  summarize(
    total_orders = n(),
    total_spent = sum(amount),
    avg_order = mean(amount),
    .groups = "drop"
  )

customers %>%
  left_join(order_summary, by = c("id" = "customer_id"))

# 3. Multiple joins
students <- tibble(id = 1:3, name = c("A", "B", "C"))
math_grades <- tibble(student_id = 1:3, math = c(85, 90, 78))
english_grades <- tibble(student_id = 1:3, english = c(88, 92, 85))

students %>%
  left_join(math_grades, by = c("id" = "student_id")) %>%
  left_join(english_grades, by = c("id" = "student_id"))

# 4. Self join
employees <- tibble(
  id = 1:4,
  name = c("Alice", "Bob", "Charlie", "David"),
  manager_id = c(NA, 1, 1, 2)
)

employees %>%
  left_join(
    employees %>% select(id, manager_name = name),
    by = c("manager_id" = "id")
  )

# 5. Rolling join (closest match)
events <- tibble(
  time = c(1, 3, 5, 7),
  event = c("A", "B", "C", "D")
)

measurements <- tibble(
  time = c(0, 2, 4, 6, 8),
  value = c(10, 20, 30, 40, 50)
)

# Nearest match (simplified example)
events %>%
  left_join(measurements, by = "time")  # Only exact matches
```
</div>

## Checking Join Results

<div class="bestpractice-box">
🎯 **Best Practice: Validate Joins**

```{r}
validate_join <- function(original, joined, join_type = "left") {
  cat("=== Join Validation ===\n\n")
  
  # Row count check
  cat("Original rows:", nrow(original), "\n")
  cat("Joined rows:", nrow(joined), "\n")
  
  if (join_type == "inner") {
    if (nrow(joined) > nrow(original)) {
      warning("Inner join resulted in MORE rows - duplicates in join key?")
    }
  } else if (join_type == "left") {
    if (nrow(joined) < nrow(original)) {
      warning("Left join resulted in FEWER rows - this shouldn't happen!")
    } else if (nrow(joined) > nrow(original)) {
      warning("Left join resulted in MORE rows - duplicates in right table?")
    }
  }
  
  # NA check
  na_cols <- names(joined)[colSums(is.na(joined)) > 0]
  if (length(na_cols) > 0) {
    cat("\nColumns with NAs:\n")
    for (col in na_cols) {
      n_na <- sum(is.na(joined[[col]]))
      cat("  ", col, ":", n_na, "NAs\n")
    }
  }
  
  # Duplicate check in key columns
  cat("\n")
}

# Test
customers <- tibble(id = 1:3, name = c("A", "B", "C"))
orders <- tibble(customer_id = c(1, 1, 2, 4), amount = c(100, 150, 200, 75))

result <- customers %>%
  left_join(orders, by = c("id" = "customer_id"))

validate_join(customers, result, "left")
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Four main joins** - inner, left, right, full
2. **Specify by argument** - When column names differ
3. **Watch for duplicates** - Can create many rows
4. **Left join creates NAs** - For unmatched rows
5. **Use filtering joins** - semi_join(), anti_join()
6. **Check join results** - Validate row counts and NAs
7. **Name conflicts** - Use suffix or rename

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| by must be supplied | No common columns | Use `by = c("x" = "y")` |
| Too many rows | Duplicate keys | Check for duplicates, aggregate |
| Unexpected NAs | Unmatched rows | Use right join type, fill NAs |
| Column conflicts | Same name, not in key | Use suffix argument |

**Join Types:**

```{r eval=FALSE}
# Mutating joins (add columns)
inner_join(x, y, by = "key")  # Only matching
left_join(x, y, by = "key")   # All x, matching y
right_join(x, y, by = "key")  # Matching x, all y
full_join(x, y, by = "key")   # All rows

# Filtering joins (filter rows)
semi_join(x, y, by = "key")   # x rows that match y
anti_join(x, y, by = "key")   # x rows that don't match y

# Different column names
left_join(x, y, by = c("x_id" = "y_id"))

# Multiple keys
left_join(x, y, by = c("id", "type"))

# Handle name conflicts
left_join(x, y, by = "id", suffix = c("_x", "_y"))
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
left_join(x, y, by = "key")              # Explicit by
Check for duplicates before join          # Avoid cartesian
Validate row counts after join            # Catch issues
Use appropriate join type                 # inner vs left vs full

# ❌ Avoid
left_join(x, y)  # No by (relies on names)
Join without checking duplicates          # Surprise row explosion
Ignore NAs after join                     # May indicate issues
Always use left_join                      # Sometimes inner is right
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Basic Joins**

Given:
```{r eval=FALSE}
products <- tibble(id = 1:5, name = c("A", "B", "C", "D", "E"))
sales <- tibble(product_id = c(1, 2, 2, 3, 6), amount = c(100, 150, 200, 75, 50))
```

1. Inner join - only products with sales
2. Left join - all products, add sales
3. Find products with no sales
4. Find sales for unknown products
</div>

<div class="exercise-box">
📝 **Exercise 2: Duplicate Handling**

Given data with duplicates:
```{r eval=FALSE}
customers <- tibble(id = c(1, 1, 2, 3), name = c("A", "A", "B", "C"))
orders <- tibble(cust_id = c(1, 2, 2, 3), amount = c(100, 150, 200, 75))
```

1. Identify the duplicates
2. Remove them appropriately
3. Perform safe join
4. Validate result
</div>

<div class="exercise-box">
📝 **Exercise 3: Safe Join Function**

Write `safe_join(x, y, by, type)` that:
1. Checks for duplicate keys
2. Validates join will work
3. Performs join
4. Checks result makes sense
5. Returns result with metadata
</div>

<div class="exercise-box">
📝 **Exercise 4: Multiple Table Join**

Join three tables efficiently:
```{r eval=FALSE}
students <- tibble(id = 1:3, name = c("A", "B", "C"))
courses <- tibble(student_id = c(1, 1, 2, 3), course = c("Math", "Eng", "Math", "Sci"))
grades <- tibble(student_id = c(1, 1, 2, 3), course = c("Math", "Eng", "Math", "Sci"), 
                grade = c(85, 90, 78, 92))
```
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
products <- tibble(id = 1:5, name = c("A", "B", "C", "D", "E"))
sales <- tibble(product_id = c(1, 2, 2, 3, 6), amount = c(100, 150, 200, 75, 50))

# 1. Inner join - only products with sales
products_with_sales <- products %>%
  inner_join(sales, by = c("id" = "product_id"))
products_with_sales

# 2. Left join - all products, add sales
all_products <- products %>%
  left_join(sales, by = c("id" = "product_id"))
all_products

# 3. Find products with no sales
no_sales <- products %>%
  anti_join(sales, by = c("id" = "product_id"))
no_sales

# 4. Find sales for unknown products
unknown_products <- sales %>%
  anti_join(products, by = c("product_id" = "id"))
unknown_products
```

**Exercise 2:**

```{r}
customers <- tibble(id = c(1, 1, 2, 3), name = c("A", "A", "B", "C"))
orders <- tibble(cust_id = c(1, 2, 2, 3), amount = c(100, 150, 200, 75))

# 1. Identify duplicates
cat("Duplicate customers:\n")
customers %>%
  group_by(id) %>%
  filter(n() > 1)

cat("\nDuplicate orders:\n")
orders %>%
  group_by(cust_id) %>%
  filter(n() > 1)

# 2. Remove duplicates
customers_unique <- customers %>%
  distinct(id, .keep_all = TRUE)

# Orders: multiple per customer is expected, so aggregate
orders_summary <- orders %>%
  group_by(cust_id) %>%
  summarize(
    n_orders = n(),
    total_amount = sum(amount),
    avg_amount = mean(amount),
    .groups = "drop"
  )

# 3. Perform safe join
result <- customers_unique %>%
  left_join(orders_summary, by = c("id" = "cust_id"))

# 4. Validate
cat("\nValidation:\n")
cat("Original customers:", nrow(customers_unique), "\n")
cat("Result rows:", nrow(result), "\n")
cat("Should be equal:", nrow(customers_unique) == nrow(result), "\n")

result
```

**Exercise 3:**

```{r}
safe_join <- function(x, y, by, type = "left") {
  # Validate inputs
  if (!type %in% c("inner", "left", "right", "full")) {
    stop("type must be inner, left, right, or full")
  }
  
  # Extract key columns
  if (is.character(by)) {
    x_keys <- if (is.null(names(by))) by else names(by)
    y_keys <- if (is.null(names(by))) by else unname(by)
  } else {
    stop("by must be a character vector")
  }
  
  # Check keys exist
  if (!all(x_keys %in% names(x))) {
    stop("Some keys not found in x: ", 
         paste(setdiff(x_keys, names(x)), collapse = ", "))
  }
  if (!all(y_keys %in% names(y))) {
    stop("Some keys not found in y: ", 
         paste(setdiff(y_keys, names(y)), collapse = ", "))
  }
  
  # Check for duplicates
  x_dups <- x %>%
    group_by(across(all_of(x_keys))) %>%
    filter(n() > 1) %>%
    nrow()
  
  y_dups <- y %>%
    group_by(across(all_of(y_keys))) %>%
    filter(n() > 1) %>%
    nrow()
  
  if (x_dups > 0) {
    warning("x has ", x_dups, " duplicate key rows")
  }
  if (y_dups > 0) {
    warning("y has ", y_dups, " duplicate key rows")
  }
  
  # Perform join
  result <- switch(type,
    inner = inner_join(x, y, by = by),
    left = left_join(x, y, by = by),
    right = right_join(x, y, by = by),
    full = full_join(x, y, by = by)
  )
  
  # Validate result
  metadata <- list(
    type = type,
    x_rows = nrow(x),
    y_rows = nrow(y),
    result_rows = nrow(result),
    x_duplicates = x_dups,
    y_duplicates = y_dups,
    na_columns = names(result)[colSums(is.na(result)) > 0]
  )
  
  # Check expectations
  if (type == "left" && nrow(result) < nrow(x)) {
    warning("Left join resulted in fewer rows than x!")
  }
  
  if (type == "inner" && nrow(result) > min(nrow(x), nrow(y))) {
    warning("Inner join resulted in more rows than either input - check for duplicates")
  }
  
  # Return with metadata
  attr(result, "join_metadata") <- metadata
  result
}

# Test
customers <- tibble(id = 1:3, name = c("A", "B", "C"))
orders <- tibble(customer_id = c(1, 1, 2, 4), amount = c(100, 150, 200, 75))

result <- safe_join(customers, orders, by = c("id" = "customer_id"), type = "left")
result
attr(result, "join_metadata")
```

**Exercise 4:**

```{r}
students <- tibble(id = 1:3, name = c("A", "B", "C"))
courses <- tibble(
  student_id = c(1, 1, 2, 3), 
  course = c("Math", "Eng", "Math", "Sci")
)
grades <- tibble(
  student_id = c(1, 1, 2, 3), 
  course = c("Math", "Eng", "Math", "Sci"), 
  grade = c(85, 90, 78, 92)
)

# Method 1: Chain joins
result1 <- students %>%
  left_join(courses, by = c("id" = "student_id")) %>%
  left_join(grades, by = c("id" = "student_id", "course" = "course"))
result1

# Method 2: Join courses and grades first
result2 <- students %>%
  left_join(
    courses %>%
      left_join(grades, by = c("student_id", "course")),
    by = c("id" = "student_id")
  )
result2

# Method 3: Using reduce (for many tables)
library(purrr)

tables <- list(
  students,
  courses,
  grades
)

keys <- list(
  c("id" = "student_id"),
  c("id" = "student_id", "course" = "course")
)

# This is complex - usually method 1 or 2 is better
# Showing for completeness

# Summary by student
summary_result <- students %>%
  left_join(
    courses %>%
      left_join(grades, by = c("student_id", "course")) %>%
      group_by(student_id) %>%
      summarize(
        n_courses = n(),
        avg_grade = mean(grade, na.rm = TRUE),
        courses = paste(course, collapse = ", "),
        .groups = "drop"
      ),
    by = c("id" = "student_id")
  )

summary_result
```
</details>
