# Introduction to tidyr {#tidyr-intro}

<div class="chapter-summary">
**What You'll Learn:**

- Tidy data principles
- Wide vs long format
- pivot_longer() and pivot_wider()
- Common reshaping errors
- When to reshape data

**Key Errors Covered:** 15+ tidyr errors

**Difficulty:** ⭐⭐ Intermediate to ⭐⭐⭐ Advanced
</div>

## Introduction

tidyr helps reshape data between wide and long formats:

```{r message=FALSE}
library(tidyr)
library(dplyr)

# Wide format
wide_data <- tibble(
  id = 1:3,
  time_1 = c(10, 20, 30),
  time_2 = c(15, 25, 35),
  time_3 = c(20, 30, 40)
)

# Convert to long format
long_data <- wide_data %>%
  pivot_longer(
    cols = starts_with("time"),
    names_to = "time",
    values_to = "value"
  )

long_data
```

Let's master reshaping!

## Tidy Data Principles

<div class="insight-box">
💡 **Key Insight: What is Tidy Data?**

```{r}
# Tidy data rules:
# 1. Each variable is a column
# 2. Each observation is a row
# 3. Each type of observational unit is a table

# Example: UNTIDY (wide format)
untidy <- tibble(
  person = c("Alice", "Bob"),
  age = c(25, 30),
  income_2020 = c(50000, 60000),
  income_2021 = c(52000, 62000),
  income_2022 = c(54000, 64000)
)

untidy

# TIDY (long format)
tidy <- untidy %>%
  pivot_longer(
    cols = starts_with("income"),
    names_to = "year",
    names_prefix = "income_",
    values_to = "income"
  )

tidy

# Now year is a variable, each row is person-year observation
```

**Why tidy data?**
- Works naturally with dplyr and ggplot2
- Easy to filter, group, and summarize
- Consistent structure for analysis
</div>

## pivot_longer() Basics

<div class="insight-box">
💡 **Key Insight: Wide to Long with pivot_longer()**

```{r}
# Basic pivot_longer
wide <- tibble(
  id = 1:3,
  q1 = c(5, 3, 4),
  q2 = c(4, 5, 3),
  q3 = c(3, 4, 5)
)

wide %>%
  pivot_longer(
    cols = q1:q3,           # Columns to pivot
    names_to = "quarter",   # New column for names
    values_to = "score"     # New column for values
  )

# Using selection helpers
wide %>%
  pivot_longer(
    cols = starts_with("q"),
    names_to = "quarter",
    values_to = "score"
  )

# Keep some columns
wide %>%
  pivot_longer(
    cols = -id,             # All except id
    names_to = "quarter",
    values_to = "score"
  )

# Multiple value types
data <- tibble(
  id = 1:2,
  x_value = c(10, 20),
  x_error = c(1, 2),
  y_value = c(30, 40),
  y_error = c(3, 4)
)

data %>%
  pivot_longer(
    cols = -id,
    names_to = c("variable", ".value"),
    names_sep = "_"
  )
```
</div>

## Error #1: `cols must select at least one column` {#pivot-no-cols}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-args">📋 ARGS</span>

### The Error

```{r error=TRUE}
data <- tibble(
  id = 1:3,
  value = c(10, 20, 30)
)

data %>%
  pivot_longer(
    cols = starts_with("time"),  # No columns match!
    names_to = "time",
    values_to = "measurement"
  )
```

<div class="error-box">
🔴 **ERROR**

```
Error in pivot_longer():
! `cols` must select at least one column.
```
</div>

### What It Means

The column selection didn't match any columns.

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Column Names**

```{r}
# Verify columns exist
names(data)

# Use correct selection
data_wide <- tibble(
  id = 1:3,
  time_1 = c(10, 20, 30),
  time_2 = c(15, 25, 35)
)

data_wide %>%
  pivot_longer(
    cols = starts_with("time"),
    names_to = "time",
    values_to = "value"
  )
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use Flexible Selection**

```{r}
# Select all numeric columns
data_wide %>%
  pivot_longer(
    cols = where(is.numeric),
    names_to = "variable",
    values_to = "value"
  )

# Or exclude specific columns
data_wide %>%
  pivot_longer(
    cols = -id,
    names_to = "variable",
    values_to = "value"
  )
```
</div>

## pivot_wider() Basics

<div class="insight-box">
💡 **Key Insight: Long to Wide with pivot_wider()**

```{r}
# Basic pivot_wider
long <- tibble(
  id = rep(1:3, each = 2),
  time = rep(c("before", "after"), 3),
  value = c(10, 15, 20, 25, 30, 35)
)

long %>%
  pivot_wider(
    names_from = time,      # Column to get names from
    values_from = value     # Column to get values from
  )

# Multiple value columns
long_multi <- tibble(
  id = rep(1:2, each = 2),
  time = rep(c("t1", "t2"), 2),
  score = c(10, 15, 20, 25),
  error = c(1, 2, 3, 4)
)

long_multi %>%
  pivot_wider(
    names_from = time,
    values_from = c(score, error)
  )

# With name formatting
long %>%
  pivot_wider(
    names_from = time,
    values_from = value,
    names_prefix = "time_"
  )
```
</div>

## Error #2: `Values are not uniquely identified` {#pivot-not-unique}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-logic">🧠 LOGIC</span>

### The Error

```{r error=TRUE}
# Data with duplicates
data <- tibble(
  id = c(1, 1, 2, 2),
  time = c("before", "before", "before", "after"),
  value = c(10, 15, 20, 25)
)

data %>%
  pivot_wider(
    names_from = time,
    values_from = value
  )
```

<div class="error-box">
🔴 **WARNING/ERROR**

```
Warning: Values from `value` are not uniquely identified; output will contain list-cols.
```
</div>

### What It Means

Multiple values exist for same id-time combination. Can't decide which to use.

### Common Causes

```{r}
# Multiple measurements per combination
measurements <- tibble(
  person = c("Alice", "Alice", "Bob", "Bob"),
  test = c("test1", "test1", "test2", "test2"),
  score = c(85, 87, 90, 92)  # Two scores for Alice-test1
)

measurements %>%
  pivot_wider(
    names_from = test,
    values_from = score
  )
# Creates list columns!
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Aggregate First**

```{r}
measurements %>%
  group_by(person, test) %>%
  summarize(score = mean(score), .groups = "drop") %>%
  pivot_wider(
    names_from = test,
    values_from = score
  )
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Add Identifier**

```{r}
# Add row number to make unique
measurements %>%
  group_by(person, test) %>%
  mutate(attempt = row_number()) %>%
  pivot_wider(
    names_from = c(test, attempt),
    values_from = score
  )
```
</div>

<div class="solution-box">
✅ **SOLUTION 3: Use values_fn**

```{r}
# Specify how to aggregate
measurements %>%
  pivot_wider(
    names_from = test,
    values_from = score,
    values_fn = list(score = mean)
  )

# Or use list for multiple aggregations
measurements %>%
  pivot_wider(
    names_from = test,
    values_from = score,
    values_fn = list(score = list)
  )
```
</div>

## separate() and unite()

<div class="insight-box">
💡 **Key Insight: Splitting and Combining Columns**

```{r}
# separate: split one column into multiple
data <- tibble(
  id = 1:3,
  name = c("Alice_Smith", "Bob_Jones", "Charlie_Brown")
)

data %>%
  separate(name, into = c("first", "last"), sep = "_")

# With different separator
data2 <- tibble(
  id = 1:3,
  date = c("2024-01-15", "2024-02-20", "2024-03-25")
)

data2 %>%
  separate(date, into = c("year", "month", "day"), sep = "-")

# Convert types
data2 %>%
  separate(date, into = c("year", "month", "day"), 
           sep = "-", convert = TRUE)

# unite: combine multiple columns
data_split <- tibble(
  id = 1:3,
  first = c("Alice", "Bob", "Charlie"),
  last = c("Smith", "Jones", "Brown")
)

data_split %>%
  unite(name, first, last, sep = " ")

# Keep original columns
data_split %>%
  unite(name, first, last, sep = " ", remove = FALSE)
```
</div>

## fill() for Missing Values

<div class="insight-box">
💡 **Key Insight: Fill Missing Values**

```{r}
# Data with implicit missing values
data <- tibble(
  group = c("A", NA, NA, "B", NA, "C"),
  value = 1:6
)

# Fill down
data %>%
  fill(group)

# Fill up
data %>%
  fill(group, .direction = "up")

# Fill both directions
data %>%
  fill(group, .direction = "downup")
```
</div>

## complete() for Implicit Missing Values

<div class="insight-box">
💡 **Key Insight: Make Implicit NAs Explicit**

```{r}
# Incomplete data
sales <- tibble(
  product = c("A", "A", "B"),
  quarter = c(1, 2, 1),
  sales = c(100, 150, 200)
)

# Missing: product B quarter 2
# Make it explicit
sales %>%
  complete(product, quarter)

# Fill with specific value
sales %>%
  complete(product, quarter, fill = list(sales = 0))

# With sequences
dates <- tibble(
  date = as.Date(c("2024-01-01", "2024-01-03", "2024-01-05")),
  value = c(10, 20, 30)
)

dates %>%
  complete(date = seq(min(date), max(date), by = "day"))
```
</div>

## Common Patterns

<div class="bestpractice-box">
🎯 **Best Practice: Reshaping Workflows**

```{r}
# 1. Survey data wide to long
survey_wide <- tibble(
  id = 1:3,
  q1_satisfaction = c(5, 4, 3),
  q1_importance = c(5, 5, 4),
  q2_satisfaction = c(4, 3, 5),
  q2_importance = c(4, 4, 5)
)

survey_long <- survey_wide %>%
  pivot_longer(
    cols = -id,
    names_to = c("question", "measure"),
    names_sep = "_",
    values_to = "rating"
  )

survey_long

# 2. Time series wide to long
stocks <- tibble(
  date = as.Date("2024-01-01") + 0:2,
  AAPL = c(150, 152, 149),
  GOOGL = c(140, 142, 138),
  MSFT = c(380, 385, 378)
)

stocks %>%
  pivot_longer(
    cols = -date,
    names_to = "stock",
    values_to = "price"
  )

# 3. Pivot wider for comparison
comparisons <- tibble(
  id = rep(1:3, each = 2),
  condition = rep(c("control", "treatment"), 3),
  outcome = c(10, 12, 15, 18, 20, 22)
)

comparisons %>%
  pivot_wider(
    names_from = condition,
    values_from = outcome
  ) %>%
  mutate(difference = treatment - control)

# 4. Complex reshape
data <- tibble(
  id = 1:2,
  treatment_1_pre = c(10, 15),
  treatment_1_post = c(12, 17),
  treatment_2_pre = c(20, 25),
  treatment_2_post = c(22, 27)
)

data %>%
  pivot_longer(
    cols = -id,
    names_to = c("treatment", "time"),
    names_pattern = "treatment_(.*)_(.*)",
    values_to = "score"
  )
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Tidy data principles** - Variables in columns, observations in rows
2. **pivot_longer()** - Wide to long format
3. **pivot_wider()** - Long to wide format
4. **Check for duplicates** - Before pivot_wider()
5. **separate() and unite()** - Split/combine columns
6. **fill()** - Handle missing values
7. **complete()** - Make implicit NAs explicit

**Quick Reference:**

| Function | Purpose | Common Use |
|----------|---------|------------|
| pivot_longer() | Wide → Long | Convert multiple columns to rows |
| pivot_wider() | Long → Wide | Spread values across columns |
| separate() | Split column | Parse combined data |
| unite() | Combine columns | Create composite keys |
| fill() | Fill NAs | Propagate values |
| complete() | Add missing combos | Explicit NAs |

**Reshaping:**

```{r eval=FALSE}
# Wide to long
data %>%
  pivot_longer(
    cols = col1:col3,
    names_to = "variable",
    values_to = "value"
  )

# Long to wide
data %>%
  pivot_wider(
    names_from = variable,
    values_from = value
  )

# Split column
data %>%
  separate(col, into = c("part1", "part2"), sep = "_")

# Combine columns
data %>%
  unite(new_col, col1, col2, sep = "_")

# Fill missing
data %>%
  fill(col, .direction = "down")

# Complete combinations
data %>%
  complete(var1, var2, fill = list(value = 0))
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
Check data structure first                # Know what you have
Aggregate before pivot_wider              # Avoid list columns
Use descriptive names_to/values_to        # Clear column names
Test on small sample first                # Verify logic

# ❌ Avoid
Pivot_wider with duplicates               # Creates list columns
Complex reshapes in one step              # Hard to debug
Forgetting to specify cols                # Error or unexpected
Not checking result structure             # May not be what you want
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Wide to Long**

Given:
```{r eval=FALSE}
grades <- tibble(
  student = c("Alice", "Bob", "Charlie"),
  math_q1 = c(85, 78, 92),
  math_q2 = c(88, 82, 95),
  english_q1 = c(90, 85, 88),
  english_q2 = c(92, 87, 90)
)
```

Convert to long format with columns: student, subject, quarter, grade
</div>

<div class="exercise-box">
📝 **Exercise 2: Long to Wide**

Given:
```{r eval=FALSE}
measurements <- tibble(
  id = rep(1:3, each = 3),
  time = rep(c("t1", "t2", "t3"), 3),
  value = c(10, 12, 14, 20, 22, 24, 30, 32, 34)
)
```

Convert to wide format with time as columns
</div>

<div class="exercise-box">
📝 **Exercise 3: Safe Pivot**

Write `safe_pivot_wider(data, names_from, values_from)` that:
1. Checks for duplicate combinations
2. Aggregates if needed
3. Performs pivot
4. Validates result
</div>

<div class="exercise-box">
📝 **Exercise 4: Complex Reshape**

Given survey data:
```{r eval=FALSE}
survey <- tibble(
  id = 1:3,
  age = c(25, 30, 35),
  q1_pre = c(3, 4, 2),
  q1_post = c(4, 5, 3),
  q2_pre = c(2, 3, 4),
  q2_post = c(3, 4, 5)
)
```

Reshape to: id, age, question, time, response
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
grades <- tibble(
  student = c("Alice", "Bob", "Charlie"),
  math_q1 = c(85, 78, 92),
  math_q2 = c(88, 82, 95),
  english_q1 = c(90, 85, 88),
  english_q2 = c(92, 87, 90)
)

# Method 1: Two-step pivot
result <- grades %>%
  pivot_longer(
    cols = -student,
    names_to = "subject_quarter",
    values_to = "grade"
  ) %>%
  separate(subject_quarter, into = c("subject", "quarter"), sep = "_")

result

# Method 2: Use names_pattern
result2 <- grades %>%
  pivot_longer(
    cols = -student,
    names_to = c("subject", "quarter"),
    names_pattern = "(.*)_(.*)",
    values_to = "grade"
  )

result2
```

**Exercise 2:**

```{r}
measurements <- tibble(
  id = rep(1:3, each = 3),
  time = rep(c("t1", "t2", "t3"), 3),
  value = c(10, 12, 14, 20, 22, 24, 30, 32, 34)
)

result <- measurements %>%
  pivot_wider(
    names_from = time,
    values_from = value
  )

result

# Verify dimensions
cat("Original: ", nrow(measurements), "rows\n")
cat("Wide: ", nrow(result), "rows,", ncol(result) - 1, "time columns\n")
```

**Exercise 3:**

```{r}
safe_pivot_wider <- function(data, names_from, values_from, 
                             aggregate_fn = mean, warn = TRUE) {
  
  names_col <- rlang::ensym(names_from)
  values_col <- rlang::ensym(values_from)
  
  # Get ID columns (everything except names_from and values_from)
  id_cols <- setdiff(names(data), c(as.character(names_col), 
                                     as.character(values_col)))
  
  # Check for duplicates
  dup_check <- data %>%
    group_by(across(all_of(c(id_cols, as.character(names_col))))) %>%
    filter(n() > 1) %>%
    ungroup()
  
  if (nrow(dup_check) > 0) {
    if (warn) {
      warning("Found ", nrow(dup_check), " duplicate rows. ",
              "Aggregating with ", deparse(substitute(aggregate_fn)))
      print(head(dup_check))
    }
    
    # Aggregate
    data <- data %>%
      group_by(across(all_of(c(id_cols, as.character(names_col))))) %>%
      summarize(
        !!values_col := aggregate_fn(!!values_col),
        .groups = "drop"
      )
  }
  
  # Perform pivot
  result <- data %>%
    pivot_wider(
      names_from = !!names_col,
      values_from = !!values_col
    )
  
  # Validate
  expected_rows <- length(unique(interaction(data[id_cols])))
  actual_rows <- nrow(result)
  
  if (expected_rows != actual_rows) {
    warning("Row count mismatch! Expected: ", expected_rows, 
            ", Got: ", actual_rows)
  }
  
  result
}

# Test with clean data
measurements <- tibble(
  id = rep(1:3, each = 2),
  time = rep(c("t1", "t2"), 3),
  value = c(10, 15, 20, 25, 30, 35)
)

safe_pivot_wider(measurements, time, value)

# Test with duplicates
measurements_dup <- tibble(
  id = c(1, 1, 2, 2, 2),
  time = c("t1", "t1", "t1", "t2", "t2"),
  value = c(10, 12, 20, 25, 27)
)

safe_pivot_wider(measurements_dup, time, value)
```

**Exercise 4:**

```{r}
survey <- tibble(
  id = 1:3,
  age = c(25, 30, 35),
  q1_pre = c(3, 4, 2),
  q1_post = c(4, 5, 3),
  q2_pre = c(2, 3, 4),
  q2_post = c(3, 4, 5)
)

# Method 1: Step by step
result <- survey %>%
  pivot_longer(
    cols = starts_with("q"),
    names_to = "question_time",
    values_to = "response"
  ) %>%
  separate(question_time, into = c("question", "time"), sep = "_")

result

# Method 2: Using names_pattern
result2 <- survey %>%
  pivot_longer(
    cols = starts_with("q"),
    names_to = c("question", "time"),
    names_pattern = "(q\\d+)_(.*)",
    values_to = "response"
  )

result2

# Verify
cat("Original questions:", (ncol(survey) - 2), "\n")
cat("Long format rows:", nrow(result), "\n")
cat("Should be:", nrow(survey) * (ncol(survey) - 2), "\n")
```
</details>
