# Data Manipulation Best Practices {#dplyr-best-practices}

<div class="chapter-summary">
**What You'll Learn:**

- Efficient dplyr workflows
- Performance optimization
- Code organization
- Common anti-patterns
- Debugging strategies

**Key Errors Covered:** 10+ workflow errors

**Difficulty:** ⭐⭐ Intermediate to ⭐⭐⭐ Advanced
</div>

## Introduction

Best practices make code faster, clearer, and more maintainable:

```{r message=FALSE}
library(dplyr)
library(tidyr)

# Good: Clear pipeline
result <- mtcars %>%
  filter(mpg > 20) %>%
  select(mpg, cyl, hp) %>%
  mutate(hp_per_cyl = hp / cyl) %>%
  arrange(desc(hp_per_cyl))

head(result, 3)
```

## Performance Optimization

<div class="bestpractice-box">
🎯 **Best Practice: Optimize for Speed**

```{r}
# 1. Filter early
# Bad: filter after heavy operations
mtcars %>%
  mutate(complex = mpg * hp / wt) %>%
  left_join(mtcars, by = c("cyl")) %>%
  filter(mpg > 20)

# Good: filter first
mtcars %>%
  filter(mpg > 20) %>%
  mutate(complex = mpg * hp / wt) %>%
  left_join(mtcars, by = c("cyl"))

# 2. Select early
# Bad: keep all columns
mtcars %>%
  mutate(new = mpg * 2) %>%
  filter(cyl == 4) %>%
  select(mpg, cyl, new)

# Good: select needed columns early
mtcars %>%
  select(mpg, cyl, hp, wt) %>%
  filter(cyl == 4) %>%
  mutate(new = mpg * 2)

# 3. Use data.table for large data
# library(dtplyr)
# lazy_dt(large_data) %>%
#   filter(...) %>%
#   as_tibble()
```
</div>

## Code Organization

<div class="bestpractice-box">
🎯 **Best Practice: Readable Pipelines**

```{r}
# Good: One operation per line
result <- mtcars %>%
  filter(mpg > 20) %>%
  mutate(
    hp_per_ton = hp / wt,
    efficiency = mpg / hp
  ) %>%
  group_by(cyl) %>%
  summarize(
    avg_efficiency = mean(efficiency),
    n = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_efficiency))

# Good: Use intermediate variables for clarity
filtered_cars <- mtcars %>%
  filter(mpg > 20, cyl %in% c(4, 6))

efficiency_metrics <- filtered_cars %>%
  mutate(
    hp_per_ton = hp / wt,
    efficiency = mpg / hp
  )

final_summary <- efficiency_metrics %>%
  group_by(cyl) %>%
  summarize(
    avg_efficiency = mean(efficiency),
    n = n(),
    .groups = "drop"
  )

# Bad: Everything in one long pipe
# result <- mtcars %>% filter(mpg > 20) %>% mutate(hp_per_ton = hp / wt, efficiency = mpg / hp) %>% group_by(cyl) %>% summarize(avg_efficiency = mean(efficiency), n = n(), .groups = "drop") %>% arrange(desc(avg_efficiency))
```
</div>

## Common Anti-Patterns

<div class="pitfall-box">
⚠️ **Avoid These Patterns**

```{r}
# Anti-pattern 1: Growing data frame in loop
# Bad
result <- tibble()
for (i in 1:100) {
  result <- bind_rows(result, tibble(x = i, y = i^2))
}

# Good
result <- tibble(
  x = 1:100,
  y = x^2
)

# Anti-pattern 2: Unnecessary grouping
# Bad
mtcars %>%
  group_by(cyl) %>%
  select(mpg, cyl, hp)  # select doesn't need groups

# Good
mtcars %>%
  select(mpg, cyl, hp) %>%
  group_by(cyl)  # Group only when needed

# Anti-pattern 3: Repeated similar code
# Bad
mtcars %>%
  summarize(
    mean_mpg = mean(mpg),
    mean_hp = mean(hp),
    mean_wt = mean(wt),
    mean_disp = mean(disp)
  )

# Good
mtcars %>%
  summarize(across(c(mpg, hp, wt, disp), mean, .names = "mean_{.col}"))

# Anti-pattern 4: Ignoring NAs
# Bad
data_with_na <- tibble(x = c(1, 2, NA, 4))
data_with_na %>%
  summarize(mean_x = mean(x))  # Returns NA!

# Good
data_with_na %>%
  summarize(mean_x = mean(x, na.rm = TRUE))
```
</div>

## Debugging Strategies

<div class="bestpractice-box">
🎯 **Best Practice: Debug Effectively**

```{r}
# 1. Check intermediate steps
mtcars %>%
  filter(mpg > 20) %>%
  {cat("After filter:", nrow(.), "rows\n"); .} %>%
  select(mpg, cyl, hp) %>%
  {cat("After select:", ncol(.), "cols\n"); .} %>%
  head(3)

# 2. Use View() in RStudio
# mtcars %>%
#   filter(mpg > 20) %>%
#   View()

# 3. Print summary statistics
debug_data <- function(data, label = "") {
  cat("\n=== Debug:", label, "===\n")
  cat("Rows:", nrow(data), "\n")
  cat("Cols:", ncol(data), "\n")
  cat("Columns:", paste(names(data), collapse = ", "), "\n")
  data
}

mtcars %>%
  debug_data("original") %>%
  filter(mpg > 20) %>%
  debug_data("after filter")

# 4. Check for common issues
check_pipeline <- function(data) {
  cat("=== Pipeline Check ===\n")
  cat("Rows:", nrow(data), "\n")
  cat("NA counts:\n")
  print(colSums(is.na(data)))
  cat("\nGrouping:", paste(group_vars(data), collapse = ", "), "\n")
  cat("Grouped:", is_grouped_df(data), "\n")
  data
}
```
</div>

## Documentation

<div class="bestpractice-box">
🎯 **Best Practice: Document Complex Operations**

```{r}
# Good: Comments explain WHY, not WHAT
clean_sales_data <- function(data) {
  data %>%
    # Remove test transactions (internal use only)
    filter(customer_id >= 1000) %>%
    # Standardize date format for downstream systems
    mutate(date = as.Date(date)) %>%
    # Calculate revenue (excluding tax per accounting rules)
    mutate(revenue = quantity * price) %>%
    # Remove cancelled orders (status code 99)
    filter(status != 99)
}

# Good: Function documentation
#' Calculate customer lifetime value
#'
#' @param orders Data frame with customer_id, date, amount
#' @param window_days Number of days to consider (default: 365)
#' @return Data frame with customer_id and lifetime_value
calculate_ltv <- function(orders, window_days = 365) {
  cutoff_date <- Sys.Date() - window_days
  
  orders %>%
    filter(date >= cutoff_date) %>%
    group_by(customer_id) %>%
    summarize(
      lifetime_value = sum(amount),
      n_orders = n(),
      .groups = "drop"
    )
}
```
</div>

## Error Handling

<div class="bestpractice-box">
🎯 **Best Practice: Handle Errors Gracefully**

```{r}
# Validate inputs
safe_summarize <- function(data, col) {
  # Check data frame
  if (!is.data.frame(data)) {
    stop("data must be a data frame")
  }
  
  # Check column exists
  if (!col %in% names(data)) {
    stop("Column '", col, "' not found. ",
         "Available: ", paste(names(data), collapse = ", "))
  }
  
  # Check numeric
  if (!is.numeric(data[[col]])) {
    stop("Column '", col, "' must be numeric")
  }
  
  # Perform operation
  data %>%
    summarize(
      mean = mean(.data[[col]], na.rm = TRUE),
      sd = sd(.data[[col]], na.rm = TRUE),
      n = sum(!is.na(.data[[col]]))
    )
}

# Test
safe_summarize(mtcars, "mpg")
```

```{r error=TRUE}
safe_summarize(mtcars, "missing_col")
```
</div>

## Testing Pipelines

<div class="bestpractice-box">
🎯 **Best Practice: Test Your Code**

```{r}
# Create test function
test_clean_data <- function() {
  # Setup test data
  test_data <- tibble(
    id = 1:5,
    value = c(10, NA, 30, 40, 50),
    group = c("A", "A", "B", "B", "C")
  )
  
  # Run pipeline
  result <- test_data %>%
    filter(!is.na(value)) %>%
    group_by(group) %>%
    summarize(mean_value = mean(value), .groups = "drop")
  
  # Check results
  stopifnot(nrow(result) == 3)  # 3 groups
  stopifnot(all(!is.na(result$mean_value)))  # No NAs
  stopifnot("group" %in% names(result))  # Has group column
  
  message("✓ All tests passed")
  TRUE
}

test_clean_data()
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Filter early** - Reduce data size first
2. **Select needed columns** - Don't carry extra data
3. **One operation per line** - Readable pipelines
4. **Use across()** - Avoid repetition
5. **Handle NAs** - Always specify na.rm
6. **Ungroup** - Clean up after group operations
7. **Document WHY** - Not what code does
8. **Test code** - Catch errors early

**Best Practices:**

```{r eval=FALSE}
# ✅ Excellent pipeline
result <- data %>%
  # Filter first (reduce data)
  filter(important_condition) %>%
  # Select needed columns
  select(needed_cols) %>%
  # Transform
  mutate(new_col = calculation) %>%
  # Aggregate
  group_by(category) %>%
  summarize(
    metric = mean(value, na.rm = TRUE),
    n = n(),
    .groups = "drop"  # Always ungroup
  ) %>%
  # Sort
  arrange(desc(metric))

# ✅ Use functions for repeated logic
calculate_stats <- function(data, group_var, value_var) {
  data %>%
    group_by({{ group_var }}) %>%
    summarize(
      across({{ value_var }},
             list(mean = ~mean(., na.rm = TRUE),
                  sd = ~sd(., na.rm = TRUE)),
             .names = "{.col}_{.fn}"),
      .groups = "drop"
    )
}

# ✅ Document complex operations
# Remove outliers using IQR method (1.5 * IQR)
clean_data <- data %>%
  group_by(category) %>%
  filter(
    value >= quantile(value, 0.25) - 1.5 * IQR(value),
    value <= quantile(value, 0.75) + 1.5 * IQR(value)
  ) %>%
  ungroup()
```

**Anti-Patterns to Avoid:**

```{r eval=FALSE}
# ❌ Bad
Growing data in loops
Unnecessary grouping
Repeated similar code
Ignoring NAs
No error checking
Complex one-liners
No comments on complex logic

# ✅ Good
Vectorized operations
Group only when needed
Use across() for repetition
Always handle NAs
Validate inputs
Break into readable steps
Document WHY not WHAT
```
</div>

## Exercise

<div class="exercise-box">
📝 **Exercise: Refactor Pipeline**

Improve this code:

```{r eval=FALSE}
result <- data %>%
  mutate(x = a + b, y = c + d, z = e + f) %>%
  filter(category == "A" | category == "B" | category == "C") %>%
  group_by(category) %>%
  summarize(mean_x = mean(x), mean_y = mean(y), mean_z = mean(z)) %>%
  filter(mean_x > 0)
```
</div>

## Exercise Answer

<details>
<summary>Click to see answer</summary>

```{r eval=FALSE}
# Refactored version
result <- data %>%
  # Filter early for better performance
  filter(category %in% c("A", "B", "C")) %>%
  # Use across() to avoid repetition
  mutate(
    across(c(a, b), .names = "sum_{.col}",
           .fns = ~ . + .)
  ) %>%
  # Group for aggregation
  group_by(category) %>%
  # Use across() in summarize
  summarize(
    across(starts_with("sum_"),
           mean,
           na.rm = TRUE,
           .names = "mean_{.col}"),
    .groups = "drop"
  ) %>%
  # Filter on summarized values
  filter(mean_sum_a > 0)
```
</details>

## Completion

<div class="chapter-summary">
**Part VIII Complete!**

You've mastered:
- dplyr fundamentals (filter, select, mutate, etc.)
- Grouping and summarizing
- Joining data
- Reshaping with tidyr
- Advanced tidy evaluation
- Best practices and optimization

**Ready for:** Part IX (Graphics) or other advanced topics!
</div>
