# Part IX: Graphics (ggplot2) {-}

# Introduction to ggplot2 {#ggplot2-intro}

<div class="chapter-summary">
**What You'll Learn:**

- Grammar of Graphics principles
- Basic ggplot2 structure
- Common plotting errors
- Aesthetics and geoms
- Layer system

**Key Errors Covered:** 20+ ggplot2 errors

**Difficulty:** ⭐⭐ Intermediate to ⭐⭐⭐ Advanced
</div>

## Introduction

ggplot2 revolutionized R graphics with the Grammar of Graphics:

```{r message=FALSE, warning=FALSE}
library(ggplot2)

# Basic plot
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point()
```

But ggplot2 has unique error patterns. Let's master them!

## ggplot2 Structure

<div class="insight-box">
💡 **Key Insight: Grammar of Graphics**

```{r}
# Three essential components:
# 1. Data
# 2. Aesthetic mappings (aes)
# 3. Geometric objects (geom)

# Basic structure
ggplot(data = mtcars, mapping = aes(x = mpg, y = hp)) +
  geom_point()

# Shortened (common)
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point()

# Can specify aes in geom instead
ggplot(mtcars) +
  geom_point(aes(x = mpg, y = hp))

# Or mix (useful for multiple layers)
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  geom_smooth()
```

**Key principle:** Build plots in layers with `+`
</div>

## Error #1: `object not found` in aes() {#ggplot-object-not-found}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-scope">🔍 SCOPE</span>

### The Error

```{r error=TRUE}
ggplot(mtcars, aes(x = mpg, y = horsepower)) +
  geom_point()
```

<div class="error-box">
🔴 **ERROR**

```
Error in FUN(X[[i]], ...) : object 'horsepower' not found
```
</div>

### What It Means

The column name doesn't exist in the data.

### Common Causes

```{r error=TRUE}
# Typo in column name
ggplot(mtcars, aes(x = mpgg, y = hp)) +
  geom_point()

# Wrong dataset
ggplot(iris, aes(x = mpg, y = hp)) +
  geom_point()

# Forgot to create column
ggplot(mtcars, aes(x = mpg, y = efficiency)) +
  geom_point()
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check Column Names**

```{r}
# Verify columns
names(mtcars)

# Use correct name
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point()
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Create Column First**

```{r}
library(dplyr)

mtcars %>%
  mutate(efficiency = mpg / hp) %>%
  ggplot(aes(x = hp, y = efficiency)) +
  geom_point()
```
</div>

## Error #2: Using `+` vs `%>%` {#ggplot-plus-vs-pipe}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-syntax">🔤 SYNTAX</span>

### The Error

```{r error=TRUE}
library(dplyr)

mtcars %>%
  filter(cyl == 4) %>%
  ggplot(aes(x = mpg, y = hp)) %>%  # Wrong operator!
  geom_point()
```

<div class="error-box">
🔴 **ERROR**

```
Error in geom_point(.) : 
  Cannot use `+` with a ggplot object. Did you accidentally use `%>%` instead of `+`?
```
</div>

### What It Means

Must use `+` to add ggplot2 layers, not `%>%`.

### Solutions

<div class="solution-box">
✅ **SOLUTION: Use + for ggplot layers**

```{r}
# Correct: + for ggplot
mtcars %>%
  filter(cyl == 4) %>%
  ggplot(aes(x = mpg, y = hp)) +  # Use +
  geom_point()

# Pipe data into ggplot, then use +
mtcars %>%
  filter(cyl == 4) %>%
  ggplot(aes(x = mpg, y = hp)) +
  geom_point() +
  theme_minimal()
```
</div>

## Aesthetics (aes)

<div class="insight-box">
💡 **Key Insight: Aesthetic Mappings**

```{r}
# Map variables to visual properties
ggplot(mtcars, aes(x = mpg, y = hp, color = factor(cyl))) +
  geom_point()

# Common aesthetics:
# x, y - position
# color - point/line color
# fill - area fill color
# size - point/line size
# shape - point shape
# alpha - transparency
# linetype - line pattern

# Multiple aesthetics
ggplot(mtcars, aes(x = mpg, y = hp, 
                   color = factor(cyl),
                   size = wt)) +
  geom_point()

# Set vs map
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(color = "blue")  # Set: all points blue

ggplot(mtcars, aes(x = mpg, y = hp, color = factor(cyl))) +
  geom_point()  # Map: color varies by cyl
```
</div>

## Error #3: Aesthetic outside aes() {#aesthetic-outside-aes}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-logic">🧠 LOGIC</span>

### The Error

```{r error=TRUE}
# Trying to map cyl to color outside aes()
ggplot(mtcars) +
  geom_point(aes(x = mpg, y = hp), color = cyl)
```

<div class="error-box">
🔴 **ERROR**

```
Error in layer(...) : object 'cyl' not found
```
</div>

### What It Means

Variable mappings must be inside `aes()`.

### Common Causes

```{r error=TRUE}
# Want color by variable
ggplot(mtcars) +
  geom_point(aes(x = mpg, y = hp), color = factor(cyl))

# Want size by variable
ggplot(mtcars) +
  geom_point(aes(x = mpg, y = hp), size = wt)
```

### Solutions

<div class="solution-box">
✅ **SOLUTION: Put Variable Mappings in aes()**

```{r}
# Correct: color mapping inside aes
ggplot(mtcars) +
  geom_point(aes(x = mpg, y = hp, color = factor(cyl)))

# Can be in ggplot() aes
ggplot(mtcars, aes(x = mpg, y = hp, color = factor(cyl))) +
  geom_point()

# Fixed values go OUTSIDE aes
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(color = "blue", size = 3)  # All points same
```
</div>

<div class="pitfall-box">
⚠️ **Common Confusion: Inside vs Outside aes()**

```{r}
# INSIDE aes(): varies by data
ggplot(mtcars, aes(x = mpg, y = hp, color = factor(cyl))) +
  geom_point()  # Color varies by cyl

# OUTSIDE aes(): fixed for all
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(color = "red")  # All points red

# Wrong: puts string in aes
ggplot(mtcars, aes(x = mpg, y = hp, color = "red")) +
  geom_point()  # Creates legend for "red"!
```
</div>

## Common geoms

<div class="insight-box">
💡 **Key Insight: Geometric Objects**

```{r}
# Points
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point()

# Lines
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line()

# Bars
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar()

# Histogram
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 10)

# Boxplot
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot()

# Smooth
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  geom_smooth(method = "lm")

# Text
ggplot(mtcars, aes(x = mpg, y = hp, label = rownames(mtcars))) +
  geom_text(size = 3)
```
</div>

## Error #4: `stat_count() requires x or y` {#stat-count-error}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-args">📋 ARGS</span>

### The Error

```{r error=TRUE}
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_bar()
```

<div class="error-box">
🔴 **ERROR**

```
Error in `geom_bar()`:
! Problem while computing stat.
ℹ Error occurred in the 1st layer.
Caused by error:
! `stat_count()` must only have an `x` or `y` aesthetic.
```
</div>

### What It Means

`geom_bar()` is for counts. For pre-computed heights, use `geom_col()`.

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use geom_col() for Heights**

```{r}
# Pre-computed values
data <- data.frame(
  category = c("A", "B", "C"),
  value = c(10, 15, 20)
)

ggplot(data, aes(x = category, y = value)) +
  geom_col()

# Or use stat = "identity" with geom_bar
ggplot(data, aes(x = category, y = value)) +
  geom_bar(stat = "identity")
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use geom_bar() for Counts**

```{r}
# Count occurrences
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar()

# Equivalent to
mtcars %>%
  count(cyl) %>%
  ggplot(aes(x = factor(cyl), y = n)) +
  geom_col()
```
</div>

## Faceting

<div class="insight-box">
💡 **Key Insight: Small Multiples**

```{r}
# Facet by one variable
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  facet_wrap(~ cyl)

# Facet by two variables
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  facet_grid(cyl ~ gear)

# Free scales
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  facet_wrap(~ cyl, scales = "free")

# Number of columns
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  facet_wrap(~ cyl, ncol = 2)
```
</div>

## Themes and Customization

<div class="bestpractice-box">
🎯 **Best Practice: Customizing Plots**

```{r}
# Built-in themes
p <- ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point()

p + theme_minimal()
p + theme_classic()
p + theme_bw()

# Custom labels
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  labs(
    title = "Fuel Efficiency vs Horsepower",
    subtitle = "Motor Trend Car Road Tests",
    x = "Miles per Gallon",
    y = "Horsepower",
    caption = "Source: mtcars dataset"
  )

# Customize theme elements
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 10),
    panel.grid.minor = element_blank()
  )
```
</div>

## Error #5: Non-numeric variable for histogram {#histogram-non-numeric}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_histogram()
```

<div class="error-box">
🔴 **ERROR**

```
Error in `geom_histogram()`:
! `stat_bin()` requires a numeric `x` variable
```
</div>

### What It Means

Histograms need continuous numeric data, not categorical.

### Solutions

<div class="solution-box">
✅ **SOLUTION: Use Appropriate Geom**

```{r}
# For categorical: use geom_bar
ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar()

# For continuous: histogram works
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 10)

# Density plot alternative
ggplot(mtcars, aes(x = mpg)) +
  geom_density()
```
</div>

## Saving Plots

<div class="bestpractice-box">
🎯 **Best Practice: Save Plots**

```{r eval=FALSE}
# Create plot
p <- ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  theme_minimal()

# Save with ggsave
ggsave("plot.png", p, width = 6, height = 4, dpi = 300)

# Or save last plot
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point()

ggsave("last_plot.png", width = 6, height = 4)

# Different formats
ggsave("plot.pdf", p)
ggsave("plot.svg", p)
ggsave("plot.jpg", p)
```
</div>

## Common Patterns

<div class="bestpractice-box">
🎯 **Best Practice: Common Plot Types**

```{r}
# Scatterplot with trend line
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()

# Grouped bar chart
mtcars %>%
  count(cyl, gear) %>%
  ggplot(aes(x = factor(cyl), y = n, fill = factor(gear))) +
  geom_col(position = "dodge")

# Boxplot with points
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.3)

# Time series
ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line() +
  theme_minimal() +
  labs(title = "US Unemployment Over Time")
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Three components** - Data, aes(), geom
2. **Use + not %>%** - Add layers with +
3. **Variables in aes()** - Fixed values outside
4. **geom_bar() vs geom_col()** - Counts vs heights
5. **Check column names** - Before plotting
6. **Histograms need numeric** - Use geom_bar() for categorical
7. **Build in layers** - Add components step by step

**Quick Reference:**

| Error | Cause | Fix |
|-------|-------|-----|
| object not found | Column doesn't exist | Check names(data) |
| Can't use + | Used %>% instead of + | Use + for ggplot layers |
| object 'var' not found | Variable outside aes | Put in aes() |
| stat_count requires x or y | geom_bar with y | Use geom_col() |
| requires numeric x | Non-numeric histogram | Use appropriate geom |

**Basic Structure:**

```{r eval=FALSE}
# Template
ggplot(data, aes(x = var1, y = var2)) +
  geom_point() +
  theme_minimal()

# With pipes
data %>%
  filter(condition) %>%
  ggplot(aes(x = var1, y = var2)) +  # Use +
  geom_point() +
  labs(title = "Plot Title")

# Common aesthetics
aes(
  x = var,           # x-axis
  y = var,           # y-axis
  color = var,       # point/line color
  fill = var,        # area fill
  size = var,        # size
  shape = var,       # point shape
  alpha = var,       # transparency
  linetype = var     # line pattern
)

# Common geoms
geom_point()         # scatter
geom_line()          # line
geom_bar()           # bar (counts)
geom_col()           # bar (heights)
geom_histogram()     # histogram
geom_boxplot()       # boxplot
geom_smooth()        # trend line
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
ggplot(data, aes(x = var1, y = var2, color = var3)) +
  geom_point() +
  theme_minimal()

data %>% filter(...) %>%
  ggplot(aes(x = var)) +  # + not %>%
  geom_histogram()

# ❌ Avoid
ggplot(data, aes(x = var1, y = var2)) %>%  # Wrong operator
  geom_point()

ggplot(data) +
  geom_point(aes(x = var), color = other_var)  # Should be in aes

geom_histogram(aes(x = factor_var))  # Need numeric
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Basic Plot**

Create a scatterplot of mtcars:
1. mpg vs hp
2. Color by cyl
3. Size by wt
4. Add title and labels
5. Use theme_minimal()
</div>

<div class="exercise-box">
📝 **Exercise 2: Error Fixing**

Fix these errors:
```{r eval=FALSE}
# Error 1
ggplot(mtcars, aes(x = mpg, y = hp)) %>%
  geom_point()

# Error 2
ggplot(mtcars) +
  geom_point(aes(x = mpg, y = hp), color = cyl)

# Error 3
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_histogram()
```
</div>

<div class="exercise-box">
📝 **Exercise 3: Multiple Geoms**

Create a plot with:
1. Points for raw data
2. Smooth line for trend
3. Faceted by cyl
4. Custom colors
</div>

<div class="exercise-box">
📝 **Exercise 4: Bar Chart**

Using mtcars:
1. Count cars by cyl
2. Fill by gear
3. Dodge position
4. Add labels
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
ggplot(mtcars, aes(x = mpg, y = hp, color = factor(cyl), size = wt)) +
  geom_point(alpha = 0.7) +
  labs(
    title = "Fuel Efficiency vs Horsepower",
    subtitle = "Motor Trend Car Road Tests",
    x = "Miles per Gallon",
    y = "Horsepower",
    color = "Cylinders",
    size = "Weight (1000 lbs)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )
```

**Exercise 2:**

```{r}
# Error 1: Use + not %>%
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point()

# Error 2: Put variable in aes()
ggplot(mtcars) +
  geom_point(aes(x = mpg, y = hp, color = factor(cyl)))

# Error 3: Use geom_boxplot for this
ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot()

# Or if want histogram of mpg
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(bins = 10)
```

**Exercise 3:**

```{r}
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl)), size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  facet_wrap(~ cyl, labeller = label_both) +
  scale_color_manual(values = c("4" = "#E41A1C", "6" = "#377EB8", "8" = "#4DAF4A")) +
  labs(
    title = "MPG vs HP by Number of Cylinders",
    x = "Miles per Gallon",
    y = "Horsepower",
    color = "Cylinders"
  ) +
  theme_bw()
```

**Exercise 4:**

```{r}
library(dplyr)

mtcars %>%
  count(cyl, gear) %>%
  ggplot(aes(x = factor(cyl), y = n, fill = factor(gear))) +
  geom_col(position = "dodge") +
  labs(
    title = "Car Count by Cylinders and Gears",
    x = "Number of Cylinders",
    y = "Count",
    fill = "Number of Gears"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```
</details>
