# Advanced ggplot2 {#ggplot2-advanced}

<div class="chapter-summary">
**What You'll Learn:**

- Scales and coordinate systems
- Advanced themes
- Annotations and labels
- Multiple plots
- Statistical transformations

**Key Errors Covered:** 18+ advanced ggplot2 errors

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Introduction

Advanced ggplot2 enables publication-quality graphics:

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

# Advanced plot
ggplot(mtcars, aes(x = mpg, y = hp, color = wt)) +
  geom_point(size = 3) +
  scale_color_gradient(low = "blue", high = "red") +
  scale_x_continuous(breaks = seq(10, 35, 5)) +
  coord_cartesian(ylim = c(50, 350)) +
  theme_minimal() +
  labs(title = "Horsepower vs MPG")
```

## Scales

<div class="insight-box">
💡 **Key Insight: Scale Functions**

```{r}
# Continuous scales
ggplot(mtcars, aes(x = mpg, y = hp, color = wt)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red")

# Discrete scales
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_boxplot() +
  scale_fill_manual(values = c("4" = "green", "6" = "blue", "8" = "red"))

# Log scales
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  scale_y_log10()
```
</div>

## Error #1: `Discrete value supplied to continuous scale` {#discrete-continuous}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
ggplot(mtcars, aes(x = mpg, y = hp, color = cyl)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red")
```

<div class="error-box">
🔴 **ERROR**

```
Error: Discrete value supplied to continuous scale
```
</div>

### Solutions

<div class="solution-box">
✅ **SOLUTION: Match Scale to Data Type**

```{r}
# For discrete: use discrete scale
ggplot(mtcars, aes(x = mpg, y = hp, color = factor(cyl))) +
  geom_point() +
  scale_color_manual(values = c("4" = "green", "6" = "blue", "8" = "red"))

# For continuous: ensure numeric
ggplot(mtcars, aes(x = mpg, y = hp, color = wt)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red")
```
</div>

## Themes

<div class="bestpractice-box">
🎯 **Best Practice: Custom Themes**

```{r}
custom_theme <- theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

ggplot(mtcars, aes(x = mpg, y = hp, color = factor(cyl))) +
  geom_point(size = 3) +
  custom_theme +
  labs(title = "Custom Themed Plot")
```
</div>

## Annotations

<div class="insight-box">
💡 **Key Insight: Adding Annotations**

```{r}
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point() +
  annotate("text", x = 30, y = 250, label = "High HP", color = "red") +
  geom_hline(yintercept = mean(mtcars$hp), linetype = "dashed")
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Match scales to data types**
2. **Customize with themes**
3. **Add annotations for context**
4. **Use appropriate coordinates**

**Quick Reference:**

```{r eval=FALSE}
# Scales
scale_x_continuous()
scale_color_manual()
scale_fill_brewer()

# Themes
theme_minimal()
theme()

# Annotations
annotate()
geom_hline/vline()
```
</div>
