# ggplot2 Extensions {#ggplot2-extensions}

<div class="chapter-summary">
**What You'll Learn:**

- Extension packages
- Special plot types  
- Interactive plots
- Combining plots

**Key Errors Covered:** 15+ extension errors

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Introduction

ggplot2 extensions add specialized functionality:

```{r message=FALSE}
library(ggplot2)
library(dplyr)
```

## patchwork

<div class="insight-box">
💡 **Key Insight: Combining Plots**

```{r message=FALSE}
library(patchwork)

p1 <- ggplot(mtcars, aes(x = mpg, y = hp)) + geom_point()
p2 <- ggplot(mtcars, aes(x = mpg, y = wt)) + geom_point()

p1 + p2
p1 / p2
```
</div>

## ggrepel

<div class="insight-box">
💡 **Key Insight: Better Labels**

```{r message=FALSE}
library(ggrepel)

ggplot(mtcars, aes(x = mpg, y = hp, label = rownames(mtcars))) +
  geom_point() +
  geom_text_repel(max.overlaps = 10)
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **patchwork** - Combine plots
2. **ggrepel** - Non-overlapping labels
3. **Many extensions available**

**Common Extensions:**

```{r eval=FALSE}
library(patchwork)  # Combine plots
library(ggrepel)    # Labels
library(ggthemes)   # Themes
library(plotly)     # Interactive
```
</div>
