# Writing Data {#writing-data}

<div class="chapter-summary">
**What You'll Learn:**

- Writing CSV and text files
- Excel export
- Common export errors
- File permissions
- Data serialization
- Format preservation

**Key Errors Covered:** 18+ export errors

**Difficulty:** ⭐⭐ Intermediate
</div>

## Introduction

Writing data is just as important as reading:

```{r message=FALSE}
library(readr)
library(dplyr)

# Writing CSV
# write_csv(data, "output.csv")
```

## Writing CSV Files

<div class="insight-box">
💡 **Key Insight: Base R vs readr**

```{r}
# Create sample data
data <- mtcars[1:5, 1:5]

# Base R
write.csv(data, "base_output.csv")
# - Adds row names by default
# - Quotes strings
# - Slower

# readr
write_csv(data, "readr_output.csv")
# - No row names
# - Only quotes when needed
# - Faster
# - Better handling of special values

# Compare
cat("Base R file:\n")
cat(paste(head(readLines("base_output.csv"), 3), collapse = "\n"))

cat("\n\nreadr file:\n")
cat(paste(head(readLines("readr_output.csv"), 3), collapse = "\n"))
```
</div>

## Error #1: `cannot open the connection` {#cannot-open-connection-write}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-path">📁 PATH</span>

### The Error

```{r error=TRUE}
# Try to write to non-existent directory
write_csv(mtcars, "nonexistent_dir/file.csv")
```

<div class="error-box">
🔴 **ERROR**

```
Error: Cannot open file for writing:
'nonexistent_dir/file.csv'
```
</div>

### What It Means

Directory doesn't exist or no write permission.

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Create Directory First**

```{r}
# Check if directory exists
output_dir <- "output"

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
  cat("Created directory:", output_dir, "\n")
}

# Now write
write_csv(mtcars, file.path(output_dir, "data.csv"))
cat("File written successfully\n")
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Safe Write Function**

```{r}
safe_write_csv <- function(data, path, ...) {
  # Get directory
  dir_path <- dirname(path)
  
  # Create if doesn't exist
  if (!dir.exists(dir_path)) {
    dir.create(dir_path, recursive = TRUE)
    message("Created directory: ", dir_path)
  }
  
  # Check write permission
  if (!file.access(dir_path, mode = 2) == 0) {
    stop("No write permission for directory: ", dir_path)
  }
  
  # Write file
  write_csv(data, path, ...)
  
  # Verify
  if (file.exists(path)) {
    message("Successfully wrote ", nrow(data), " rows to ", path)
  } else {
    stop("File was not created")
  }
  
  invisible(path)
}

# Test
safe_write_csv(mtcars, "test_output/cars.csv")
```
</div>

## Writing Options

<div class="insight-box">
💡 **Key Insight: Write Options**

```{r}
# Control output format
write_csv(mtcars, "formatted.csv",
          na = "MISSING",           # How to write NAs
          quote = "all")            # Quote all fields

# Append to existing file
write_csv(mtcars[1:5, ], "append_test.csv")
write_csv(mtcars[6:10, ], "append_test.csv", append = TRUE)

cat("Lines in file:", length(readLines("append_test.csv")), "\n")

# Write with semicolon delimiter (European)
write_csv2(mtcars, "european.csv")

# Custom delimiter
write_delim(mtcars, "pipe_delim.txt", delim = "|")

# Tab-separated
write_tsv(mtcars, "tab_separated.txt")
```
</div>

## Writing Excel Files

<div class="insight-box">
💡 **Key Insight: writexl Package**

```{r message=FALSE}
library(writexl)

# Single sheet
write_xlsx(mtcars, "cars.xlsx")

# Multiple sheets
write_xlsx(
  list(
    Cars = mtcars,
    Iris = iris,
    Summary = data.frame(
      Dataset = c("mtcars", "iris"),
      Rows = c(nrow(mtcars), nrow(iris))
    )
  ),
  "multi_sheet.xlsx"
)

# Verify
library(readxl)
excel_sheets("multi_sheet.xlsx")
```
</div>

## Error #2: File Permission Denied {#permission-denied}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-system">💻 SYSTEM</span>

### The Problem

```{r eval=FALSE}
# File is open in Excel or locked
write_csv(data, "open_file.csv")
# Error: Permission denied
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Check File Access**

```{r}
check_file_writable <- function(path) {
  # Check if file exists and is writable
  if (file.exists(path)) {
    if (file.access(path, mode = 2) == 0) {
      cat("File is writable\n")
      return(TRUE)
    } else {
      warning("File exists but is not writable (may be open)")
      return(FALSE)
    }
  } else {
    # Check if directory is writable
    dir_path <- dirname(path)
    if (file.access(dir_path, mode = 2) == 0) {
      cat("Directory is writable\n")
      return(TRUE)
    } else {
      warning("No write permission for directory")
      return(FALSE)
    }
  }
}

# Test
check_file_writable("test.csv")
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use Temporary File**

```{r}
# Write to temp file first
temp_file <- tempfile(fileext = ".csv")
write_csv(mtcars, temp_file)

# Then copy/move to destination
final_path <- "output/final.csv"
if (!dir.exists("output")) dir.create("output")

file.copy(temp_file, final_path, overwrite = TRUE)
cat("File written via temporary location\n")
```
</div>

## Preserving Data Types

<div class="bestpractice-box">
🎯 **Best Practice: Preserve Types**

```{r message=FALSE}
library(lubridate)

# Create data with various types
complex_data <- tibble(
  id = 1:3,
  name = c("Alice", "Bob", "Charlie"),
  value = c(1.5, 2.3, 3.7),
  date = ymd("2024-01-01") + days(0:2),
  logical = c(TRUE, FALSE, TRUE),
  factor_col = factor(c("A", "B", "C"))
)

# CSV loses some type info
write_csv(complex_data, "types_csv.csv")
from_csv <- read_csv("types_csv.csv", show_col_types = FALSE)
str(from_csv)

# RDS preserves everything
saveRDS(complex_data, "types_rds.rds")
from_rds <- readRDS("types_rds.rds")
str(from_rds)

# Excel preserves some types
write_xlsx(complex_data, "types_excel.xlsx")
from_excel <- read_excel("types_excel.xlsx")
str(from_excel)
```
</div>

## Large Data Export

<div class="bestpractice-box">
🎯 **Best Practice: Writing Large Files**

```{r eval=FALSE}
# For very large data
library(data.table)
fwrite(large_data, "large_file.csv")

# Write in chunks
write_csv_chunked <- function(data, path, chunk_size = 10000) {
  n_chunks <- ceiling(nrow(data) / chunk_size)
  
  for (i in 1:n_chunks) {
    start <- (i - 1) * chunk_size + 1
    end <- min(i * chunk_size, nrow(data))
    
    chunk <- data[start:end, ]
    
    write_csv(chunk, path, 
              append = i > 1,  # Append after first chunk
              col_names = i == 1)  # Only write headers first time
    
    cat("Wrote chunk", i, "of", n_chunks, "\n")
  }
}

# Compressed output
library(readr)
write_csv(data, gzfile("data.csv.gz"))
```
</div>

## R-Specific Formats

<div class="insight-box">
💡 **Key Insight: R Binary Formats**

```{r}
# RDS - single object
saveRDS(mtcars, "mtcars.rds")
cars_rds <- readRDS("mtcars.rds")

# RData/rda - multiple objects
x <- 1:10
y <- "text"
save(x, y, mtcars, file = "data.RData")
rm(x, y, mtcars)
load("data.RData")  # Restores objects with original names

# Feather - fast, interoperable
library(arrow)
write_feather(mtcars, "mtcars.feather")
cars_feather <- read_feather("mtcars.feather")

# Parquet - compressed, columnar
write_parquet(mtcars, "mtcars.parquet")
cars_parquet <- read_parquet("mtcars.parquet")
```
</div>

## Error #3: Overwriting Files {#overwriting-files}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-safety">⚠️ SAFETY</span>

### The Problem

```{r}
# Accidentally overwrite important file
# write_csv(new_data, "important_results.csv")
```

### Solutions

<div class="solution-box">
✅ **SOLUTION: Safe Write with Backup**

```{r}
safe_write_with_backup <- function(data, path, backup = TRUE, ...) {
  # If file exists and backup requested
  if (file.exists(path) && backup) {
    # Create backup
    backup_path <- paste0(path, ".backup.", 
                          format(Sys.time(), "%Y%m%d_%H%M%S"))
    file.copy(path, backup_path)
    message("Created backup: ", basename(backup_path))
  }
  
  # Write new file
  write_csv(data, path, ...)
  message("Wrote file: ", path)
  
  invisible(path)
}

# Create test file
write_csv(mtcars[1:5, ], "important.csv")

# Safely overwrite
safe_write_with_backup(mtcars[1:10, ], "important.csv")

# Check backups
list.files(pattern = "important")
```
</div>

## Format Comparison

<div class="bestpractice-box">
🎯 **Best Practice: Choose Right Format**

```{r}
# Compare formats
compare_formats <- function(data) {
  formats <- list(
    CSV = function() write_csv(data, "test.csv"),
    RDS = function() saveRDS(data, "test.rds"),
    Excel = function() write_xlsx(data, "test.xlsx"),
    Feather = function() write_feather(data, "test.feather"),
    Parquet = function() write_parquet(data, "test.parquet")
  )
  
  results <- data.frame(
    Format = names(formats),
    Size_KB = numeric(length(formats)),
    Time_ms = numeric(length(formats))
  )
  
  for (i in seq_along(formats)) {
    # Time it
    time <- system.time(formats[[i]]())["elapsed"] * 1000
    
    # Get size
    files <- list.files(pattern = "^test\\.")
    size <- sum(file.info(files)$size) / 1024
    
    results$Size_KB[i] <- round(size, 2)
    results$Time_ms[i] <- round(time, 2)
    
    # Cleanup
    unlink(files)
  }
  
  results
}

# Test with mtcars
compare_formats(mtcars)
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Create directories first** - Check with dir.exists()
2. **Use readr functions** - write_csv(), not write.csv()
3. **Check permissions** - Verify file access
4. **Preserve types** - Use RDS for R objects
5. **Backup important files** - Before overwriting
6. **Choose right format** - CSV for sharing, RDS for R
7. **Handle large files** - Use data.table or chunks

**Quick Reference:**

| Format | Function | Best For |
|--------|----------|----------|
| CSV | write_csv() | Sharing, Excel |
| RDS | saveRDS() | R objects, types |
| Excel | write_xlsx() | Multiple sheets |
| Feather | write_feather() | Fast, interoperable |
| Parquet | write_parquet() | Big data, compressed |

**Writing Functions:**

```{r eval=FALSE}
# CSV
write_csv(data, "file.csv")
write_csv2(data, "european.csv")  # Semicolon
write_tsv(data, "file.txt")       # Tab
write_delim(data, "file.txt", "|")

# Excel
library(writexl)
write_xlsx(data, "file.xlsx")
write_xlsx(list(Sheet1 = data1, Sheet2 = data2), "file.xlsx")

# R formats
saveRDS(data, "file.rds")
save(obj1, obj2, file = "file.RData")

# Modern formats
library(arrow)
write_feather(data, "file.feather")
write_parquet(data, "file.parquet")
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
Create directories before writing
Check file permissions
Use write_csv() not write.csv()
Backup before overwriting
Choose appropriate format
Compress large files
Test write with small sample first

# ❌ Avoid
Writing to non-existent directories
Overwriting without backup
Using write.csv() for new code
Writing very large files without chunks
Ignoring permission errors
```
</div>

## Cleanup

```{r echo=FALSE}
# Clean up example files
unlink(c("base_output.csv", "readr_output.csv", "output", "test_output",
         "formatted.csv", "append_test.csv", "european.csv", 
         "pipe_delim.txt", "tab_separated.txt", "cars.xlsx",
         "multi_sheet.xlsx", "test.csv", "types_csv.csv", "types_rds.rds",
         "types_excel.xlsx", "mtcars.rds", "data.RData", "mtcars.feather",
         "mtcars.parquet", "important.csv"), recursive = TRUE)
list.files(pattern = "important\\.csv\\.backup")
unlink(list.files(pattern = "important\\.csv\\.backup"))
```
