# Part XIII: Programming Patterns {-}

# apply Family & Iteration {#apply-family}

<div class="chapter-summary">
**What You'll Learn:**

- apply, lapply, sapply, mapply
- Common iteration errors
- When to use each function
- Performance considerations
- Alternative approaches

**Key Errors Covered:** 20+ iteration errors

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Introduction

R's apply family provides vectorized operations:

```{r}
# Instead of loops
result <- numeric(nrow(mtcars))
for (i in 1:nrow(mtcars)) {
  result[i] <- mean(as.numeric(mtcars[i, ]))
}

# Use apply
result <- apply(mtcars, 1, mean)
head(result)
```

Let's master the apply family!

## apply() - Arrays and Matrices

<div class="insight-box">
💡 **Key Insight: apply() for Matrices**

```{r}
# Create matrix
mat <- matrix(1:12, nrow = 3, ncol = 4)
mat

# Apply to rows (MARGIN = 1)
row_sums <- apply(mat, 1, sum)
row_sums

# Apply to columns (MARGIN = 2)
col_means <- apply(mat, 2, mean)
col_means

# Custom function
apply(mat, 1, function(x) max(x) - min(x))

# With additional arguments
apply(mat, 2, sum, na.rm = TRUE)

# Both dimensions
apply(mat, c(1, 2), sqrt)
```
</div>

## Error #1: `dim(X) must have positive length` {#dim-positive}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
# Try apply on vector
vec <- 1:10
apply(vec, 1, sum)
```

<div class="error-box">
🔴 **ERROR**

```
Error in apply(vec, 1, sum) : dim(X) must have a positive length
```
</div>

### What It Means

`apply()` requires a matrix/array, not a vector.

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use lapply() or sapply()**

```{r}
# For vectors, use lapply/sapply
vec <- 1:10

# Wrong: apply(vec, 1, sqrt)

# Right: use sapply
sapply(vec, sqrt)

# Or vectorized operation
sqrt(vec)
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Convert to Matrix**

```{r}
# If you really need apply
vec <- 1:10
mat <- matrix(vec, ncol = 1)
apply(mat, 1, function(x) x * 2)

# But this is unnecessary - use vectorization
vec * 2
```
</div>

## lapply() - Lists

<div class="insight-box">
💡 **Key Insight: lapply() Always Returns List**

```{r}
# Create list
my_list <- list(a = 1:5, b = 6:10, c = 11:15)

# Apply function to each element
result <- lapply(my_list, mean)
result

# With custom function
lapply(my_list, function(x) sum(x^2))

# On data frame (df is a list of columns)
lapply(mtcars[, 1:3], mean)

# Extract elements
lapply(my_list, `[`, 1:2)

# With multiple arguments
lapply(my_list, sum, na.rm = TRUE)
```
</div>

## sapply() - Simplified

<div class="insight-box">
💡 **Key Insight: sapply() Simplifies Output**

```{r}
my_list <- list(a = 1:5, b = 6:10, c = 11:15)

# lapply returns list
lapply(my_list, mean)

# sapply returns vector
sapply(my_list, mean)

# sapply with matrix output
sapply(my_list, function(x) c(mean = mean(x), sd = sd(x)))

# When can't simplify, returns list
sapply(my_list, summary)  # Returns list
```
</div>

## Error #2: Unexpected Output Type {#unexpected-type}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Problem

```{r}
# sapply behavior depends on output
result1 <- sapply(1:3, function(x) x)
result1
class(result1)  # vector

result2 <- sapply(1:3, function(x) c(x, x^2))
result2
class(result2)  # matrix

result3 <- sapply(1:3, function(x) list(x, x^2))
result3
class(result3)  # list
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use vapply() for Type Safety**

```{r}
# Specify output type
result <- vapply(1:3, function(x) x * 2, FUN.VALUE = numeric(1))
result

# Will error if output doesn't match
# vapply(1:3, function(x) c(x, x^2), FUN.VALUE = numeric(1))

# For multiple values
result <- vapply(1:3, function(x) c(x, x^2), FUN.VALUE = numeric(2))
result
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Use lapply() and Post-Process**

```{r}
# Always get list, then convert
result <- lapply(1:3, function(x) x * 2)
unlist(result)

# Or use do.call
do.call(c, result)
```
</div>

## mapply() - Multiple Arguments

<div class="insight-box">
💡 **Key Insight: mapply() for Parallel Iteration**

```{r}
# Apply function to multiple vectors in parallel
mapply(function(x, y) x + y, 
       x = 1:5, 
       y = 6:10)

# Multiple arguments
mapply(rep, 
       x = 1:4, 
       times = 4:1)

# With data frames
df1 <- data.frame(a = 1:3, b = 4:6)
df2 <- data.frame(c = 7:9, d = 10:12)

mapply(function(x, y) x + y, df1$a, df2$c)

# MoreArgs for constant arguments
mapply(function(x, y, z) x + y + z,
       x = 1:3,
       y = 4:6,
       MoreArgs = list(z = 10))
```
</div>

## tapply() - Grouped Apply

<div class="insight-box">
💡 **Key Insight: tapply() for Split-Apply-Combine**

```{r}
# Apply function by groups
tapply(mtcars$mpg, mtcars$cyl, mean)

# Multiple grouping variables
tapply(mtcars$mpg, 
       list(Cyl = mtcars$cyl, Gear = mtcars$gear), 
       mean)

# With custom function
tapply(mtcars$mpg, mtcars$cyl, 
       function(x) c(mean = mean(x), sd = sd(x)))

# Like dplyr group_by + summarize
library(dplyr)
mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg))
```
</div>

## Common Patterns

<div class="bestpractice-box">
🎯 **Best Practice: Choose Right Function**

```{r}
# Pattern 1: Apply to each column of data frame
lapply(mtcars[, 1:3], mean)
sapply(mtcars[, 1:3], mean)

# Pattern 2: Apply to each row
apply(mtcars[, 1:3], 1, sum)

# Pattern 3: Apply with multiple inputs
mapply(function(x, y) x / y,
       x = mtcars$hp,
       y = mtcars$wt)

# Pattern 4: Apply by groups
tapply(mtcars$mpg, mtcars$cyl, mean)

# Pattern 5: Nested lists
nested <- list(
  a = list(x = 1:3, y = 4:6),
  b = list(x = 7:9, y = 10:12)
)

# Get all 'x' elements
lapply(nested, `[[`, "x")

# Apply to nested structure
lapply(nested, function(sublist) {
  lapply(sublist, mean)
})
```
</div>

## Performance Considerations

<div class="bestpractice-box">
🎯 **Best Practice: Vectorize When Possible**

```{r}
# Compare performance
n <- 10000

# Loop (slow)
system.time({
  result <- numeric(n)
  for (i in 1:n) {
    result[i] <- sqrt(i)
  }
})

# sapply (better)
system.time({
  result <- sapply(1:n, sqrt)
})

# Vectorized (best)
system.time({
  result <- sqrt(1:n)
})

# When apply family is appropriate
df <- data.frame(matrix(rnorm(1000), ncol = 10))

# Row-wise operations (apply is good)
system.time({
  apply(df, 1, mean)
})

# Column-wise operations (vectorized is better)
system.time({
  colMeans(df)
})
```
</div>

## Error #3: Function Not Vectorized {#not-vectorized}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-logic">🧠 LOGIC</span>

### The Problem

```{r error=TRUE}
# Custom function expecting single value
my_function <- function(x) {
  if (x > 0) {
    return("positive")
  } else {
    return("negative")
  }
}

# Doesn't work with vectors
my_function(c(-1, 2, -3, 4))
```

<div class="warning-box">
⚠️ **WARNING**

```
Warning: the condition has length > 1
```
</div>

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: Use sapply/vapply**

```{r}
# Apply to each element
sapply(c(-1, 2, -3, 4), my_function)

# Type-safe version
vapply(c(-1, 2, -3, 4), my_function, FUN.VALUE = character(1))
```
</div>

<div class="solution-box">
✅ **SOLUTION 2: Vectorize the Function**

```{r}
# Make function vectorized
my_function_vec <- Vectorize(my_function)
my_function_vec(c(-1, 2, -3, 4))

# Or rewrite using ifelse
my_function_better <- function(x) {
  ifelse(x > 0, "positive", "negative")
}

my_function_better(c(-1, 2, -3, 4))
```
</div>

## Alternative: purrr Package

<div class="insight-box">
💡 **Key Insight: purrr for Modern Iteration**

```{r message=FALSE}
library(purrr)

# map() family (like lapply/sapply)
map(1:3, ~ . * 2)           # Returns list
map_dbl(1:3, ~ . * 2)       # Returns numeric
map_chr(1:3, ~ as.character(.))  # Returns character

# map2() for two inputs (like mapply)
map2_dbl(1:3, 4:6, ~ .x + .y)

# pmap() for multiple inputs
pmap_dbl(list(x = 1:3, y = 4:6, z = 7:9), 
         function(x, y, z) x + y + z)

# Useful helpers
list(a = 1:3, b = 4:6, c = 7:9) %>%
  map_dbl(mean)

# Safe operations
map(c("1", "2", "not_a_number"), 
    possibly(as.numeric, otherwise = NA))
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **apply()** - For matrices/arrays
2. **lapply()** - Always returns list
3. **sapply()** - Simplifies output
4. **vapply()** - Type-safe sapply
5. **mapply()** - Multiple inputs
6. **tapply()** - Grouped operations
7. **Vectorize when possible** - Faster than apply

**Quick Reference:**

| Function | Input | Output | Use Case |
|----------|-------|--------|----------|
| apply() | Matrix/array | Vector/list | Row/column operations |
| lapply() | List/vector | List | Any operation |
| sapply() | List/vector | Vector/matrix | When simplified OK |
| vapply() | List/vector | Specified type | Type safety |
| mapply() | Multiple vectors | Vector/list | Parallel iteration |
| tapply() | Vector + groups | Array | Split-apply-combine |

**Usage Patterns:**

```{r eval=FALSE}
# Matrices
apply(matrix, 1, function)    # By row
apply(matrix, 2, function)    # By column

# Lists
lapply(list, function)        # Returns list
sapply(list, function)        # Simplified
vapply(list, function, type)  # Type-safe

# Multiple inputs
mapply(function, x, y)

# Grouped
tapply(values, groups, function)

# Modern alternative
library(purrr)
map(list, function)           # Like lapply
map_dbl(list, function)       # Like sapply with numeric
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
Use vectorized operations when possible
Use vapply() for type safety
Choose right function for task
Consider purrr for complex operations

# ❌ Avoid
Using apply() on vectors
Using sapply() when type matters
Growing objects in loops
Unnecessary apply when vectorized solution exists
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Apply Practice**

Using mtcars:
1. Calculate row means using apply
2. Calculate column medians using apply
3. Find max value in each row
4. Compare performance with vectorized versions
</div>

<div class="exercise-box">
📝 **Exercise 2: Custom Function with Apply**

Write function to:
1. Take a data frame
2. For each numeric column, calculate mean, sd, min, max
3. Return as data frame
4. Use appropriate apply function
</div>

<div class="exercise-box">
📝 **Exercise 3: mapply Practice**

Create two vectors and:
1. Add them element-wise with mapply
2. Use custom function with multiple arguments
3. Compare with vectorized approach
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
# 1. Row means
row_means_apply <- apply(mtcars, 1, mean)
row_means_vec <- rowMeans(mtcars)

all.equal(row_means_apply, row_means_vec)

# 2. Column medians
col_medians <- apply(mtcars, 2, median)
col_medians

# 3. Max in each row
row_max <- apply(mtcars, 1, max)
head(row_max)

# 4. Performance comparison
library(microbenchmark)

microbenchmark(
  apply = apply(mtcars, 1, mean),
  rowMeans = rowMeans(mtcars),
  times = 100
)

# Vectorized is much faster!
```

**Exercise 2:**

```{r}
summarize_numeric <- function(df) {
  # Get numeric columns
  numeric_cols <- sapply(df, is.numeric)
  df_numeric <- df[, numeric_cols]
  
  # Calculate statistics for each column
  stats <- lapply(df_numeric, function(col) {
    c(
      mean = mean(col, na.rm = TRUE),
      sd = sd(col, na.rm = TRUE),
      min = min(col, na.rm = TRUE),
      max = max(col, na.rm = TRUE)
    )
  })
  
  # Convert to data frame
  result <- do.call(rbind, stats)
  result <- as.data.frame(result)
  result$variable <- rownames(result)
  rownames(result) <- NULL
  
  result[, c("variable", "mean", "sd", "min", "max")]
}

# Test
summarize_numeric(mtcars)

# Alternative using vapply for type safety
summarize_numeric_safe <- function(df) {
  numeric_cols <- sapply(df, is.numeric)
  df_numeric <- df[, numeric_cols]
  
  stats <- vapply(df_numeric, function(col) {
    c(mean = mean(col, na.rm = TRUE),
      sd = sd(col, na.rm = TRUE),
      min = min(col, na.rm = TRUE),
      max = max(col, na.rm = TRUE))
  }, FUN.VALUE = numeric(4))
  
  result <- as.data.frame(t(stats))
  result$variable <- rownames(result)
  rownames(result) <- NULL
  
  result[, c("variable", "mean", "sd", "min", "max")]
}

summarize_numeric_safe(iris)
```

**Exercise 3:**

```{r}
# Create vectors
x <- 1:10
y <- 11:20

# 1. Add with mapply
result_mapply <- mapply(function(a, b) a + b, x, y)
result_mapply

# 2. Custom function
weighted_sum <- function(a, b, weight = 0.5) {
  a * weight + b * (1 - weight)
}

result_custom <- mapply(weighted_sum, x, y, MoreArgs = list(weight = 0.3))
result_custom

# 3. Compare with vectorized
result_vec <- x + y
all.equal(result_mapply, result_vec)

# Performance
microbenchmark(
  mapply = mapply(`+`, x, y),
  vectorized = x + y,
  times = 1000
)

# Vectorized is MUCH faster
```
</details>
