# purrr Package {#purrr-package}

<div class="chapter-summary">
**What You'll Learn:**

- map() family functions
- Type-safe iteration
- Error handling in iteration
- Advanced patterns
- Advantages over apply family

**Key Errors Covered:** 15+ purrr errors

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Introduction

purrr provides modern, type-safe iteration:

```{r message=FALSE}
library(purrr)

# map() instead of lapply()
map(1:3, ~ . * 2)

# Type-safe variants
map_dbl(1:3, ~ . * 2)
```

## map() Family

<div class="insight-box">
💡 **Key Insight: Type-Safe Mapping**

```{r}
library(purrr)

# map() returns list (like lapply)
map(1:3, sqrt)

# Type-specific variants
map_dbl(1:3, sqrt)      # numeric vector
map_chr(1:3, as.character)  # character vector
map_int(1:3, ~ .x)      # integer vector
map_lgl(c(TRUE, FALSE, TRUE), ~ .x)  # logical vector

# Error if wrong type
# map_dbl(1:3, as.character)  # Error!

# map_df for data frames
map_df(1:3, ~ data.frame(x = .x, y = .x^2))
```
</div>

## Formula Syntax

<div class="insight-box">
💡 **Key Insight: Convenient Formula Syntax**

```{r}
# Three ways to write functions

# 1. Regular function
map(1:3, function(x) x * 2)

# 2. Formula with ~ (one argument: .x or .)
map(1:3, ~ .x * 2)
map(1:3, ~ . * 2)

# 3. Formula with ~ (two arguments: .x and .y)
map2(1:3, 4:6, ~ .x + .y)

# Complex expressions
map(1:3, ~ {
  squared <- .x^2
  sqrt(squared)
})

# Accessing list elements
lst <- list(
  list(a = 1, b = 2),
  list(a = 3, b = 4)
)

map(lst, ~ .x$a)
# Or simpler:
map(lst, "a")
```
</div>

## Error #1: Type Mismatch {#purrr-type-mismatch}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

### The Error

```{r error=TRUE}
# Function returns character, but expecting numeric
map_dbl(1:3, ~ as.character(.x))
```

<div class="error-box">
🔴 **ERROR**

```
Error: Can't coerce element 1 from a character to a double
```
</div>

### Solutions

<div class="solution-box">
✅ **SOLUTION: Use Correct Type**

```{r}
# Use map_chr for character output
map_chr(1:3, ~ as.character(.x))

# Or use map() and get list
map(1:3, ~ as.character(.x))

# Convert inside function
map_dbl(1:3, ~ as.numeric(as.character(.x)))
```
</div>

## map2() and pmap()

<div class="insight-box">
💡 **Key Insight: Multiple Inputs**

```{r}
# map2() for two inputs
map2_dbl(1:3, 4:6, ~ .x + .y)

map2_chr(c("a", "b", "c"), 1:3, ~ paste(.x, .y))

# pmap() for multiple inputs (list of inputs)
inputs <- list(
  x = 1:3,
  y = 4:6,
  z = 7:9
)

pmap_dbl(inputs, function(x, y, z) x + y + z)

# With formula
pmap_dbl(inputs, ~ ..1 + ..2 + ..3)

# Named arguments
pmap_chr(
  list(
    name = c("Alice", "Bob"),
    age = c(25, 30)
  ),
  ~ paste(.x, "is", .y, "years old")
)
```
</div>

## Error Handling

<div class="bestpractice-box">
🎯 **Best Practice: Safe Iteration**

```{r}
# Data with potential errors
data <- list("1", "2", "not_a_number", "4")

# map() will fail
# map_dbl(data, as.numeric)  # Error!

# possibly() returns default on error
safe_numeric <- possibly(as.numeric, otherwise = NA_real_)
map_dbl(data, safe_numeric)

# safely() returns list of result and error
safe_parse <- safely(as.numeric)
map(data, safe_parse)

# quietly() captures messages/warnings
quiet_sqrt <- quietly(sqrt)
map(c(4, -1, 9), quiet_sqrt)

# Try each until one works
try_functions <- list(
  as.numeric,
  function(x) 0
)

map(data, ~ reduce(try_functions, function(val, f) {
  if (!is.na(val)) return(val)
  try(f(.x), silent = TRUE)
}, .init = NA))
```
</div>

## Advanced Patterns

<div class="bestpractice-box">
🎯 **Best Practice: Complex Patterns**

```{r}
# Keep/discard based on condition
numbers <- list(1, "a", 3, "b", 5)

keep(numbers, is.numeric)
discard(numbers, is.character)

# Detect if any/all meet condition
some(numbers, is.numeric)
every(numbers, is.numeric)

# Find position
detect(numbers, is.character)
detect_index(numbers, is.character)

# Reduce (fold)
reduce(1:5, `+`)
reduce(1:5, `*`)

# Accumulate (show intermediate steps)
accumulate(1:5, `+`)

# Modify elements
modify(list(1, 2, 3), ~ .x * 2)
modify_if(list(1, "a", 3), is.numeric, ~ .x * 2)
modify_at(list(a = 1, b = 2, c = 3), "b", ~ .x * 10)
```
</div>

## Nested Data

<div class="insight-box">
💡 **Key Insight: Working with Nested Lists**

```{r}
# Nested list
nested <- list(
  person1 = list(name = "Alice", age = 25, scores = c(90, 85, 92)),
  person2 = list(name = "Bob", age = 30, scores = c(88, 92, 87)),
  person3 = list(name = "Charlie", age = 35, scores = c(95, 89, 91))
)

# Extract single element
map(nested, "name")
map(nested, "age")

# Extract nested element
map(nested, list("scores", 1))  # First score

# Complex extraction
map(nested, ~ mean(.x$scores))

# Modify nested structure
map(nested, ~ {
  .x$avg_score <- mean(.x$scores)
  .x
})

# Flatten nested lists
nested_scores <- map(nested, "scores")
flatten_dbl(nested_scores)
```
</div>

## Comparison with Base R

<div class="bestpractice-box">
🎯 **Best Practice: purrr vs Base R**

```{r}
# Get means of each column
df <- mtcars[, 1:3]

# Base R
lapply(df, mean)
sapply(df, mean)

# purrr (type-safe)
map(df, mean)
map_dbl(df, mean)

# Multiple inputs
# Base R
mapply(function(x, y) x + y, 1:3, 4:6)

# purrr (cleaner)
map2_dbl(1:3, 4:6, ~ .x + .y)

# Error handling
# Base R
tryCatch(as.numeric("a"), error = function(e) NA)

# purrr (composable)
possibly(as.numeric, NA)("a")

# Advantages of purrr:
# - Type safety
# - Consistent interface
# - Better error handling
# - Cleaner syntax
# - Composable functions
```
</div>

## Real-World Examples

<div class="bestpractice-box">
🎯 **Best Practice: Practical Uses**

```{r message=FALSE}
library(dplyr)

# 1. Read multiple files
# files <- list.files(pattern = "\\.csv$")
# data <- map_df(files, read_csv, .id = "file")

# 2. Fit multiple models
models <- mtcars %>%
  split(.$cyl) %>%
  map(~ lm(mpg ~ hp, data = .))

# Get R-squared from each
map_dbl(models, ~ summary(.)$r.squared)

# 3. Extract nested information
results <- list(
  model1 = list(coef = c(1.5, 2.3), r2 = 0.85),
  model2 = list(coef = c(1.8, 2.1), r2 = 0.90),
  model3 = list(coef = c(1.2, 2.5), r2 = 0.78)
)

map_dbl(results, "r2")
map(results, "coef")

# 4. Validate data
data_list <- list(
  df1 = data.frame(x = 1:3, y = 4:6),
  df2 = data.frame(x = 1:3),  # Missing y
  df3 = data.frame(x = 1:3, y = 4:6)
)

# Check all have 'y' column
map_lgl(data_list, ~ "y" %in% names(.))

# Filter to valid ones
valid_data <- keep(data_list, ~ "y" %in% names(.))
length(valid_data)
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Type-safe mapping** - Use map_dbl(), map_chr(), etc.
2. **Formula syntax** - Clean with ~ .x
3. **Error handling** - possibly(), safely(), quietly()
4. **Multiple inputs** - map2(), pmap()
5. **Powerful helpers** - keep(), discard(), reduce()
6. **Nested data** - Easy extraction with list indexing
7. **Consistent interface** - All functions work similarly

**Quick Reference:**

| Function | Purpose | Returns |
|----------|---------|---------|
| map() | Apply to each | List |
| map_dbl() | Apply to each | Numeric vector |
| map_chr() | Apply to each | Character vector |
| map_int() | Apply to each | Integer vector |
| map_lgl() | Apply to each | Logical vector |
| map2() | Two inputs | List/vector |
| pmap() | Multiple inputs | List/vector |
| walk() | Side effects only | Input invisibly |

**Common Patterns:**

```{r eval=FALSE}
# Basic mapping
map(list, function)
map_dbl(list, ~ .x * 2)

# Multiple inputs
map2(x, y, ~ .x + .y)
pmap(list(x, y, z), function(x, y, z) x + y + z)

# Error handling
possibly(function, otherwise = NA)
safely(function)

# Filtering
keep(list, is.numeric)
discard(list, is.na)

# Detection
some(list, is.numeric)
every(list, is.numeric)

# Reduction
reduce(list, `+`)
accumulate(list, `+`)

# Extraction
map(nested_list, "element_name")
map(nested_list, list("level1", "level2"))
```

**Best Practices:**

```{r eval=FALSE}
# ✅ Good
Use type-specific map variants (map_dbl, etc.)
Use formula syntax for clarity
Handle errors with possibly/safely
Use descriptive function names

# ❌ Avoid
Using map() when type matters
Ignoring potential errors
Complex nested anonymous functions
```
</div>
