# Documentation & Testing {#package-documentation-testing}

<div class="chapter-summary">
**What You'll Learn:**

- roxygen2 documentation
- testthat framework
- Unit testing best practices
- Code coverage
- Vignettes

**Key Errors Covered:** 15+ testing errors

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Introduction

Documentation and testing ensure package quality:

```{r eval=FALSE}
library(testthat)
library(devtools)

# Document
document()

# Test
test()
```

## roxygen2 Documentation

<div class="insight-box">
💡 **Key Insight: Complete Function Documentation**

```{r eval=FALSE}
#' Add Two Numbers
#'
#' This function adds two numbers together.
#' It handles NA values appropriately.
#'
#' @param x A numeric value
#' @param y A numeric value
#' @param na.rm Logical; should NA values be removed?
#'
#' @return The sum of x and y
#' @export
#'
#' @examples
#' add_numbers(2, 3)
#' add_numbers(2, NA, na.rm = TRUE)
#'
#' @seealso \code{\link{sum}}
add_numbers <- function(x, y, na.rm = FALSE) {
  if (na.rm) {
    x <- ifelse(is.na(x), 0, x)
    y <- ifelse(is.na(y), 0, y)
  }
  x + y
}
```

**Key Tags:**
- `@param` - Parameter description
- `@return` - What function returns
- `@export` - Make function available
- `@examples` - Usage examples
- `@seealso` - Related functions
- `@importFrom` - Import external functions
</div>

## testthat Framework

<div class="bestpractice-box">
🎯 **Best Practice: Comprehensive Testing**

```{r eval=FALSE}
# Set up testing
use_testthat()

# Create test file
use_test("add_numbers")

# In tests/testthat/test-add_numbers.R:
test_that("add_numbers works correctly", {
  expect_equal(add_numbers(2, 3), 5)
  expect_equal(add_numbers(0, 0), 0)
  expect_equal(add_numbers(-1, 1), 0)
})

test_that("add_numbers handles NA", {
  expect_true(is.na(add_numbers(2, NA)))
  expect_equal(add_numbers(2, NA, na.rm = TRUE), 2)
})

test_that("add_numbers validates input", {
  expect_error(add_numbers("a", 2))
  expect_error(add_numbers(2, "b"))
})

# Run tests
test()
```

**Common Expectations:**
- `expect_equal(x, y)` - Values are equal
- `expect_true(x)` - x is TRUE
- `expect_false(x)` - x is FALSE
- `expect_error(code)` - Code produces error
- `expect_warning(code)` - Code produces warning
- `expect_message(code)` - Code produces message
- `expect_length(x, n)` - x has length n
- `expect_type(x, type)` - x is of type
</div>

## Code Coverage

<div class="insight-box">
💡 **Key Insight: Measure Test Coverage**

```{r eval=FALSE}
# Install covr
install.packages("covr")

# Check coverage
library(covr)
coverage <- package_coverage()
coverage

# View in browser
report(coverage)

# Target: >80% coverage
```
</div>

## Vignettes

<div class="bestpractice-box">
🎯 **Best Practice: Write Vignettes**

```{r eval=FALSE}
# Create vignette
use_vignette("introduction")

# In vignettes/introduction.Rmd:
---
title: "Introduction to mypackage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to mypackage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Build vignettes
devtools::build_vignettes()
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Document completely** - All parameters, returns, examples
2. **Test thoroughly** - Cover all cases
3. **Use expectations** - Clear, specific tests
4. **Check coverage** - Aim for >80%
5. **Write vignettes** - Long-form documentation

**Quick Reference:**

```{r eval=FALSE}
# Documentation
devtools::document()

# Testing
usethis::use_testthat()
usethis::use_test("function")
devtools::test()

# Coverage
covr::package_coverage()

# Vignettes
usethis::use_vignette("name")
```
</div>
