# Part XVII: Advanced R Programming {-}

# S3 Object System {#s3-oop}

<div class="chapter-summary">
**What You'll Learn:**

- S3 classes and objects
- Generic functions and method dispatch
- Inheritance
- Common S3 errors

**Key Errors Covered:** 15+ S3 errors

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Introduction

S3 is R's simplest OOP system, used throughout base R and many packages.

```{r}
# S3 objects are just lists with a class attribute
person <- list(name = "Alice", age = 30)
class(person) <- "person"

# Method dispatch based on class
class(mtcars)  # "data.frame"
class(lm(mpg ~ wt, mtcars))  # "lm"
```

## Creating S3 Objects

<div class="insight-box">
💡 **Key Insight: Constructor Pattern**

```{r}
# Constructor
new_person <- function(name, age) {
  structure(
    list(name = name, age = age),
    class = "person"
  )
}

# Validator
validate_person <- function(x) {
  if (!is.character(x$name)) stop("name must be character")
  if (!is.numeric(x$age) || x$age < 0) stop("age must be positive")
  x
}

# Helper (user-facing)
person <- function(name, age) {
  validate_person(new_person(name, age))
}

alice <- person("Alice", 30)
```
</div>

## Generic Functions

```{r}
# Create generic
greet <- function(x) {
  UseMethod("greet")
}

# Define methods
greet.person <- function(x) {
  paste("Hello", x$name)
}

greet.default <- function(x) {
  "Hello!"
}

# Test
greet(alice)  # Calls greet.person
greet(123)    # Calls greet.default
```

## Common Errors

### Error: no applicable method

```{r error=TRUE}
# Forgot default method
process <- function(x) UseMethod("process")
process.person <- function(x) x$name

process(123)  # Error!
```

**Solution:** Always provide default method

```{r}
process.default <- function(x) {
  stop("No method for class ", class(x)[1], call. = FALSE)
}
```

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **S3 = simple OOP** - Just lists with class attribute
2. **UseMethod()** - Enables dispatch
3. **Generic.class** - Method naming
4. **Always provide default** - Avoid errors
5. **Validate inputs** - Use constructor pattern

**Quick Reference:**

```{r eval=FALSE}
# Create class
new_class <- function(...) {
  structure(list(...), class = "myclass")
}

# Create generic
generic <- function(x) UseMethod("generic")

# Create method
generic.myclass <- function(x) {
  # Implementation
}

# Default
generic.default <- function(x) {
  stop("No method")
}
```
</div>
