# Non-Standard Evaluation {#nse-metaprogramming}

<div class="chapter-summary">
**What You'll Learn:**

- Non-standard evaluation (NSE)
- Tidy evaluation
- Quasiquotation
- Common metaprogramming errors

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Introduction

NSE allows functions to capture and manipulate code:

```{r message=FALSE}
library(dplyr)
library(rlang)

# Standard evaluation
filter(mtcars, cyl == 4)

# The 'cyl == 4' is captured as an expression
# Not evaluated immediately
```

## Tidy Evaluation

<div class="insight-box">
💡 **Key Insight: Embrace and Inject**

```{r}
# Problem: variables don't work
my_filter <- function(df, condition) {
  filter(df, condition)  # Error!
}

# Solution: embrace with {{}}
my_filter <- function(df, condition) {
  filter(df, {{ condition }})
}

my_filter(mtcars, cyl == 4)

# For column names
my_select <- function(df, col) {
  select(df, {{ col }})
}

my_select(mtcars, mpg)
```
</div>

## Quasiquotation

```{r}
# Inject values with !!
threshold <- 20
mtcars %>%
  filter(mpg > !!threshold)

# Inject names with :=
name_col <- "efficiency"
mtcars %>%
  mutate(!!name_col := mpg / wt)

# Splice multiple arguments with !!!
group_vars <- c("cyl", "gear")
mtcars %>%
  group_by(!!!syms(group_vars)) %>%
  summarize(mean_mpg = mean(mpg))
```

## Common Errors

### Error: object not found

```{r error=TRUE}
# Problem: forgot to embrace
my_mutate <- function(df, new_col, expr) {
  mutate(df, new_col = expr)
}

my_mutate(mtcars, efficiency, mpg / wt)
```

**Solution:** Use {{}} and :=

```{r}
my_mutate <- function(df, new_col, expr) {
  mutate(df, {{new_col}} := {{expr}})
}

my_mutate(mtcars, efficiency, mpg / wt)
```

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **{{}}** - Embrace and inject
2. **!!** - Unquote single value
3. **!!!** - Unquote-splice multiple
4. **:=** - Dynamic names
5. **NSE** - Enables dplyr's clean syntax

**Quick Reference:**

```{r eval=FALSE}
# Embrace columns
function(df, col) {
  df %>% select({{ col }})
}

# Inject values
x <- 5
filter(df, value > !!x)

# Dynamic names
name <- "new_col"
mutate(df, !!name := expression)

# Splice multiple
cols <- c("a", "b")
select(df, !!!syms(cols))
```
</div>
