# Rcpp Basics {#rcpp-basics}

<div class="chapter-summary">
**What You'll Learn:**

- When to use C++
- Basic Rcpp syntax
- Common errors
- Performance gains

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Introduction

Rcpp allows you to write C++ code called from R:

```{r}
library(Rcpp)

# Simple C++ function
cppFunction('
int add(int x, int y) {
  return x + y;
}
')

add(2, 3)
```

## When to Use Rcpp

✅ **Use Rcpp when:**
- Loops that can't be vectorized
- Recursive functions
- Need maximum speed
- Processing large data

❌ **Don't use Rcpp when:**
- R solution is fast enough
- Can vectorize in R
- Maintenance burden too high

## Basic Syntax

<div class="insight-box">
💡 **Key Insight: Rcpp Sugar**

```{r}
# R-like syntax in C++
cppFunction('
NumericVector compute(NumericVector x) {
  return sqrt(x * 2 + 1);
}
')

compute(1:5)

# Loops in C++
cppFunction('
double sum_cpp(NumericVector x) {
  double total = 0;
  for(int i = 0; i < x.size(); i++) {
    total += x[i];
  }
  return total;
}
')

sum_cpp(1:1000000)
```
</div>

## Performance Example

```{r}
# R version
mean_r <- function(x) {
  sum(x) / length(x)
}

# C++ version
cppFunction('
double mean_cpp(NumericVector x) {
  int n = x.size();
  double total = 0;
  for(int i = 0; i < n; i++) {
    total += x[i];
  }
  return total / n;
}
')

# Compare
x <- rnorm(1000000)

library(microbenchmark)
microbenchmark(
  mean(x),
  mean_r(x),
  mean_cpp(x),
  times = 100
)
```

## Common Errors

### Error: No matching function

```{r error=TRUE, eval=FALSE}
# Wrong: R syntax in C++
cppFunction('
NumericVector add_one(NumericVector x) {
  return x + 1  # Needs semicolon!
}
')
```

**Solution:** C++ requires semicolons

```{r}
cppFunction('
NumericVector add_one(NumericVector x) {
  return x + 1;
}
')
```

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Rcpp** - Integrate C++ for speed
2. **Use for loops** - When can't vectorize
3. **Rcpp Sugar** - R-like syntax
4. **Test performance** - Not always faster
5. **Maintenance cost** - Consider complexity

**Quick Reference:**

```{r eval=FALSE}
# Inline C++
library(Rcpp)

cppFunction('
double my_function(NumericVector x) {
  // C++ code here
  return result;
}
')

# Common types:
# NumericVector - numeric vector
# IntegerVector - integer vector
# CharacterVector - character vector
# NumericMatrix - matrix
# List - list
```
</div>
