# Function Factories & Operators {#function-factories}

<div class="chapter-summary">
**What You'll Learn:**

- Function factories
- Function operators
- Closures
- Practical applications

**Difficulty:** ⭐⭐⭐ Advanced
</div>

## Function Factories

Functions that create functions:

```{r}
# Factory: returns a function
power_factory <- function(exp) {
  function(x) {
    x ^ exp
  }
}

# Create specific functions
square <- power_factory(2)
cube <- power_factory(3)

square(4)  # 16
cube(4)   # 64

# The exponent is "captured" in the closure
```

## Practical Example

<div class="insight-box">
💡 **Key Insight: Custom Validators**

```{r}
# Create validators
in_range <- function(min, max) {
  function(x) {
    x >= min & x <= max
  }
}

is_percentage <- in_range(0, 100)
is_probability <- in_range(0, 1)

is_percentage(50)   # TRUE
is_percentage(150)  # FALSE

is_probability(0.5)  # TRUE
is_probability(1.5)  # FALSE
```
</div>

## Function Operators

Functions that take functions and return modified functions:

```{r}
# Add logging to any function
add_logging <- function(f) {
  function(...) {
    cat("Calling function\n")
    result <- f(...)
    cat("Function returned:", result, "\n")
    result
  }
}

# Wrap existing function
logged_sqrt <- add_logging(sqrt)
logged_sqrt(4)

# Add error handling
safe_function <- function(f, default = NULL) {
  function(...) {
    tryCatch(
      f(...),
      error = function(e) {
        message("Error occurred: ", e$message)
        default
      }
    )
  }
}

safe_log <- safe_function(log, default = NA)
safe_log(10)
safe_log(-1)  # Returns NA instead of error
```

## Memoization

<div class="bestpractice-box">
🎯 **Best Practice: Cache Results**

```{r}
# Memoize slow functions
memoize <- function(f) {
  cache <- new.env(parent = emptyenv())
  
  function(...) {
    key <- paste(..., sep = "_")
    
    if (!exists(key, envir = cache)) {
      cache[[key]] <- f(...)
    }
    
    cache[[key]]
  }
}

# Slow function
fibonacci <- function(n) {
  if (n <= 1) return(n)
  fibonacci(n - 1) + fibonacci(n - 2)
}

# Fast version with memoization
fibonacci_memo <- memoize(fibonacci)

system.time(fibonacci(30))
system.time(fibonacci_memo(30))  # Much faster on repeat
```
</div>

## Real-World Applications

```{r}
# 1. Create family of similar functions
make_adder <- function(n) {
  function(x) x + n
}

add_10 <- make_adder(10)
add_100 <- make_adder(100)

add_10(5)
add_100(5)

# 2. Configure behavior
make_filter <- function(pattern) {
  function(x) {
    grep(pattern, x, value = TRUE)
  }
}

filter_r <- make_filter("^r")
filter_r(c("apple", "rose", "banana", "ruby"))

# 3. Delayed computation
make_lazy <- function(expr) {
  computed <- FALSE
  value <- NULL
  
  function() {
    if (!computed) {
      value <<- expr
      computed <<- TRUE
    }
    value
  }
}

# Computation only happens when called
get_data <- make_lazy({
  cat("Computing...\n")
  rnorm(1000)
})

# First call computes
data <- get_data()

# Subsequent calls use cached value
data2 <- get_data()  # No "Computing..." message
```

## Common Patterns

<div class="bestpractice-box">
🎯 **Best Practice: Common Factory Patterns**

```{r}
# 1. Partial application
partial <- function(f, ...) {
  args <- list(...)
  function(...) {
    do.call(f, c(args, list(...)))
  }
}

# Create specialized version
divide_by_10 <- partial(`/`, e2 = 10)
divide_by_10(100)

# 2. Compose functions
compose <- function(f, g) {
  function(...) {
    f(g(...))
  }
}

# sqrt(abs(x))
sqrt_abs <- compose(sqrt, abs)
sqrt_abs(-16)

# 3. Negate predicate
negate <- function(f) {
  function(...) {
    !f(...)
  }
}

is_not_na <- negate(is.na)
is_not_na(c(1, NA, 3))
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Function factories** - Functions that create functions
2. **Closures** - Capture environment
3. **Function operators** - Modify function behavior
4. **Practical uses** - Validation, logging, memoization
5. **Composition** - Build complex from simple

**Quick Reference:**

```{r eval=FALSE}
# Factory
make_power <- function(n) {
  function(x) x^n
}

# Operator
add_logging <- function(f) {
  function(...) {
    cat("Calling\n")
    f(...)
  }
}

# Composition
compose <- function(f, g) {
  function(...) f(g(...))
}

# Partial application
partial <- function(f, ...) {
  args <- list(...)
  function(...) do.call(f, c(args, list(...)))
}

# Memoization
memoize <- function(f) {
  cache <- new.env()
  function(...) {
    key <- paste(...)
    if (!exists(key, cache)) cache[[key]] <- f(...)
    cache[[key]]
  }
}
```

**Common Applications:**

```{r eval=FALSE}
# ✅ Good uses
Validators (in_range, matches_pattern)
Adders (add_10, multiply_by)
Loggers (add_logging)
Error handlers (make_safe)
Memoizers (cache results)

# ❌ Avoid
Over-complicating simple code
When direct function is clearer
Excessive abstraction
```
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Temperature Converter Factory**

Create a factory that makes temperature converters:
- `to_fahrenheit()` - Celsius to Fahrenheit
- `to_celsius()` - Fahrenheit to Celsius
- `to_kelvin()` - Celsius to Kelvin
</div>

<div class="exercise-box">
📝 **Exercise 2: Timing Function Operator**

Create a function operator that times how long a function takes:
- Prints execution time
- Returns the result
- Works with any function
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

```{r}
# Temperature converter factory
make_temp_converter <- function(formula) {
  function(temp) {
    formula(temp)
  }
}

# Create converters
to_fahrenheit <- make_temp_converter(function(c) c * 9/5 + 32)
to_celsius <- make_temp_converter(function(f) (f - 32) * 5/9)
to_kelvin <- make_temp_converter(function(c) c + 273.15)

# Test
to_fahrenheit(0)   # 32
to_celsius(32)     # 0
to_kelvin(0)       # 273.15
```

**Exercise 2:**

```{r}
# Timing operator
add_timing <- function(f) {
  function(...) {
    start <- Sys.time()
    result <- f(...)
    end <- Sys.time()
    
    cat("Execution time:", 
        format(difftime(end, start, units = "secs")), "\n")
    
    result
  }
}

# Test
slow_sum <- function(n) {
  total <- 0
  for (i in 1:n) total <- total + i
  total
}

timed_sum <- add_timing(slow_sum)
timed_sum(1000000)
```
</details>

## Completion

<div class="chapter-summary">
**Part XVII Complete!**

You've mastered advanced R programming:
- S3 object system
- R6 and S4
- Non-standard evaluation
- Rcpp integration
- Function factories

**Ready for:** Production R development!
</div>
