# (APPENDIX) Appendix {-}

# Complete Error Dictionary (A-Z) {#error-dictionary}

This appendix provides an alphabetical quick reference of all errors covered in the book. Each entry includes:

- Error message
- Chapter reference
- Quick fix
- Difficulty level

Use Ctrl+F (Cmd+F on Mac) to search for specific error text.

---

## A

### `All aesthetics have length 1, but data has X rows`

**Package:** ggplot2  
**Chapter:** \@ref(ggplot2-basics)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Pass a data frame to ggplot(), not individual vectors  
**Example:**
```{r eval=FALSE}
# Wrong:
ggplot() + geom_point(aes(x = 1:10, y = 1:10))

# Right:
df <- data.frame(x = 1:10, y = 1:10)
ggplot(df) + geom_point(aes(x, y))
```

### `argument "x" is missing, with no default`

**Package:** base  
**Chapter:** \@ref(function-definition)  
**Difficulty:** ⭐  
**Quick Fix:** Provide the required argument or set a default value  
**Example:**
```{r eval=FALSE}
# Function needs argument:
my_func <- function(x) { x + 1 }
my_func()  # Error!

# Fix: provide it:
my_func(5)

# Or give it a default:
my_func <- function(x = 0) { x + 1 }
```

### `argument is not interpretable as logical`

**Package:** base  
**Chapter:** \@ref(control-flow)  
**Difficulty:** ⭐  
**Quick Fix:** Ensure condition evaluates to TRUE/FALSE  
**Example:**
```{r eval=FALSE}
# Wrong:
if ("yes") { }  # Character not logical

# Right:
if (TRUE) { }
```

### `argument is of length zero`

**Package:** base  
**Chapter:** \@ref(control-flow)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Check for empty vectors before using in if()  
**Example:**
```{r eval=FALSE}
x <- numeric(0)
if (x > 5) { }  # Error!

# Fix:
if (length(x) > 0 && x > 5) { }
```

### `arguments imply differing number of rows: X, Y`

**Package:** base  
**Chapter:** \@ref(dataframe-construction)  
**Difficulty:** ⭐  
**Quick Fix:** Ensure all columns have same length or use recycling explicitly  
**Example:**
```{r eval=FALSE}
# Wrong:
data.frame(a = 1:5, b = 1:3)  # Error!

# Right:
data.frame(a = 1:5, b = rep(1:3, length.out = 5))
```

---

## B

### `bytecode version mismatch`

**Package:** base  
**Chapter:** \@ref(package-loading)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** Reinstall packages after R update  

### `by must specify column(s) as numbers, names or logical`

**Package:** base  
**Chapter:** \@ref(base-manipulation)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Check merge/aggregate by argument format  

---

## C

### `Can't combine <type1> and <type2>`

**Package:** dplyr  
**Chapter:** \@ref(dplyr)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Convert columns to compatible types before combining  
**Example:**
```{r eval=FALSE}
# If binding rows with incompatible types:
df1 <- data.frame(x = 1:5)
df2 <- data.frame(x = letters[1:5])
bind_rows(df1, df2)  # Error!

# Fix: make types consistent
df2$x <- as.character(df2$x)
```

### `Can't rename columns that don't exist`

**Package:** dplyr  
**Chapter:** \@ref(dplyr)  
**Difficulty:** ⭐  
**Quick Fix:** Check column name spelling  

### `Can't subset columns that don't exist`

**Package:** dplyr  
**Chapter:** \@ref(dplyr)  
**Difficulty:** ⭐  
**Quick Fix:** Verify column names with `names()` or `colnames()`  

### `cannot allocate vector of size X`

**Package:** base  
**Chapter:** \@ref(memory-errors)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** Reduce data size, use data.table, or add RAM  
**See also:** Memory management techniques in Chapter \@ref(memory-errors)

### `cannot change value of locked binding for 'X'`

**Package:** base  
**Chapter:** \@ref(objects-variables)  
**Difficulty:** ⭐  
**Quick Fix:** Don't try to reassign T, F, or other protected objects  

### `cannot change working directory`

**Package:** base  
**Chapter:** \@ref(working-directory-paths)  
**Difficulty:** ⭐  
**Quick Fix:** Check path exists and you have permission  

### `cannot coerce class "X" to a data.frame`

**Package:** base  
**Chapter:** \@ref(type-mismatch)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Convert object to appropriate type first  

### `cannot open file 'X': No such file or directory`

**Package:** base  
**Chapter:** \@ref(working-directory-paths)  
**Difficulty:** ⭐  
**Quick Fix:** Check file path and working directory with `getwd()`  
**See also:** File path best practices in Chapter \@ref(working-directory-paths)

### `cannot open the connection`

**Package:** base  
**Chapter:** \@ref(working-directory-paths)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Close file in other programs, check file path  

### `character string is not in a standard unambiguous format`

**Package:** base  
**Chapter:** \@ref(datetime-parsing)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Specify date format explicitly with `format` argument  

### `column 'X' not found`

**Package:** various  
**Chapter:** \@ref(column-manipulation)  
**Difficulty:** ⭐  
**Quick Fix:** Check spelling and use `names(data)` to verify  

### `contexts stack overflow`

**Package:** base  
**Chapter:** \@ref(scoping-environment)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** Check for infinite recursion  

### `contrasts can be applied only to factors with 2 or more levels`

**Package:** stats  
**Chapter:** \@ref(model-fitting)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Check factor has multiple levels or convert to numeric  

### `could not find function "X"`

**Package:** base  
**Chapter:** \@ref(objects-variables)  
**Difficulty:** ⭐  
**Quick Fix:** Load required package with `library()` or use `package::function()`  
**See also:** Package management in Chapter \@ref(package-loading)

---

## D

### `do not know how to convert 'X' to class "Date"`

**Package:** base  
**Chapter:** \@ref(datetime-parsing)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Use proper date parsing function like `as.Date()` with format  

### `duplicate row.names are not allowed`

**Package:** base  
**Chapter:** \@ref(dataframe-construction)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Ensure row names are unique or use `row.names = FALSE`  

---

## E

### `embedded nul in string`

**Package:** base  
**Chapter:** \@ref(working-directory-paths)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** File is corrupted or binary; re-download or clean data  

### `ERROR: dependency 'X' is not available`

**Package:** utils  
**Chapter:** \@ref(installation-environment)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Install with `dependencies = TRUE` or install dependency first  

### `Error in library(X) : there is no package called 'X'`

**Package:** base  
**Chapter:** \@ref(installation-environment)  
**Difficulty:** ⭐  
**Quick Fix:** Install package first: `install.packages("X")`  

### `evaluation nested too deeply`

**Package:** base  
**Chapter:** \@ref(scoping-environment)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** Increase `options(expressions = ...)` or fix infinite recursion  

---

## F

### `Faceting variables must have at least one value`

**Package:** ggplot2  
**Chapter:** \@ref(ggplot2-advanced)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Remove empty factor levels or NA values  

### `figure margins too large`

**Package:** graphics  
**Chapter:** \@ref(base-graphics)  
**Difficulty:** ⭐  
**Quick Fix:** Increase plot window size or adjust `par(mar = ...)`  

### `formal argument "X" matched by multiple actual arguments`

**Package:** base  
**Chapter:** \@ref(function-definition)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Don't provide same argument twice (by name and position)  

---

## G

### `geom_X requires the following missing aesthetics: Y`

**Package:** ggplot2  
**Chapter:** \@ref(ggplot2-basics)  
**Difficulty:** ⭐  
**Quick Fix:** Provide required aesthetic mapping  

### `glm.fit: algorithm did not converge`

**Package:** stats  
**Chapter:** \@ref(model-fitting)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** Increase iterations, check data scaling, simplify model  

---

## I

### `incomplete final line found`

**Package:** base  
**Chapter:** \@ref(working-directory-paths)  
**Difficulty:** ⭐  
**Quick Fix:** Usually harmless; add newline to file or use readr  

### `incorrect number of dimensions`

**Package:** base  
**Chapter:** \@ref(subscript-dimension)  
**Difficulty:** ⭐  
**Quick Fix:** Check object dimensions with `dim()` or `str()`  

### `incorrect number of subscripts on matrix`

**Package:** base  
**Chapter:** \@ref(subscript-dimension)  
**Difficulty:** ⭐  
**Quick Fix:** Use correct number of indices [row, col]  

### `installation of package 'X' had non-zero exit status`

**Package:** utils  
**Chapter:** \@ref(installation-environment)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Check error details, ensure dependencies installed  

### `invalid factor level, NA generated`

**Package:** base  
**Chapter:** \@ref(factor-creation)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Add new levels to factor before assigning  

### `invalid multibyte string`

**Package:** base  
**Chapter:** \@ref(working-directory-paths)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Specify file encoding: `fileEncoding = "UTF-8"`  

---

## L

### `lazy-load database 'X' is corrupt`

**Package:** base  
**Chapter:** \@ref(installation-environment)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Reinstall the package  

### `longer object length is not a multiple of shorter object length`

**Package:** base  
**Chapter:** \@ref(vectors-recycling)  
**Difficulty:** ⭐  
**Quick Fix:** Ensure vectors have compatible lengths  

---

## M

### `missing value where TRUE/FALSE needed`

**Package:** base  
**Chapter:** \@ref(na-null-nan-inf)  
**Difficulty:** ⭐  
**Quick Fix:** Handle NA values before using in if() or logic  

### `more columns than column names`

**Package:** base  
**Chapter:** \@ref(reading-data)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Check CSV delimiter, specify `sep` argument  

---

## N

### `NAs introduced by coercion`

**Package:** base  
**Chapter:** \@ref(type-mismatch)  
**Difficulty:** ⭐  
**Quick Fix:** Check data before converting; can't convert "abc" to number  

### `need finite 'xlim' values`

**Package:** graphics  
**Chapter:** \@ref(base-graphics)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Remove infinite or NA values from data  

### `no applicable method for 'X' applied to an object of class "Y"`

**Package:** base  
**Chapter:** \@ref(s3-s4-methods)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Object type doesn't work with that function; convert or use different function  

### `non-conformable arrays`

**Package:** base  
**Chapter:** \@ref(matrix-operations)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Ensure matrix dimensions match for operation  

### `non-numeric argument to binary operator`

**Package:** base  
**Chapter:** \@ref(type-mismatch)  
**Difficulty:** ⭐  
**Quick Fix:** Convert character to numeric before math operations  

---

## O

### `object 'X' not found`

**Package:** base  
**Chapter:** \@ref(objects-variables)  
**Difficulty:** ⭐  
**Quick Fix:** Check spelling, ensure object created, load package if needed  
**Most Common Error:** See Chapter \@ref(objects-variables) for comprehensive guide

### `object of type 'closure' is not subsettable`

**Package:** base  
**Chapter:** \@ref(objects-variables)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Add () to call function, or you referenced function name instead of object  

---

## P

### `package 'X' is not available for R version Y`

**Package:** utils  
**Chapter:** \@ref(installation-environment)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Update R or install from archive  

### `package 'X' was built before R 4.0.0`

**Package:** base  
**Chapter:** \@ref(package-loading)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Reinstall package  

### `pandoc document conversion failed`

**Package:** rmarkdown  
**Chapter:** \@ref(rmarkdown-quarto)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Check pandoc installation, update RStudio  

### `plot.new has not been called yet`

**Package:** graphics  
**Chapter:** \@ref(base-graphics)  
**Difficulty:** ⭐  
**Quick Fix:** Create plot before adding elements  

### `promise already under evaluation`

**Package:** base  
**Chapter:** \@ref(scoping-environment)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** Circular dependency in default arguments; break the cycle  

### `protect(): protection stack overflow`

**Package:** base  
**Chapter:** \@ref(memory-errors)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** Simplify code, reduce nested calls, restart R  

---

## R

### `rank-deficient fit may be misleading`

**Package:** stats  
**Chapter:** \@ref(model-fitting)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** Remove collinear predictors, check for perfect correlation  

### `recursive indexing failed at level X`

**Package:** base  
**Chapter:** \@ref(dollar-double-bracket)  
**Difficulty:** ⭐⭐  
**Quick Fix:** List element doesn't exist or check index depth  

### `Removed X rows containing missing values`

**Package:** ggplot2  
**Chapter:** \@ref(ggplot2-basics)  
**Difficulty:** ⭐  
**Quick Fix:** Handle NA values or acknowledge they're being dropped  

### `replacement has X rows, data has Y`

**Package:** base  
**Chapter:** \@ref(subscript-dimension)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Ensure replacement vector matches subset size  

### `Rtools is required to build R packages but is not currently installed`

**Package:** utils  
**Chapter:** \@ref(installation-environment)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Install Rtools (Windows) or use binary packages  

---

## S

### `stat_count() must not be used with a y aesthetic`

**Package:** ggplot2  
**Chapter:** \@ref(ggplot2-basics)  
**Difficulty:** ⭐  
**Quick Fix:** Use `geom_col()` instead of `geom_bar()` when providing y values  

### `subscript out of bounds`

**Package:** base  
**Chapter:** \@ref(subscript-dimension)  
**Difficulty:** ⭐  
**Quick Fix:** Index is too large; check object length/dimensions  

### `system is computationally singular`

**Package:** base  
**Chapter:** \@ref(matrix-operations)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** Matrix not invertible; check for zero determinant  

---

## T

### `the condition has length > 1`

**Package:** base  
**Chapter:** \@ref(control-flow)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Use `any()` or `all()`, or use `&&` instead of `&`  

### `there is no package called 'X'`

**Package:** base  
**Chapter:** \@ref(installation-environment)  
**Difficulty:** ⭐  
**Quick Fix:** Install package first  

---

## U

### `unable to access index for repository`

**Package:** utils  
**Chapter:** \@ref(installation-environment)  
**Difficulty:** ⭐  
**Quick Fix:** Check internet connection, try different CRAN mirror  

### `undefined columns selected`

**Package:** base  
**Chapter:** \@ref(column-manipulation)  
**Difficulty:** ⭐  
**Quick Fix:** Check column name spelling  

### `unexpected '=' in "X"`

**Package:** base  
**Chapter:** \@ref(objects-variables)  
**Difficulty:** ⭐  
**Quick Fix:** Use `<-` for assignment or `==` for comparison  

### `unexpected symbol in "X"`

**Package:** base  
**Chapter:** \@ref(objects-variables)  
**Difficulty:** ⭐  
**Quick Fix:** Remove spaces from variable names, add missing operators  

### `unused argument (X = Y)`

**Package:** base  
**Chapter:** \@ref(function-definition)  
**Difficulty:** ⭐  
**Quick Fix:** Remove incorrect argument or check function documentation  

---

## V

### `variable 'X' has no visible binding`

**Package:** base  
**Chapter:** \@ref(scoping-environment)  
**Difficulty:** ⭐⭐⭐  
**Quick Fix:** R CMD check issue; use `.data$` pronoun or declare global  

### `variable lengths differ`

**Package:** stats  
**Chapter:** \@ref(model-fitting)  
**Difficulty:** ⭐⭐  
**Quick Fix:** Ensure all variables in formula have same length  

---

## W

### `Warning in install.packages : unable to access index`

**Package:** utils  
**Chapter:** \@ref(installation-environment)  
**Difficulty:** ⭐  
**Quick Fix:** Check internet, try different mirror  

---

## X-Z

[Additional errors as needed]

---

## Notes on Using This Dictionary

1. **Error text matching**: Some errors have variables (like 'X', 'Y') that will show actual values
2. **Package column**: Indicates which package generates the error
3. **Difficulty**: ⭐ = Common beginner, ⭐⭐⭐ = Advanced/rare
4. **Quick Fix**: First thing to try; full details in referenced chapter
5. **See also**: Links to related errors and concepts

## Can't Find Your Error?

1. Use the index at the beginning of book
2. Search within relevant chapter (e.g., all ggplot errors in ggplot chapter)
3. Check Appendix D for package-specific errors
4. Search complete book text (Ctrl+F in HTML version)
5. Post on Stack Overflow with `[r]` tag
6. Ask on RStudio Community

## Contributing Error Definitions

Found an error not listed? Submit it via GitHub issues at [repo-link]
