# CHAPTER_TITLE {#chapter-id}

<div class="chapter-summary">
**What You'll Learn:**

- Key concept 1
- Key concept 2
- Key concept 3

**Key Errors Covered:** X+ errors

**Difficulty:** ⭐ Beginner / ⭐⭐ Intermediate / ⭐⭐⭐ Advanced
</div>

## Introduction

[Brief introduction to the chapter topic. Why are these errors important? What context do they appear in?]

[1-2 paragraphs setting up the chapter]

## Error #1: Error Message Text {#error-id-1}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-syntax">🔤 CATEGORY</span>

### The Error

```{r error=TRUE}
# Code that produces the error
example_code_here()
```

<div class="error-box">
🔴 **ERROR**

```
Exact error message text here
```
</div>

### What It Means

[Plain English explanation of what R is actually telling you]

[What R was trying to do and why it failed]

### Common Causes

#### Cause 1: Descriptive Name

[Explanation of this cause]

```{r error=TRUE}
# Example demonstrating this cause
bad_code()
```

#### Cause 2: Another Cause

[Explanation]

```{r error=TRUE}
# Another example
more_bad_code()
```

#### Cause 3: Third Common Cause

[Explanation]

```{r error=TRUE}
# Yet another example
```

### Solutions

<div class="solution-box">
✅ **SOLUTION 1: First Approach**

```{r}
# Correct code
good_code()
```

[Explanation of why this works]
</div>

<div class="solution-box">
✅ **SOLUTION 2: Alternative Approach**

```{r}
# Different correct approach
alternative_code()
```

[When to use this approach vs. solution 1]
</div>

<div class="solution-box">
✅ **SOLUTION 3: Advanced/Best Practice**

```{r}
# Best practice code
best_practice_code()
```

[Why this is the best approach]
</div>

### Deep Dive (Optional)

<div class="insight-box">
💡 **Key Insight: Related Concept**

[Deeper explanation of underlying R behavior]

```{r}
# Illustrative example
```

[Further explanation]
</div>

<div class="pitfall-box">
⚠️ **Common Pitfall**

[Description of a common mistake related to this error]

```{r eval=FALSE}
# Bad code example
```

**Why this is problematic:**
- Reason 1
- Reason 2
- Reason 3
</div>

<div class="bestpractice-box">
🎯 **Best Practice**

[Recommended approach to avoid this error]

```{r}
# Example of best practice
```
</div>

## Error #2: Second Error Message {#error-id-2}

<span class="difficulty-intermediate">⭐⭐ INTERMEDIATE</span> <span class="category-badge cat-type">🔢 TYPE</span>

[Follow same structure as Error #1]

### The Error

```{r error=TRUE}
# Code
```

<div class="error-box">
🔴 **ERROR**

```
Error message
```
</div>

### What It Means

[Explanation]

### Common Causes

#### Cause 1: Name

[...]

### Solutions

[...]

## Error #3: Third Error {#error-id-3}

[Continue pattern...]

## Warning #1: Warning Message {#warning-id-1}

<span class="difficulty-beginner">⭐ BEGINNER</span> <span class="category-badge cat-math">🧮 MATH</span>

### The Warning

```{r}
# Code that produces warning
code_with_warning()
```

<div class="warning-box">
🟡 **WARNING**

```
Warning message text
```
</div>

### What It Means

[Explanation]

### Should You Worry?

**Usually NO** - [When it's safe to ignore]

**Maybe YES** - [When you should investigate]

**Definitely YES** - [When it's a serious problem]

### Solutions

[If action needed]

## Advanced Topic: Related Concept {#advanced-topic}

[Optional deeper dive section]

## Debugging Strategy

<div class="insight-box">
💡 **Systematic Approach**

When encountering [this type of error]:

1. **Step 1** - [Check this]
2. **Step 2** - [Then check this]  
3. **Step 3** - [Finally try this]

```{r eval=FALSE}
# Diagnostic code
check_this()
```
</div>

## Summary

<div class="chapter-summary">
**Key Takeaways:**

1. **Point 1** - Brief explanation
2. **Point 2** - Brief explanation
3. **Point 3** - Brief explanation
4. **Point 4** - Brief explanation
5. **Point 5** - Brief explanation

**Quick Reference:**

| Error | Likely Cause | First Thing to Try |
|-------|--------------|-------------------|
| Error 1 | Cause | Solution |
| Error 2 | Cause | Solution |
| Error 3 | Cause | Solution |
| Warning 1 | Cause | Solution |

**Remember:**
- Key principle 1
- Key principle 2
- Key principle 3
</div>

## Exercises

<div class="exercise-box">
📝 **Exercise 1: Understanding Errors**

[Question that tests understanding]

```{r eval=FALSE}
# Code to analyze
code_here()
```

**Questions:**
1. What will happen?
2. Why?
3. How would you fix it?
</div>

<div class="exercise-box">
📝 **Exercise 2: Fix the Code**

Debug this script:

```{r eval=FALSE}
# Broken code
intentionally_broken()
```

Identify and fix all errors.
</div>

<div class="exercise-box">
📝 **Exercise 3: Best Practices**

Rewrite this code following chapter best practices:

```{r eval=FALSE}
# Suboptimal code
needs_improvement()
```
</div>

<div class="exercise-box">
📝 **Exercise 4: Real World**

[Practical scenario-based exercise]

Create a function that:
1. Requirement 1
2. Requirement 2
3. Handles errors gracefully

Test it with various inputs.
</div>

## Exercise Answers

<details>
<summary>Click to see answers</summary>

**Exercise 1:**

[Detailed answer with explanation]

```{r eval=FALSE}
# Corrected code
fixed_code()
```

**Why this works:**
- Reason 1
- Reason 2

**Exercise 2:**

```{r eval=FALSE}
# Original (broken)
broken_code()

# Fixed
fixed_code()
```

**Fixes made:**
1. Fix 1 - explanation
2. Fix 2 - explanation
3. Fix 3 - explanation

**Exercise 3:**

```{r eval=FALSE}
# Improved code following best practices
improved_function <- function(x) {
  # Input validation
  stopifnot(is.numeric(x))
  
  # Clear logic
  result <- process(x)
  
  # Meaningful return
  return(result)
}
```

**Improvements:**
- Used meaningful names
- Added input validation
- Clear structure
- Proper error handling

**Exercise 4:**

```{r eval=FALSE}
# Example solution
robust_function <- function(input_data) {
  # Validate input
  if (!is.data.frame(input_data)) {
    stop("input_data must be a data.frame")
  }
  
  # Handle potential errors
  tryCatch({
    # Process data
    result <- input_data %>%
      filter(!is.na(value)) %>%
      mutate(processed = value * 2)
    
    return(result)
  }, error = function(e) {
    message("Error processing data: ", e$message)
    return(NULL)
  })
}

# Test cases
robust_function(data.frame(value = 1:5))  # Works
robust_function("not a dataframe")        # Handled error
```
</details>

## Additional Resources

- [Link to related documentation]
- [Stack Overflow tag or common questions]
- [Relevant package vignettes]

## Cross-References

See also:
- Chapter X: [Related Topic] (Section \@ref(section-id))
- Appendix A: Error Dictionary for complete listing
- Chapter Y: [Another Related Topic]

---

**Next Chapter:** [Chapter Title] - [Brief description]
